### 2017-11-03 inst/preprint/luxsdt.tex
wersja z preprintem i skryptem po angielsku

% TODO Bardziej szczegółowo przedstawić wyniku testów na symulacji

% TODO Wspomnieć o gamma_to_crit

% TODO wyjaśnić sens prioru ln(100) dla gamma

% TODO opisać domyślne priory dla zeta

% mixing properties?

% constrained czy restricted?

\documentclass[oneside,a4paper]{article}
% \usepackage[MeX]{polski} %pakiety m.in. umozliwia rozne ciekawe rzeczy
% \usepackage[latin2]{inputenc}
% Bedzie osobnym punktem, na osobnej stronie
\usepackage[sectionbib,bibnewpage]{apacite}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{float}
\usepackage{graphicx}
\usepackage{listings}

\title{The bhsdtr package: flexible bayesian inference for
  hierarchical or non-hierarchical equal variance normal Signal
  Detection Theory models with or without ratings.}

\begin{document}
\tableofcontents{}

\section{Abstract}

We describe a novel method of bayesian inference for hierarchical or
non-hierarchical equal variance normal Signal Detection Theory
models. The method is implemented as an open-source R package that
uses state of the art platform Stan for sampling from the posterior
distribution. Our method can accommodate binary responses as well as
additional ratings, the user can specify an arbitrary number of random
grouping factors, sensitivity and criteria parameters can be regressed
on additional predictors within the same model via intermediate
unconstrained parameters and the model can be extended by using an
automatically generated human-readable Stan code as a template. We
explain how our method improves other similar available methods, we
also provide an overview of the package and demonstrate the ease of
use and correctness of implementation by providing a real study data
analysis walk-through and by showing that the model successfully
recovers known parameter values when fitted to simulated data.

\section{Introduction}

Many tasks used in psychology studies are essentially classification
tasks. For example, in a memory study a test item may be either old or
new or in a perceptual study an object may either be a letter or a
digit. If a task requires classification it is always possible that
conclusions based on accuracy or percent correct are invalid because
the ability to discriminate between stimulus classes (i.e.,
sensitivity) is confounded with bias, which is the tendency to
classify stimuli as belonging to a particular class
\cite{GreenSwets66}. In principle, any effect that manifests itself in
the differences in classification accuracy may reflect differences in
sensitivity, bias or both. Signal Detection Theory provides a simple
and popular solution to this common and important problem. However,
SDT models are non-linear and variability in SDT parameters due to
factors such as participants or items has to be accounted for. When
effects of such factors are not accounted for in non-linear models
parameter estimates are biased and the associated uncertainty may be
underestimated.

We have created the \texttt{bhsdtr} package because in our opinion
available general purpose methods of fitting hierarchical SDT models
suffer from important limitations. The goal was to create a tool that
would be useful in a wide range of typical applications, that would
cover a broad spectrum of possible hierarchical structures defined on
SDT parameters and that would be relatively easy to use. We believe
that the \texttt{bhsdtr} package fulfills all of these goals. The
package is publicly available at
https://github.com/boryspaulewicz/bhsdtr, together with annotated R
code and data that was used to produce all the plots and perform all
the analyses described in this paper.

Our method is meant to be used by researchers already familiar with
Signal Detection Theory, hierarchical modelling and bayesian inference
so we begin by providing only a brief refreshment on certain issues in
SDT modelling. To those who do not yet possess the required expertise
we recommend \citeA{GelmanHill2007} for an introduction to bayesian
inference, hierarchical modelling and R and
\citeA{macmillan2004detection} for an introduction to SDT. Before we
go any further, however, a note on terminology seems in order.

In the context of hierarchical modelling factors such as participants,
items or replications are often referred to as groups, a naming
convention that we find confusing at times. For example, using this
terminology a single participant is both a group and a member of some
group, at the same time the term "group" is perhaps most strongly
associated with study conditions, as in "experimental group". In this
paper we use the term "sampled factor" instead, as it seems to capture
all the essential properties of such variables, i.e., the nominal
scale, the fact that values are sampled from a larger population and
are usually not of direct interest, as in "this is only a sample", and
that conclusions of statistical analysis are meant to apply to the
whole population of possible values (e.g., all the people or all the
items of a certain kind).

\section{Equal variance normal Signal Detection Theory model with
  additional criteria}

A generalization of the basic SDT model that we focus on here is shown
in Fig.~\ref{fig:1} below.

\begin{figure}[H]
  \centering
  \includegraphics[width=.8\linewidth]{plot_sdtr.pdf}
  \caption{Equal variance normal Signal Detection Theory model with
    additional criteria}
  \label{fig:1}
\end{figure}
% Czułość i kryterium

The stimulus is assumed to give rise, by some unspecified cognitive
process, to unidimensional internal evidence value $s$ sampled from a
distribution that depends on the stimulus class. For historical
reasons the stimulus classess are often referred to as "noise" and
"signal" and task performance is described in terms of hits, correct
rejections, omissions and false alarms, but this terminology is
appropriate only when the model is applied to tasks that require
detection, which is by far not always the case. In the most widely
used version of the model the two evidence distributions are normal
with the same variance which is usually fixed at unity to make the
model identifiable. Sensitivity is represented by the distance $d'$
between the means of evidence distributions. Because normal
distributions are unbounded evidence is always ambiguous and a
criterion $c$ placed on the evidence axis has to be used to reach a
discrete decision. The location of the decision criterion represents
the direction and magnitude of bias. The generalized version of the
model shown in Fig.~\ref{fig:1} is applicable in studies where
participants are asked to rate their binary classification decisions
on a performance or stimulus related dimension, confidence being a
prime example. The ratings and binary classification decisions can be
provided together (e.g., "I am almost certain that it was a digit"),
or in an arbitrary order (ratings before or after the decision).

The ratings are accommodated by modelling combined response $y$, which
represents both the binary classification decision and rating. The
value of $y$ increases with the strength of evidence in favor of the
second stimulus class. For example, if confidence is rated on a
four-point scale, then $y = 1$ when the subject decides that the
stimulus belongs to the first class (e.g., an old item) with certainty
4, $y = 2$ corresponds to first stimulus class with certainty 3,
$y = 5$ corresponds to second stimulus class (e.g., a new item) with
certainty 1, and $y = 8$ corresponds to second stimulus class with
certainty 4. More formally, the subject is assumed to give response
$k$ if $k = \min \{i : s <= c_i\}$, where
$s \sim \text{Normal}(\mu, 1)$ and $\mu = -d'/2$ for the first
stimulus class and $\mu = d'/2$ for the second stimulus class, $K$ is
the number of possible responses and $c_i$ are the decision criteria,
with $c_0$ and $c_K$ fixed at $-\infty$ and $+\infty$ respectively.

When $K = 2$ (no ratings) the model always fits perfectly, because the
data and the model have the same dimensionality. This makes
generalization to the $K > 2$ case particularly important, as it is
only when $K > 2$ that formal assumptions of the model (e.g., equal or
unequal variance) can be tested\footnote{In contrast to formal
  assumptions the \emph{psychological interpretation} of SDT
  parameters can be tested even when ratings are not available, e.g.,
  by means of selective influence \cite{sternberg2001separate}}, often
by comparing the observed and the theoretical ROC
curves. % , although to the best of our knowledge there exists no direct
% test of the ROC curve shape.  Off course, depending on research
% questions these additional criteria can be of independent interest,
% as can be the ratio of evidence distribution variances or their
% functional form, but here we concentrate on more basic applications.

SDT is a non-linear model, and when non-linear models are used in
psychology studies researchers are usually interested in the
relationships between the free parameters of such models and certain
additional measured or manipulated variables. A good example in the
context of classification performance is the dependence of $d'$ (a
free parameter of a non-linear SDT model) on stimulus strength
(usually a manipulated variable). Also, in great many psychology
studies data have hierarchical structure, e.g., there are repeated
measures for each participant or item and participants or items are
only samples from the target population. It seems worthwhile to
explain why it is important that our method accommodates both kinds of
situations.

\subsection{Importance of hierarchical structure}

If data have hierarchical structure, as is typically the case when SDT
models are used, and assuming that the SDT model is true, it is only
when the requirement of accounting for variability due to subjects,
items or other such factors is met that the estimates of the average
(fixed) effects are guaranteed to be unbiased and the conclusions are
guaranteed to generalize to the target population.

A not so uncommon practice of analysing data aggregated over subjects,
items or other sampled factors represents an extreme case of ignoring
hierarchical data structure. Invalidity of this approach in the
context of SDT was clearly illustrated with results of simulational
studies by \citeA{morey2008problematic}, although strictly speaking it
requires such demonstrations only in order to make a point about a
specific case. SDT is a non-linear model, therefore, by definition
expected values of estimates of its parameters based on data
aggregated over sampled factors are not in general true population
averages. Such estimates are biased and inference about the target
population based on such estimates is simply not valid. To give a
concrete example, concider two unbiased participants, one with
$d' = 2$ and one with $d' = 4$. Their expected average accuracy is
given by $(\Phi(1) + \Phi(2)) / 2 = .91$, which corresponds to
$d' = 2.68$, whereas their true average $d'$ is $3$.

Aggregation is not the only way of ignoring hierarchical data
structure. Sometimes non-aggregated data are analyzed by using
separate estimates for every participant $\times$ item $\times$
condition combination but uncertainty due to subject or item effects
\emph{distributions} is not accounted for by means of hierarchical
model structure. In this case the conclusions, at least with respect
to the uncertainty in the estimates, are guaranteed to be valid only
for the given sample, not the target population. This is rarely, if
ever, the purpose of a study.

\subsection{Importance of including regression structure}

Another common source of problems is separation of estimation of
non-linear model parameters from regression analysis relating these
parameters to other measured or manipulated variables. When SDT
parameters are estimated separately for each subject and condition and
only later regressed on predictors of interest a number of additional
issues may arise.

Firstly, standard errors or credible intervals associated with
regression coefficients do not reflect the uncertainty in SDT
parameter estimates, because the latter are treated as mere data
points. This may lead to overly optimistic inference on regression
coefficients, since what this amounts to, among other things, is
ignoring part of the residual variance.

Secondly, including the regression part in the same model makes the
SDT parameter estimates dependent on the common regression structure
and so also on each other. This can improve their quality and affect
estimates of their uncertainty, just like assuming that random effects
are samples from a common distribution may improve their estimates by
making them dependent in the model.

Thirdly, the precision of parameter estimates often varies between
participants, items or conditions but when these estimates are treated
as mere data points in regression analysis no use is made of this
information. When regression structure is included in the same model
the regression coefficients are less influenced by the parameter
estimates that have lower precision.

\section{A critical summary of avaiable methods of bayesian inference
  for hierarchical SDT models}

To the best of our knowledge only two attempts at creating something
akin to the general purpose method of inference for hierarchical SDT
models with ratings can be found in literature. We will not consider
the ROC Toolbox \cite{Koen2016} or other similar tools here because as
of this writing these tools cannot be used to fit hierarchical SDT
models. The two available methods that can be used to fit hierarchical
SDT models are the Gibbs sampler proposed by
\citeA{morey2008problematic} and the Hierarchical Meta-d' model
(HMeta-d') proposed by \citeA{hmetad}. In our summary of these two
methods we mainly concentrate on two issues, namely the correctness of
hierarchical structure implementation and possibility to model SDT
parameters as dependent on additional variables.

\citeA{morey2008problematic} proposed a Gibbs sampling algorithm for
inference on hierarchical unequal variance normal SDT models with
ratings. A slightly modified version of the algorithm was later made
available as part of the \texttt{hbmem} R package. The authors adopted
what they called a fixed-criteria parametrization. Under this
parametrization the outermost criteria $\bm{c}_1$ and $\bm{c}_{K-1}$
are fixed at $0$ and $1$ respectively, while the remaining criteria
together with means and variances of the evidence distributions are
free to vary. The hierarchical part of evidence distribution means is
represented by additive independent normal effects of at most two
sampled factors which, following the authors, we label as subject and
item effects for convenience. However, the criteria are represented as
subject specific \emph{fixed}, not random effects. Items can have an
indirect and somewhat limited random effect on the criteria, because
they can have an effect on evidence distribution means by shifting
them independently. As the authors clarify, shifting both means by the
same amount in the same direction does not change the $d'$ but
simultaneously shifts all the criteria relative to the evidence
distributions. This allows an item to have the same effect on all the
criteria but it does not allow an item to have different effects on
different criteria. Random effects correlations are not accounted
for. Criteria are order-restricted, but the means of evidence
distributions are order-restricted only in the version implemented in
the \texttt{hbmem} package. SDT parameters can be regressed on
additional predictors to some extent by specifying a model
matrix. Lastly, because this is a Gibbs sampler, authors rely on
conjugate priors, including the problematic \cite{gelman2004prior}
inverse gamma prior placed on variances.

Fleming's \citeyear{hmetad} HMeta-d' model is a hierarchical version
of the meta-d' model \cite{maniscalco2012signal}, which in turn is a
generalization of the SDT model with additional criteria that allows
for a separate "meta-sensitivity" in order to account for possible
discrepancies between binary stimulus classification (referred to as
type 1 task) and a rating task (referred to as type 2 or
meta-cognitive task). We consider HMeta-d' here because it reduces to
the SDT model with ratings when type 1 and type 2 sensitivities are
equal. In HMeta-d' evidence distributions are normal with equal
variance. Normally distributed random effects of one factor are
allowed for sensitivities and parameters indirectly representing the
criteria, for lack of a better name from now on referred to as raw
criteria. The actual criteria are subject-specific sorted values of
unconstrained raw criteria and it is these unconstrained parameters,
not the actual criteria, that are given independent hierarchical
normal distributions representing variability due to subjects. Since
sorting is not injective the space of raw criteria is not isomorphic
to the space of actual criteria and each element of the raw criteria
vector can represent a value of any criterion, depending on the values
of other elements of the raw criteria vector. The model does not
account for possible correlations between random
effects. Sensitivities are represented as unrestricted scalars even
though they are non-negative by definition. Out of the box the code
provided by the author allows only for a very limited regression
structure to be placed on SDT parameters. Software implementation
relies on ready-made model specifications expressed in JAGS language
that are aimed at several simple types of experimental design (e.g.,
single subject or single group of subjects). The models can be fitted
using convenient MATLAB functions.

\section{Hierarchical Signal Detection Theory in unconstrained
  parameter space}

As we have already stressed, in a typical study both sensitivity and
criteria parameters have to be supplemented with hierarchical
structure for inference to be guaranteed to generalize to the target
population. In our critical summary of the two available methods we
have listed some reasons why we believe that without substantial
modifications neither method correctly addresses this issue. The main
reason why both methods fail in correctly specifying the hierarchical
structure on SDT parameters is that random effects are represented by
normal distributions but parameters are constrained, i.e.,
sensitivities are non-negative and criteria are ordered.

In the \texttt{bhsdtr} package sensitivities and criteria are derived
from unconstrained parameters by convenient isomorphisms. This is
easily achieved in case of sensitivity by deriving $d'$ from
$\delta = \ln(d')$ and modelling random effects on $d'$, if any, by
assuming that $\delta$ is normally distributed. Difficulties in
representing the hierarchical structure of the criteria stem from the
fact that even though each criterion considered in isolation is an
unconstrained parameter, the vector of all the criteria is ordered.
% The solution used in the \texttt{bhsdtr} package is based on
% the way ordered categorical variables with two or more categories,
% such as confidence ratings, are modelled in ordered logistic
% regression.
We propose to map the $R^{K-1}$ space, representing unconstrained
criteria, to the $K$ dimensional probability simplex space using the
softmax function and to map the simplex space to the ordered criteria
space by means of the inverse normal CDF. Specifically, assume that a
normal distribution is centered on the midpoint between the two
evidence distributions. The value of criterion $i$ is then given by:

\begin{align}
  \bm{c}_i &= \Phi^{-1}(\sum_{k = 1}^i(e^{\bm{\gamma}_k}) / \sum_{j=1}^K(e^{\bm{\gamma}_j}))
\label{eq:1}
\end{align}

\noindent where $\bm{\gamma} \in R^K$ with $\bm{\gamma}_K$ fixed at
$0$ for identifiability. The idea is illustrated in
Fig.~\ref{fig:2} below:

\begin{figure}[H]
  \centering
  \includegraphics[width=.8\linewidth]{plot.pdf}
  \caption{Mapping between unconstrained $\gamma$ vector and criteria}
  \label{fig:2}
\end{figure}

The mapping in Eq.~\ref{eq:1} is an isomorphism between the $R^{K-1}$
space and the space of $K-1$ ordered criteria vectors. Elements
$i = 1...K$ of the $\bm{\gamma}$ vector represent relative distances
between adjacent criteria because their exponents represent relative
magnitudes of areas under the normal curve delineated by the pairs of
adjacent criteria:
$e^{\bm{\gamma}_i} / e^{\bm{\gamma}_j} = (\Phi(\bm{c}_i) -
\Phi(\bm{c}_{i-1})) / (\Phi(\bm{c}_j) - \Phi(\bm{c}_{j-1}))$.
% Other things being equal, the greater the $\bm{\gamma}_i$ value the
% more the $\bm{c}_{j \geq i}$ criteria are shifted to the right.
When $K=2$ only $\bm{\gamma}_1$ is free to vary and its value directly
represents the direction and magnitude of bias: $\bm{\gamma}_1$ is $0$
when the criterion is placed at the midpoint between the means of
evidence distributions, and the more negative (positive) it is the
more the criterion is placed to the left (right) of midpoint. Our
experience indicates that it is often a good idea to multiply all the
criteria by a value greater than $1$ which is equivalent to making the
mapping distribution in Eq.~\ref{eq:1} wider. This tends to even out
the values of $\bm{\gamma}$, especially when criteria are widely
spread as can happen for moderate to large $d'$ values. This feature
is implemented in the \texttt{bhsdtr} package by introducing a scaling
factor.

Once the $d'$ and $c$ are represented as derived from unconstrained
$\delta$ and $\gamma$ parameters the SDT model can be supplemented
with hierarchical linear regression structure. To avoid having to deal
with even more complicated index notation, below we have presented
only the simple case of one sampled factor.

\begin{align*}
  \bm{\delta} &= \bm{X}^{(\delta)} \bm{\beta}^{(\delta)} + \bm{Z}^{(\delta)} \bm{\theta}^{(\delta)} \\
  \bm{d'}_i &= e^{\bm{\delta}_i} \\
  \bm{\gamma}_{i,.} &= \bm{X}^{(\gamma)}_{i,.} \bm{\beta}^{(\gamma)} + \bm{Z}^{(\gamma)}_{i,.}
                      \bm{\theta}^{(\gamma)} \\
  \bm{c}_{i,k} &= s \ \Phi^{-1}(\sum_{l = 1}^i(e^{\bm{\gamma}_{i,l}}) /
                 \sum_{m=1}^K(e^{\bm{\gamma}_{i,m}})) \\
  p(\bm{y}_i = k|\bm{stim}_i = 1) &= \Phi(\bm{c}_{i,k} + \bm{d'}_i / 2) - \Phi(\bm{c}_{i,k-1} + \bm{d'}_i / 2) \\
  p(\bm{y}_i = k|\bm{stim}_i = 2) &= \Phi(\bm{c}_{i,k} - \bm{d'}_i / 2) - \Phi(\bm{c}_{i,k-1} - \bm{d'}_i / 2)
\end{align*}

\noindent Here $i$ is the observation number, $\bm{X}$ is the fixed
effects model matrix for the respective parameter, $\bm{Z}$ is the
random effects model matrix, $\bm{\beta}$ and $\bm{\theta}$ are the
fixed and random effects, $\bm{c}_{i,k}$ is the $k$-th criterion for
the $i$-th observation, $s$ is the criteria scaling factor, and
$\bm{y}$ is the combined response. Note that $\bm{d'}_i$ is a scalar,
but $\bm{\gamma}_{i,.}$ is in general a vector and so
$\bm{\beta}^{(\gamma)}$ and $\bm{\theta}^{(\gamma)}$ are matrices. The
$j$-th rows of the $\bm{\beta}^{(\gamma)}$ and
$\bm{\theta}^{(\gamma)}$ matrices represent fixed and random effects
for the $j$-th element of the $\bm{\gamma}$ vector.

Following \citeA{sorensen2015bayesian} we make use of the Cholesky
decomposition of the correlation matrices because it improves
efficiency and admits a convenient prior on random effects'
correlations:

\begin{align*}
  \text{vectorized}(\bm{\theta}^{(\gamma)}) &= \text{diag}(\bm{\tau}^{(\gamma)}) \bm{L}^{(\gamma)} \bm{z}^{(\gamma)} \\
  \bm{\theta}^{(\delta)} &= \text{diag}(\bm{\tau}^{(\delta)}) \bm{L}^{(\delta)} \bm{z}^{(\delta)} \\
  \bm{z}^{(\delta)}_i &\sim \text{Normal}(0, 1) \\
  \bm{z}^{(\gamma)}_j &\sim \text{Normal}(0, 1)
\end{align*}

\noindent where $\bm{\tau}$ is a vector of standard deviations of
random effects and $\bm{L}$ is the Cholesky decomposition of the
correlation matrix, i.e., $\bm{C} = \bm{L} \bm{L}'$. This way
$\bm{\theta}$ is multivariate normal with covariance matrix
$\text{diag}(\bm{\tau}) \bm{L}$.

We use weakly informative proper priors because they provide
regularization and help stabilize computation. Fixed effects
($\bm{\beta}^{(\delta)}$ and $\bm{\beta}^{(\gamma)}$) are given
independent normal priors, random effects standard deviations
($\bm{\tau}^{(\delta)}$ and $\bm{\tau}^{(\gamma)}$) are given
independent half-Cauchy priors as recommended by
\citeA{gelman2004prior} and each $\bm{L}$ is given an independent lkj
prior:

\begin{align*}
  \bm{\beta}^{(\delta)}_i &\sim \text{Normal}(\bm{\mu}^{(\delta)}_i, \bm{\sigma}^{(\delta)}_i) \\
  \bm{\beta}^{(\gamma)}_{k,l} &\sim \text{Normal}(\bm{\mu}^{(\gamma)}_{k,l}, \bm{\sigma}^{(\gamma)}_{k,l}) \\
  \bm{\tau}^{(\delta)}_i &\sim \text{half-Cauchy}(0, \bm{\zeta}^{(\delta)}_i) \\
  \bm{\tau}^{(\gamma)}_{k,l} &\sim \text{half-Cauchy}(0, \bm{\zeta}^{(\gamma)}_{k,l}) \\
  \bm{L}^{(\delta)} &\sim \text{lkj}(\nu^{(\delta)}) \\
  \bm{L}^{(\gamma)} &\sim \text{lkj}(\nu^{(\gamma)}) \\
\end{align*}

\subsection{Specifying prior distributions}

Specifying the priors on sensitivity effects does not pose any special
difficulties. Sensitivity of an unbiased classifier given percent
correct is given by $2 \Phi^{-1}(\text{pc})$. When both stimulus
classes are equally likely the greater the bias the lower the accuracy
meaning that unbiased sensitivity represents the lower bound on
sensitivity given percent correct. Let us assume that a great majority
of subjects are expected to achieve percent correct within the $.51$
to $.99$ range with negligible bias. Since
$\ln(2\Phi^{-1}(.51)) = -2.99$ and $\ln(2 \Phi^{-1}(.99)) = 1.54$ a
reasonable weakly informative prior on delta is normal with mean
$(1.54 - 2.99) / 2$ and standard deviation $(1.54 + 2.99) / 2$ which
is the default prior on delta effects in the \texttt{bhsdtr}
package.

Specifying priors on the criteria can be challenging because of the
ordering restriction. In addition, complexity of mapping in
Eq.~\ref{eq:1} makes it difficult to reason about the priors on
$\bm{\gamma}$ in terms of the criteria. By default in the
\texttt{bhsdtr} package each entry in $\bm{\sigma}^{(\gamma)}$ and
$\bm{\zeta}^{(\gamma)}$ matrices is set to $\ln(100)$ and the criteria
scaling factor is fixed at 2. The meaning of $\ln(100)$ depends on
model matrix parametrization to some degree, but usually it means that
a priori areas under the mapping distribution between adjacent
criteria may easily differ from the area between criterion $K-1$ and
$K$ by a factor of 100 but differences by a factor of $2 \ln(100)$ or
more are highly unlikely. Off course, when in doubt it is a good idea
to perform some sort of sensitivity analysis, remembering that strong
sensitivity to priors indicates that the study may be underpowered.

The prior on random effects standard deviations is parametrized by
$\bm{\zeta}$, which represents half-width at half-maximum of the
half-Cauchy distribution. In our opinion a not-unreasonable starting
point is to set $\bm{\zeta}$ at the value that is greater or equal to
the most likely value of the random effects standard deviation.

By default $\nu^{(\delta)} = \nu^{(\gamma)} = 1$ implying a uniform
prior on random effects correlation matrices. Because the greater the
value of $\nu$ the more emphasis is put on zero off-diagonal
correlations the researcher can force the correlations to be near-zero
by choosing large $\nu$ value.

\section{Overview of the software implementation}

We have chosen the Stan modelling language \cite{carpenter2016stan}
because it uses state-of-the-art adaptive Hamiltonian Monte Carlo
sampling algorithm which often handles high-dimensional correlated
posteriors better than the Gibbs sampler. The \texttt{bhsdtr} package
is essentially a collection of documented functions. The
\texttt{aggregate\_responses} function serves to aggregate the data as
much as possible for efficiency, but without distorting the
hierarchical structure. The \texttt{make\_stan\_model} function
creates the model code based on the provided model formulae for random
effects. The Stan code produced by the \texttt{make\_stan\_model}
function can be fitted as is or modified by the user if needed, e.g.,
to change the prior distributions or drop the equal variance
assumption. The \texttt{make\_stan\_data} function creates regression
model matrices and other data structures required by the model created
with the \texttt{make\_stan\_model} function. Finally, the
\texttt{plot\_sdt\_fit} function can be used to assess the model fit
visually. This function creates publication-ready ROC curve plots or
response distribution plots with posterior predictive intervals
calculated for the chosen $\alpha$ level.

\subsection{Usage example: installing the package and testing the
  model on real data}

% Instalacja

In order to make full use of the \texttt{bhsdtr} package functionality
three non-standard R packages are required, namely \texttt{rstan},
\texttt{plyr} and \texttt{ggplot2}. We recommend using the
\texttt{devtools} package to install the \texttt{bhsdtr} package
directly from the github repository. This will also automatically
install any missing required packages:

\begin{lstlisting}{language = R}
  devtools::install_git('boryspaulewicz/bhsdtr')
  library(bhsdtr)
\end{lstlisting}

% gabor 13

We tested our method on the \texttt{gabor} dataset that is part of the
\texttt{bhsdtr} package. The \texttt{gabor} dataset comes from an
unpublished study in which on each trial participants had to classify
the briefly presented gabor patch as either oriented to the left or to
the right using the arrow keys. The participants were also asked to
rate the stimuli on a 4 point Perceptual Awareness Scale
\cite{ramsoy2004introspection} presented at the bottom of the
screen. The gabor patch was immediately followed by a mask. The PAS
ratings ranged from "no experience" to "absolutely clear image" and
were provided either before (RD condition) or after (DR condition)
pressing the arrow keys. On each trial the gabor patch was equally
likely to be presented for 32ms or 64ms. Decision-rating order was a
between-subject variable and gabor patch presentation duration was a
within-subject variable. There were 47 participants and 48 trials per
condition.

We will now explain how to perform some of the essential steps of a
typical data analysis process, which at the very least usually
involves preparing the data, creating the model code, fitting the
model, assessing the fit, and possibly converting the unconstrained
$\gamma$ and $\delta$ parameters to $d'$ and $c$.

\subsubsection{Step 1: preparing the data}
% Combined response

In the study in question the response was originally encoded as
accuracy and rating, so the first step was to create an appropriate
response variable using the \texttt{combined\_response} function. This
function needs three variables, one encoding the stimulus class, one
encoding the rating (as an integer) and one encoding decision
accuracy. This step is required only if ratings are available and the
combined response variable is not already present in the data. In the
$K=2$ case the combined response variable is simply the binary
classification decision.

\begin{lstlisting}{language=R}
  gabor$resp = combined_response(gabor$stim, 
                                 gabor$rating, 
                                 gabor$acc)
\end{lstlisting}

% aggregate_responses

Next, the data was aggregated by computing combined response counts,
but only to the extent that preserved all the participant effects. The
\texttt{aggregate\_responses} function that does exactly this requires
as arguments a data frame containing all relevant variables, a name of
the stimulus class variable, a name of the combined response variable
and a vector of names of all variables that are to be preserved in a
resulting aggregated dataset, i.e., those encoding sampled factors and
those representing independent variables used in the regression part
of the model:

\begin{lstlisting}{language=R}
adata = aggregate_responses(gabor, 'stim', 'resp', 
                            c('id', 'duration', 'order'))  
\end{lstlisting}

The main purpose of the aggregation step is to improve the efficiency
of sampling from the posterior distribution. When the data are
aggregated in this way the likelihood for each condition $\times$
participant combination has to be computed only once rather than as
many times as there are trials per condition per participant. Note
that if there are other sampled factors present in the data (e.g.,
items, replications, etc.) then these factors have to be specified at
this stage as well to preserve hierarchical structure. The
\texttt{aggregate\_responses} function creates a list with three
components. The \texttt{data} component is a data frame containing
additional preserved variables, the \texttt{stimulus} component is the
stimulus class variable and the \texttt{counts} component is a matrix
of counts with $K$ columns, where $K$ is the highest possible combined
response.

\subsubsection{Step 2: creating the model code}
% stan requirements

The model is fitted using the \texttt{stan} function from the
\texttt{rstan} package which needs a special list of data structures
used by the model as well as a model specification expressed in the
Stan language. Every SDT model has some fixed effects structure, since
even when there are no predictors the model parameters can be
expressed as regressed on a vector of ones (i.e., an
intercept). However, many models also have a hierarchical structure
and, if that is the case, this hierarchical structure has to be
specified when using the \texttt{make\_stan\_model} function. This is
done by providing a list of lists of R model formulae. Each list of
model formulae is composed of at least three elements and specifies
correlated random effects of one sampled factor. The element named
\texttt{group} specifies the sampled factor and the elements named
\texttt{delta} and \texttt{gamma} specify which effects are assumed to
vary between the levels of this sampled factor. % Four other optional
% elements can be used to specify the priors on random effects
% distributions. The elements named \texttt{delta\_nu} and
% \texttt{gamma\_nu} correspond to $\nu^{(\bm{\delta})}$ and
% $\nu^{(\bm{\gamma})}$ parameters and the elements named
% \texttt{delta\_scale} and \texttt{gamma\_scale} correspond to
% $\tau^{(\bm{\delta})}$ and $\tau^{(\bm{\delta})}$ parameters.
%
% Of course, random effects must be estimable within the levels of the
% sampled factor and must already be present in the fixed effects
% specification for the respective parameter, at least implicitly.
When \texttt{make\_stan\_model} is used without any arguments it
specifies the SDT model without any random effects. Fixed effects
model matrices are specified by providing a list with at least two
model formulae, named \texttt{delta} and \texttt{gamma}, to the
\texttt{make\_stan\_data} function, described later in this
paper. Non-default priors can be specified by adding optional elements
to the random and fixed effects specification lists, as described in
the \texttt{make\_stan\_data} function documentation.

In the study in question there was only one sampled factor (id) which
represents participants. Because duration was a within-participant
variable in principle its effect could vary between participants for
all the SDT parameters. However, preliminary data analysis indicated
that the barely noticeable 32ms difference in duration seemed to
affect only the sensitivity. Thus, it was assumed that duration
affects only the $\delta$ parameter and that this effect may vary
between participants:

\begin{lstlisting}{language=R}
fixed = list(delta = ~ -1 + duration:order, 
             gamma = ~ order)
random = list(list(group = ~ id, 
              delta = ~ -1 + duration, 
              gamma = ~ 1))
model = make_stan_model(random)
\end{lstlisting}

% KOD Specyfikacja i dopasowanie modelu

The \texttt{make\_stan\_data} function creates fixed and random effect
model matrices based on the respective model formulae using dummy
contrast coding. Note that the implicit intercept was suppressed for
the $\delta$ model matrix. This way $\delta$ was estimated for every
duration $\times$ order condition\footnote{R uses dummy contrast
  coding by default}. The resulting separate intercepts and slopes
parametrization makes it easier to calculate arbitrary contrasts based
on the posterior samples. A more standard parametrization was used for
the $\gamma$ parameter, because it was initially assumed that criteria
depend only on order and so there was only one contrast of interest
for every element of the $\gamma$ vector. On the other hand even in
such case the separate intercepts and slopes parametrization may be
more convenient if the researcher is interested in the actual
criteria. It is perhaps worth emphasising how the separation of
$\delta$ and $\gamma$ regression structures makes it possible to test
a broad class of linear models representing the dependence of
sensitivity and criteria on additional variables.

\subsubsection{Step 3: fitting the model}
% make_stan_data

A separate data structure used by the Stan sampler has to be created
using the \texttt{make\_stan\_data} function. The obligatory arguments
to this function are the aggregated data object created by
\texttt{aggregate\_responses} function and fixed effects
specification. Because in our example study duration was a
within-subject variable here we also provide the random effects
specification. Importantly, if random effects are modelled the same
random effects specification has to be provided to the
\texttt{make\_stan\_model} and \texttt{make\_stan\_data} functions.

\begin{lstlisting}{language=R}
sdata = make_stan_data(adata, fixed, random)
\end{lstlisting}

Finally, a vector of names of parameters of interest has to be
specified when calling the \texttt{stan} function:

\begin{lstlisting}{language=R}
fit = stan(model_code = model,
           data = sdata,
           pars = c('delta_fixed', 'gamma_fixed',
                    'delta_sd_1', 'gamma_sd_1',
                    'delta_random_1', 'gamma_random_1',
                    'Corr_delta_1', 'Corr_gamma_1',
                    'counts_new'),
           iter = 8000,
           chains = 4)
\end{lstlisting}

Note that since more than one sampled factor is allowed the names of
parameters defining the hierarchical part are indexed (e.g.,
\texttt{delta\_sd\_1}, \texttt{delta\_random\_1},
\texttt{Corr\_delta\_1}). The name \texttt{counts\_new} refers to the
posterior predictive samples that are required by the
\texttt{plot\_sdt\_fit} function, names starting with \texttt{Corr}
refer to random effects correlation matrices.

\subsubsection{Step 4: assessing the model fit}
% Jak się zbiega

Four chains of 8000 iterations each were run simultaneously and the
first half of the posterior samples serving as a warm-up period for
tuning the parameters of the sampling algorithm was discarded. Part of
the resulting stan output is presented in Table~\ref{tab:fitsum}
below.

\begin{table}[H]
\caption{Model fit summary statistics}
\label{tab:fitsum}
\begin{tabular}{rrrrrrrr}
  \hline
  & mean & $SE_{mean}$ & $SD$ & $2.5\%$ & $97.5\%$ & Effective samples & $\hat{R}$ \\ 
  \hline
  delta\_fixed[1] & -0.10 & 0.00 & 0.15 & -0.41 & 0.18 & 4626 & 1.00 \\ 
  delta\_fixed[2] & 1.12 & 0.00 & 0.09 & 0.95 & 1.29 & 4764 & 1.00 \\ 
  delta\_fixed[3] & -0.39 & 0.00 & 0.20 & -0.80 & -0.03 & 5427 & 1.00 \\ 
  delta\_fixed[4] & 1.28 & 0.00 & 0.11 & 1.06 & 1.50 & 5464 & 1.00 \\ 
  gamma\_fixed[1,1] & -0.14 & 0.00 & 0.06 & -0.27 & -0.02 & 2925 & 1.00 \\ 
  gamma\_fixed[1,2] & -0.22 & 0.00 & 0.10 & -0.42 & -0.01 & 8199 & 1.00 \\ 
  gamma\_fixed[2,1] & -0.70 & 0.00 & 0.18 & -1.05 & -0.35 & 3060 & 1.00 \\ 
  gamma\_fixed[2,2] & 0.49 & 0.00 & 0.29 & -0.06 & 1.04 & 3452 & 1.00 \\ 
  gamma\_fixed[3,1] & -0.54 & 0.00 & 0.22 & -0.96 & -0.11 & 3103 & 1.00 \\ 
  gamma\_fixed[3,2] & 0.83 & 0.01 & 0.35 & 0.14 & 1.51 & 3086 & 1.00 \\ 
  gamma\_fixed[4,1] & 0.28 & 0.00 & 0.25 & -0.20 & 0.76 & 3090 & 1.00 \\ 
  gamma\_fixed[4,2] & 0.42 & 0.01 & 0.40 & -0.37 & 1.22 & 3291 & 1.00 \\ 
  gamma\_fixed[5,1] & -0.21 & 0.01 & 0.30 & -0.80 & 0.37 & 3440 & 1.00 \\ 
  gamma\_fixed[5,2] & 0.83 & 0.01 & 0.48 & -0.10 & 1.78 & 3637 & 1.00 \\ 
  gamma\_fixed[6,1] & -0.77 & 0.00 & 0.24 & -1.23 & -0.31 & 3378 & 1.00 \\ 
  gamma\_fixed[6,2] & 0.80 & 0.01 & 0.38 & 0.05 & 1.56 & 3316 & 1.00 \\ 
  gamma\_fixed[7,1] & -0.32 & 0.00 & 0.16 & -0.64 & 0.01 & 3346 & 1.00 \\ 
  gamma\_fixed[7,2] & 0.42 & 0.00 & 0.28 & -0.11 & 0.98 & 3333 & 1.00 \\ 
  delta\_sd\_1[1] & 0.66 & 0.00 & 0.11 & 0.48 & 0.90 & 6373 & 1.00 \\ 
  delta\_sd\_1[2] & 0.44 & 0.00 & 0.05 & 0.35 & 0.56 & 5463 & 1.00 \\ 
  gamma\_sd\_1[1] & 0.16 & 0.00 & 0.08 & 0.02 & 0.31 & 2071 & 1.00 \\ 
  gamma\_sd\_1[2] & 0.82 & 0.00 & 0.11 & 0.62 & 1.06 & 5852 & 1.00 \\ 
  gamma\_sd\_1[3] & 1.08 & 0.00 & 0.12 & 0.86 & 1.33 & 4293 & 1.00 \\ 
  gamma\_sd\_1[4] & 1.27 & 0.00 & 0.14 & 1.03 & 1.57 & 3723 & 1.00 \\ 
  gamma\_sd\_1[5] & 1.55 & 0.00 & 0.17 & 1.24 & 1.91 & 4297 & 1.00 \\ 
  gamma\_sd\_1[6] & 1.18 & 0.00 & 0.13 & 0.95 & 1.47 & 4348 & 1.00 \\ 
  gamma\_sd\_1[7] & 0.78 & 0.00 & 0.10 & 0.60 & 1.00 & 4399 & 1.00 \\ 
  Corr\_delta\_1[1,1] & 1.00 & 0.00 & 0.00 & 1.00 & 1.00 & 16000 &  \\ 
  Corr\_delta\_1[1,2] & 0.76 & 0.00 & 0.10 & 0.52 & 0.91 & 3871 & 1.00 \\ 
  Corr\_delta\_1[2,1] & 0.76 & 0.00 & 0.10 & 0.52 & 0.91 & 3871 & 1.00 \\ 
  Corr\_delta\_1[2,2] & 1.00 & 0.00 & 0.00 & 1.00 & 1.00 & 15199 & 1.00 \\ 
  Corr\_gamma\_1[1,1] & 1.00 & 0.00 & 0.00 & 1.00 & 1.00 & 16000 &  \\ 
  Corr\_gamma\_1[1,2] & -0.20 & 0.01 & 0.27 & -0.66 & 0.39 & 840 & 1.00 \\ 
  Corr\_gamma\_1[1,3] & -0.17 & 0.01 & 0.28 & -0.67 & 0.41 & 820 & 1.01 \\ 
  Corr\_gamma\_1[1,4] & -0.04 & 0.01 & 0.28 & -0.56 & 0.51 & 840 & 1.00 \\ 
  Corr\_gamma\_1[1,5] & -0.06 & 0.01 & 0.28 & -0.60 & 0.49 & 836 & 1.00 \\ 
  Corr\_gamma\_1[1,6] & -0.27 & 0.01 & 0.27 & -0.75 & 0.32 & 838 & 1.00 \\ 
  Corr\_gamma\_1[1,7] & -0.19 & 0.01 & 0.28 & -0.69 & 0.39 & 871 & 1.01 \\ 
  Corr\_gamma\_1[2,1] & -0.20 & 0.01 & 0.27 & -0.66 & 0.39 & 840 & 1.00 \\ 
  Corr\_gamma\_1[2,2] & 1.00 & 0.00 & 0.00 & 1.00 & 1.00 & 16000 & 1.00 \\ 
  Corr\_gamma\_1[2,3] & 0.58 & 0.00 & 0.13 & 0.29 & 0.79 & 3716 & 1.00 \\ 
  Corr\_gamma\_1[2,4] & 0.39 & 0.00 & 0.15 & 0.06 & 0.65 & 2806 & 1.00 \\ 
  Corr\_gamma\_1[2,5] & 0.22 & 0.00 & 0.16 & -0.11 & 0.52 & 2712 & 1.00 \\ 
  Corr\_gamma\_1[2,6] & 0.46 & 0.00 & 0.14 & 0.17 & 0.70 & 3968 & 1.00 \\ 
  Corr\_gamma\_1[2,7] & 0.80 & 0.00 & 0.09 & 0.60 & 0.93 & 5171 & 1.00 \\ 
  \dots{} & \dots{} & \dots{} & \dots{} & \dots{} & \dots{} & \dots{}
                                                                       & \dots{} \\
  \hline
\end{tabular}
\end{table}

As can be seen, the chains exhibited good mixing properties given the
complexity of the model and seemed to have reached apparent
convergence: there were enough effective samples of fixed effect
parameters to estimate $95\%$ credible intervals well and none of the
Gelman-Rubin $\hat{R}$ statistics crossed the conventional $1.01$
threshold suggesting negligible sensitivity to initial values.

Fig.~\ref{fig:6} and \ref{fig:7} below contain normal
quantile-quantile plots of random effects showing that normally
distributed random $\delta$ and $\gamma$ effects seem to be good
candidates for representing variability in sensitivities and criteria
due to sampled factors.

\begin{figure}[H]
  \centering
  \includegraphics[width=.8\linewidth]{delta_qq_plot.pdf}
  \caption{Normal quantile-quantile plots of $\delta$ random effects}
  \label{fig:6}
\end{figure}

\begin{figure}[H]
  \centering
  \includegraphics[width=.8\linewidth]{gamma_qq_plot.pdf}
  \caption{Normal quantile-quantile plots of $\gamma$ random effects}
  \label{fig:7}
\end{figure}

Even when stan output summary does not indicate any obvious sampler
convergence issues before drawing any further conclusions the
researcher should first check if the model fits the data. We recommend
using the \texttt{plot\_sdt\_fit} function for that purpose:

\begin{lstlisting}{language = R}
plot_sdt_fit(fit, adata, c('order', 'duration')))
\end{lstlisting}

This function needs at least three agruments; a stanfit object, an
aggregated data list produced by the \texttt{aggregate\_responses}
function that was used to produce the stanfit object and a vector of
names of variables that define how the data will be partitioned before
plotting. Even though we highly recommend assessing the fit at the
individual level we did not include the participant identification
number in the list of conditioning variables because the resulting
plot would take up too much space. As can be seen in Fig.~\ref{fig:3}
showing the ROC curves produced by the code above the model seemed to
fit the data well in all but one condition (decision after rating,
64ms duration) where three out of seven relevant\footnote{The point in
  the upper-right corner of the ROC curve is always in the $(1,1)$
  position} points were outside the two-dimensional 95\% posterior
predictive regions.

\begin{figure}[H]
  \centering
  \includegraphics[width=.8\linewidth]{roc_fit.pdf}
  \caption{ROC curve fit}
  \label{fig:3}
\end{figure}

% The reader should be warned that ROC point posterior predictive
% intervals must be interpreted with care, because ROC points are not
% independent and it is in general not correct to expect 95\% coverage
% from 95\% ROC point posterior predictive intervals when the model is
% true.

Another way to assess the model fit visually is by inspecting the
conditional response distributions ($p(y|stim)$) shown in
Fig.~\ref{fig:4} which was also created using the
\texttt{plot\_sdt\_fit} function.

\begin{figure}[H]
  \centering
  \includegraphics[width=.8\linewidth]{response_fit.pdf}
  \caption{Response distribution fit}
  \label{fig:4}
\end{figure}

This kind of plot can sometimes be informative about the reasons why
the model does not fit the data. In this particular case the plot
seems to suggest that it may be a good idea to inspect the fit at the
individual level and see if there are some participants with unusual
$p(y|stim=1)$ distributions in the RD $\times$ 64ms condition. On the
other hand it is also possible that the lack of fit was mainly a
consequence of the assumption that duration had zero effect on
$\bm{\gamma}$ or maybe the model is simply incorrect and more
substantial modifications are necessary.

\subsubsection{Step 5: converting unconstrained $\delta$ and $\gamma$
  parameters to sensitivities and criteria}

The posterior $\delta$ and $\gamma$ samples have to be transformed in
order to work with $d'$ and $c$ parameters. This is straightforward
only when fixed effects represent separate conditions, not the
differences between conditions or the regression slopes. In our
example, because separate intercepts and slopes parametrization was
used for the $\delta$ fixed effects model matrix all four
\texttt{delta\_fixed} parameters can easily be transformed to
sensitivities by applying the logarithm function to the posterior
samples. It is important to remember that the $\delta$ to $d'$
conversion should usually be done first, before applying any other
transformations to the samples, for example, the logarithm of point
and interval $\delta$ estimate is not equal to the point and interval
$d'$ estimate.

In case of the \texttt{gamma\_fixed} matrix the first column (the
intercept) corresponds to the values of the $\bm{\gamma}$ vector
\emph{in} the DR condition but the second column corresponds to
\emph{the effect} of the order variable on the $\bm{\gamma}$
vector. For this reason in this particular case posterior criteria
samples can be obtained using the \texttt{gamma\_to\_crit} function
only for the first column of the \texttt{gamma\_fixed} matrix.

\subsection{Testing the model on simulated data}

To test if the model correctly recovers known parameter values we have
simulated the data from a hypothetical exact replication of the
previously described experiment using point estimates from the
previous fit as known realistic parameter values. Mixing performance
and apparent convergence were similar as in the real data case. All
the model parameters were correctly recovered in a sense that the true
values were outside the 95\% credible intervals no more than 5\% of
the time.

For illustration purposes the SDT model that differed from the true
model only in that it did not have the hierarchical structure
representing participant effects was fitted to the same simulated
dataset. Since the model was much simpler and the data consisted of
only eight vectors of response counts apparent convergence was quickly
reached and chain mixing was excellent.

Point and interval estimates of fixed effects based on both models are
compared in Fig.~\ref{fig:5} below. The estimates were centered at the
true values to simplify the presentation. As can be seen, the true
model correctly recovered known parameter values, but estimates based
on the non-hierarchical model were clearly biased; $95\%$ credible
intervals were not only about three times shorter on average but also
failed to contain most of the true values.

\begin{figure}[H]
  \centering
  \includegraphics[width=.8\linewidth]{true_vs_nonhier.pdf}
  \caption{Comparison of point and interval posterior estimates based
    on the true and simplified non-hierarchical models}
  \label{fig:5}
\end{figure}

However, as can be seen on Fig.~\ref{fig:8} below in this case the
observed ROC curves seemed to fit the simplified model's predictions
quite well, giving a false impression of model validity.

\begin{figure}[H]
  \centering
  \includegraphics[width=.8\linewidth]{roc_sim_aggr_fit.pdf}
  \caption{ROC curve fit for the non-hierarchical model}
  \label{fig:8}
\end{figure}

\section{Concluding remarks}

The great importance of SDT to psychology stems from the fact that
given some fairly weak assumptions about the underlying decision
process it promises to deconfound sensitivity from bias in arbitrary
classification tasks. To the best of our knowledge at present the
\texttt{bhsdtr} package provides the only method of bayesian inference
for SDT models with or without ratings that can be recommended as a
default choice in typical applications. Our method allows for great
flexibility that requires minimal input from the user. Parametrization
forces sensitivities to be non-negative and criteria to be ordered
while the isomorphisms between unconstrained $\delta$ and $\gamma$ and
derived $d'$ abd $c$ parameters ensure that hierarchical structure is
valid. There is no limit to the number of sampled factors except for
the one imposed by available computational resources, correlations of
random effects of the same sampled factor are accounted for, all the
SDT parameters can be modelled by linear regression within the same
model and all the effects on all the SDT parameters estimable within
the levels of sampled factors can have associated correlated random
effects. In case a need arises to relax a built-in restriction
experienced users can extend the model in arbitrary ways by using
automatically generated human-readable Stan model code as a
template. We hope that researchers with basic understanding of Signal
Detection Theory, bayesian inference and hierarchical modelling will
find our package useful and adopt it as a method of analysis of
classification performance in their future studies. In the near future
we are planning to extend the package by relaxing the equal variance
assumption and adding the ability to fit other generalizations of SDT,
such as the meta-d' model, that allow for different information to be
used by classification decision and rating.

\bibliography{/home/borys/cs/literatura}
\bibliographystyle{apacite}
\end{document}

### 2017-11-03 inst/preprint/analysis_script.R

@@ -1,5 +1,11 @@
 ## -*- coding: utf-8 -*-
 
+## Change this to FALSE once all the stanfit objects are stored to
+## save time
+fresh_start = TRUE
+## Where do you want to store all those large stanfit objects?
+temp_path = '~/temp'
+
 library(rstan)
 library(ggplot2)
 library(Hmisc) # rMultinom for simulations
@@ -12,6 +18,12 @@ rstan_options(auto_write = TRUE)
 options(mc.cores = parallel::detectCores())
 library(plyr)
 
+######################################################################
+## Fitting the model to real study data
+
+## We will be using the dataset provided with the package
+?gabor
+
 ## Combined response has to be calculated for this dataset
 gabor$r = combined_response(gabor$stim, gabor$rating, gabor$acc)
 
@@ -22,29 +34,35 @@ adata = aggregate_responses(gabor, 'stim', 'r', c('duration', 'id', 'order'))
 fixed = list(delta = ~ -1 + duration:order, gamma = ~ order)
 random = list(list(group = ~ id, delta = ~ -1 + duration, gamma = ~ 1))
 model = make_stan_model(random)
+## Let's take a look at the model code
+cat(model)
 
 ## Stan data structure
 sdata = make_stan_data(adata, fixed, random)
 
 ## Main model fit
-fit = stan(model_code = model,
-           data = sdata,
-           pars = c('delta_fixed', 'gamma_fixed',
-                    'delta_sd_1', 'gamma_sd_1',
-                    'delta_random_1', 'gamma_random_1',
-                    'Corr_delta_1', 'Corr_gamma_1',
-                    ## we want counts_new for plotting
-                    'counts_new'),
-           iter = 8000,
-           chains = 4)
-## save(fit, file = '~/temp/fit')
-load('~/temp/fit')
+if(fresh_start){
+    fit = stan(model_code = model,
+               data = sdata,
+               pars = c('delta_fixed', 'gamma_fixed',
+                        'delta_sd_1', 'gamma_sd_1',
+                        'delta_random_1', 'gamma_random_1',
+                        'Corr_delta_1', 'Corr_gamma_1',
+                        ## we need counts_new for plotting
[…]

### 2017-11-03 inst/preprint/luxsdt.tex
strona tytulowa i wyciete komentarze

@@
-% TODO Bardziej szczegółowo przedstawić wyniku testów na symulacji
-
-% TODO Wspomnieć o gamma_to_crit
-
-% TODO wyjaśnić sens prioru ln(100) dla gamma
-
-% TODO opisać domyślne priory dla zeta
-
-% mixing properties?
-
-% constrained czy restricted?
+% mixing properties, constrained czy restricted
 
 \documentclass[oneside,a4paper]{article}
-% \usepackage[MeX]{polski} %pakiety m.in. umozliwia rozne ciekawe rzeczy
-% \usepackage[latin2]{inputenc}
-% Bedzie osobnym punktem, na osobnej stronie
 \usepackage[sectionbib,bibnewpage]{apacite}
@@
 \begin{document}
+\maketitle
 \tableofcontents{}
@@
 \end{figure}
-% Czułość i kryterium
 
@@
 requires such demonstrations only in order to make a point about a
-specific case. SDT is a non-linear model, therefore, by definition
+specific case. Because SDT is a non-linear model, by definition
 expected values of estimates of its parameters based on data
@@
 unconstrained parameter, the vector of all the criteria is ordered.
-% The solution used in the \texttt{bhsdtr} package is based on
-% the way ordered categorical variables with two or more categories,
-% such as confidence ratings, are modelled in ordered logistic
-% regression.
 We propose to map the $R^{K-1}$ space, representing unconstrained
@@
 $e^{\bm{\gamma}_i} / e^{\bm{\gamma}_j} = (\Phi(\bm{c}_i) -
-\Phi(\bm{c}_{i-1})) / (\Phi(\bm{c}_j) - \Phi(\bm{c}_{j-1}))$.
-% Other things being equal, the greater the $\bm{\gamma}_i$ value the
-% more the $\bm{c}_{j \geq i}$ criteria are shifted to the right.
-When $K=2$ only $\bm{\gamma}_1$ is free to vary and its value directly
+\Phi(\bm{c}_{i-1})) / (\Phi(\bm{c}_j) - \Phi(\bm{c}_{j-1}))$. When
+$K=2$ only $\bm{\gamma}_1$ is free to vary and its value directly
 represents the direction and magnitude of bias: $\bm{\gamma}_1$ is $0$
@@
 
-% Instalacja
-
 In order to make full use of the \texttt{bhsdtr} package functionality
@@
 \end{lstlisting}
-
-% gabor 13
 
@@
 \subsubsection{Step 1: preparing the data}
-% Combined response
 
@@
 \end{lstlisting}
-
-% aggregate_responses
 
@@
 \subsubsection{Step 2: creating the model code}
-% stan requirements
 
@@
 \texttt{delta} and \texttt{gamma} specify which effects are assumed to
-vary between the levels of this sampled factor. % Four other optional
-% elements can be used to specify the priors on random effects
-% distributions. The elements named \texttt{delta\_nu} and
-% \texttt{gamma\_nu} correspond to $\nu^{(\bm{\delta})}$ and
-% $\nu^{(\bm{\gamma})}$ parameters and the elements named
-% \texttt{delta\_scale} and \texttt{gamma\_scale} correspond to
-% $\tau^{(\bm{\delta})}$ and $\tau^{(\bm{\delta})}$ parameters.
-%
-% Of course, random effects must be estimable within the levels of the
-% sampled factor and must already be present in the fixed effects
-% specification for the respective parameter, at least implicitly.
-When \texttt{make\_stan\_model} is used without any arguments it
-specifies the SDT model without any random effects. Fixed effects
-model matrices are specified by providing a list with at least two
-model formulae, named \texttt{delta} and \texttt{gamma}, to the
+vary between the levels of this sampled factor. When
+\texttt{make\_stan\_model} is used without any arguments it specifies
+the SDT model without any random effects. Fixed effects model matrices
+are specified by providing a list with at least two model formulae,
+named \texttt{delta} and \texttt{gamma}, to the
 \texttt{make\_stan\_data} function, described later in this
@@
 
-% KOD Specyfikacja i dopasowanie modelu
-
 The \texttt{make\_stan\_data} function creates fixed and random effect
@@
 \subsubsection{Step 3: fitting the model}
-% make_stan_data
 
@@
 \subsubsection{Step 4: assessing the model fit}
-% Jak się zbiega
 
@@
 \end{figure}
-
-% The reader should be warned that ROC point posterior predictive
-% intervals must be interpreted with care, because ROC points are not
-% independent and it is in general not correct to expect 95\% coverage
-% from 95\% ROC point posterior predictive intervals when the model is
-% true.
 

### 2017-11-03 inst/preprint/analysis_script.R
kosmetyka skryptu

@@ -1,5 +1,8 @@
 ## -*- coding: utf-8 -*-
 
+######################################################################
+## IMPORTANT GLOBAL VARIABLES
+##
 ## Change this to FALSE once all the stanfit objects are stored to
 ## save time
 fresh_start = TRUE
@@ -9,10 +12,7 @@ temp_path = '~/temp'
 library(rstan)
 library(ggplot2)
 library(Hmisc) # rMultinom for simulations
-devtools::document('~/cs/code/r/bhsdtr')
-devtools::install('~/cs/code/r/bhsdtr')
 library(bhsdtr)
-data(gabor)
 ## Standard Stan optimizations
 rstan_options(auto_write = TRUE)
 options(mc.cores = parallel::detectCores())
@@ -22,6 +22,7 @@ library(plyr)
 ## Fitting the model to real study data
 
 ## We will be using the dataset provided with the package
+data(gabor)
 ?gabor
 
 ## Combined response has to be calculated for this dataset

### 2017-11-03 inst/preprint/luxsdt.tex
added single-criterion SDT test

@@
 encoding the rating (as an integer) and one encoding decision
-accuracy. This step is required only if ratings are available and the
-combined response variable is not already present in the data. In the
-$K=2$ case the combined response variable is simply the binary
-classification decision.
+accuracy.
 
@@
                                  gabor$acc)
+\end{lstlisting}
+
+This step is required only if ratings are available and the combined
+response variable is not already present in the data. In the single
+criterion case the combined response variable is simply the binary
+classification decision. To fit the single-criterion SDT model to this
+dataset the code above would have to be replaced with the following:
+
+\begin{lstlisting}{language=R}
+  gabor$resp = combined_response(gabor$stim, accuracy = gabor$acc)
 \end{lstlisting}

### 2017-11-03 inst/preprint/analysis_script.R

@@ -4,7 +4,7 @@
 ## IMPORTANT GLOBAL VARIABLES
 ##
 ## Change this to FALSE once all the stanfit objects are stored to
-## save time
+## save time and memory
 fresh_start = TRUE
 ## Where do you want to store all those large stanfit objects?
 temp_path = '~/temp'
@@ -80,6 +80,68 @@ ggsave(p2, file = 'response_fit.pdf')
 ## The criteria are placed fairly symmetrically
 crit = gamma_to_c(as.data.frame(fit))
 round(apply(crit, 2, mean), 2)
+
+######################################################################
+## Fitting the baic single criterion SDT model
+
+fixed = list(delta = ~ -1 + duration:order, gamma = ~ order)
+random = list(list(group = ~ id, delta = ~ -1 + duration, gamma = ~ 1))
+gabor$r = combined_response(gabor$stim, accuracy = gabor$acc)
+if(fresh_start){
+    fit.s = stan(model_code = make_stan_model(random),
+                 data = make_stan_data(aggregate_responses(gabor, 'stim', 'r', c('duration', 'order', 'id')),
+                                       fixed, random),
+                 pars = c('delta_fixed', 'gamma_fixed',
+                          'delta_sd_1', 'gamma_sd_1',
+                          'delta_random_1', 'gamma_random_1',
+                          'Corr_delta_1', 'Corr_gamma_1',
+                          'counts_new'),
+                 iter = 8000,
+                 chains = 4)
+    save(fit.s, file = paste(temp_path, 'fit.s', sep = '/'))
+}
+load(paste(temp_path, 'fit.s', sep = '/'))
+## Single criterion version is much faster to sample
+
+print(fit.s, probs = c(.025, .957),
+      pars = c('delta_fixed', 'gamma_fixed',
+               'delta_sd_1', 'gamma_sd_1',
+               'Corr_delta_1', 'Corr_gamma_1'))
+## All is fine
+
+## It does not make sense to use plot_sdt_fit here
+
+## Compare delta estimates
+smr1 = as.data.frame(summary(fit)$summary[,c(1, 2, 3, 4, 8, 9, 10)])
+smr2 = as.data.frame(summary(fit.s)$summary[,c(1, 2, 3, 4, 8, 9, 10)])
+smr = rbind(smr1[grep('delta', rownames(smr1)),],
[…]

### 2017-11-03 inst/preprint/luxsdt.tex
mniej in order to

@@
 on additional predictors within the same model via intermediate
-unconstrained parameters and the model can be extended by using an
+unconstrained parameters, and the model can be extended by using an
 automatically generated human-readable Stan code as a template. We
@@
 generalization of the SDT model with additional criteria that allows
-for a separate "meta-sensitivity" in order to account for possible
+for a separate "meta-sensitivity" to account for possible
 discrepancies between binary stimulus classification (referred to as
@@
 
-In order to make full use of the \texttt{bhsdtr} package functionality
-three non-standard R packages are required, namely \texttt{rstan},
+To make full use of the \texttt{bhsdtr} package functionality three
+non-standard R packages are required, namely \texttt{rstan},
 \texttt{plyr} and \texttt{ggplot2}. We recommend using the

### 2017-11-03 inst/preprint/luxsdt.tex
black and white option in plot_sdt_fit

@@
 variance. Normally distributed random effects of one factor are
-allowed for sensitivities and parameters indirectly representing the
-criteria, for lack of a better name from now on referred to as raw
+allowed for sensitivities and for parameters indirectly representing
+the criteria, for lack of a better name from now on referred to as raw
 criteria. The actual criteria are subject-specific sorted values of
@@
 provided by the author allows only for a very limited regression
-structure to be placed on SDT parameters. Software implementation
-relies on ready-made model specifications expressed in JAGS language
-that are aimed at several simple types of experimental design (e.g.,
-single subject or single group of subjects). The models can be fitted
-using convenient MATLAB functions.
+structure to be placed on SDT parameters.
+% Software implementation relies on ready-made model specifications
+% expressed in JAGS language that are aimed at several simple types of
+% experimental design (e.g., single subject or single group of
+% subjects). The models can be fitted using convenient MATLAB
+% functions.
 

### 2017-11-03 R/plot_sdt_fit.R

@@ -5,19 +5,23 @@
 #'
 #' @param fit a stanfit object.
 #' @param adata an aggregated data object produced by the
-#'   \code{aggregate_responses} function. It must be the same aggregated data
-#'   object that was used to produce the stanfit object.
-#' @param variables an optional vector of names of variables. These variables,
-#'   if specified, define the subsets of data that will be represented by
-#'   separate plots.
-#' @param type a type of plot required. If type = 'roc' then ROC curve plots are
-#'   created, if type != 'roc' then response distribution plots are created.
-#' @param alpha an alpha level for the posterior predictive intervals, e.g., if
-#'   alpha = .05 then 100\% - alpha = 95\% posterior predictive intervals will
-#'   be calculated.
+#'     \code{aggregate_responses} function. It must be the same
+#'     aggregated data object that was used to produce the stanfit
+#'     object.
+#' @param variables an optional vector of names of variables. These
+#'     variables, if specified, define the subsets of data that will
+#'     be represented by separate plots.
+#' @param type a type of plot required. If type = 'roc' then ROC curve
+#'     plots are created, if type != 'roc' then response distribution
+#'     plots are created.
+#' @param alpha an alpha level for the posterior predictive intervals,
+#'     e.g., if alpha = .05 then 100\% - alpha = 95\% posterior
+#'     predictive intervals will be calculated.
+#' @param bw if TRUE the black-and-white version of the response
+#'     distribution plot will be created. The default is FALSE.
 #' @return a plot object.
 #' @export
-plot_sdt_fit = function(fit, adata, variables = NULL, type = 'roc', alpha = .05){
+plot_sdt_fit = function(fit, adata, variables = NULL, type = 'roc', alpha = .05, bw = FALSE){
     s = as.data.frame(fit)
     rm(fit)
     cnt_new = t(s[,grep('counts_new', names(s))])
@@ -89,12 +93,21 @@ plot_sdt_fit = function(fit, adata, variables = NULL, type = 'roc', alpha = .05)
         dfa[,c('count.lo', 'count.hi', 'count.fit')] = cbind(apply(cnt_new_a, 2, function(x)stats::quantile(x, alpha / 2)),
                                                              apply(cnt_new_a, 2, function(x)stats::quantile(x, 1 - alpha / 2)),
[…]

### 2017-12-06 inst/preprint/luxsdt.tex
wersja ktora poszla do Mikea

--- 2017-11-03 inst/preprint/luxsdt.tex
@@
-% mixing properties, constrained czy restricted
+% WĄTPLIWOŚCI JĘZYKOWE oznaczone przez %?? !! oznacza miejsce, gdzie
+% skończyłem ostatnio poprawiać
+
+% przykłady w sieci value of <> [directly] represents the <>
+
+% jak się pisze 16ms czy 16 ms
+
+% argumenty funkcji są an, gdy piszemy ogólnie o działaniu tych
+% funkcji
 
@@
 \usepackage{listings}
-
-\title{The bhsdtr package: flexible bayesian inference for
-  hierarchical or non-hierarchical equal variance normal Signal
-  Detection Theory models with or without ratings.}
+\usepackage{caption}
+
+\title{The bhsdtr package: a general purpose method of bayesian
+  inference for Signal Detection Theory models}
 
@@
 We describe a novel method of bayesian inference for hierarchical or
-non-hierarchical equal variance normal Signal Detection Theory
-models. The method is implemented as an open-source R package that
-uses state of the art platform Stan for sampling from the posterior
-distribution. Our method can accommodate binary responses as well as
-additional ratings, the user can specify an arbitrary number of random
-grouping factors, sensitivity and criteria parameters can be regressed
-on additional predictors within the same model via intermediate
-unconstrained parameters, and the model can be extended by using an
-automatically generated human-readable Stan code as a template. We
-explain how our method improves other similar available methods, we
-also provide an overview of the package and demonstrate the ease of
-use and correctness of implementation by providing a real study data
-analysis walk-through and by showing that the model successfully
-recovers known parameter values when fitted to simulated data.
+non-hierarchical equal variance normal Signal Detection Theory models
+with one or more criteria. The method is implemented as an open-source
+R package that uses the state-of-the-art platform Stan for sampling
+from posterior distributions. Our method can accommodate binary
+responses as well as additional ratings and an arbitrary number of
+nested
+% ?? WSZYSTKIE sdt parameters CAN be regressed, więc the
+or crossed random grouping factors. The SDT parameters can be
+regressed on additional predictors within the same model via
+intermediate unconstrained parameters, and the model can be extended
+by using automatically generated human-readable Stan code as a
+% ?? czy to jest przypadek other -> bez przedimka?
+template. In the paper we explain how our method improves on other
+similar available methods, we give an overview of the package,
+demonstrate the ease of use by providing a real-study data analysis
+walk-through, and show that the model successfully recovers known
+parameter values when fitted to simulated data.
 
@@
 Many tasks used in psychology studies are essentially classification
-tasks. For example, in a memory study a test item may be either old or
-new or in a perceptual study an object may either be a letter or a
-digit. If a task requires classification it is always possible that
-conclusions based on accuracy or percent correct are invalid because
-the ability to discriminate between stimulus classes (i.e.,
-sensitivity) is confounded with bias, which is the tendency to
-classify stimuli as belonging to a particular class
-\cite{GreenSwets66}. In principle, any effect that manifests itself in
-the differences in classification accuracy may reflect differences in
-sensitivity, bias or both. Signal Detection Theory provides a simple
-and popular solution to this common and important problem. However,
-SDT models are non-linear and variability in SDT parameters due to
-factors such as participants or items has to be accounted for. When
-effects of such factors are not accounted for in non-linear models
-parameter estimates are biased and the associated uncertainty may be
-underestimated.
-
-We have created the \texttt{bhsdtr} package because in our opinion
-available general purpose methods of fitting hierarchical SDT models
-suffer from important limitations. The goal was to create a tool that
-would be useful in a wide range of typical applications, that would
-cover a broad spectrum of possible hierarchical structures defined on
-SDT parameters and that would be relatively easy to use. We believe
-that the \texttt{bhsdtr} package fulfills all of these goals. The
-package is publicly available at
-https://github.com/boryspaulewicz/bhsdtr, together with annotated R
-code and data that was used to produce all the plots and perform all
-the analyses described in this paper.
-
-Our method is meant to be used by researchers already familiar with
-Signal Detection Theory, hierarchical modelling and bayesian inference
-so we begin by providing only a brief refreshment on certain issues in
-SDT modelling. To those who do not yet possess the required expertise
-we recommend \citeA{GelmanHill2007} for an introduction to bayesian
-inference, hierarchical modelling and R and
-\citeA{macmillan2004detection} for an introduction to SDT. Before we
-go any further, however, a note on terminology seems in order.
+tasks. For example, in a memory study participants may be required to
+decide if a given test item is old or new, or, in a perceptual study,
+an object may either be a letter or a digit. If a task requires
+classification it is always possible that conclusions based on
+% ? an ability mi tu nie pasuje
+accuracy or percent correct are invalid because ability to
+discriminate between stimulus classes (i.e., sensitivity) is
+confounded with bias, which is a tendency to classify stimuli as
+belonging to a particular class \cite{GreenSwets66}. In principle, any
+effect that manifests itself in differences in classification accuracy
+% ?? ani the sensitivity ani a sensitivity nie pasuje. Chodzi o
+% ogólnie rozumiane zjawisko.
+may reflect differences in sensitivity, bias, or both. Signal
+Detection Theory provides a simple and popular solution to this common
+and important problem.
+
+However, because the SDT model is non-linear any variability in its
+parameters due to factors such as participants or items has to be
+accounted for, otherwise the estimates of SDT parameters are
+biased. As we explain later in this paper, neither of the two
+available methods of inference for hierarchical SDT models that we are
+aware of addresses this problem correctly. Hence, our main goal was to
+create a correct implementation of the general hierarchical linear
+regression structure defined on SDT parameters. We argue that the
+\texttt{bhsdtr} package for R \cite{rstatistical} provides exactly
+such an implementation and we have made it publicly available at
+% ?? tylko jeden, konkretny kod i zbiór danych, o którym czytelnik
+% dopiero się dowiaduje, ale o którym nie będziemy więcej mówić, czyli
+% the
+https://github.com/boryspaulewicz/bhsdtr, together with the annotated
+source code that was used to perform all the analyses and produce all
+the figures presented in this paper.
+
+% OPIS STRUKTURY ARTYKUŁU
+
+In what follows, after introducing the most common version of the SDT
+model, we describe its generalization which can accomodate data from
+rating experiments. Next, we explain briefly why, if a method of
+inference for SDT models were to be of general use in psychology
+studies, it is essential that it was based on a model equipped with
+the general hierarchical linear regression structure. The
+\texttt{bhsdtr} package meets this requirement thanks to a novel
+parametrization. We describe this novel parametrization in great
+detail and we motivate it by explaining how the reliance on a
+relatively standard parametrisation leads to problems in the two other
+available implementations. We end the first part of this paper with a
+formal definition of the model as implemented in \texttt{bhsdtr}. The
+second part contains an overview of the package and a tutorial in
+which we demonstrate how to use our method in practice. Before we go
+any further, however, a note on terminology seems in order.
+
+% SAMPLED FACTOR
 
 In the context of hierarchical modelling factors such as participants,
-items or replications are often referred to as groups, a naming
-convention that we find confusing at times. For example, using this
-terminology a single participant is both a group and a member of some
-group, at the same time the term "group" is perhaps most strongly
-associated with study conditions, as in "experimental group". In this
-paper we use the term "sampled factor" instead, as it seems to capture
-all the essential properties of such variables, i.e., the nominal
-scale, the fact that values are sampled from a larger population and
-are usually not of direct interest, as in "this is only a sample", and
-that conclusions of statistical analysis are meant to apply to the
-whole population of possible values (e.g., all the people or all the
-items of a certain kind).
+items, or replications are often referred to as groups. In our opinion
+this naming convention may be confusing; a single participant is both
+a group and a member of some group, while at the same time the term
+"group" is perhaps most strongly associated with study conditions, as
+in "experimental group". In this paper we use the term "sampled
+factor" instead, because, by virtue of being a new term, it is
+unambiguous, and because it seems descriprively correct: the term
+"sampled factor" captures all the essential properties of such
+variables, i.e., a nominal scale, the fact that values are sampled
+from a larger population and are usually not of direct interest, as in
+"this is only a sample", and that conclusions
+% ?? jeżeli już mamy jakąś analizę statystyczną, to zakładamy, w
+% uproszczeniu, że target population jest określona
+of statistical analysis are meant to apply to the whole population of
+possible values.
 
@@
 
-A generalization of the basic SDT model that we focus on here is shown
-in Fig.~\ref{fig:1} below.
+According to Signal Detection Theory each stimulus in a classification
+task gives rise, by some unspecified cognitive process, to an
+unidimensional internal evidence value $s$ sampled from
+% ?? stimulus już był wprowadzony, więc jego klasa jest (jakaś, ale)
+% określona
+a distribution that depends on the stimulus class. For historical
+reasons the two stimulus classess are often referred to as "noise" and
+"signal", and task performance is described in terms of hits, correct
+rejections, omissions, and false alarms, but this terminology is
+appropriate only when the model is applied to tasks that require
+detection, which is by far not always the case. In the most widely
+used version of the model, shown in Fig.~\ref{basicsdt}, the two
+evidence distributions are normal with the same variance, which is
+usually fixed at unity to make the model identifiable. The distance
+$d'$ between the means of the
+% ?? sensitivity bez przedimka?
+evidence distributions represents sensitivity. Because normal
+distributions are unbounded, $s$ is always ambiguous, and so a
+criterion $c$ placed on the evidence axis has to be used to reach a
+discrete
+% ?? bias bez przedimka?
+decision. A participant is assumed to decide that a stimulus belongs
+to the first class (e.g., an old item) if $s < c$, or that it belongs
+to the second class (e.g., a new item) if $s \geq c$. The location of
+the decision criterion $c$ represents bias.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=.8\linewidth]{plot_basicsdt.pdf}
+  \caption{Equal variance normal Signal Detection Theory model}
+  \label{basicsdt}
+\end{figure}
+
+% Najprostsze użycie
+
+Perhaps the simplest way of using this model is to fit it to observed
+response counts and use the estimated $d'$ values in place of the
+percent correct scores; If the model is correct, the resulting
+performance measure is not contaminated by bias. Obviously, the model
+may not be correct, which is one of the reasons why we focus more on
+the generalized version shown in Fig.~\ref{sdtr} below.
+
+% Model K >= 2
 
@@
     additional criteria}
-  \label{fig:1}
+  \label{sdtr}
 \end{figure}
 
-The stimulus is assumed to give rise, by some unspecified cognitive
-process, to unidimensional internal evidence value $s$ sampled from a
-distribution that depends on the stimulus class. For historical
-reasons the stimulus classess are often referred to as "noise" and
-"signal" and task performance is described in terms of hits, correct
-rejections, omissions and false alarms, but this terminology is
-appropriate only when the model is applied to tasks that require
-detection, which is by far not always the case. In the most widely
-used version of the model the two evidence distributions are normal
-with the same variance which is usually fixed at unity to make the
-model identifiable. Sensitivity is represented by the distance $d'$
-between the means of evidence distributions. Because normal
-distributions are unbounded evidence is always ambiguous and a
-criterion $c$ placed on the evidence axis has to be used to reach a
-discrete decision. The location of the decision criterion represents
-the direction and magnitude of bias. The generalized version of the
-model shown in Fig.~\ref{fig:1} is applicable in studies where
-participants are asked to rate their binary classification decisions
-on a performance or stimulus related dimension, confidence being a
-prime example. The ratings and binary classification decisions can be
+This
+% ?? applicable to czy applicable in
+generalized model is applicable to studies where participants are
+asked to rate their binary classification decisions on confidence or
+some other performance or stimulus related
+% the ratings, bo study już określone, podobnie jak (rodzaj)
+% klasyfikacji binarnej, ale czy the ratings and THE binary ...?
+dimension. The ratings and the binary classification decisions can be
 provided together (e.g., "I am almost certain that it was a digit"),
-or in an arbitrary order (ratings before or after the decision).
-
-The ratings are accommodated by modelling combined response $y$, which
+or in an arbitrary order.
+
+% MODELOWANIE RATINGÓW ZA POMOCĄ KRYTERIÓW
+
+% ?? a combined response? Skoro mamy ratings (i w domyśle mamy też
+% binary class.), to czy nie the combined response? A może
+% jakiekolwiek ratingi są załatwione za pomocą (ogólnego rozwiązania)
+% modelowania czegoś, co jest "odpowiedzią łączną"?
+Ratings are accommodated by introducing additional criteria and
+modelling a combined response $y$, which
+% ?? znowu kwestia przedimków vs. and
 represents both the binary classification decision and rating. The
@@
 second stimulus class. For example, if confidence is rated on a
-four-point scale, then $y = 1$ when the subject decides that the
-stimulus belongs to the first class (e.g., an old item) with certainty
-4, $y = 2$ corresponds to first stimulus class with certainty 3,
-$y = 5$ corresponds to second stimulus class (e.g., a new item) with
-certainty 1, and $y = 8$ corresponds to second stimulus class with
-certainty 4. More formally, the subject is assumed to give response
-$k$ if $k = \min \{i : s <= c_i\}$, where
-$s \sim \text{Normal}(\mu, 1)$ and $\mu = -d'/2$ for the first
-stimulus class and $\mu = d'/2$ for the second stimulus class, $K$ is
-the number of possible responses and $c_i$ are the decision criteria,
-with $c_0$ and $c_K$ fixed at $-\infty$ and $+\infty$ respectively.
-
-When $K = 2$ (no ratings) the model always fits perfectly, because the
-data and the model have the same dimensionality. This makes
+four-point scale, then $y = 1$ when a subject decides
+% ?? bodziec, który zobaczył badany, więc the? bo chodzi o decyzję o
+% określonym dla decydenta bodźcu, a nie nieokreślonym dla decydenta?
+that a stimulus belongs to the first class with certainty $4$, and
+$y = 5$ when a subject decides that a stimulus belongs to the second
+class with certainty $1$. More formally, if $K$ is the number of
+possible combined responses then a subject is assumed to give response
+$k$ if $s \in (\bm{c}_{k-1},\bm{c}_k]$, where $\bm{c}$ are the $K+1$
+decision criteria, with $\bm{c}_0$ and $\bm{c}_K$ fixed at $-\infty$
+and $+\infty$ respectively.
+
+% UŻYWAĆ SDTR CHOĆBY PO TO, ŻEBY SPRAWDZIĆ JEGO ZAŁOŻENIA
+
+There is a good reason to collect the ratings and use the generalized
+SDT model from Fig.~\ref{sdtr} even when neither the ratings nor the
+placement of criteria are relevant to the research problem.
+% ?? If K = 2 to jakby: w (jakimś) przypadku, w którym, When K = 2 to
+% jakby: zawsze, gdy ...? Czuję, że lepiej the data i the model potem
+When $K = 2$ (no ratings), the SDT model fits perfectly, regardless of
+it being a reasonably good approximation to reality or not, because
+the data and the model have the same dimensionality. This makes the
 generalization to the $K > 2$ case particularly important, as it is
-only when $K > 2$ that formal assumptions of the model (e.g., equal or
-unequal variance) can be tested\footnote{In contrast to formal
-  assumptions the \emph{psychological interpretation} of SDT
+% ?? niektóre lub wszystkie formal assumptions, ale mamy na myśli
+% konkretny zbiór formal assumptions, który czytelnik już zna, a
+% nawet, jeśli nie, to te założenia są unikalnie wyznaczone przed
+% model - the?
+only when $K > 2$, that the formal assumptions of the model (e.g.,
+equal or unequal variance) can be tested\footnote{In contrast to the
+  formal assumptions, a psychological interpretation of the SDT
   parameters can be tested even when ratings are not available, e.g.,
-  by means of selective influence \cite{sternberg2001separate}}, often
-by comparing the observed and the theoretical ROC
-curves. % , although to the best of our knowledge there exists no direct
-% test of the ROC curve shape.  Off course, depending on research
-% questions these additional criteria can be of independent interest,
-% as can be the ratio of evidence distribution variances or their
-% functional form, but here we concentrate on more basic applications.
-
-SDT is a non-linear model, and when non-linear models are used in
-psychology studies researchers are usually interested in the
-relationships between the free parameters of such models and certain
-additional measured or manipulated variables. A good example in the
-context of classification performance is the dependence of $d'$ (a
-free parameter of a non-linear SDT model) on stimulus strength
-(usually a manipulated variable). Also, in great many psychology
-studies data have hierarchical structure, e.g., there are repeated
-measures for each participant or item and participants or items are
-only samples from the target population. It seems worthwhile to
-explain why it is important that our method accommodates both kinds of
-situations.
-
-\subsection{Importance of hierarchical structure}
-
-If data have hierarchical structure, as is typically the case when SDT
-models are used, and assuming that the SDT model is true, it is only
-when the requirement of accounting for variability due to subjects,
-items or other such factors is met that the estimates of the average
-(fixed) effects are guaranteed to be unbiased and the conclusions are
-guaranteed to generalize to the target population.
-
-A not so uncommon practice of analysing data aggregated over subjects,
-items or other sampled factors represents an extreme case of ignoring
+  by means of selective influence \cite{sternberg2001separate}}, which
+is often done by comparing observed and predicted ROC curves.
+
+When SDT models are used in psychology studies, researchers are
+usually interested in relationships between SDT parameters and
+additional measured or manipulated
+% ?? the dependence, bo ogólnie rozumiana kwestia zależności, stimulus
+% strength - ogólne zjawisko, a nie jakaś konkretna, albo pojedyncza,
+% ale bliżej nieokreślona
+variables; A good example is the dependence of $d'$ on stimulus
+strength. Also, in a great majority of psychology studies in which
+classification tasks are used the data have hierarchical structure,
+i.e., there are repeated measures for each participant or item and
+participants or items are only samples from the target population. It
+seems worthwhile to explain why a general-purpose method of inference
+for SDT models should accomodate both kinds of situations.
+
+% ?? Ważność hierarchical structure jako takiej
+\subsection{The importance of hierarchical regression structure}
+
+If data have hierarchical structure, but variability due to subjects,
+items, or other sampled factors is not accounted for, estimates of
+average (fixed) effects are not guaranteed to be unbiased and
+conclusions are not guaranteed to generalize to a target population. A
+not so uncommon practice of analysing data aggregated over sampled
+factors represents an extreme case of ignoring
+% ?? nieokreślona hier. structure, bo tu mówimy o externe case
+% ogólnego zjawiska ignorowania ogólnej kwestii hierarchiczności
 hierarchical data structure. Invalidity of this approach in the
-context of SDT was clearly illustrated with results of simulational
-studies by \citeA{morey2008problematic}, although strictly speaking it
-requires such demonstrations only in order to make a point about a
-specific case. Because SDT is a non-linear model, by definition
-expected values of estimates of its parameters based on data
-aggregated over sampled factors are not in general true population
-averages. Such estimates are biased and inference about the target
-population based on such estimates is simply not valid. To give a
-concrete example, concider two unbiased participants, one with
-$d' = 2$ and one with $d' = 4$. Their expected average accuracy is
-given by $(\Phi(1) + \Phi(2)) / 2 = .91$, which corresponds to
+% ?? Jeżeli illustrated w czasie przeszłym dokonanym i mówimy zaraz
+% czyje rezultaty, to chyba the results?
+context of SDT was clearly illustrated with the results of
+simulational studies by \citeA{morey2008problematic}, although,
+strictly speaking, it requires such demonstrations only in order to
+make a point about a specific case. Because SDT is a non-linear model,
+% ?? data jest unikalne dla tokenów estimates, więc the data?
+by definition, when estimates of its parameters are based on data
+aggregated over sampled factors the expected values of these estimates
+are not in general true population averages. Such estimates are
+% ?? zakładamy (w uproszczeniu), że target population jest ustalona
+% dla każdego estimate
+biased and inference about a target population based on such estimates
+is simply not valid. To give a concrete example, consider two unbiased
+participants, one with $d'_1 = 2$ and one with $d'_2 = 4$. Their
+expected average accuracy is given by
+$(\Phi(d'_1/2) + \Phi(d'_2/2)) / 2 = .91$, which corresponds to
 $d' = 2.68$, whereas their true average $d'$ is $3$.
@@
 separate estimates for every participant $\times$ item $\times$
-condition combination but uncertainty due to subject or item effects
-\emph{distributions} is not accounted for by means of hierarchical
-model structure. In this case the conclusions, at least with respect
-to the uncertainty in the estimates, are guaranteed to be valid only
-for the given sample, not the target population. This is rarely, if
-ever, the purpose of a study.
-
-\subsection{Importance of including regression structure}
-
-Another common source of problems is separation of estimation of
-non-linear model parameters from regression analysis relating these
-parameters to other measured or manipulated variables. When SDT
-parameters are estimated separately for each subject and condition and
-only later regressed on predictors of interest a number of additional
-issues may arise.
-
-Firstly, standard errors or credible intervals associated with
-regression coefficients do not reflect the uncertainty in SDT
+% ?? mamy już dane i estimates, to (zwykle nieznany) (poziom)
+% uncertainty jest określony
+condition combination, but the uncertainty due to subject or item
+effects \emph{distributions} is not accounted for by means of a
+hierarchical model
+% dalej jedziemy z nieokreślonymi, bo wyjaśniamy ogólne zasady:
+% wnioski mogą być jakiekolwiek, próbki i populacje są bliżej
+% nieokreślone
+structure. In such cases conclusions, at least with respect to
+% ?? może chodzić o różne population-average różnie estymowalne
+%
+% ?? population-average - or fixed - effects? chodzi mi o to, że
+% effects ma się też "łączyć" z population-average
+the uncertainties in estimates of population-average (fixed)
+% ?? sample i population jest dana dla tokenów
+effects, are guaranteed to be valid only for the given sample, not the
+target population.
+
+% mówimy o ogólnym co prawda, ale zarazem pojedynczym/unikalnym
+% problemie: the separation? Chętnie bym się pozbył przedimka
+Another common source of problems is the separation of estimation of
+non-linear model parameters from regression analysis whichs aim to
+relate these parameters to measured or manipulated variables. When the
+SDT parameters are estimated separately for each subject and
+condition, and only later regressed on predictors of interest, a
+number of additional issues may arise.
+
+% tutaj jakaś regression analysis już jest wcześniej wprowadzona
+Firstly, the standard errors or credible intervals associated with the
+regression coefficients do not reflect the uncertainty in the SDT
 parameter estimates, because the latter are treated as mere data
-points. This may lead to overly optimistic inference on regression
-coefficients, since what this amounts to, among other things, is
-ignoring part of the residual variance.
-
-Secondly, including the regression part in the same model makes the
-SDT parameter estimates dependent on the common regression structure
-and so also on each other. This can improve their quality and affect
-estimates of their uncertainty, just like assuming that random effects
-are samples from a common distribution may improve their estimates by
-making them dependent in the model.
-
-Thirdly, the precision of parameter estimates often varies between
-participants, items or conditions but when these estimates are treated
-as mere data points in regression analysis no use is made of this
-information. When regression structure is included in the same model
-the regression coefficients are less influenced by the parameter
-estimates that have lower precision.
-
-\section{A critical summary of avaiable methods of bayesian inference
-  for hierarchical SDT models}
-
-To the best of our knowledge only two attempts at creating something
-akin to the general purpose method of inference for hierarchical SDT
-models with ratings can be found in literature. We will not consider
-the ROC Toolbox \cite{Koen2016} or other similar tools here because as
-of this writing these tools cannot be used to fit hierarchical SDT
-models. The two available methods that can be used to fit hierarchical
-SDT models are the Gibbs sampler proposed by
-\citeA{morey2008problematic} and the Hierarchical Meta-d' model
-(HMeta-d') proposed by \citeA{hmetad}. In our summary of these two
-methods we mainly concentrate on two issues, namely the correctness of
-hierarchical structure implementation and possibility to model SDT
-parameters as dependent on additional variables.
-
-\citeA{morey2008problematic} proposed a Gibbs sampling algorithm for
-inference on hierarchical unequal variance normal SDT models with
-ratings. A slightly modified version of the algorithm was later made
-available as part of the \texttt{hbmem} R package. The authors adopted
-what they called a fixed-criteria parametrization. Under this
-parametrization the outermost criteria $\bm{c}_1$ and $\bm{c}_{K-1}$
-are fixed at $0$ and $1$ respectively, while the remaining criteria
-together with means and variances of the evidence distributions are
-free to vary. The hierarchical part of evidence distribution means is
-represented by additive independent normal effects of at most two
-sampled factors which, following the authors, we label as subject and
-item effects for convenience. However, the criteria are represented as
-subject specific \emph{fixed}, not random effects. Items can have an
-indirect and somewhat limited random effect on the criteria, because
-they can have an effect on evidence distribution means by shifting
-them independently. As the authors clarify, shifting both means by the
-same amount in the same direction does not change the $d'$ but
-simultaneously shifts all the criteria relative to the evidence
-distributions. This allows an item to have the same effect on all the
-criteria but it does not allow an item to have different effects on
-different criteria. Random effects correlations are not accounted
-for. Criteria are order-restricted, but the means of evidence
-distributions are order-restricted only in the version implemented in
-the \texttt{hbmem} package. SDT parameters can be regressed on
-additional predictors to some extent by specifying a model
-matrix. Lastly, because this is a Gibbs sampler, authors rely on
-conjugate priors, including the problematic \cite{gelman2004prior}
-inverse gamma prior placed on variances.
-
-Fleming's \citeyear{hmetad} HMeta-d' model is a hierarchical version
-of the meta-d' model \cite{maniscalco2012signal}, which in turn is a
-generalization of the SDT model with additional criteria that allows
-for a separate "meta-sensitivity" to account for possible
-discrepancies between binary stimulus classification (referred to as
-type 1 task) and a rating task (referred to as type 2 or
+points. Precision of parameter estimates often varies between
+participants, items, or conditions, but when the estimates are treated
+as data points no use is made of this information. Secondly,
+regressing parameters on numeric predictors makes their estimates
+dependent on the common regression structure, and so also on each
+other, which can improve the quality of the estimates, just as
+assuming that random effects are samples from a common distribution
+may improve their estimates.
+
+The aforementioned problems can be dealt with by supplementing the SDT
+model with the general hierarchical linear regression structure. For
+the reasons that we will now explain, this can only be achieved if the
+model is substantially reparametrised.
+
+% dopiero wprowadzamy ową unconstrained parameter space
+\section{Hierarchical Signal Detection Theory in a constrained
+  parameter space}
+
+Both $d'$ and $c$ have the virtue of being directly interpretable in
+terms of sensitivity and bias. However, both $d'$ and $c$ are
+constrained: $d'$ is non-negative and, when there is more than one
+criterion, the elements of the $\bm{c}$ vector are order-restricted
+($\bm{c}_{i+1} > \bm{c}_i$). Because normal distribution is unbounded,
+hierarchical linear regression structure, in which random effects are
+normally distributed and fixed effects are represented by
+unconstrained parameters, can only be defined on unconstrained
+parameters. We provide examples of the problems that may otherwise
+arise in the following summary of hierarchical SDT implementations.
+
+We are aware of only two attempts at creating a general purpose method
+of inference for hierarchical SDT models with ratings. One is the
+Gibbs sampler proposed by \citeA{morey2008problematic} and the other
+is the Hierarchical Meta-d' model (HMeta-d') proposed by
+\citeA{hmetad}.
+% We do not consider the ROC Toolbox \cite{Koen2016} or other similar
+% tools here because, as of this writing, these tools cannot be used to
+% fit hierarchical SDT models. The two available methods that can be
+% used to fit hierarchical SDT models are
+The HMeta-d' model is a hierarchical version of the meta-d' model
+\cite{maniscalco2012signal}, which in turn is a generalization of the
+SDT model that allows for a separate "meta-sensitivity" to account for
+possible discrepancies
+% referred to as ... bez przyimka, prawda?
+between a binary stimulus classification (referred to as type 1 task)
+% ?? wcześniej było a binary i a rating, jakby to były (potencjalnie)
+% niezwiązane zadania
+and the associated rating task (referred to as type 2 or
 meta-cognitive task). We consider HMeta-d' here because it reduces to
-the SDT model with ratings when type 1 and type 2 sensitivities are
-equal. In HMeta-d' evidence distributions are normal with equal
-variance. Normally distributed random effects of one factor are
-allowed for sensitivities and parameters indirectly representing the
-criteria, for lack of a better name from now on referred to as raw
-criteria. The actual criteria are subject-specific sorted values of
-unconstrained raw criteria and it is these unconstrained parameters,
-not the actual criteria, that are given independent hierarchical
-normal distributions representing variability due to subjects. Since
-sorting is not injective the space of raw criteria is not isomorphic
-to the space of actual criteria and each element of the raw criteria
-vector can represent a value of any criterion, depending on the values
-of other elements of the raw criteria vector. The model does not
-account for possible correlations between random
-effects. Sensitivities are represented as unrestricted scalars even
-though they are non-negative by definition. Out of the box the code
-provided by the author allows only for a very limited regression
-structure to be placed on SDT parameters. Software implementation
-relies on ready-made model specifications expressed in JAGS language
-that are aimed at several simple types of experimental design (e.g.,
-single subject or single group of subjects). The models can be fitted
-using convenient MATLAB functions.
-
-\section{Hierarchical Signal Detection Theory in unconstrained
+the SDT model with
+% ?? powtórzenie the
+ratings when the type 1 and the type 2 sensitivities are equal.
+
+The Gibbs sampler created by \citeA{morey2008problematic} allows for
+at most two sampled factors to have independent normally distributed
+random effects on the evidence distribution means. Unlike $d'$, each
+evidence distribution mean considered in isolation is an unconstrained
+parameter, but the mean of the second one is by definition greater
+($d' > 0$), or equal ($d' = 0$), than the mean of the first one. The
+authors explicitly admit that their algorithm does not enforce this
+restriction, because, as they claim in the paper, it would greatly
+complicate analysis. At the same time the authors fail to mention that
+an immediate consequence of allowing for negative $d'$ values is that
+the resulting posterior distribution looses its intended
+interpretation, since it can have a non-zero mass when $d' < 0$. The
+outermost criteria are fixed at $0$ and $1$, and the oredering
+restriction is enforced by assuming that the likelihood is $0$
+whenever $\bm{c}_{i+1} \leq \bm{c}_i$. As the authors explain, because
+a sampled factor can have independent random effects on the evidence
+distribution means, it can have an effect on all the criteria:
+shifting both means by the same amount in the same direction is
+equivalent to keeping the sensitivity intact, while shifting the
+criteria relative to the evidence distributions. However, the elements
+of the criteria vector cannot be affected differently by the same
+sampled factor, which is an unrealistic restriction.
+
+In HMeta-d' the hierarchical structure is restricted to normally
+distributed random intercepts of one sampled factor. Perhaps the most
+problematic aspect of the HMeta-d' model, apart from the fact that
+also in this model the $d'$ parameter is allowed to assume negative
+values, is the representation of the criteria. Each element of the
+criteria vector has an associated independent normal distribution,
+which does allow for criteria random effects, but does not enforce the
+necessary ordering restriction. The elements of this vector are sorted
+to obtain another criteria vector, and it is this sorted criteria
+vector that is used to model the conditional combined response
+distributions. Consequently, the model does contain parameters
+representing the actual, order-restricted criteria, but, because
+sorting is not injective, the
+% ?? (i.e., not isomorphic to)
+space of the actual criteria is only loosely related (i.e., not
+isomorphic) to the space of the unrestricted criteria vectors that are
+associated with a hierarchical structure.
+
+% Hierarchical structure does not have to be linear, but when it is, the
+% parameters associated with unconstrained fixed effects and normally
+% distributed random effects of sampled factors cannot be non-negative
+% or order-restricted. We propose to substantially reparametrise the SDT
+% model.
+
+\section{Hierarchical Signal Detection Theory in an unconstrained
   parameter space}
 
-As we have already stressed, in a typical study both sensitivity and
-criteria parameters have to be supplemented with hierarchical
-structure for inference to be guaranteed to generalize to the target
-population. In our critical summary of the two available methods we
-have listed some reasons why we believe that without substantial
-modifications neither method correctly addresses this issue. The main
-reason why both methods fail in correctly specifying the hierarchical
-structure on SDT parameters is that random effects are represented by
-normal distributions but parameters are constrained, i.e.,
-sensitivities are non-negative and criteria are ordered.
-
-In the \texttt{bhsdtr} package sensitivities and criteria are derived
-from unconstrained parameters by convenient isomorphisms. This is
-easily achieved in case of sensitivity by deriving $d'$ from
-$\delta = \ln(d')$ and modelling random effects on $d'$, if any, by
-assuming that $\delta$ is normally distributed. Difficulties in
-representing the hierarchical structure of the criteria stem from the
-fact that even though each criterion considered in isolation is an
-unconstrained parameter, the vector of all the criteria is ordered.
-We propose to map the $R^{K-1}$ space, representing unconstrained
-criteria, to the $K$ dimensional probability simplex space using the
-softmax function and to map the simplex space to the ordered criteria
-space by means of the inverse normal CDF. Specifically, assume that a
-normal distribution is centered on the midpoint between the two
-evidence distributions. The value of criterion $i$ is then given by:
+The general hierarchical linear regression structure can be defined on
+SDT parameters only if the latter are derived from unconstrained
+parameters. In the \texttt{bhsdtr} package $d'$ is derived from
+$\delta = \ln(d')$, this way random effects on $d'$ can be modelled by
+assuming that $\delta$ is normally distributed. The problem of
+representing the criteria by unconstrained parameters
+% ?? skoro the space of, to potem the... ? w końcu chodzi nam o
+% *wszystkie*, takie, że...?
+is solved by mapping the $R^{K-1}$ space of unconstrained criteria
+vectors to the $K$ dimensional probability simplex space using the
+softmax function, and mapping the simplex space to the space of
+order-restricted criteria vectors by means of the inverse normal CDF:
 
 \begin{align}
-  \bm{c}_i &= \Phi^{-1}(\sum_{k = 1}^i(e^{\bm{\gamma}_k}) / \sum_{j=1}^K(e^{\bm{\gamma}_j}))
+  \bm{c}_i &= \Phi^{-1}(\sum_{k = 1}^i(e^{\bm{\gamma}_k}) /
+             \sum_{j=1}^K(e^{\bm{\gamma}_j}))
 \label{eq:1}
@@
 
-\noindent where $\bm{\gamma} \in R^K$ with $\bm{\gamma}_K$ fixed at
-$0$ for identifiability. The idea is illustrated in
-Fig.~\ref{fig:2} below:
+\noindent where $\Phi$ is the CDF of the standard normal distrubution
+and $\bm{\gamma} \in R^K$, with $\bm{\gamma}_K$ fixed at $0$ for
+identifiability. The idea is illustrated in Fig.~\ref{fig:2} below:
 
@@
   \includegraphics[width=.8\linewidth]{plot.pdf}
-  \caption{Mapping between unconstrained $\gamma$ vector and criteria}
+  \caption{Mapping between the unconstrained $\bm{\gamma}$ vector and
+    the criteria}
   \label{fig:2}
@@
 
-The mapping in Eq.~\ref{eq:1} is an isomorphism between the $R^{K-1}$
-space and the space of $K-1$ ordered criteria vectors. Elements
-$i = 1...K$ of the $\bm{\gamma}$ vector represent relative distances
-between adjacent criteria because their exponents represent relative
-magnitudes of areas under the normal curve delineated by the pairs of
-adjacent criteria:
+Note that the normal distribution centered at the midpoint is merely a
+mapping device, not a third evidence distribution, and that, for the
+reasons that will soon become clear, it is wider than the two evidence
+distributions. The mapping expressed by Eq.~\ref{eq:1} is an
+isomorphism between the $R^{K-1}$ space and the space of
+order-restricted criteria vectors. It's inverse is given by
+$\gamma_i =\ln{(\int_{c_{i-1}}^{c_i} f(s) \,ds /
+  \int_{c_{K-1}}^{\infty} f(s) \, ds)}$, where $f$ is the standard
+normal probability density function. The elements of the $\bm{\gamma}$
+% ?? jeżeli correspond to <nieokreślonych> to represent też
+% <nieokreślonych>?
+vector correspond to relative distances between pairs of adjacent
+criteria, because their exponents represent the relative
+% cały czas mówimy ogólnie i dopiero teraz wprowadzamy pojęcie pair of
+% adjacent criteria, więc rodzajnik nieokreślony?
+magnitudes of areas under the standard normal curve, delineated by the
+pairs of adjacent criteria:
 $e^{\bm{\gamma}_i} / e^{\bm{\gamma}_j} = (\Phi(\bm{c}_i) -
 \Phi(\bm{c}_{i-1})) / (\Phi(\bm{c}_j) - \Phi(\bm{c}_{j-1}))$. When
-$K=2$ only $\bm{\gamma}_1$ is free to vary and its value directly
+$K=2$, only $\bm{\gamma}_1$ is free to vary, and its value directly
 represents the direction and magnitude of bias: $\bm{\gamma}_1$ is $0$
-when the criterion is placed at the midpoint between the means of
-evidence distributions, and the more negative (positive) it is the
-more the criterion is placed to the left (right) of midpoint. Our
-experience indicates that it is often a good idea to multiply all the
-criteria by a value greater than $1$ which is equivalent to making the
-mapping distribution in Eq.~\ref{eq:1} wider. This tends to even out
-the values of $\bm{\gamma}$, especially when criteria are widely
-spread as can happen for moderate to large $d'$ values. This feature
-is implemented in the \texttt{bhsdtr} package by introducing a scaling
-factor.
-
-Once the $d'$ and $c$ are represented as derived from unconstrained
-$\delta$ and $\gamma$ parameters the SDT model can be supplemented
-with hierarchical linear regression structure. To avoid having to deal
-with even more complicated index notation, below we have presented
-only the simple case of one sampled factor.
+when the criterion is placed at the midpoint between the evidence
+distributions and the more negative (positive) $\bm{\gamma}_1$ is, the
+more the criterion is shifted to the left (right) of the midpoint. It
+is often a good idea to multiply all the criteria by a value greater
+than $1$, which is equivalent to making the mapping distribution
+wider. This
+% ?? skoro już mówimy, że coś zwykle wpływa na wartości gamma, to nie
+% the wartości?
+tends to even out values of $\bm{\gamma}$, by preventing the outermost
+areas under the mapping distribution curve from becoming very small,
+especially when the criteria are widely spread, as can happen for
+moderate to large $d'$ values. This feature is implemented in the
+\texttt{bhsdtr} package by introducing a scaling factor.
+
+Once $d'$ and $c$ are derived from the unconstrained $\delta$ and
+$\gamma$ parameters, the SDT model can be supplemented with a
+hierarchical linear regression structure. To avoid having to deal with
+an even more complicated index notation\footnote{The reader familiar
+  with hierarchical models may be surprised by our use of superscrip
+  parenthesised greek letters to express hierarchical
+  relationships. We chose this convention because it allowed us to use
+  subscripts to denote elements of vectors and matrices while
+  minimizing the number of nested sub- or superscripts.}, below we
+have presented only the simple case of one sampled factor.
 
@@
   \bm{d'}_i &= e^{\bm{\delta}_i} \\
-  \bm{\gamma}_{i,.} &= \bm{X}^{(\gamma)}_{i,.} \bm{\beta}^{(\gamma)} + \bm{Z}^{(\gamma)}_{i,.}
+  \bm{\gamma}_{i,\cdot} &= \bm{X}^{(\gamma)}_{i,\cdot} \bm{\beta}^{(\gamma)} + \bm{Z}^{(\gamma)}_{i,\cdot}
                       \bm{\theta}^{(\gamma)} \\
-  \bm{c}_{i,k} &= s \ \Phi^{-1}(\sum_{l = 1}^i(e^{\bm{\gamma}_{i,l}}) /
+  \bm{c}_{i,k} &= s \ \Phi^{-1}(\sum_{l = 1}^k(e^{\bm{\gamma}_{i,l}}) /
                  \sum_{m=1}^K(e^{\bm{\gamma}_{i,m}})) \\
@@
 
-\noindent Here $i$ is the observation number, $\bm{X}$ is the fixed
-effects model matrix for the respective parameter, $\bm{Z}$ is the
-random effects model matrix, $\bm{\beta}$ and $\bm{\theta}$ are the
-fixed and random effects, $\bm{c}_{i,k}$ is the $k$-th criterion for
-the $i$-th observation, $s$ is the criteria scaling factor, and
-$\bm{y}$ is the combined response. Note that $\bm{d'}_i$ is a scalar,
-but $\bm{\gamma}_{i,.}$ is in general a vector and so
+\noindent Here $i = 1\dots N$ is the observation number, $\bm{X}$ is
+the fixed effects model matrix for the respective parameter, $\bm{Z}$
+is the random effects model matrix, $\bm{\beta}$ and $\bm{\theta}$ are
+the
+% opisać raczej jako macierz c, itd. - jest macierzą != jest efektem
+% losowym
+fixed and random effects, $\bm{c}$ is a $N \times K-1$ matrix, $s$ is
+the criteria scaling factor, and $\bm{y}$ is the combined
+response. Note that $\bm{d'}_i$ is a scalar, but
+$\bm{\gamma}_{i,\cdot}$ is in general a vector, and so
 $\bm{\beta}^{(\gamma)}$ and $\bm{\theta}^{(\gamma)}$ are matrices. The
 $j$-th rows of the $\bm{\beta}^{(\gamma)}$ and
-$\bm{\theta}^{(\gamma)}$ matrices represent fixed and random effects
-for the $j$-th element of the $\bm{\gamma}$ vector.
-
+$\bm{\theta}^{(\gamma)}$ matrices
+% ?? effects on ... ?
+represent fixed and random effects on the $j$-th element of the
+$\bm{\gamma}$ vector.
+
+% ?? tutaj coś dziwnego dzieje się z liczbą: ta dekompozycja tych
+% macierzy, i ta (jedna) dekompozycja poprawia próbkowanie itd.
 Following \citeA{sorensen2015bayesian} we make use of the Cholesky
-decomposition of the correlation matrices because it improves
-efficiency and admits a convenient prior on random effects'
+decomposition of the correlation matrices, because it improves
+efficiency and admits a convenient prior on random effects
 correlations:
@@
 
-\noindent where $\bm{\tau}$ is a vector of standard deviations of
-random effects and $\bm{L}$ is the Cholesky decomposition of the
-correlation matrix, i.e., $\bm{C} = \bm{L} \bm{L}'$. This way
-$\bm{\theta}$ is multivariate normal with covariance matrix
-$\text{diag}(\bm{\tau}) \bm{L}$.
-
-We use weakly informative proper priors because they provide
-regularization and help stabilize computation. Fixed effects
-($\bm{\beta}^{(\delta)}$ and $\bm{\beta}^{(\gamma)}$) are given
-independent normal priors, random effects standard deviations
-($\bm{\tau}^{(\delta)}$ and $\bm{\tau}^{(\gamma)}$) are given
+% ?? problem z przedimkami: tau odpowiada obydwu tau, żeby było
+% krócej, ale przez to jest "a" wektorem sd efektów losowych, chociaż
+% niezupełnie dowolnym
+\noindent where each $\bm{\tau}$ is a vector of standard deviations of
+random effects and each $\bm{L}$ is a Cholesky decomposition of a
+random effects correlation matrix, i.e., $\bm{C} = \bm{L}
+\bm{L}'$. This way $\bm{\theta}$ is multivariate normal with
+covariance matrix $\text{diag}(\bm{\tau}) \bm{L}$.
+
+Finally, we use weakly informative proper priors because they provide
+regularization and help stabilize computation. The fixed effects
+$\bm{\beta}^{(\delta)}$ and $\bm{\beta}^{(\gamma)}$ are given
+independent normal priors, the random effects standard deviations
+$\bm{\tau}^{(\delta)}$ and $\bm{\tau}^{(\gamma)}$ are given
 independent half-Cauchy priors as recommended by
-\citeA{gelman2004prior} and each $\bm{L}$ is given an independent lkj
+\citeA{gelman2004prior}, and each $\bm{L}$ is given an independent lkj
 prior:
@@
 
-\subsection{Specifying prior distributions}
-
-Specifying the priors on sensitivity effects does not pose any special
-difficulties. Sensitivity of an unbiased classifier given percent
-correct is given by $2 \Phi^{-1}(\text{pc})$. When both stimulus
-classes are equally likely the greater the bias the lower the accuracy
-meaning that unbiased sensitivity represents the lower bound on
-sensitivity given percent correct. Let us assume that a great majority
-of subjects are expected to achieve percent correct within the $.51$
-to $.99$ range with negligible bias. Since
-$\ln(2\Phi^{-1}(.51)) = -2.99$ and $\ln(2 \Phi^{-1}(.99)) = 1.54$ a
-reasonable weakly informative prior on delta is normal with mean
-$(1.54 - 2.99) / 2$ and standard deviation $(1.54 + 2.99) / 2$ which
-is the default prior on delta effects in the \texttt{bhsdtr}
-package.
-
-Specifying priors on the criteria can be challenging because of the
-ordering restriction. In addition, complexity of mapping in
-Eq.~\ref{eq:1} makes it difficult to reason about the priors on
-$\bm{\gamma}$ in terms of the criteria. By default in the
-\texttt{bhsdtr} package each entry in $\bm{\sigma}^{(\gamma)}$ and
-$\bm{\zeta}^{(\gamma)}$ matrices is set to $\ln(100)$ and the criteria
-scaling factor is fixed at 2. The meaning of $\ln(100)$ depends on
-model matrix parametrization to some degree, but usually it means that
-a priori areas under the mapping distribution between adjacent
-criteria may easily differ from the area between criterion $K-1$ and
-$K$ by a factor of 100 but differences by a factor of $2 \ln(100)$ or
-more are highly unlikely. Off course, when in doubt it is a good idea
-to perform some sort of sensitivity analysis, remembering that strong
-sensitivity to priors indicates that the study may be underpowered.
+% ?? the? Tytuły rządzą się specjalnymi prawami, np. to nie są zdania,
+% ani nawet clauses chyba
+\subsection{Specifying the prior distributions}
+
+% !! TUTAJ skończyłem wstawiać intuicyjne przecinki
+The formal definition of a bayesian model is not complete without
+providing fixed values of all the parameters that define prior
+distributions. Specifying the priors on sensitivity effects does not
+pose any special difficulties. Sensitivity of an unbiased classifier
+given percent correct is given by $2 \Phi^{-1}(\text{pc})$. When
+$p(stim = 1) = 0.5$ the greater the bias the lower the accuracy
+meaning that an unbiased sensitivity is a lower bound on sensitivity
+given percent correct. Let us assume that the majority of subjects are
+expected to achieve
+% ?? with negligible bias, bez niczego?
+percent correct within the $.51$ to $.99$ range with negligible
+bias. Since $\ln(2\Phi^{-1}(.51)) = -2.99$ and
+$\ln(2 \Phi^{-1}(.99)) = 1.54$, a reasonable weakly informative prior
+on $\delta$ is normal with mean $(1.54 - 2.99) / 2$ and standard
+deviation $(1.54 + 2.99) / 2$, which is the default prior on delta
+effects in the \texttt{bhsdtr} package.
+
+Specifying the priors on criteria effects can be challenging because
+the criteria are order-restricted and because the complexity of the
+mapping expressed by Eq.~\ref{eq:1} makes it difficult to reason about
+priors on $\bm{\gamma}$ in terms of criteria effects. By default in
+the \texttt{bhsdtr} package each entry in the $\bm{\sigma}^{(\gamma)}$
+and $\bm{\zeta}^{(\gamma)}$ matrices is set to $\ln(100)$ and the
+criteria scaling factor is fixed at 2.
+% The interpretation of $\ln(100)$ depends on the model matrix, but
+% usually it means that a priori the areas under the mapping
+% distribution curve between adjacent criteria may easily differ from
+% the area between the criteria $K-1$ and $K$ by a factor of 100 but
+% differences by a factor of $2 \ln(100)$ or highly unlikely.
+
+% When in doubt, it is a good idea to perform some sort of sensitivity
+% analysis, remembering that strong sensitivity to priors indicates
+% that the study may be underpowered.
 
@@
 
-By default $\nu^{(\delta)} = \nu^{(\gamma)} = 1$ implying a uniform
-prior on random effects correlation matrices. Because the greater the
-value of $\nu$ the more emphasis is put on zero off-diagonal
-correlations the researcher can force the correlations to be near-zero
-by choosing large $\nu$ value.
+Finally, by default $\nu^{(\delta)} = \nu^{(\gamma)} = 1$, which
+implies a uniform prior on random effects correlation
+matrices. Because the greater the value of $\nu$ the more emphasis is
+put on zero
+% ?? the researcher - the one specifying the priors
+off-diagonal correlations, the researcher can force the correlations
+to be near-zero by choosing a large $\nu$ value.
 
@@
 
-We have chosen the Stan modelling language \cite{carpenter2016stan}
-because it uses state-of-the-art adaptive Hamiltonian Monte Carlo
-sampling algorithm which often handles high-dimensional correlated
-posteriors better than the Gibbs sampler. The \texttt{bhsdtr} package
-is essentially a collection of documented functions. The
-\texttt{aggregate\_responses} function serves to aggregate the data as
-much as possible for efficiency, but without distorting the
+% ?? tutaj wprowadzamy zaraz the, tak jakbyśmy mówili o sekwencji
+% czynności objaśniając funkcje pakietu.
+The \texttt{bhsdtr} package implements the model described in the
+previous section in the Stan modelling language
+\cite{carpenter2016stan} because it uses a state-of-the-art adaptive
+Hamiltonian Monte Carlo sampling algorithm which often handles
+high-dimensional correlated posteriors better than a Gibbs
+sampler. Our package is essentially a collection of documented
+functions. The \texttt{aggregate\_responses} function aggregates data
+as much as possible for efficiency, but without distorting the
 hierarchical structure. The \texttt{make\_stan\_model} function
-creates the model code based on the provided model formulae for random
-effects. The Stan code produced by the \texttt{make\_stan\_model}
-function can be fitted as is or modified by the user if needed, e.g.,
-to change the prior distributions or drop the equal variance
-assumption. The \texttt{make\_stan\_data} function creates regression
-model matrices and other data structures required by the model created
-with the \texttt{make\_stan\_model} function. Finally, the
-\texttt{plot\_sdt\_fit} function can be used to assess the model fit
-visually. This function creates publication-ready ROC curve plots or
-response distribution plots with posterior predictive intervals
-calculated for the chosen $\alpha$ level.
+creates a model definition in the Stan language. The Stan code
+produced by the \texttt{make\_stan\_model} function can be fitted as
+is or modified by
+% ?? the user - of this function
+the user if needed, e.g., to change the prior distributions or to drop
+the equal variance assumption. The \texttt{make\_stan\_data} function
+creates regression model matrices and other data structures required
+% ?? już mówiliśmy o tym modelu (kodzie modelu)
+by the model created using the \texttt{make\_stan\_model}
+function. Finally, the \texttt{plot\_sdt\_fit} function can be used to
+visually assess the fit of the model by creating publication-ready ROC
+curve plots or response distribution plots with posterior predictive
+intervals calculated for the chosen $\alpha$ level.
 
@@
 To make full use of the \texttt{bhsdtr} package functionality three
-non-standard R packages are required, namely \texttt{rstan},
-\texttt{plyr} and \texttt{ggplot2}. We recommend using the
+non-standard R packages are required, namely
+\texttt{rstan}, %?? , czy średnik - namely
+\texttt{plyr}, and \texttt{ggplot2}. We recommend using the
 \texttt{devtools} package to install the \texttt{bhsdtr} package
-directly from the github repository. This will also automatically
-install any missing required packages:
+directly from the github repository. This will automatically install
+any missing required packages:
 
@@
 
-We tested our method on the \texttt{gabor} dataset that is part of the
-\texttt{bhsdtr} package. The \texttt{gabor} dataset comes from an
-unpublished study in which on each trial participants had to classify
-the briefly presented gabor patch as either oriented to the left or to
+We will now explain how to perform some of the essential steps of a
+typical data analysis process, which at the very least will usually
+involve preparing the data, creating the model code, fitting the
+model, assessing the fit, and possibly converting the unconstrained
+$\gamma$ and $\delta$ parameters to $d'$ and $c$.
+
+\subsubsection{Preparing the data}
+
+The \texttt{bhsdtr} package contains the \texttt{gabor} dataset which
+comes from an unpublished study. In this study on each trial the
+participants had to
+% ?? Gabor patch presented on that trial - the?
+classify a briefly presented Gabor patch as tilted to the left or to
 the right using the arrow keys. The participants were also asked to
@@
 \cite{ramsoy2004introspection} presented at the bottom of the
-screen. The gabor patch was immediately followed by a mask. The PAS
-ratings ranged from "no experience" to "absolutely clear image" and
-were provided either before (RD condition) or after (DR condition)
-pressing the arrow keys. On each trial the gabor patch was equally
-likely to be presented for 32ms or 64ms. Decision-rating order was a
-between-subject variable and gabor patch presentation duration was a
-within-subject variable. There were 47 participants and 48 trials per
-condition.
-
-We will now explain how to perform some of the essential steps of a
-typical data analysis process, which at the very least usually
-involves preparing the data, creating the model code, fitting the
-model, assessing the fit, and possibly converting the unconstrained
-$\gamma$ and $\delta$ parameters to $d'$ and $c$.
-
-\subsubsection{Step 1: preparing the data}
-
-In the study in question the response was originally encoded as
-accuracy and rating, so the first step was to create an appropriate
-response variable using the \texttt{combined\_response} function. This
-function needs three variables, one encoding the stimulus class, one
-encoding the rating (as an integer) and one encoding decision
+screen. The Gabor patch was immediately followed by a mask. The PAS
+ratings ranged from "no experience" to "absolutely clear image"
+and % przecinek?
+were provided either before (RD order condition) or after (DR order
+condition) % , or?
+pressing the arrow keys. On each trial the Gabor patch was equally
+% ?? nazwy zmiennych bez przedimka?
+likely to be presented for 32 ms or 64 ms. Order was a between-subject
+variable and duration was a within-subject variable. There were 47
+participants and 48 trials per condition.
+
+In the study in question the response was originally encoded using
+separate variables for accuracy and rating, so the first step was to
+create an appropriate response variable using the
+\texttt{combined\_response} function. This function requires three
+variables, one encoding the stimulus class, one encoding the rating
+(as an integer), and one binary variable encoding the decision
 accuracy.
@@
 
-This step is required only if ratings are available and the combined
+This step is required only if the ratings are available and a combined
 response variable is not already present in the data. In the single
 criterion case the combined response variable is simply the binary
-classification decision. To fit the single-criterion SDT model to this
+classification decision. To fit a single-criterion SDT model to this
 dataset the code above would have to be replaced with the following:
@@
 \begin{lstlisting}{language=R}
-  gabor$resp = combined_response(gabor$stim, accuracy = gabor$acc)
+  gabor$resp = combined_response(gabor$stim, 
+                                 accuracy = gabor$acc)
 \end{lstlisting}
 
-Next, the data was aggregated by computing combined response counts,
-but only to the extent that preserved all the participant effects. The
-\texttt{aggregate\_responses} function that does exactly this requires
-as arguments a data frame containing all relevant variables, a name of
-the stimulus class variable, a name of the combined response variable
-and a vector of names of all variables that are to be preserved in a
-resulting aggregated dataset, i.e., those encoding sampled factors and
-those representing independent variables used in the regression part
-of the model:
+% tutaj czuję, że the trzeba prawie wszędzie
+Next, the data has to be aggregated using the
+\texttt{aggregate\_responses} function, but only to the extent that
+preserves all the random effects. This function requires as arguments
+a data frame containing
+% ?? all the relevant variables?
+all the relevant variables, a name of the stimulus class variable, a
+name of the combined response variable and a vector of names of all
+the variables that are to be preserved in the resulting aggregated
+dataset (apart from the stimulus class variable and the combined
+response variable), i.e., those encoding the sampled factors and those
+representing the independent variables used in the regression part of
+the model:
 
@@
 The main purpose of the aggregation step is to improve the efficiency
-of sampling from the posterior distribution. When the data are
-aggregated in this way the likelihood for each condition $\times$
-participant combination has to be computed only once rather than as
-many times as there are trials per condition per participant. Note
-that if there are other sampled factors present in the data (e.g.,
-items, replications, etc.) then these factors have to be specified at
-this stage as well to preserve hierarchical structure. The
+of sampling from the posterior distribution. When data are aggregated
+in this way the likelihood for each condition $\times$ participant
+combination has to be computed only once rather than as many times as
+there are trials per condition per participant. Note that if there are
+% ?? czuję, że the data, bo mówimy o (jakimś), ale nadal tym samym
+% hipotetycznym przypadku
+other sampled factors present in the data (e.g., items, replications,
+etc.) then these factors have to be specified at this stage as well to
+preserve the hierarchical data structure. The
 \texttt{aggregate\_responses} function creates a list with three
 components. The \texttt{data} component is a data frame containing
+% ?? additional zawsze bez przedimka?
 additional preserved variables, the \texttt{stimulus} component is the
-stimulus class variable and the \texttt{counts} component is a matrix
-of counts with $K$ columns, where $K$ is the highest possible combined
-response.
-
-\subsubsection{Step 2: creating the model code}
-
-The model is fitted using the \texttt{stan} function from the
-\texttt{rstan} package which needs a special list of data structures
-used by the model as well as a model specification expressed in the
-Stan language. Every SDT model has some fixed effects structure, since
-even when there are no predictors the model parameters can be
-expressed as regressed on a vector of ones (i.e., an
+stimulus class variable and the \texttt{counts} component is a
+$N\times K$ matrix of combined response counts, where $N$ is the
+number of data points and $K$ is the number of possible combined
+response values.
+
+\subsubsection{Creating the model code}
+
+A model is fitted using the \texttt{stan} function from the
+\texttt{rstan} package. This function requires a special list of data
+structures used by the model as well as a model specification
+expressed in the Stan language. Every model has some fixed effects
+structure, since even when there are no predictors the model
+parameters can be expressed as regressed on a vector of ones (i.e., an
 intercept). However, many models also have a hierarchical structure
@@
 specified when using the \texttt{make\_stan\_model} function. This is
+% ?? This is done by nie jest zbyt kolokwialne?
 done by providing a list of lists of R model formulae. Each list of
@@
 vary between the levels of this sampled factor. When
+% ?? a model, bo jest niedookreślony bez dodatkowej informacji z
+% make_stan_data?
 \texttt{make\_stan\_model} is used without any arguments it specifies
-the SDT model without any random effects. Fixed effects model matrices
-are specified by providing a list with at least two model formulae,
-named \texttt{delta} and \texttt{gamma}, to the
-\texttt{make\_stan\_data} function, described later in this
-paper. Non-default priors can be specified by adding optional elements
-to the random and fixed effects specification lists, as described in
-the \texttt{make\_stan\_data} function documentation.
-
-In the study in question there was only one sampled factor (id) which
-represents participants. Because duration was a within-participant
-variable in principle its effect could vary between participants for
-all the SDT parameters. However, preliminary data analysis indicated
-that the barely noticeable 32ms difference in duration seemed to
-affect only the sensitivity. Thus, it was assumed that duration
-affects only the $\delta$ parameter and that this effect may vary
-between participants:
+% ?? <bez przedimka> Fixed effects model matrices nieokreślone -
+% czuję, że jestem niekonsekwentny w tych opisach
+a model without any random effects. Fixed effects model matrices are
+specified by providing a list with at least two model formulae, named
+\texttt{delta} and \texttt{gamma}, to the \texttt{make\_stan\_data}
+function, described later in this paper. Non-default priors can be
+% ?? czuję, że the random and fixed effects specification lists
+specified by adding optional elements to the random and fixed effects
+specification lists, as described in the \texttt{make\_stan\_data}
+function documentation.
+
+% ?? the participants?
+In the study in question there was only one sampled factor, i.e., the
+participants. Because duration was a within-participant variable in
+principle its effect could vary between the participants for all the
+SDT parameters. However, a preliminary data analysis indicated that
+the barely noticeable 32 ms difference in duration seemed to affect
+% ?? teraz nagle the sensitivity? jako parametr SDT to jest chyba
+% jednak the
+only the sensitivity. Thus, it was assumed that $\delta$ may depend on
+duration and order, but $\gamma$ may only be affected by
+order. Because duration was a within-subjects variable its effect on
+$\delta$ was assumed to vary between the participants, but the only
+random effect associated with $\gamma$ was the intercept:
 
@@
 The \texttt{make\_stan\_data} function creates fixed and random effect
+% ?? dummy contrast coding jest niepoliczalne i nie chodzi nam o żadne
+% pojedyncze, tylko o ogólnie rozumiany sposób kodowania
 model matrices based on the respective model formulae using dummy
@@
 the $\delta$ model matrix. This way $\delta$ was estimated for every
-duration $\times$ order condition\footnote{R uses dummy contrast
-  coding by default}. The resulting separate intercepts and slopes
-parametrization makes it easier to calculate arbitrary contrasts based
-on the posterior samples. A more standard parametrization was used for
-the $\gamma$ parameter, because it was initially assumed that criteria
-depend only on order and so there was only one contrast of interest
-for every element of the $\gamma$ vector. On the other hand even in
-such case the separate intercepts and slopes parametrization may be
-more convenient if the researcher is interested in the actual
-criteria. It is perhaps worth emphasising how the separation of
-$\delta$ and $\gamma$ regression structures makes it possible to test
-a broad class of linear models representing the dependence of
-sensitivity and criteria on additional variables.
-
-\subsubsection{Step 3: fitting the model}
-
-A separate data structure used by the Stan sampler has to be created
-using the \texttt{make\_stan\_data} function. The obligatory arguments
-to this function are the aggregated data object created by
-\texttt{aggregate\_responses} function and fixed effects
-specification. Because in our example study duration was a
-within-subject variable here we also provide the random effects
+duration $\times$ order condition. The resulting separate intercepts
+and slopes parametrization makes it easier to calculate arbitrary
+% ?? contrasts ON posterior samples?
+contrasts on posterior samples. A more standard parametrization was
+used for the $\gamma$ parameter, because it was initially assumed that
+the criteria depend only on order, and so there was only one contrast
+of interest for every element of the $\gamma$ vector. On the other
+% ?? the separate ?
+hand, even in such cases the separate intercepts and slopes
+parametrization may be more convenient if a researcher is interested
+in the actual criteria, as we will later explain when introducing the
+\texttt{gamma\_to\_crit} function. This example also illustrates how
+the separation of the $\delta$ and $\gamma$ regression structures
+makes it possible to test a broad class of linear models representing
+% ?? the dependence bo zależność we wszystkich jej przejawach?
+the dependence of the SDT parameters on additional variables.
+
+\subsubsection{Fitting the model}
+
+In order to fit the model a separate data structure used by the Stan
+sampler has to be created using the \texttt{make\_stan\_data}
+function. The obligatory arguments to this function are an aggregated
+data object created by the \texttt{aggregate\_responses} function and
+a fixed effects specification. Because in our example study duration
+was a within-subject variable here we also provide a random effects
 specification. Importantly, if random effects are modelled the same
@@
 
+% ?? tutaj wydaje mi się oczywiste, że <bliżej nieokreślonych>
+% parameters of interest, a więc relevant variables wcześniej też
+% powinno być <nieokreślonych>?
 Finally, a vector of names of parameters of interest has to be
@@
 Note that since more than one sampled factor is allowed the names of
-parameters defining the hierarchical part are indexed (e.g.,
-\texttt{delta\_sd\_1}, \texttt{delta\_random\_1},
-\texttt{Corr\_delta\_1}). The name \texttt{counts\_new} refers to the
-posterior predictive samples that are required by the
-\texttt{plot\_sdt\_fit} function, names starting with \texttt{Corr}
-refer to random effects correlation matrices.
-
-\subsubsection{Step 4: assessing the model fit}
+hierarchical parameters are indexed (e.g., \texttt{delta\_sd\_1},
+\texttt{delta\_random\_1}, \texttt{Corr\_delta\_1}). The name
+\texttt{counts\_new} refers to posterior predictive samples that are
+required by the \texttt{plot\_sdt\_fit} function.
+% ?? Tu już mówimy ogólnie o parametrach o takich nazwach
+Names starting with \texttt{Corr} refer to random effects correlation
+matrices, which are computed from the Cholesky decompositions.
+
+\subsubsection{Assessing the model fit}
 
@@
 tuning the parameters of the sampling algorithm was discarded. Part of
-the resulting stan output is presented in Table~\ref{tab:fitsum}
+the resulting Stan output is presented in Table~\ref{tab:fitsum}
 below.
@@
   \hline
-  & mean & $SE_{mean}$ & $SD$ & $2.5\%$ & $97.5\%$ & Effective samples & $\hat{R}$ \\ 
+  & mean & $SE_{mean}$ & $SD$ & $2.5\%$ & $97.5\%$ & No. eff. samples & $\hat{R}$ \\ 
   \hline
@@
 
-As can be seen, the chains exhibited good mixing properties given the
-complexity of the model and seemed to have reached apparent
-convergence: there were enough effective samples of fixed effect
-parameters to estimate $95\%$ credible intervals well and none of the
-Gelman-Rubin $\hat{R}$ statistics crossed the conventional $1.01$
-threshold suggesting negligible sensitivity to initial values.
-
-Fig.~\ref{fig:6} and \ref{fig:7} below contain normal
-quantile-quantile plots of random effects showing that normally
-distributed random $\delta$ and $\gamma$ effects seem to be good
-candidates for representing variability in sensitivities and criteria
-due to sampled factors.
+As can be seen, the chains exhibited good mixing given the complexity
+of the model and seemed to have converged: there were enough effective
+samples of the fixed effect parameters to estimate $95\%$ credible
+intervals well and none of the Gelman-Rubin $\hat{R}$ statistics
+crossed the conventional $1.01$ threshold suggesting negligible
+sensitivity to the initial values.
+
+Figs.~\ref{fig:6} and \ref{fig:7} below contain normal
+quantile-quantile plots of random effects. The plots indicate that
+normally distributed random $\delta$ and $\gamma$ effects seem to be
+good candidates for representing variability in the sensitivity and
+criteria parameters due to sampled factors.
 
@@
 
-Even when stan output summary does not indicate any obvious sampler
-convergence issues before drawing any further conclusions the
-researcher should first check if the model fits the data. We recommend
-using the \texttt{plot\_sdt\_fit} function for that purpose:
+Once enough good quality posterior samples are obtained for the
+parameters of interest the inference process can be carried out by
+calculating credible intervals, HPD intervals, or bayes factors for
+any function of the parameters. However, even when the stan output
+summary does not indicate sampler convergence issues, before drawing
+any further conclusions the researcher should first check if the model
+fits the data. The \texttt{plot\_sdt\_fit} function can be used for
+that purpose:
 
@@
 
-This function needs at least three agruments; a stanfit object, an
+This function requires at least three agruments; a stanfit object, an
 aggregated data list produced by the \texttt{aggregate\_responses}
-function that was used to produce the stanfit object and a vector of
-names of variables that define how the data will be partitioned before
-plotting. Even though we highly recommend assessing the fit at the
-individual level we did not include the participant identification
-number in the list of conditioning variables because the resulting
-plot would take up too much space. As can be seen in Fig.~\ref{fig:3}
-showing the ROC curves produced by the code above the model seemed to
-fit the data well in all but one condition (decision after rating,
-64ms duration) where three out of seven relevant\footnote{The point in
-  the upper-right corner of the ROC curve is always in the $(1,1)$
-  position} points were outside the two-dimensional 95\% posterior
-predictive regions.
+function and that was used to produce the stanfit object, and a vector
+of names of variables that will determine how the data will be
+partitioned before plotting. We recommend assessing the fit at the
+individual level, but we did not include the participant
+identification number in the list of conditioning variables, because
+the resulting plot would take up too much space.
+
+As can be seen in Fig.~\ref{fig:3} showing the ROC curves produced by
+the code above the model seemed to fit the data well in all but one
+condition (decision after rating, 64 ms duration) where three out of
+seven relevant\footnote{The point in the upper-right corner of a ROC
+  curve is always in the $(1,1)$ position} points were outside the
+two-dimensional 95\% posterior predictive regions.
 
@@
 
-Another way to assess the model fit visually is by inspecting the
-conditional response distributions ($p(y|stim)$) shown in
-Fig.~\ref{fig:4} which was also created using the
+Another way to assess a model fit visually is by inspecting the
+% ?? to mają być conditional response distr. dla tego model fita
+conditional response distributions ($p(y|stim)$) such as those shown
+in Fig.~\ref{fig:4}, which was also created using the
 \texttt{plot\_sdt\_fit} function.
@@
 
-This kind of plot can sometimes be informative about the reasons why
-the model does not fit the data. In this particular case the plot
-seems to suggest that it may be a good idea to inspect the fit at the
+% ?? przedimek reasons? niby the model, bo dla plotu model jest już
+% określony, ale tu jest this kind of plot
+This kind of plot can be informative about the reasons why a model
+does not fit the data. In this particular case the plot seems to
+suggest that it may be a good idea to inspect the fit at the
 individual level and see if there are some participants with unusual
-$p(y|stim=1)$ distributions in the RD $\times$ 64ms condition. On the
-other hand it is also possible that the lack of fit was mainly a
+$p(y|stim=1)$ distributions in the RD $\times$ 64 ms condition. On the
+other hand, it is also possible that the lack of fit is mainly a
 consequence of the assumption that duration had zero effect on
-$\bm{\gamma}$ or maybe the model is simply incorrect and more
-substantial modifications are necessary.
-
-\subsubsection{Step 5: converting unconstrained $\delta$ and $\gamma$
+$\bm{\gamma}$, or that more substantial modifications are necessary,
+such as dropping the equal variance assumption.
+
+\subsubsection{Converting unconstrained $\delta$ and $\gamma$
   parameters to sensitivities and criteria}
 
-The posterior $\delta$ and $\gamma$ samples have to be transformed in
-order to work with $d'$ and $c$ parameters. This is straightforward
-only when fixed effects represent separate conditions, not the
-differences between conditions or the regression slopes. In our
-example, because separate intercepts and slopes parametrization was
-used for the $\delta$ fixed effects model matrix all four
-\texttt{delta\_fixed} parameters can easily be transformed to
-sensitivities by applying the logarithm function to the posterior
-samples. It is important to remember that the $\delta$ to $d'$
-conversion should usually be done first, before applying any other
-transformations to the samples, for example, the logarithm of point
-and interval $\delta$ estimate is not equal to the point and interval
-$d'$ estimate.
-
-In case of the \texttt{gamma\_fixed} matrix the first column (the
-intercept) corresponds to the values of the $\bm{\gamma}$ vector
-\emph{in} the DR condition but the second column corresponds to
-\emph{the effect} of the order variable on the $\bm{\gamma}$
-vector. For this reason in this particular case posterior criteria
-samples can be obtained using the \texttt{gamma\_to\_crit} function
-only for the first column of the \texttt{gamma\_fixed} matrix.
+Posterior $\delta$ and $\gamma$ samples have to be transformed in
+order to work with the $d'$ and $c$ parameters. This is
+straightforward only when fixed effects represent average parameter
+values in separate conditions, not differences between conditions or
+regression slopes. In our example, because separate intercepts and
+slopes parametrization was used for the $\delta$ fixed effects model
+matrix, all four \texttt{delta\_fixed} parameters can easily be
+transformed to sensitivities by applying the natural logarithm
+function. It is important to remember that, because the logarithm is a
+non-linear transformation, the $\delta$ to $d'$ conversion step should
+be done first, before
+% ?? nie wiem czemu the posterior samples, ale tak czuję
+applying any other transformations to the posterior samples, for
+example, the logarithm of a point and interval $\delta$ estimate is
+not equal to the point and interval $d'$ estimate calculated after
+transforming the $\delta$ posterior samples to the $d'$ samples.
+
+In this case the first column of the \texttt{gamma\_fixed} parameter
+matrix (the intercept) corresponds to the values of the $\bm{\gamma}$
+vector in the DR condition but the second column corresponds to the
+\emph{effect} of order on $\bm{\gamma}$. For this reason in this
+particular case the posterior criteria samples can be obtained using
+the \texttt{gamma\_to\_crit} function only for the first column of the
+\texttt{gamma\_fixed} matrix.
 
@@
 simulated the data from a hypothetical exact replication of the
-previously described experiment using point estimates from the
+previously described experiment using the point estimates from the
 previous fit as known realistic parameter values. Mixing performance
-and apparent convergence were similar as in the real data case. All
-the model parameters were correctly recovered in a sense that the true
-values were outside the 95\% credible intervals no more than 5\% of
-the time.
-
-For illustration purposes the SDT model that differed from the true
-model only in that it did not have the hierarchical structure
-representing participant effects was fitted to the same simulated
-dataset. Since the model was much simpler and the data consisted of
-only eight vectors of response counts apparent convergence was quickly
-reached and chain mixing was excellent.
-
-Point and interval estimates of fixed effects based on both models are
-compared in Fig.~\ref{fig:5} below. The estimates were centered at the
-true values to simplify the presentation. As can be seen, the true
-model correctly recovered known parameter values, but estimates based
-on the non-hierarchical model were clearly biased; $95\%$ credible
-intervals were not only about three times shorter on average but also
-failed to contain most of the true values.
+was similar to the real data case. All the model parameters were
+correctly recovered in a sense that the true values were outside the
+95\% credible intervals no more than 5\% of the time.
+
+For illustration purposes an SDT model that differed from the true
+model only in that it did not have any hierarchical structure was
+fitted to the same simulated dataset. Since the non-hierarchical model
+was much simpler and the data consisted of only eight vectors of
+response counts the mixing of the chains was excellent.
+
+The 95\% credible intervals calculated for the fixed effects based on
+each model are compared in Fig.~\ref{fig:5} below. The estimates were
+centered on the true values to simplify the presentation. As can be
+seen, the true model correctly recovered the known parameter values,
+but the estimates based on the simplified, non-hierarchical model were
+biased; the $95\%$ credible intervals were not only about three times
+shorter on average but also failed to contain most of the true values.
 
@@
   \includegraphics[width=.8\linewidth]{true_vs_nonhier.pdf}
-  \caption{Comparison of point and interval posterior estimates based
-    on the true and simplified non-hierarchical models}
+  \caption{Comparison of the point and interval posterior estimates
+    based on the true hierarchical and the simplified non-hierarchical
+    models}
   \label{fig:5}
@@
 
-However, as can be seen on Fig.~\ref{fig:8} below in this case the
+However, as can be seen in Fig.~\ref{fig:8} below in this case the
 observed ROC curves seemed to fit the simplified model's predictions
@@
 The great importance of SDT to psychology stems from the fact that
-given some fairly weak assumptions about the underlying decision
-process it promises to deconfound sensitivity from bias in arbitrary
+given weak assumptions about an underlying decision process it
+promises to deconfound sensitivity from bias in arbitrary
 classification tasks. To the best of our knowledge at present the
@@
 for SDT models with or without ratings that can be recommended as a
-default choice in typical applications. Our method allows for great
-flexibility that requires minimal input from the user. Parametrization
-forces sensitivities to be non-negative and criteria to be ordered
-while the isomorphisms between unconstrained $\delta$ and $\gamma$ and
-derived $d'$ abd $c$ parameters ensure that hierarchical structure is
-valid. There is no limit to the number of sampled factors except for
-the one imposed by available computational resources, correlations of
-random effects of the same sampled factor are accounted for, all the
-SDT parameters can be modelled by linear regression within the same
-model and all the effects on all the SDT parameters estimable within
-the levels of sampled factors can have associated correlated random
-effects. In case a need arises to relax a built-in restriction
-experienced users can extend the model in arbitrary ways by using
-automatically generated human-readable Stan model code as a
+default choice in typical applications. Our parametrization forces the
+sensitivity to be non-negative and the criteria to be order-restricted
+while the isomorphisms between the $d'$ and $c$ parameters and the
+unconstrained $\delta$ and $\gamma$ parameters make it possible to
+supplement the SDT model with the general hierarchical linear
+regression structure. There is no limit to the number of sampled
+factors except for the one imposed by available computational
+resources, correlations of random effects of the same sampled factor
+are accounted for, all the SDT parameters can be modelled by linear
+regression within the same model and all the effects on all the SDT
+parameters estimable within the levels of the sampled factors can have
+associated random effects. In case a need arises to relax a built-in
+restriction experienced users can extend the model in arbitrary ways
+by using automatically generated human-readable Stan code as a
+% ?? basic understanding niepoliczalne
 template. We hope that researchers with basic understanding of Signal
-Detection Theory, bayesian inference and hierarchical modelling will
+Detection Theory, bayesian inference, and hierarchical modelling will
 find our package useful and adopt it as a method of analysis of
-classification performance in their future studies. In the near future
-we are planning to extend the package by relaxing the equal variance
-assumption and adding the ability to fit other generalizations of SDT,
-such as the meta-d' model, that allow for different information to be
-used by classification decision and rating.
+classification performance in their future studies.
 

### 2017-12-06 R/acc_to_delta.R

@@ -10,4 +10,3 @@ acc_to_delta = function(acc){
         stop('Some accuracy scores were to extreme')
     log(2 * stats::qnorm(acc))
 }
-## ok

### 2017-12-06 R/aggregate_responses.R

@@ -43,4 +43,3 @@ aggregate_responses = function(data, stimulus, response, variables = NULL){
        stimulus = res[[stimulus[1]]],
        counts = counts)
 }
-## ok

### 2017-12-06 R/combined_response.R

@@ -69,4 +69,3 @@ combined_response = function(stimulus, rating = NULL, accuracy = NULL, decision 
         (1 - decision) * (max.rating + 1 - rating) + decision * (max.rating + rating)
     }
 }
-## ok

### 2017-12-06 R/gabor.R

@@ -30,4 +30,3 @@
 #'   experience', 3 = 'absolutely clear image'} \item{stim}{Stimulus class: 0 =
 #'   gabor tilted to the left, 1 = gabor tilted to the right.} }
 "gabor"
-## ok

### 2017-12-06 R/gamma_to_crit.R

@@ -65,4 +65,3 @@ gamma_to_crit = function(samples, beta_index = 1, s = 2){
     colnames(crit) = gsub('gamma', 'criteria', colnames(crit))
     crit
 }
-## ok

### 2017-12-06 R/make_stan_model.R

@@ -46,7 +46,6 @@ make_stan_model = function(random = NULL){
     close(f)
     paste(model, collapse = '\n')
 }
-## ok
 
 rmatch = function (pattern, vector){
   res = TRUE

### 2017-12-06 R/plot_sdt_fit.R

@@ -17,11 +17,11 @@
 #' @param alpha an alpha level for the posterior predictive intervals,
 #'     e.g., if alpha = .05 then 100\% - alpha = 95\% posterior
 #'     predictive intervals will be calculated.
-#' @param bw if TRUE the black-and-white version of the response
-#'     distribution plot will be created. The default is FALSE.
+#' @param bw if TRUE (the default), a black-and-white publication-friendly version
+#'     of the response distribution plot will be created.
 #' @return a plot object.
 #' @export
-plot_sdt_fit = function(fit, adata, variables = NULL, type = 'roc', alpha = .05, bw = FALSE){
+plot_sdt_fit = function(fit, adata, variables = NULL, type = 'roc', alpha = .05, bw = TRUE){
     s = as.data.frame(fit)
     rm(fit)
     cnt_new = t(s[,grep('counts_new', names(s))])
@@ -80,7 +80,7 @@ plot_sdt_fit = function(fit, adata, variables = NULL, type = 'roc', alpha = .05,
         dfrocs = dfroc[dfroc$stimulus == '1',]
         dfrocs$stim2 = dfroc[dfroc$stimulus == '2',]
         rm(dfroc)
-        ggplot(dfrocs, aes(cumfr, stim2$cumfr)) +
+        p = ggplot(dfrocs, aes(cumfr, stim2$cumfr)) +
             geom_line(aes(x = cumfr.fit, y = stim2$cumfr.fit), lty = 2) +
             geom_errorbar(aes(ymin = stim2$cumfr.lo, ymax = stim2$cumfr.hi, x = cumfr.fit), width = 0.02) +
             geom_errorbarh(aes(xmin = cumfr.lo, xmax = cumfr.hi, y = stim2$cumfr.fit), height = 0.02) +
@@ -88,6 +88,11 @@ plot_sdt_fit = function(fit, adata, variables = NULL, type = 'roc', alpha = .05,
             labs(x = 'p(Hit)', y = 'p(FA)') +
             coord_fixed() +
             facet_wrap(~f)
+        if(bw){
+            p + theme_minimalist()
+        }else{
+            p
+        }
     }else{
         ## Wykres rozkładów odpowiedzi
         dfa[,c('count.lo', 'count.hi', 'count.fit')] = cbind(apply(cnt_new_a, 2, function(x)stats::quantile(x, alpha / 2)),
@@ -99,7 +104,7 @@ plot_sdt_fit = function(fit, adata, variables = NULL, type = 'roc', alpha = .05,
[…]

### 2017-12-06 inst/preprint/analysis_script.R

@@ -19,6 +19,25 @@ options(mc.cores = parallel::detectCores())
 library(plyr)
 
 ######################################################################
+## Combining parameter estimation and regression analysis relating
+## these parameters to additional variables
+
+I = 10
+J = 10
+d = expand.grid(i = 1:I, j = 1:J)
+## We have a strong linear relationship
+d$x = (1:I)[d$i]
+d$y = rnorm(nrow(d)) + d$x
+## except for the subset with a lower number of data points that look like outliers
+d = d[!(d$i == I & d$j <= (J / 2)),]
+d$y[d$x == I] = rnorm(sum(d$x == I))
+d$y[d$x == I] = d$y[d$x == I] - mean(d$y[d$x == I])
+est = aggregate(y ~ x, d, mean)
+names(est) = c('x_aggregated', 'par_estimate')
+res = stan(file = 'regression.stan', data = append(d, append(est, list(N = nrow(d), I = I))),
+           pars = c('a0', 'b0', 'a1', 'b1', 'sigma', 'par'), chains = 1)
+
+######################################################################
 ## Fitting the model to real study data
 
 ## We will be using the dataset provided with the package
@@ -54,8 +73,9 @@ if(fresh_start){
                iter = 8000,
                chains = 4)
     save(fit, file = paste(temp_path, 'fit', sep = '/'))
+}else{
+    load(paste(temp_path, 'fit', sep = '/'))
 }
-load(paste(temp_path, 'fit', sep = '/'))
 
 ## Model fit summary does not indicate any convergence issues
 print(fit, probs = c(.025, .957),
@@ -82,7 +102,7 @@ crit = gamma_to_c(as.data.frame(fit))
 round(apply(crit, 2, mean), 2)
 
 ######################################################################
-## Fitting the baic single criterion SDT model
+## Fitting the basic single criterion SDT model
 
 fixed = list(delta = ~ -1 + duration:order, gamma = ~ order)
 random = list(list(group = ~ id, delta = ~ -1 + duration, gamma = ~ 1))
@@ -99,8 +119,9 @@ if(fresh_start){
                  iter = 8000,
                  chains = 4)
     save(fit.s, file = paste(temp_path, 'fit.s', sep = '/'))
+}else{
+    load(paste(temp_path, 'fit.s', sep = '/'))
 }
[…]

### 2018-02-12 R/aggregate_responses.R
kosmetyka

@@ -34,12 +34,12 @@
 #' @export
 aggregate_responses = function(data, stimulus, response, variables = NULL){
   ## stimulus classes encoded as 1 or 2
-  data[[stimulus[1]]] = as.numeric(as.factor(as.character(data[[stimulus[1]]])))
-  K = max(data[[response[1]]], na.rm = T)
-  res = plyr::ddply(data, unique(c(variables, stimulus[1])),
-                    function(df)table(c(df[[response[1]]], 1:K)) - 1)
-  counts = res[, c((ncol(res)-K+1):ncol(res))]
-  list(data = res[, setdiff(variables, c(stimulus[1], response[1]))],
-       stimulus = res[[stimulus[1]]],
-       counts = counts)
+    data[[stimulus[1]]] = as.numeric(as.factor(as.character(data[[stimulus[1]]])))
+    K = max(data[[response[1]]], na.rm = T)
+    res = plyr::ddply(data, unique(c(variables, stimulus[1])),
+                      function(df)table(c(df[[response[1]]], 1:K)) - 1)
+    counts = res[, c((ncol(res)-K+1):ncol(res))]
+    list(data = res[, setdiff(variables, c(stimulus[1], response[1]))],
+         stimulus = res[[stimulus[1]]],
+         counts = counts)
 }

### 2018-02-12 R/make_stan_data.R

@@ -114,8 +114,8 @@
 #' @param criteria_scale a scaling factor corresponding to mapping
 #'     distribution's standard deviation. The default is 2. See also
 #'     'Details'.
-#' @param metad [experimental] when TRUE data for the meta-d' model are created, the
-#'     default if FALSE.
+#' @param metad when TRUE additional data required by the meta-d'
+#'     model are created, the default if FALSE.
 #' @return a list with response and stimulus data, model matrices,
 #'     prior parameter values, and other data required by the stan
 #'     model generated using \code{make_stan_model}.

### 2018-02-12 R/make_stan_model.R

@@ -11,8 +11,8 @@
 #'
 #' @param random an optional list specifying random effects
 #'     structure. See \code{\link{make_stan_data}} for details.
-#' @param metad [experimental] if TRUE meta-d' model code is created,
-#'     default is FALSE.
+#' @param metad if TRUE meta-d' model code is created, default is
+#'     FALSE.
 #' @return a string containing the full model definition in the stan
 #'     modelling language.
 #' @examples

### 2018-02-12 inst/preprint/analysis_script.R

@@ -5,7 +5,7 @@
 ##
 ## Change this to FALSE once all the stanfit objects are stored to
 ## save time and memory
-fresh_start = TRUE
+fresh_start = FALSE
 ## Where do you want to store all those large stanfit objects?
 temp_path = '~/temp'
 

### 2018-02-12 inst/preprint/paper.tex
paper added

--- 2017-12-06 inst/preprint/luxsdt.tex
@@
 In what follows, after introducing the most common version of the SDT
-model, we describe its generalization which can accomodate data from
+model, we describe its generalization which can accommodate data from
 rating experiments. Next, we explain briefly why, if a method of
@@
 factor" instead, because, by virtue of being a new term, it is
-unambiguous, and because it seems descriprively correct: the term
+unambiguous, and because it seems descriptively correct: the term
 "sampled factor" captures all the essential properties of such
@@
 a distribution that depends on the stimulus class. For historical
-reasons the two stimulus classess are often referred to as "noise" and
+reasons the two stimulus classes are often referred to as "noise" and
 "signal", and task performance is described in terms of hits, correct
@@
 seems worthwhile to explain why a general-purpose method of inference
-for SDT models should accomodate both kinds of situations.
+for SDT models should accommodate both kinds of situations.
 
@@
 Another common source of problems is the separation of estimation of
-non-linear model parameters from regression analysis whichs aim to
+non-linear model parameters from regression analysis which aim to
 relate these parameters to measured or manipulated variables. When the
@@
 interpretation, since it can have a non-zero mass when $d' < 0$. The
-outermost criteria are fixed at $0$ and $1$, and the oredering
+outermost criteria are fixed at $0$ and $1$, and the ordering
 restriction is enforced by assuming that the likelihood is $0$
@@
 
-\noindent where $\Phi$ is the CDF of the standard normal distrubution
+\noindent where $\Phi$ is the CDF of the standard normal distribution
 and $\bm{\gamma} \in R^K$, with $\bm{\gamma}_K$ fixed at $0$ for
@@
 an even more complicated index notation\footnote{The reader familiar
-  with hierarchical models may be surprised by our use of superscrip
-  parenthesised greek letters to express hierarchical
+  with hierarchical models may be surprised by our use of superscript
+  parenthesised Greek letters to express hierarchical
   relationships. We chose this convention because it allowed us to use
@@
 
-This function requires at least three agruments; a stanfit object, an
+This function requires at least three arguments; a stanfit object, an
 aggregated data list produced by the \texttt{aggregate\_responses}

### 2018-02-12 inst/preprint/analysis_script.R
kosmetyka skryptu

@@ -6,7 +6,7 @@
 ## Change this to FALSE once all the stanfit objects are stored to
 ## save time and memory
 fresh_start = FALSE
-## Where do you want to store all those large stanfit objects?
+## Where do you want to store all these large stanfit objects?
 temp_path = '~/temp'
 
 library(rstan)
@@ -21,7 +21,7 @@ library(plyr)
 ######################################################################
 ## Fitting the model to real study data
 
-## We will be using the dataset provided with the package
+## We will be using the gabor dataset provided with the package
 data(gabor)
 ?gabor
 
@@ -35,10 +35,10 @@ adata = aggregate_responses(gabor, 'stim', 'r', c('duration', 'id', 'order'))
 fixed = list(delta = ~ -1 + duration:order, gamma = ~ order)
 random = list(list(group = ~ id, delta = ~ -1 + duration, gamma = ~ 1))
 model = make_stan_model(random)
-## Let's take a look at the model code
+## The model code is human-readable
 cat(model)
 
-## Stan data structure
+## Required Stan data structure
 sdata = make_stan_data(adata, fixed, random)
 
 ## Main model fit
@@ -72,7 +72,7 @@ smr = smr[-grep('random', rownames(smr)),]
 smr = smr[-grep('counts_new', rownames(smr)),]
 print(xtable(smr[-nrow(smr),]), file = 'fit_table.tex')
 
-## Let's make some plots
+## Some plots
 (p1 = plot_sdt_fit(fit, adata, c('order', 'duration')))
 ggsave(p1, file = 'roc_fit.pdf')
 (p2 = plot_sdt_fit(fit, adata, c('order', 'duration'), type = ''))
@@ -83,10 +83,11 @@ crit = gamma_to_c(as.data.frame(fit))
 round(apply(crit, 2, mean), 2)
 
 ######################################################################
-## Fitting the basic single criterion SDT model
+## Fitting the single criterion hierarchical SDT model
 
 fixed = list(delta = ~ -1 + duration:order, gamma = ~ order)
 random = list(list(group = ~ id, delta = ~ -1 + duration, gamma = ~ 1))
+## this time we ignore the ratings
 gabor$r = combined_response(gabor$stim, accuracy = gabor$acc)
 if(fresh_start){
[…]

### 2018-02-12 inst/preprint/analysis_script.R
plots

@@ -75,8 +75,10 @@ print(xtable(smr[-nrow(smr),]), file = 'fit_table.tex')
 ## Some plots
 (p1 = plot_sdt_fit(fit, adata, c('order', 'duration')))
 ggsave(p1, file = 'roc_fit.pdf')
-(p2 = plot_sdt_fit(fit, adata, c('order', 'duration'), type = ''))
+ggsave(p1, file = 'roc_fit.png')
+(p2 = plot_sdt_fit(fit, adata, c('order', 'duration'), type = 'response'))
 ggsave(p2, file = 'response_fit.pdf')
+ggsave(p2, file = 'response_fit.png')
 
 ## The criteria are placed fairly symmetrically
 crit = gamma_to_c(as.data.frame(fit))

### 2018-02-13 inst/preprint/paper.tex
apa formatted paper

@@
-% WĄTPLIWOŚCI JĘZYKOWE oznaczone przez %?? !! oznacza miejsce, gdzie
-% skończyłem ostatnio poprawiać
-
-% przykłady w sieci value of <> [directly] represents the <>
-
-% jak się pisze 16ms czy 16 ms
-
-% argumenty funkcji są an, gdy piszemy ogólnie o działaniu tych
-% funkcji
-
-\documentclass[oneside,a4paper]{article}
-\usepackage[sectionbib,bibnewpage]{apacite}
+% -*- coding: utf-8 -*-
+
+\documentclass[a4paper,man,apacite,floatsintext]{apa6}
+\usepackage[english]{babel}
+\usepackage[utf8x]{inputenc}
 \usepackage{amsmath}
@@
 
-\title{The bhsdtr package: a general purpose method of bayesian
+\title{The bhsdtr package: a general purpose method of Bayesian
   inference for Signal Detection Theory models}
+
+\shorttitle{The bhsdtr package}
+
+\abstract{We describe a novel method of Bayesian inference for
+  hierarchical or non-hierarchical equal-variance normal Signal
+  Detection Theory models with one or more criteria. The method is
+  implemented as an open-source R package that uses the
+  state-of-the-art platform, Stan, for sampling from posterior
+  distributions. Our method can accommodate binary responses as well
+  as additional ratings and an arbitrary number of nested or crossed
+  random grouping factors. The SDT parameters can be regressed on
+  additional predictors within the same model via intermediate
+  unconstrained parameters, and the model can be extended by using
+  automatically generated human-readable Stan code as a template. In
+  the paper we explain how our method improves on other similar
+  available methods, we give an overview of the package, demonstrate
+  its ease of use by providing a real-study data analysis
+  walk-through, and show that the model successfully recovers known
+  parameter values when fitted to simulated data.}
+
+\keywords{Signal Detection Theory, Bayesian inference, hierarchical
+  models}
+
+\twoauthors{Borysław Paulewicz}{Agata Blaut}
+
+\twoaffiliations{University of Social Sciences and Humanities, Faculty
+  in Katowice, Poland}{Department of Psychology, Jagiellonian
+  University, Krakow, Poland}
+
+\authornote{This research was supported in part by the National
+  Science Centre, Poland HARMONIA grant given to Michal Wierzchon
+  (2014/14/M/HS6/00911).
+
+  Correspondence concerning this article should be addressed to
+  Borysław Paulewicz, Department of Psychology, University of Social
+  Sciences and Humanities, Faculty in Katowice, ul. Techników 9,
+  40-326 Katowice, Poland.
+
+Contact: bpaulewicz@swps.edu.pl}
+
+\newcommand{\code}[1]{\texttt{#1}}
 
@@
 \maketitle
-\tableofcontents{}
-
-\section{Abstract}
-
-We describe a novel method of bayesian inference for hierarchical or
-non-hierarchical equal variance normal Signal Detection Theory models
-with one or more criteria. The method is implemented as an open-source
-R package that uses the state-of-the-art platform Stan for sampling
-from posterior distributions. Our method can accommodate binary
-responses as well as additional ratings and an arbitrary number of
-nested
-% ?? WSZYSTKIE sdt parameters CAN be regressed, więc the
-or crossed random grouping factors. The SDT parameters can be
-regressed on additional predictors within the same model via
-intermediate unconstrained parameters, and the model can be extended
-by using automatically generated human-readable Stan code as a
-% ?? czy to jest przypadek other -> bez przedimka?
-template. In the paper we explain how our method improves on other
-similar available methods, we give an overview of the package,
-demonstrate the ease of use by providing a real-study data analysis
-walk-through, and show that the model successfully recovers known
-parameter values when fitted to simulated data.
-
-\section{Introduction}
-
 Many tasks used in psychology studies are essentially classification
@@
 decide if a given test item is old or new, or, in a perceptual study,
-an object may either be a letter or a digit. If a task requires
-classification it is always possible that conclusions based on
-% ? an ability mi tu nie pasuje
-accuracy or percent correct are invalid because ability to
+an object may be either a letter or a digit. If a task requires
+classification, it is always possible that conclusions based on
+accuracy or percent correct are invalid because the ability to
 discriminate between stimulus classes (i.e., sensitivity) is
@@
 effect that manifests itself in differences in classification accuracy
-% ?? ani the sensitivity ani a sensitivity nie pasuje. Chodzi o
-% ogólnie rozumiane zjawisko.
 may reflect differences in sensitivity, bias, or both. Signal
@@
 
-However, because the SDT model is non-linear any variability in its
+However, because the SDT model is non-linear, variability in its
 parameters due to factors such as participants or items has to be
@@
 regression structure defined on SDT parameters. We argue that the
-\texttt{bhsdtr} package for R \cite{rstatistical} provides exactly
-such an implementation and we have made it publicly available at
-% ?? tylko jeden, konkretny kod i zbiór danych, o którym czytelnik
-% dopiero się dowiaduje, ale o którym nie będziemy więcej mówić, czyli
-% the
-https://github.com/boryspaulewicz/bhsdtr, together with the annotated
-source code that was used to perform all the analyses and produce all
-the figures presented in this paper.
-
-% OPIS STRUKTURY ARTYKUŁU
+\code{bhsdtr} package for R \cite{rstatistical} provides exactly such
+an implementation and we have made it publicly available at
+\code{https://github.com/boryspaulewicz/bhsdtr}, together with the
+annotated source code that was used to perform all the analyses and
+produce all the figures presented in this paper.
 
 In what follows, after introducing the most common version of the SDT
-model, we describe its generalization which can accommodate data from
+model, we describe its generalization, which can accommodate data from
 rating experiments. Next, we explain briefly why, if a method of
 inference for SDT models were to be of general use in psychology
-studies, it is essential that it was based on a model equipped with
-the general hierarchical linear regression structure. The
-\texttt{bhsdtr} package meets this requirement thanks to a novel
-parametrization. We describe this novel parametrization in great
-detail and we motivate it by explaining how the reliance on a
-relatively standard parametrisation leads to problems in the two other
-available implementations. We end the first part of this paper with a
-formal definition of the model as implemented in \texttt{bhsdtr}. The
-second part contains an overview of the package and a tutorial in
-which we demonstrate how to use our method in practice. Before we go
-any further, however, a note on terminology seems in order.
-
-% SAMPLED FACTOR
-
-In the context of hierarchical modelling factors such as participants,
-items, or replications are often referred to as groups. In our opinion
-this naming convention may be confusing; a single participant is both
-a group and a member of some group, while at the same time the term
-"group" is perhaps most strongly associated with study conditions, as
-in "experimental group". In this paper we use the term "sampled
-factor" instead, because, by virtue of being a new term, it is
-unambiguous, and because it seems descriptively correct: the term
+studies, it is essential that it is based on a model equipped with the
+general hierarchical linear regression structure. The \code{bhsdtr}
+package meets this requirement thanks to a novel parametrization. We
+describe this novel parametrization in great detail and explain how
+reliance on relatively standard parametrization leads to problems in
+the two other available implementations. We end the first part of this
+paper with a formal definition of the model as implemented in
+\code{bhsdtr}. The second part contains an overview of the package and
+a tutorial in which we demonstrate how to use our method in
+practice. Before we go any further, however, a note on terminology
+seems in order.
+
+In the context of hierarchical modelling, factors such as
+participants, items, or replications are often referred to as
+groups. In our opinion this naming convention may be confusing; a
+single participant is both a group and a member of some group, while
+at the same time the term "group" is perhaps most strongly associated
+with study conditions, as in "experimental group". In this paper we
+use the term "sampled factor" instead because, by virtue of being a
+new term, it is unambiguous and seems descriptively correct: the term
 "sampled factor" captures all the essential properties of such
@@
 from a larger population and are usually not of direct interest, as in
-"this is only a sample", and that conclusions
-% ?? jeżeli już mamy jakąś analizę statystyczną, to zakładamy, w
-% uproszczeniu, że target population jest określona
-of statistical analysis are meant to apply to the whole population of
-possible values.
-
-\section{Equal variance normal Signal Detection Theory model with
-  additional criteria}
-
-According to Signal Detection Theory each stimulus in a classification
-task gives rise, by some unspecified cognitive process, to an
-unidimensional internal evidence value $s$ sampled from
-% ?? stimulus już był wprowadzony, więc jego klasa jest (jakaś, ale)
-% określona
-a distribution that depends on the stimulus class. For historical
-reasons the two stimulus classes are often referred to as "noise" and
-"signal", and task performance is described in terms of hits, correct
-rejections, omissions, and false alarms, but this terminology is
-appropriate only when the model is applied to tasks that require
-detection, which is by far not always the case. In the most widely
-used version of the model, shown in Fig.~\ref{basicsdt}, the two
-evidence distributions are normal with the same variance, which is
-usually fixed at unity to make the model identifiable. The distance
-$d'$ between the means of the
-% ?? sensitivity bez przedimka?
-evidence distributions represents sensitivity. Because normal
-distributions are unbounded, $s$ is always ambiguous, and so a
-criterion $c$ placed on the evidence axis has to be used to reach a
-discrete
-% ?? bias bez przedimka?
-decision. A participant is assumed to decide that a stimulus belongs
-to the first class (e.g., an old item) if $s < c$, or that it belongs
-to the second class (e.g., a new item) if $s \geq c$. The location of
-the decision criterion $c$ represents bias.
+"this is only a sample", and that conclusions of statistical analysis
+are meant to apply to the whole population of possible values.
+
+\section{Equal-variance normal SDT model with additional criteria}
+
+According to SDT, each stimulus in a classification task gives rise,
+by some unspecified cognitive process, to a unidimensional internal
+evidence value $s$ sampled from a distribution that depends on the
+stimulus class. For historical reasons the two stimulus classes are
+often referred to as "noise" and "signal", and task performance is
+described in terms of hits, correct rejections, omissions, and false
+alarms, but this terminology is appropriate only when the model is
+applied to tasks that require detection, which is far from always
+being the case. In the most widely used version of the model, shown in
+Fig.~\ref{basicsdt}, the two evidence distributions are normal with
+the same variance, which is usually fixed at unity to make the model
+identifiable. The distance $d'$ between the means of the evidence
+distributions represents sensitivity. Because normal distributions are
+unbounded, $s$ is always ambiguous, and so a criterion $c$ placed on
+the evidence axis has to be used to reach a binary decision. A
+participant is assumed to decide that a stimulus belongs to the first
+class (e.g., an old item) if $s < c$, or that it belongs to the second
+class (e.g., a new item) if $s \geq c$. The location of the decision
+criterion represents bias.
 
@@
   \includegraphics[width=.8\linewidth]{plot_basicsdt.pdf}
-  \caption{Equal variance normal Signal Detection Theory model}
+  \caption{Equal-variance normal Signal Detection Theory model}
   \label{basicsdt}
@@
 
-% Najprostsze użycie
-
 Perhaps the simplest way of using this model is to fit it to observed
 response counts and use the estimated $d'$ values in place of the
-percent correct scores; If the model is correct, the resulting
+percent correct scores; if the model is correct, the resulting
 performance measure is not contaminated by bias. Obviously, the model
@@
   \includegraphics[width=.8\linewidth]{plot_sdtr.pdf}
-  \caption{Equal variance normal Signal Detection Theory model with
+  \caption{Equal-variance normal Signal Detection Theory model with
     additional criteria}
@@
 
-This
-% ?? applicable to czy applicable in
-generalized model is applicable to studies where participants are
-asked to rate their binary classification decisions on confidence or
-some other performance or stimulus related
-% the ratings, bo study już określone, podobnie jak (rodzaj)
-% klasyfikacji binarnej, ale czy the ratings and THE binary ...?
-dimension. The ratings and the binary classification decisions can be
-provided together (e.g., "I am almost certain that it was a digit"),
-or in an arbitrary order.
-
-% MODELOWANIE RATINGÓW ZA POMOCĄ KRYTERIÓW
-
-% ?? a combined response? Skoro mamy ratings (i w domyśle mamy też
-% binary class.), to czy nie the combined response? A może
-% jakiekolwiek ratingi są załatwione za pomocą (ogólnego rozwiązania)
-% modelowania czegoś, co jest "odpowiedzią łączną"?
+This generalized model is applicable to studies in which participants
+are asked to rate their binary classification decisions on confidence
+or some other performance- or stimulus-related dimension. The ratings
+and the binary classification decisions can be provided together
+(e.g., "I am almost certain that it was a digit"), or in an arbitrary
+order.
+
 Ratings are accommodated by introducing additional criteria and
-modelling a combined response $y$, which
-% ?? znowu kwestia przedimków vs. and
-represents both the binary classification decision and rating. The
-value of $y$ increases with the strength of evidence in favor of the
-second stimulus class. For example, if confidence is rated on a
-four-point scale, then $y = 1$ when a subject decides
-% ?? bodziec, który zobaczył badany, więc the? bo chodzi o decyzję o
-% określonym dla decydenta bodźcu, a nie nieokreślonym dla decydenta?
-that a stimulus belongs to the first class with certainty $4$, and
-$y = 5$ when a subject decides that a stimulus belongs to the second
-class with certainty $1$. More formally, if $K$ is the number of
-possible combined responses then a subject is assumed to give response
-$k$ if $s \in (\bm{c}_{k-1},\bm{c}_k]$, where $\bm{c}$ are the $K+1$
-decision criteria, with $\bm{c}_0$ and $\bm{c}_K$ fixed at $-\infty$
-and $+\infty$ respectively.
-
-% UŻYWAĆ SDTR CHOĆBY PO TO, ŻEBY SPRAWDZIĆ JEGO ZAŁOŻENIA
+modelling a combined response $y$, which represents both the binary
+classification decision and rating. The value of $y$ increases with
+the strength of evidence in favor of the second stimulus class. For
+example, if confidence is rated on a four-point scale, then $y = 1$
+when a subject decides that a stimulus belongs to the first class with
+certainty $4$, and $y = 5$ when a subject decides that a stimulus
+belongs to the second class with certainty $1$. More formally, if $K$
+is the number of possible combined responses, then a subject is
+assumed to give response $k$ if $s \in (\bm{c}_{k-1},\bm{c}_k]$, where
+$\bm{c}$ are the decision criteria, with $\bm{c}_0$ and $\bm{c}_K$
+fixed at $-\infty$ and $+\infty$ respectively.
 
 There is a good reason to collect the ratings and use the generalized
-SDT model from Fig.~\ref{sdtr} even when neither the ratings nor the
-placement of criteria are relevant to the research problem.
-% ?? If K = 2 to jakby: w (jakimś) przypadku, w którym, When K = 2 to
-% jakby: zawsze, gdy ...? Czuję, że lepiej the data i the model potem
-When $K = 2$ (no ratings), the SDT model fits perfectly, regardless of
-it being a reasonably good approximation to reality or not, because
-the data and the model have the same dimensionality. This makes the
-generalization to the $K > 2$ case particularly important, as it is
-% ?? niektóre lub wszystkie formal assumptions, ale mamy na myśli
-% konkretny zbiór formal assumptions, który czytelnik już zna, a
-% nawet, jeśli nie, to te założenia są unikalnie wyznaczone przed
-% model - the?
-only when $K > 2$, that the formal assumptions of the model (e.g.,
-equal or unequal variance) can be tested\footnote{In contrast to the
-  formal assumptions, a psychological interpretation of the SDT
-  parameters can be tested even when ratings are not available, e.g.,
-  by means of selective influence \cite{sternberg2001separate}}, which
-is often done by comparing observed and predicted ROC curves.
+SDT model from Fig.~\ref{sdtr}, even when neither the ratings nor the
+placement of criteria are relevant to the research problem. When
+$K = 2$ (no ratings), the SDT model fits perfectly, regardless of
+whether it is a reasonably good approximation to reality, because the
+data and the model have the same dimensionality. This makes the
+generalization to the case particularly important, as it is only when
+$K > 2$ that the formal assumptions of the model (e.g., equal or
+unequal variance) can be tested\footnote{In contrast to the formal
+  assumptions, a psychological interpretation of the SDT parameters
+  can be tested even when ratings are not available, e.g., by means of
+  selective influence \cite{sternberg2001separate}}, which is often
+done by comparing observed and predicted ROC curves.
 
@@
 usually interested in relationships between SDT parameters and
-additional measured or manipulated
-% ?? the dependence, bo ogólnie rozumiana kwestia zależności, stimulus
-% strength - ogólne zjawisko, a nie jakaś konkretna, albo pojedyncza,
-% ale bliżej nieokreślona
-variables; A good example is the dependence of $d'$ on stimulus
-strength. Also, in a great majority of psychology studies in which
-classification tasks are used the data have hierarchical structure,
-i.e., there are repeated measures for each participant or item and
-participants or items are only samples from the target population. It
-seems worthwhile to explain why a general-purpose method of inference
-for SDT models should accommodate both kinds of situations.
-
-% ?? Ważność hierarchical structure jako takiej
-\subsection{The importance of hierarchical regression structure}
-
-If data have hierarchical structure, but variability due to subjects,
-items, or other sampled factors is not accounted for, estimates of
-average (fixed) effects are not guaranteed to be unbiased and
-conclusions are not guaranteed to generalize to a target population. A
-not so uncommon practice of analysing data aggregated over sampled
-factors represents an extreme case of ignoring
-% ?? nieokreślona hier. structure, bo tu mówimy o externe case
-% ogólnego zjawiska ignorowania ogólnej kwestii hierarchiczności
-hierarchical data structure. Invalidity of this approach in the
-% ?? Jeżeli illustrated w czasie przeszłym dokonanym i mówimy zaraz
-% czyje rezultaty, to chyba the results?
+additional measured or manipulated variables; a good example is the
+dependence of $d'$ on stimulus strength. Also, in a great majority of
+psychology studies in which classification tasks are used, the data
+have a hierarchical structure, i.e., there are repeated measures for
+each participant or item, and participants or items are only samples
+from the target population. It seems worthwhile to explain why a
+general-purpose method of inference for SDT models should accommodate
+both kinds of situations.
+
+\section{The importance of hierarchical regression structure}
+
+If data have a hierarchical structure, but variability due to
+subjects, items, or other sampled factors is not accounted for,
+estimates of average (fixed) effects are not guaranteed to be unbiased
+and conclusions are not guaranteed to generalize to the target
+population. The not uncommon practice of analyzing data aggregated
+over sampled factors represents an extreme case of ignoring
+hierarchical data structure. The invalidity of this approach in the
 context of SDT was clearly illustrated with the results of
-simulational studies by \citeA{morey2008problematic}, although,
+simulational studies by \citeA{morey2008problematic}; however,
 strictly speaking, it requires such demonstrations only in order to
 make a point about a specific case. Because SDT is a non-linear model,
-% ?? data jest unikalne dla tokenów estimates, więc the data?
 by definition, when estimates of its parameters are based on data
-aggregated over sampled factors the expected values of these estimates
-are not in general true population averages. Such estimates are
-% ?? zakładamy (w uproszczeniu), że target population jest ustalona
-% dla każdego estimate
-biased and inference about a target population based on such estimates
-is simply not valid. To give a concrete example, consider two unbiased
+aggregated over sampled factors, the expected values of these
+estimates are not in general true population averages. Such estimates
+are biased and inference about a target population based on them is
+simply not valid. To give a concrete example, consider two unbiased
 participants, one with $d'_1 = 2$ and one with $d'_2 = 4$. Their
@@
 
-Aggregation is not the only way of ignoring hierarchical data
+Aggregation is not the only way of ignoring a hierarchical data
 structure. Sometimes non-aggregated data are analyzed by using
 separate estimates for every participant $\times$ item $\times$
-% ?? mamy już dane i estimates, to (zwykle nieznany) (poziom)
-% uncertainty jest określony
-condition combination, but the uncertainty due to subject or item
-effects \emph{distributions} is not accounted for by means of a
-hierarchical model
-% dalej jedziemy z nieokreślonymi, bo wyjaśniamy ogólne zasady:
-% wnioski mogą być jakiekolwiek, próbki i populacje są bliżej
-% nieokreślone
-structure. In such cases conclusions, at least with respect to
-% ?? może chodzić o różne population-average różnie estymowalne
-%
-% ?? population-average - or fixed - effects? chodzi mi o to, że
-% effects ma się też "łączyć" z population-average
-the uncertainties in estimates of population-average (fixed)
-% ?? sample i population jest dana dla tokenów
-effects, are guaranteed to be valid only for the given sample, not the
-target population.
-
-% mówimy o ogólnym co prawda, ale zarazem pojedynczym/unikalnym
-% problemie: the separation? Chętnie bym się pozbył przedimka
+condition combination, but uncertainty due to distributions of subject
+or item effects is not accounted for by means of a hierarchical model
+structure. In such cases, conclusions -- at least with respect to the
+uncertainties in estimates of population-average (fixed) effects are
+guaranteed to be valid only for the given sample, not the target
+population.
+
 Another common source of problems is the separation of estimation of
-non-linear model parameters from regression analysis which aim to
+non-linear model parameters from regression analyses which aim to
 relate these parameters to measured or manipulated variables. When the
@@
 
-% tutaj jakaś regression analysis już jest wcześniej wprowadzona
 Firstly, the standard errors or credible intervals associated with the
 regression coefficients do not reflect the uncertainty in the SDT
-parameter estimates, because the latter are treated as mere data
-points. Precision of parameter estimates often varies between
+parameter estimates because the latter are treated as mere data
+points. The precision of parameter estimates often varies between
 participants, items, or conditions, but when the estimates are treated
-as data points no use is made of this information. Secondly,
+as data points, no use is made of this information. Secondly,
 regressing parameters on numeric predictors makes their estimates
@@
 model with the general hierarchical linear regression structure. For
-the reasons that we will now explain, this can only be achieved if the
-model is substantially reparametrised.
-
-% dopiero wprowadzamy ową unconstrained parameter space
+reasons that we will now explain, this can only be achieved if the
+model is substantially reparametrized.
+
 \section{Hierarchical Signal Detection Theory in a constrained
@@
 constrained: $d'$ is non-negative and, when there is more than one
-criterion, the elements of the $\bm{c}$ vector are order-restricted
+criterion, the elements of the $\bm{c}$ vector are order restricted
 ($\bm{c}_{i+1} > \bm{c}_i$). Because normal distribution is unbounded,
-hierarchical linear regression structure, in which random effects are
-normally distributed and fixed effects are represented by
-unconstrained parameters, can only be defined on unconstrained
+the hierarchical linear regression structure in which random effects
+are normally distributed and fixed effects are represented by
+unconstrained parameters can only be defined on unconstrained
 parameters. We provide examples of the problems that may otherwise
@@
 
-We are aware of only two attempts at creating a general purpose method
+We are aware of only two attempts at creating a general-purpose method
 of inference for hierarchical SDT models with ratings. One is the
@@
 is the Hierarchical Meta-d' model (HMeta-d') proposed by
-\citeA{hmetad}.
-% We do not consider the ROC Toolbox \cite{Koen2016} or other similar
-% tools here because, as of this writing, these tools cannot be used to
-% fit hierarchical SDT models. The two available methods that can be
-% used to fit hierarchical SDT models are
-The HMeta-d' model is a hierarchical version of the meta-d' model
-\cite{maniscalco2012signal}, which in turn is a generalization of the
-SDT model that allows for a separate "meta-sensitivity" to account for
-possible discrepancies
-% referred to as ... bez przyimka, prawda?
-between a binary stimulus classification (referred to as type 1 task)
-% ?? wcześniej było a binary i a rating, jakby to były (potencjalnie)
-% niezwiązane zadania
-and the associated rating task (referred to as type 2 or
-meta-cognitive task). We consider HMeta-d' here because it reduces to
-the SDT model with
-% ?? powtórzenie the
-ratings when the type 1 and the type 2 sensitivities are equal.
+\citeA{hmetad}. The HMeta-d' model is a hierarchical version of the
+meta-d' model \cite{maniscalco2012signal}, which in turn is a
+generalization of the SDT model that allows for a separate
+"meta-sensitivity" to account for possible discrepancies between a
+binary stimulus classification (referred to as a type 1 task) and the
+associated rating task (referred to as a type 2 or meta-cognitive
+task). We consider HMeta-d' here because it reduces to the SDT model
+with ratings when the type 1 and type 2 sensitivities are equal.
 
@@
 evidence distribution mean considered in isolation is an unconstrained
-parameter, but the mean of the second one is by definition greater
-($d' > 0$), or equal ($d' = 0$), than the mean of the first one. The
-authors explicitly admit that their algorithm does not enforce this
-restriction, because, as they claim in the paper, it would greatly
-complicate analysis. At the same time the authors fail to mention that
-an immediate consequence of allowing for negative $d'$ values is that
-the resulting posterior distribution looses its intended
-interpretation, since it can have a non-zero mass when $d' < 0$. The
+parameter, but the mean of the second evidence distribution is by
+definition greater than ($d' > 0$) or equal to ($d' = 0$) the mean of
+the first. The authors explicitly admit that their algorithm does not
+enforce this restriction because, as they claim in the paper, it would
+greatly complicate analysis. At the same time the authors fail to
+mention that an immediate consequence of allowing for negative $d'$
+values is that the resulting posterior distribution loses its intended
+interpretation since it can have a non-zero mass when $d' < 0$. The
 outermost criteria are fixed at $0$ and $1$, and the ordering
@@
 distributed random intercepts of one sampled factor. Perhaps the most
-problematic aspect of the HMeta-d' model, apart from the fact that
-also in this model the $d'$ parameter is allowed to assume negative
-values, is the representation of the criteria. Each element of the
+problematic aspect of the HMeta-d' model, apart from the fact that the
+$d'$ parameter is allowed to assume negative values in this model
+also, is the representation of the criteria. Each element of the
 criteria vector has an associated independent normal distribution,
@@
 representing the actual, order-restricted criteria, but, because
-sorting is not injective, the
-% ?? (i.e., not isomorphic to)
-space of the actual criteria is only loosely related (i.e., not
-isomorphic) to the space of the unrestricted criteria vectors that are
-associated with a hierarchical structure.
-
-% Hierarchical structure does not have to be linear, but when it is, the
-% parameters associated with unconstrained fixed effects and normally
-% distributed random effects of sampled factors cannot be non-negative
-% or order-restricted. We propose to substantially reparametrise the SDT
-% model.
+sorting is not injective, the space of the actual criteria is only
+loosely related (i.e., not isomorphic) to the space of the
+unrestricted criteria vectors that are associated with the
+hierarchical structure.
 
@@
 SDT parameters only if the latter are derived from unconstrained
-parameters. In the \texttt{bhsdtr} package $d'$ is derived from
-$\delta = \ln(d')$, this way random effects on $d'$ can be modelled by
+parameters. In the \code{bhsdtr} package, $d'$ is derived from
+$\delta = \ln(d')$, thus random effects on $d'$ can be modelled by
 assuming that $\delta$ is normally distributed. The problem of
-representing the criteria by unconstrained parameters
-% ?? skoro the space of, to potem the... ? w końcu chodzi nam o
-% *wszystkie*, takie, że...?
-is solved by mapping the $R^{K-1}$ space of unconstrained criteria
-vectors to the $K$ dimensional probability simplex space using the
-softmax function, and mapping the simplex space to the space of
-order-restricted criteria vectors by means of the inverse normal CDF:
+representing the criteria by unconstrained parameters is solved by
+mapping the $R^{K-1}$ space of unconstrained criteria vectors to the
+$K$ dimensional probability simplex space using the softmax function,
+and mapping the simplex space to the space of order-restricted
+criteria vectors by means of the inverse normal CDF:
 
@@
 normal probability density function. The elements of the $\bm{\gamma}$
-% ?? jeżeli correspond to <nieokreślonych> to represent też
-% <nieokreślonych>?
 vector correspond to relative distances between pairs of adjacent
-criteria, because their exponents represent the relative
-% cały czas mówimy ogólnie i dopiero teraz wprowadzamy pojęcie pair of
-% adjacent criteria, więc rodzajnik nieokreślony?
-magnitudes of areas under the standard normal curve, delineated by the
-pairs of adjacent criteria:
+criteria because their exponents represent the relative magnitudes of
+areas under the standard normal curve, delineated by the pairs of
+adjacent criteria:
 $e^{\bm{\gamma}_i} / e^{\bm{\gamma}_j} = (\Phi(\bm{c}_i) -
@@
 when the criterion is placed at the midpoint between the evidence
-distributions and the more negative (positive) $\bm{\gamma}_1$ is, the
-more the criterion is shifted to the left (right) of the midpoint. It
-is often a good idea to multiply all the criteria by a value greater
-than $1$, which is equivalent to making the mapping distribution
-wider. This
-% ?? skoro już mówimy, że coś zwykle wpływa na wartości gamma, to nie
-% the wartości?
-tends to even out values of $\bm{\gamma}$, by preventing the outermost
-areas under the mapping distribution curve from becoming very small,
-especially when the criteria are widely spread, as can happen for
-moderate to large $d'$ values. This feature is implemented in the
-\texttt{bhsdtr} package by introducing a scaling factor.
+distributions; the more negative (positive) $\bm{\gamma}_1$ is, the
+more the criterion is shifted to the left (right) of the midpoint.
+
+It is often a good idea to multiply all the criteria by a value
+greater than $1$, which is equivalent to making the mapping
+distribution wider. This tends to even out values of $\bm{\gamma}$ by
+preventing the outermost areas under the mapping distribution curve
+from becoming very small, especially when the criteria are widely
+spread, as can happen for moderate to large $d'$ values. This feature
+is implemented in the \code{bhsdtr} package by introducing a scaling
+factor.
 
@@
   with hierarchical models may be surprised by our use of superscript
-  parenthesised Greek letters to express hierarchical
+  parenthesized Greek letters to express hierarchical
   relationships. We chose this convention because it allowed us to use
@@
   minimizing the number of nested sub- or superscripts.}, below we
-have presented only the simple case of one sampled factor.
+present only the simple case of one sampled factor.
 
@@
 is the random effects model matrix, $\bm{\beta}$ and $\bm{\theta}$ are
-the
-% opisać raczej jako macierz c, itd. - jest macierzą != jest efektem
-% losowym
-fixed and random effects, $\bm{c}$ is a $N \times K-1$ matrix, $s$ is
-the criteria scaling factor, and $\bm{y}$ is the combined
+the fixed and random effects, $\bm{c}$ is an $N \times K-1$ matrix,
+$s$ is the criteria scaling factor, and $\bm{y}$ is the combined
 response. Note that $\bm{d'}_i$ is a scalar, but
@@
 $j$-th rows of the $\bm{\beta}^{(\gamma)}$ and
-$\bm{\theta}^{(\gamma)}$ matrices
-% ?? effects on ... ?
-represent fixed and random effects on the $j$-th element of the
-$\bm{\gamma}$ vector.
-
-% ?? tutaj coś dziwnego dzieje się z liczbą: ta dekompozycja tych
-% macierzy, i ta (jedna) dekompozycja poprawia próbkowanie itd.
+$\bm{\theta}^{(\gamma)}$ matrices represent fixed and random effects
+on the $j$-th element of the $\bm{\gamma}$ vector.
+
 Following \citeA{sorensen2015bayesian} we make use of the Cholesky
-decomposition of the correlation matrices, because it improves
+decomposition of the correlation matrices because it improves
 efficiency and admits a convenient prior on random effects
@@
 
-% ?? problem z przedimkami: tau odpowiada obydwu tau, żeby było
-% krócej, ale przez to jest "a" wektorem sd efektów losowych, chociaż
-% niezupełnie dowolnym
 \noindent where each $\bm{\tau}$ is a vector of standard deviations of
@@
 random effects correlation matrix, i.e., $\bm{C} = \bm{L}
-\bm{L}'$. This way $\bm{\theta}$ is multivariate normal with
-covariance matrix $\text{diag}(\bm{\tau}) \bm{L}$.
+\bm{L}'$. Thus, $\bm{\theta}$ is multivariate normal with covariance
+matrix $\text{diag}(\bm{\tau}) \bm{L}$.
 
@@
 $\bm{\tau}^{(\delta)}$ and $\bm{\tau}^{(\gamma)}$ are given
-independent half-Cauchy priors as recommended by
+independent half-Cauchy priors, as recommended by
 \citeA{gelman2004prior}, and each $\bm{L}$ is given an independent lkj
@@
 
-% ?? the? Tytuły rządzą się specjalnymi prawami, np. to nie są zdania,
-% ani nawet clauses chyba
-\subsection{Specifying the prior distributions}
-
-% !! TUTAJ skończyłem wstawiać intuicyjne przecinki
-The formal definition of a bayesian model is not complete without
+\section{Specifying the prior distributions}
+
+The formal definition of a Bayesian model is not complete without
 providing fixed values of all the parameters that define prior
 distributions. Specifying the priors on sensitivity effects does not
-pose any special difficulties. Sensitivity of an unbiased classifier
-given percent correct is given by $2 \Phi^{-1}(\text{pc})$. When
-$p(stim = 1) = 0.5$ the greater the bias the lower the accuracy
-meaning that an unbiased sensitivity is a lower bound on sensitivity
-given percent correct. Let us assume that the majority of subjects are
-expected to achieve
-% ?? with negligible bias, bez niczego?
-percent correct within the $.51$ to $.99$ range with negligible
-bias. Since $\ln(2\Phi^{-1}(.51)) = -2.99$ and
-$\ln(2 \Phi^{-1}(.99)) = 1.54$, a reasonable weakly informative prior
-on $\delta$ is normal with mean $(1.54 - 2.99) / 2$ and standard
-deviation $(1.54 + 2.99) / 2$, which is the default prior on delta
-effects in the \texttt{bhsdtr} package.
+pose any special difficulties. The sensitivity of an unbiased
+classifier given percent correct (pc) is given by
+$2 \Phi^{-1}(\text{pc})$. When $p(stim = 1) = 0.5$, the greater the
+bias, the lower the accuracy, meaning that an unbiased sensitivity is
+a lower bound on sensitivity given percent correct. Let us assume that
+the majority of subjects are expected to achieve percent correct
+within the $.51$ to $.99$ range, with negligible bias. Since
+$\ln(2\Phi^{-1}(.51)) = -2.99$ and $\ln(2 \Phi^{-1}(.99)) = 1.54$, a
+reasonable weakly informative prior on $\delta$ is normal with mean
+$(1.54 - 2.99) / 2$ and standard deviation $(1.54 + 2.99) / 2$, which
+is the default prior on delta effects in the \code{bhsdtr} package.
 
 Specifying the priors on criteria effects can be challenging because
-the criteria are order-restricted and because the complexity of the
-mapping expressed by Eq.~\ref{eq:1} makes it difficult to reason about
-priors on $\bm{\gamma}$ in terms of criteria effects. By default in
-the \texttt{bhsdtr} package each entry in the $\bm{\sigma}^{(\gamma)}$
-and $\bm{\zeta}^{(\gamma)}$ matrices is set to $\ln(100)$ and the
-criteria scaling factor is fixed at 2.
-% The interpretation of $\ln(100)$ depends on the model matrix, but
-% usually it means that a priori the areas under the mapping
-% distribution curve between adjacent criteria may easily differ from
-% the area between the criteria $K-1$ and $K$ by a factor of 100 but
-% differences by a factor of $2 \ln(100)$ or highly unlikely.
-
-% When in doubt, it is a good idea to perform some sort of sensitivity
-% analysis, remembering that strong sensitivity to priors indicates
-% that the study may be underpowered.
+the criteria are order-restricted and the complexity of the mapping
+expressed by Eq.~\ref{eq:1} makes it difficult to reason about priors
+on $\bm{\gamma}$ in terms of criteria effects. By default, in the
+\code{bhsdtr} package each entry in the $\bm{\sigma}^{(\gamma)}$ and
+$\bm{\zeta}^{(\gamma)}$ matrices is set to $\ln(100)$ and the criteria
+scaling factor is fixed at $2$.
 
@@
 $\bm{\zeta}$, which represents half-width at half-maximum of the
-half-Cauchy distribution. In our opinion a not-unreasonable starting
+half-Cauchy distribution. In our opinion, a not-unreasonable starting
 point is to set $\bm{\zeta}$ at the value that is greater or equal to
@@
 implies a uniform prior on random effects correlation
-matrices. Because the greater the value of $\nu$ the more emphasis is
-put on zero
-% ?? the researcher - the one specifying the priors
-off-diagonal correlations, the researcher can force the correlations
-to be near-zero by choosing a large $\nu$ value.
+matrices. Because the greater the value of $\nu$, the more emphasis is
+put on zero off-diagonal correlations, the researcher can force the
+correlations to be near-zero by choosing a large $\nu$ value.
 
@@
 
-% ?? tutaj wprowadzamy zaraz the, tak jakbyśmy mówili o sekwencji
-% czynności objaśniając funkcje pakietu.
-The \texttt{bhsdtr} package implements the model described in the
-previous section in the Stan modelling language
-\cite{carpenter2016stan} because it uses a state-of-the-art adaptive
-Hamiltonian Monte Carlo sampling algorithm which often handles
-high-dimensional correlated posteriors better than a Gibbs
-sampler. Our package is essentially a collection of documented
-functions. The \texttt{aggregate\_responses} function aggregates data
-as much as possible for efficiency, but without distorting the
-hierarchical structure. The \texttt{make\_stan\_model} function
-creates a model definition in the Stan language. The Stan code
-produced by the \texttt{make\_stan\_model} function can be fitted as
-is or modified by
-% ?? the user - of this function
-the user if needed, e.g., to change the prior distributions or to drop
-the equal variance assumption. The \texttt{make\_stan\_data} function
-creates regression model matrices and other data structures required
-% ?? już mówiliśmy o tym modelu (kodzie modelu)
-by the model created using the \texttt{make\_stan\_model}
-function. Finally, the \texttt{plot\_sdt\_fit} function can be used to
-visually assess the fit of the model by creating publication-ready ROC
-curve plots or response distribution plots with posterior predictive
-intervals calculated for the chosen $\alpha$ level.
+The \code{bhsdtr} package implements the model described in the
+previous section in the Stan modelling language because it uses a
+state-of-the-art adaptive Hamiltonian Monte Carlo sampling algorithm
+which often handles high-dimensional correlated posteriors better than
+a Gibbs sampler. Our package is essentially a collection of documented
+functions: The \code{aggregate\_responses} function aggregates data as
+much as possible for efficiency, but without distorting the
+hierarchical structure. The \code{make\_stan\_model} function creates a
+model definition in the Stan language. The Stan code produced by the
+\code{make\_stan\_model} function can be fitted as is or modified by the
+user if needed, e.g., to change the prior distributions or to drop the
+equal variance assumption. The \code{make\_stan\_data} function creates
+regression model matrices and other data structures required by the
+model created using the \code{make\_stan\_model} function. Finally, the
+\code{plot\_sdt\_fit} function can be used to visually assess the fit of
+the model by creating publication-ready ROC curve plots or response
+distribution plots with posterior predictive intervals calculated for
+the chosen $\alpha$ level.
 
@@
 
-To make full use of the \texttt{bhsdtr} package functionality three
-non-standard R packages are required, namely
-\texttt{rstan}, %?? , czy średnik - namely
-\texttt{plyr}, and \texttt{ggplot2}. We recommend using the
-\texttt{devtools} package to install the \texttt{bhsdtr} package
-directly from the github repository. This will automatically install
-any missing required packages:
+To make full use of the bhsdtr package functionality, three
+non-standard R packages are required, namely \code{rstan},
+\code{plyr}, and \code{ggplot2}. We recommend using the
+\code{devtools} package to install the \code{bhsdtr} package directly
+from the github repository. This will automatically install any
+missing required packages:
 
 \begin{lstlisting}{language = R}
-  devtools::install_git('boryspaulewicz/bhsdtr')
+  devtools::install_git('git://github.com/boryspaulewicz/bhsdtr')
   library(bhsdtr)
@@
 model, assessing the fit, and possibly converting the unconstrained
-$\gamma$ and $\delta$ parameters to $d'$ and $c$.
+$\delta$ and $\gamma$ parameters to $d'$ and $c$.
 
@@
 
-The \texttt{bhsdtr} package contains the \texttt{gabor} dataset which
-comes from an unpublished study. In this study on each trial the
-participants had to
-% ?? Gabor patch presented on that trial - the?
+The \code{bhsdtr} package contains a dataset, \code{gabor}, from an
+unpublished study in which on each trial the participants had to
 classify a briefly presented Gabor patch as tilted to the left or to
 the right using the arrow keys. The participants were also asked to
-rate the stimuli on a 4 point Perceptual Awareness Scale
+rate the stimuli on a 4-point Perceptual Awareness Scale
 \cite{ramsoy2004introspection} presented at the bottom of the
 screen. The Gabor patch was immediately followed by a mask. The PAS
-ratings ranged from "no experience" to "absolutely clear image"
-and % przecinek?
+ratings ranged from "no experience" to "absolutely clear image" and
 were provided either before (RD order condition) or after (DR order
-condition) % , or?
-pressing the arrow keys. On each trial the Gabor patch was equally
-% ?? nazwy zmiennych bez przedimka?
-likely to be presented for 32 ms or 64 ms. Order was a between-subject
-variable and duration was a within-subject variable. There were 47
-participants and 48 trials per condition.
-
-In the study in question the response was originally encoded using
+condition) the arrow keys were pressed. On each trial the Gabor patch
+was equally likely to be presented for 32 ms or 64 ms. Order was a
+between-subject variable and duration was a within-subject
+variable. There were 47 participants and 48 trials per condition.
+
+In the study in question, the response was originally encoded using
 separate variables for accuracy and rating, so the first step was to
 create an appropriate response variable using the
-\texttt{combined\_response} function. This function requires three
+\code{combined\_response} function. This function requires three
 variables, one encoding the stimulus class, one encoding the rating
@@
 response variable is not already present in the data. In the single
-criterion case the combined response variable is simply the binary
+criterion case, the combined response variable is simply the binary
 classification decision. To fit a single-criterion SDT model to this
-dataset the code above would have to be replaced with the following:
+dataset, the code above would have to be replaced with the following:
 
@@
 
-% tutaj czuję, że the trzeba prawie wszędzie
 Next, the data has to be aggregated using the
-\texttt{aggregate\_responses} function, but only to the extent that
+\code{aggregate\_responses} function, but only to an extent that
 preserves all the random effects. This function requires as arguments
-a data frame containing
-% ?? all the relevant variables?
-all the relevant variables, a name of the stimulus class variable, a
-name of the combined response variable and a vector of names of all
-the variables that are to be preserved in the resulting aggregated
-dataset (apart from the stimulus class variable and the combined
-response variable), i.e., those encoding the sampled factors and those
-representing the independent variables used in the regression part of
-the model:
+a data frame containing all the relevant variables, the name of the
+stimulus class variable, the name of the combined response variable,
+and the vector of the names of all the variables that are to be
+preserved in the resulting aggregated dataset (apart from the stimulus
+class variable and the combined response variable), i.e., those
+encoding the sampled factors and those representing the independent
+variables used in the regression part of the model:
 
@@
 of sampling from the posterior distribution. When data are aggregated
-in this way the likelihood for each condition $\times$ participant
+in this way, the likelihood for each condition $\times$ participant
 combination has to be computed only once rather than as many times as
 there are trials per condition per participant. Note that if there are
-% ?? czuję, że the data, bo mówimy o (jakimś), ale nadal tym samym
-% hipotetycznym przypadku
 other sampled factors present in the data (e.g., items, replications,
-etc.) then these factors have to be specified at this stage as well to
+etc.), then these factors also have to be specified at this stage to
 preserve the hierarchical data structure. The
-\texttt{aggregate\_responses} function creates a list with three
-components. The \texttt{data} component is a data frame containing
-% ?? additional zawsze bez przedimka?
-additional preserved variables, the \texttt{stimulus} component is the
-stimulus class variable and the \texttt{counts} component is a
+\code{aggregate\_responses} function creates a list with three
+components. The \code{data} component is a data frame containing
+additional preserved variables, the \code{stimulus} component is the
+stimulus class variable, and the \code{counts} component is an
 $N\times K$ matrix of combined response counts, where $N$ is the
@@
 
-A model is fitted using the \texttt{stan} function from the
-\texttt{rstan} package. This function requires a special list of data
-structures used by the model as well as a model specification
-expressed in the Stan language. Every model has some fixed effects
-structure, since even when there are no predictors the model
-parameters can be expressed as regressed on a vector of ones (i.e., an
-intercept). However, many models also have a hierarchical structure
-and, if that is the case, this hierarchical structure has to be
-specified when using the \texttt{make\_stan\_model} function. This is
-% ?? This is done by nie jest zbyt kolokwialne?
-done by providing a list of lists of R model formulae. Each list of
-model formulae is composed of at least three elements and specifies
-correlated random effects of one sampled factor. The element named
-\texttt{group} specifies the sampled factor and the elements named
-\texttt{delta} and \texttt{gamma} specify which effects are assumed to
-vary between the levels of this sampled factor. When
-% ?? a model, bo jest niedookreślony bez dodatkowej informacji z
-% make_stan_data?
-\texttt{make\_stan\_model} is used without any arguments it specifies
-% ?? <bez przedimka> Fixed effects model matrices nieokreślone -
-% czuję, że jestem niekonsekwentny w tych opisach
-a model without any random effects. Fixed effects model matrices are
-specified by providing a list with at least two model formulae, named
-\texttt{delta} and \texttt{gamma}, to the \texttt{make\_stan\_data}
-function, described later in this paper. Non-default priors can be
-% ?? czuję, że the random and fixed effects specification lists
-specified by adding optional elements to the random and fixed effects
-specification lists, as described in the \texttt{make\_stan\_data}
-function documentation.
-
-% ?? the participants?
+A model is fitted using the \code{stan} function from the \code{rstan}
+package. This function requires a special list of data structures used
+by the model as well as a model specification expressed in the Stan
+language. Every model has some fixed effects structure since, even
+when there are no predictors, the model parameters can be expressed as
+regressed on a vector of ones (i.e., an intercept). However, many
+models also have a hierarchical structure and, if that is the case,
+this hierarchical structure has to be specified when using the
+\code{make\_stan\_model} function. This is done by providing a list of
+lists of R model formulae. Each list of model formulae is composed of
+at least three elements and specifies the correlated random effects of
+one sampled factor. The \code{group} element specifies the sampled
+factor; the \code{delta} and \code{gamma} elements specify which
+effects are assumed to vary between the levels of this sampled
+factor. When \code{make\_stan\_model} is used without any arguments, it
+specifies a model without any random effects. Fixed effects model
+matrices are specified by providing a list with at least two model
+formulae, named \code{delta} and \code{gamma}, to the
+\code{make\_stan\_data} function that is described later in this
+paper. Non-default priors can be specified by adding optional elements
+to the random and fixed effects specification lists, as described in
+the \code{make\_stan\_data} function documentation.
+
 In the study in question there was only one sampled factor, i.e., the
-participants. Because duration was a within-participant variable in
+participants. Because duration was a within-participant variable, in
 principle its effect could vary between the participants for all the
@@
 the barely noticeable 32 ms difference in duration seemed to affect
-% ?? teraz nagle the sensitivity? jako parametr SDT to jest chyba
-% jednak the
 only the sensitivity. Thus, it was assumed that $\delta$ may depend on
 duration and order, but $\gamma$ may only be affected by
-order. Because duration was a within-subjects variable its effect on
+order. Because duration was a within-subjects variable, its effect on
 $\delta$ was assumed to vary between the participants, but the only
@@
 
-The \texttt{make\_stan\_data} function creates fixed and random effect
-% ?? dummy contrast coding jest niepoliczalne i nie chodzi nam o żadne
-% pojedyncze, tylko o ogólnie rozumiany sposób kodowania
+The \code{make\_stan\_data} function creates fixed and random effect
 model matrices based on the respective model formulae using dummy
 contrast coding. Note that the implicit intercept was suppressed for
-the $\delta$ model matrix. This way $\delta$ was estimated for every
-duration $\times$ order condition. The resulting separate intercepts
-and slopes parametrization makes it easier to calculate arbitrary
-% ?? contrasts ON posterior samples?
-contrasts on posterior samples. A more standard parametrization was
-used for the $\gamma$ parameter, because it was initially assumed that
-the criteria depend only on order, and so there was only one contrast
-of interest for every element of the $\gamma$ vector. On the other
-% ?? the separate ?
-hand, even in such cases the separate intercepts and slopes
-parametrization may be more convenient if a researcher is interested
-in the actual criteria, as we will later explain when introducing the
-\texttt{gamma\_to\_crit} function. This example also illustrates how
-the separation of the $\delta$ and $\gamma$ regression structures
-makes it possible to test a broad class of linear models representing
-% ?? the dependence bo zależność we wszystkich jej przejawach?
-the dependence of the SDT parameters on additional variables.
+the $\delta$ model matrix. In this way, $\delta$ was estimated for
+every duration $\times$ order condition. The resulting separate
+intercepts and slopes parametrization makes it easier to calculate
+arbitrary contrasts on posterior samples. A more standard
+parametrization was used for the $\gamma$ parameter because it was
+initially assumed that the criteria depend only on order, and so there
+was only one contrast of interest for every element of the $\gamma$
+vector. On the other hand, in such cases nested parametrization (with
+separate intercepts and slopes for every condition) may be more
+convenient if a researcher is interested in the actual criteria, as we
+will later explain when introducing the \code{gamma\_to\_crit}
+function. This example also illustrates how the separation of the
+$\delta$ and $\gamma$ regression structures makes it possible to test
+a broad class of linear models representing the dependence of the SDT
+parameters on additional variables.
 
@@
 
-In order to fit the model a separate data structure used by the Stan
-sampler has to be created using the \texttt{make\_stan\_data}
+In order to fit the model, a separate data structure used by the Stan
+sampler has to be created using the \code{make\_stan\_data}
 function. The obligatory arguments to this function are an aggregated
-data object created by the \texttt{aggregate\_responses} function and
-a fixed effects specification. Because in our example study duration
-was a within-subject variable here we also provide a random effects
-specification. Importantly, if random effects are modelled the same
-random effects specification has to be provided to the
-\texttt{make\_stan\_model} and \texttt{make\_stan\_data} functions.
+data object created by the \code{aggregate\_responses} function and a
+fixed effects specification. Importantly, if random effects are
+modelled, the same specification of random effects has to be provided
+to the \code{make\_stan\_model} and \code{make\_stan\_data} functions.
 
@@
 
-% ?? tutaj wydaje mi się oczywiste, że <bliżej nieokreślonych>
-% parameters of interest, a więc relevant variables wcześniej też
-% powinno być <nieokreślonych>?
 Finally, a vector of names of parameters of interest has to be
-specified when calling the \texttt{stan} function:
+specified when calling the \code{stan} function:
 
@@
 
-Note that since more than one sampled factor is allowed the names of
-hierarchical parameters are indexed (e.g., \texttt{delta\_sd\_1},
-\texttt{delta\_random\_1}, \texttt{Corr\_delta\_1}). The name
-\texttt{counts\_new} refers to posterior predictive samples that are
-required by the \texttt{plot\_sdt\_fit} function.
-% ?? Tu już mówimy ogólnie o parametrach o takich nazwach
-Names starting with \texttt{Corr} refer to random effects correlation
-matrices, which are computed from the Cholesky decompositions.
+Note that since more than one sampled factor is allowed, the names of
+all the hierarchical parameters are indexed (e.g., \code{delta\_sd\_1},
+\code{delta\_random\_1}, \code{Corr\_delta\_1}). The name
+\code{counts\_new} refers to posterior predictive samples that are
+required by the \code{plot\_sdt\_fit} function. Names starting with
+\code{Corr} refer to random effects correlation matrices, which are
+computed from Cholesky decompositions.
 
@@
 
-Four chains of 8000 iterations each were run simultaneously and the
-first half of the posterior samples serving as a warm-up period for
-tuning the parameters of the sampling algorithm was discarded. Part of
-the resulting Stan output is presented in Table~\ref{tab:fitsum}
-below.
+Four chains of 8,000 iterations each were run simultaneously; the
+first half of the posterior samples, which served as a warm-up period
+for tuning the parameters of the sampling algorithm, was
+discarded. Part of the resulting Stan output is presented in
+Table~\ref{fitsum} below.
 
@@
 \caption{Model fit summary statistics}
-\label{tab:fitsum}
 \begin{tabular}{rrrrrrrr}
-  \hline
+  % \hline
   & mean & $SE_{mean}$ & $SD$ & $2.5\%$ & $97.5\%$ & No. eff. samples & $\hat{R}$ \\ 
@@
   gamma\_sd\_1[5] & 1.55 & 0.00 & 0.17 & 1.24 & 1.91 & 4297 & 1.00 \\ 
-  gamma\_sd\_1[6] & 1.18 & 0.00 & 0.13 & 0.95 & 1.47 & 4348 & 1.00 \\ 
-  gamma\_sd\_1[7] & 0.78 & 0.00 & 0.10 & 0.60 & 1.00 & 4399 & 1.00 \\ 
-  Corr\_delta\_1[1,1] & 1.00 & 0.00 & 0.00 & 1.00 & 1.00 & 16000 &  \\ 
-  Corr\_delta\_1[1,2] & 0.76 & 0.00 & 0.10 & 0.52 & 0.91 & 3871 & 1.00 \\ 
-  Corr\_delta\_1[2,1] & 0.76 & 0.00 & 0.10 & 0.52 & 0.91 & 3871 & 1.00 \\ 
-  Corr\_delta\_1[2,2] & 1.00 & 0.00 & 0.00 & 1.00 & 1.00 & 15199 & 1.00 \\ 
-  Corr\_gamma\_1[1,1] & 1.00 & 0.00 & 0.00 & 1.00 & 1.00 & 16000 &  \\ 
-  Corr\_gamma\_1[1,2] & -0.20 & 0.01 & 0.27 & -0.66 & 0.39 & 840 & 1.00 \\ 
-  Corr\_gamma\_1[1,3] & -0.17 & 0.01 & 0.28 & -0.67 & 0.41 & 820 & 1.01 \\ 
-  Corr\_gamma\_1[1,4] & -0.04 & 0.01 & 0.28 & -0.56 & 0.51 & 840 & 1.00 \\ 
-  Corr\_gamma\_1[1,5] & -0.06 & 0.01 & 0.28 & -0.60 & 0.49 & 836 & 1.00 \\ 
-  Corr\_gamma\_1[1,6] & -0.27 & 0.01 & 0.27 & -0.75 & 0.32 & 838 & 1.00 \\ 
-  Corr\_gamma\_1[1,7] & -0.19 & 0.01 & 0.28 & -0.69 & 0.39 & 871 & 1.01 \\ 
-  Corr\_gamma\_1[2,1] & -0.20 & 0.01 & 0.27 & -0.66 & 0.39 & 840 & 1.00 \\ 
-  Corr\_gamma\_1[2,2] & 1.00 & 0.00 & 0.00 & 1.00 & 1.00 & 16000 & 1.00 \\ 
-  Corr\_gamma\_1[2,3] & 0.58 & 0.00 & 0.13 & 0.29 & 0.79 & 3716 & 1.00 \\ 
-  Corr\_gamma\_1[2,4] & 0.39 & 0.00 & 0.15 & 0.06 & 0.65 & 2806 & 1.00 \\ 
-  Corr\_gamma\_1[2,5] & 0.22 & 0.00 & 0.16 & -0.11 & 0.52 & 2712 & 1.00 \\ 
-  Corr\_gamma\_1[2,6] & 0.46 & 0.00 & 0.14 & 0.17 & 0.70 & 3968 & 1.00 \\ 
-  Corr\_gamma\_1[2,7] & 0.80 & 0.00 & 0.09 & 0.60 & 0.93 & 5171 & 1.00 \\ 
-  \dots{} & \dots{} & \dots{} & \dots{} & \dots{} & \dots{} & \dots{}
-                                                                       & \dots{} \\
+  \dots{} & \dots{} & \dots{} & \dots{} & \dots{} & \dots{} & \dots{} & \dots{} \\
   \hline
 \end{tabular}
+\label{fitsum}
 \end{table}
 
-As can be seen, the chains exhibited good mixing given the complexity
-of the model and seemed to have converged: there were enough effective
-samples of the fixed effect parameters to estimate $95\%$ credible
-intervals well and none of the Gelman-Rubin $\hat{R}$ statistics
-crossed the conventional $1.01$ threshold suggesting negligible
-sensitivity to the initial values.
+As can be seen, given the complexity of the model, the chains
+exhibited good mixing and seemed to have converged; there were enough
+effective samples of the fixed effect parameters to estimate $95\%$
+credible intervals well and none of the Gelman-Rubin $\hat{R}$
+statistics crossed the conventional $1.01$ threshold, suggesting
+negligible sensitivity to the initial values.
 
 Figs.~\ref{fig:6} and \ref{fig:7} below contain normal
-quantile-quantile plots of random effects. The plots indicate that
-normally distributed random $\delta$ and $\gamma$ effects seem to be
-good candidates for representing variability in the sensitivity and
-criteria parameters due to sampled factors.
+quantile-quantile plots of random effects. The plots indicate that the
+normally distributions of random $\delta$ and $\gamma$ effects can be
+approximated by normal distributions and that these parameters seem to
+be good candidates for representing variability in the sensitivity and
+criteria parameters due to the sampled factors.
 
@@
 Once enough good quality posterior samples are obtained for the
-parameters of interest the inference process can be carried out by
-calculating credible intervals, HPD intervals, or bayes factors for
+parameters of interest, the inference process can be carried out by
+calculating credible intervals, HPD intervals, or Bayes factors for
 any function of the parameters. However, even when the stan output
@@
 any further conclusions the researcher should first check if the model
-fits the data. The \texttt{plot\_sdt\_fit} function can be used for
-that purpose:
+fits the data. The \code{plot\_sdt\_fit} function can be used for this
+purpose:
 
@@
 
-This function requires at least three arguments; a stanfit object, an
-aggregated data list produced by the \texttt{aggregate\_responses}
-function and that was used to produce the stanfit object, and a vector
-of names of variables that will determine how the data will be
-partitioned before plotting. We recommend assessing the fit at the
-individual level, but we did not include the participant
-identification number in the list of conditioning variables, because
-the resulting plot would take up too much space.
-
-As can be seen in Fig.~\ref{fig:3} showing the ROC curves produced by
-the code above the model seemed to fit the data well in all but one
-condition (decision after rating, 64 ms duration) where three out of
-seven relevant\footnote{The point in the upper-right corner of a ROC
-  curve is always in the $(1,1)$ position} points were outside the
-two-dimensional 95\% posterior predictive regions.
+This function requires at least three arguments: a \code{stanfit}
+object, an aggregated data list produced by the
+\code{aggregate\_responses} function that was used to produce the
+stanfit object, and a vector of names of variables that will determine
+how the data will be partitioned before plotting. We recommend
+assessing the fit at the individual level, but we did not include the
+participant identification number in the list of conditioning
+variables because the resulting plot would take up too much space.
+
+As can be seen in Fig.~\ref{fig:3}, which shows the ROC curves
+produced by the code above, the model seemed to fit the data well in
+all but one condition (decision after rating, 64 ms duration), in
+which three out of seven relevant\footnote{The point in the
+  upper-right corner of a ROC curve is always in the $(1,1)$ position}
+points were outside the two-dimensional $95\%$ posterior predictive
+regions.
 
@@
 
-Another way to assess a model fit visually is by inspecting the
-% ?? to mają być conditional response distr. dla tego model fita
-conditional response distributions ($p(y|stim)$) such as those shown
+Another way to assess model fit visually is by inspecting the
+conditional response distributions ($p(y|stim)$), such as those shown
 in Fig.~\ref{fig:4}, which was also created using the
-\texttt{plot\_sdt\_fit} function.
+\code{plot\_sdt\_fit} function.
 
@@
 
-% ?? przedimek reasons? niby the model, bo dla plotu model jest już
-% określony, ale tu jest this kind of plot
 This kind of plot can be informative about the reasons why a model
-does not fit the data. In this particular case the plot seems to
+does not fit the data. In this particular case, the plot seems to
 suggest that it may be a good idea to inspect the fit at the
@@
 values in separate conditions, not differences between conditions or
-regression slopes. In our example, because separate intercepts and
-slopes parametrization was used for the $\delta$ fixed effects model
-matrix, all four \texttt{delta\_fixed} parameters can easily be
-transformed to sensitivities by applying the natural logarithm
-function. It is important to remember that, because the logarithm is a
-non-linear transformation, the $\delta$ to $d'$ conversion step should
-be done first, before
-% ?? nie wiem czemu the posterior samples, ale tak czuję
-applying any other transformations to the posterior samples, for
-example, the logarithm of a point and interval $\delta$ estimate is
-not equal to the point and interval $d'$ estimate calculated after
-transforming the $\delta$ posterior samples to the $d'$ samples.
-
-In this case the first column of the \texttt{gamma\_fixed} parameter
+regression slopes. In our example, because nested parametrization was
+used for the $\delta$ fixed effects model matrix, all four
+\code{delta\_fixed} parameters can easily be transformed to
+sensitivities by applying the exponential function. It is important to
+remember that because this is a non-linear transformation, the
+$\delta$ to $d'$ conversion step should be done first before applying
+any other transformations to the posterior samples; for example, the
+exponential of a point and interval estimate is not equal to the point
+and interval estimate calculated after transforming the $\delta$
+posterior samples to the $d'$ samples.
+
+In this case the first column of the \code{gamma\_fixed} parameter
 matrix (the intercept) corresponds to the values of the $\bm{\gamma}$
-vector in the DR condition but the second column corresponds to the
-\emph{effect} of order on $\bm{\gamma}$. For this reason in this
+vector in the DR condition, but the second column corresponds to the
+\emph{effect} of order on $\bm{\gamma}$. For this reason, in this
 particular case the posterior criteria samples can be obtained using
-the \texttt{gamma\_to\_crit} function only for the first column of the
-\texttt{gamma\_fixed} matrix.
+the \code{gamma\_to\_crit} function only for the first column of the
+\code{gamma\_fixed} matrix.
 
@@
 
-To test if the model correctly recovers known parameter values we have
+To test if the model correctly recovers known parameter values, we
 simulated the data from a hypothetical exact replication of the
@@
 correctly recovered in a sense that the true values were outside the
-95\% credible intervals no more than 5\% of the time.
-
-For illustration purposes an SDT model that differed from the true
+$95\%$ credible intervals no more than $5\%$ of the time.
+
+For illustration purposes, an SDT model that differed from the true
 model only in that it did not have any hierarchical structure was
@@
 was much simpler and the data consisted of only eight vectors of
-response counts the mixing of the chains was excellent.
-
-The 95\% credible intervals calculated for the fixed effects based on
-each model are compared in Fig.~\ref{fig:5} below. The estimates were
-centered on the true values to simplify the presentation. As can be
-seen, the true model correctly recovered the known parameter values,
-but the estimates based on the simplified, non-hierarchical model were
-biased; the $95\%$ credible intervals were not only about three times
-shorter on average but also failed to contain most of the true values.
+response counts, the mixing of the chains was excellent.
+
+The $95\%$ credible intervals calculated for the fixed effects based
+on each model are compared in Fig.~\ref{fig:5} below. The estimates
+were centered on the true values to simplify the presentation. As can
+be seen, the true model correctly recovered the known parameter
+values, but the estimates based on the simplified, non-hierarchical
+model were biased; the credible intervals were not only about three
+times shorter on average, but also failed to contain most of the true
+values.
 
@@
 
-However, as can be seen in Fig.~\ref{fig:8} below in this case the
+However, as can be seen in Fig.~\ref{fig:8} below, in this case the
 observed ROC curves seemed to fit the simplified model's predictions
@@
 The great importance of SDT to psychology stems from the fact that
-given weak assumptions about an underlying decision process it
-promises to deconfound sensitivity from bias in arbitrary
-classification tasks. To the best of our knowledge at present the
-\texttt{bhsdtr} package provides the only method of bayesian inference
+given weak assumptions about an underlying decision process, it
+promises to deconfound sensitivity from bias in arbitrary binary
+classification tasks. To the best of our knowledge, at present the
+\code{bhsdtr} package provides the only method of Bayesian inference
 for SDT models with or without ratings that can be recommended as a
 default choice in typical applications. Our parametrization forces the
-sensitivity to be non-negative and the criteria to be order-restricted
-while the isomorphisms between the $d'$ and $c$ parameters and the
-unconstrained $\delta$ and $\gamma$ parameters make it possible to
-supplement the SDT model with the general hierarchical linear
-regression structure. There is no limit to the number of sampled
-factors except for the one imposed by available computational
-resources, correlations of random effects of the same sampled factor
+sensitivity to be non-negative and the criteria to be
+order-restricted, while the isomorphisms between the $d'$ and $c$
+parameters and the unconstrained $\delta$ and $\gamma$ parameters make
+it possible to supplement the SDT model with the general hierarchical
+linear regression structure. There is no limit to the number of
+sampled factors except for the one imposed by available computational
+resources; correlations of random effects of the same sampled factor
 are accounted for, all the SDT parameters can be modelled by linear
-regression within the same model and all the effects on all the SDT
+regression within the same model, and all the effects on all the SDT
 parameters estimable within the levels of the sampled factors can have
-associated random effects. In case a need arises to relax a built-in
-restriction experienced users can extend the model in arbitrary ways
+associated random effects. If the need arises to relax a built-in
+restriction, experienced users can extend the model in arbitrary ways
 by using automatically generated human-readable Stan code as a
-% ?? basic understanding niepoliczalne
-template. We hope that researchers with basic understanding of Signal
-Detection Theory, bayesian inference, and hierarchical modelling will
-find our package useful and adopt it as a method of analysis of
-classification performance in their future studies.
+template. We hope that researchers with a basic understanding of
+Signal Detection Theory, Bayesian inference, and hierarchical
+modelling will find our package useful and adopt it as a method of
+analysis of binary classification performance in future studies.
 
 \bibliography{/home/borys/cs/literatura}
-\bibliographystyle{apacite}
+
 \end{document}

### 2019-03-12 inst/preprint/paper.tex
submitted version
Corrected author note in the manuscript

@@
 
-\keywords{Signal Detection Theory, Bayesian inference, hierarchical
-  models}
+\keywords{Signal Detection Theory, rating experiments, Bayesian
+  inference, hierarchical models}
 
@@
 
-\authornote{This research was supported in part by the National
-  Science Centre, Poland HARMONIA grant given to Michal Wierzchon
-  (2014/14/M/HS6/00911).
+\authornote{
+
+  % This research was supported in part by the National
+  % Science Centre, Poland HARMONIA grant given to Michal Wierzchon
+  % (2014/14/M/HS6/00911).
 

### 2019-03-13 inst/preprint/paper.tex
kosmetyczne zmiany w tekscie

@@
 However, because the SDT model is non-linear, variability in its
-parameters due to factors such as participants or items has to be
-accounted for, otherwise the estimates of SDT parameters are
-biased. As we explain later in this paper, neither of the two
-available methods of inference for hierarchical SDT models that we are
-aware of addresses this problem correctly. Hence, our main goal was to
-create a correct implementation of the general hierarchical linear
-regression structure defined on SDT parameters. We argue that the
-\code{bhsdtr} package for R \cite{rstatistical} provides exactly such
-an implementation and we have made it publicly available at
-\code{https://github.com/boryspaulewicz/bhsdtr}, together with the
-annotated source code that was used to perform all the analyses and
-produce all the figures presented in this paper.
+parameters due to factors which are usually study-specific and of no
+direct interest to the researcher such as participants or items has to
+be accounted for, otherwise the estimates of SDT parameters are
+biased. As we explain later in this paper, none of the available
+methods of inference for hierarchical SDT models that we are aware of
+addresses this problem correctly in it's generality. Hence, our main
+goal was to create a correct implementation of the general
+hierarchical linear regression structure defined on SDT parameters. We
+argue that the \code{bhsdtr} package for R \cite{rstatistical}
+provides exactly such an implementation and we have made it publicly
+available at \code{https://github.com/boryspaulewicz/bhsdtr}, together
+with the annotated source code that was used to perform all the
+analyses and produce all the figures presented in this paper.
 
@@
 studies, it is essential that it is based on a model equipped with the
-general hierarchical linear regression structure. The \code{bhsdtr}
-package meets this requirement thanks to a novel parametrization. We
-describe this novel parametrization in great detail and explain how
-reliance on relatively standard parametrization leads to problems in
-the two other available implementations. We end the first part of this
-paper with a formal definition of the model as implemented in
-\code{bhsdtr}. The second part contains an overview of the package and
-a tutorial in which we demonstrate how to use our method in
-practice. Before we go any further, however, a note on terminology
-seems in order.
+hierarchical linear regression structure. The \code{bhsdtr} package
+meets this requirement thanks to a novel parametrization. We describe
+this novel parametrization and explain how reliance on some popular
+parametrizations leads to problems in the two other available
+implementations. We end the first part of this paper with a formal
+definition of the model as implemented in \code{bhsdtr}. The second
+part contains an overview of the package and a tutorial in which we
+demonstrate how to use our method in practice. Before we go any
+further, however, a note on terminology seems in order.
 
@@
 new term, it is unambiguous and seems descriptively correct: the term
-"sampled factor" captures all the essential properties of such
+"sampled factor" seems to capture essential properties of such
 variables, i.e., a nominal scale, the fact that values are sampled
@@
 modelling a combined response $y$, which represents both the binary
-classification decision and rating. The value of $y$ increases with
-the strength of evidence in favor of the second stimulus class. For
-example, if confidence is rated on a four-point scale, then $y = 1$
-when a subject decides that a stimulus belongs to the first class with
-certainty $4$, and $y = 5$ when a subject decides that a stimulus
-belongs to the second class with certainty $1$. More formally, if $K$
-is the number of possible combined responses, then a subject is
-assumed to give response $k$ if $s \in (\bm{c}_{k-1},\bm{c}_k]$, where
-$\bm{c}$ are the decision criteria, with $\bm{c}_0$ and $\bm{c}_K$
-fixed at $-\infty$ and $+\infty$ respectively.
+classification decision and rating as a single number. The value of
+$y$ increases with the strength of evidence in favor of the second
+stimulus class. For example, if confidence is rated on a four-point
+scale, then $y = 1$ when a subject decides that a stimulus belongs to
+the first class with certainty $4$, and $y = 5$ when a subject decides
+that a stimulus belongs to the second class with certainty $1$. More
+formally, if $K$ is the number of possible combined responses, then a
+subject is assumed to give response $k$ if
+$s \in (\bm{c}_{k-1},\bm{c}_k]$, where $\bm{c}$ are the decision
+criteria, with $\bm{c}_0$ and $\bm{c}_K$ fixed at $-\infty$ and
+$+\infty$ respectively.
 
@@
 data and the model have the same dimensionality. This makes the
-generalization to the case particularly important, as it is only when
-$K > 2$ that the formal assumptions of the model (e.g., equal or
-unequal variance) can be tested\footnote{In contrast to the formal
-  assumptions, a psychological interpretation of the SDT parameters
-  can be tested even when ratings are not available, e.g., by means of
-  selective influence \cite{sternberg2001separate}}, which is often
-done by comparing observed and predicted ROC curves.
+generalization to the $K > 2$ case particularly important, as it is
+only when $K > 2$ that the formal assumptions of the model (e.g.,
+equal or unequal variance) can be tested\footnote{In contrast to the
+  formal assumptions, a psychological interpretation of the SDT
+  parameters can be tested even when ratings are not available, e.g.,
+  by means of selective influence \cite{sternberg2001separate}}, which
+is often done by comparing observed and predicted ROC curves.
 
 When SDT models are used in psychology studies, researchers are
-usually interested in relationships between SDT parameters and
-additional measured or manipulated variables; a good example is the
-dependence of $d'$ on stimulus strength. Also, in a great majority of
-psychology studies in which classification tasks are used, the data
-have a hierarchical structure, i.e., there are repeated measures for
-each participant or item, and participants or items are only samples
-from the target population. It seems worthwhile to explain why a
-general-purpose method of inference for SDT models should accommodate
-both kinds of situations.
+usually interested not in the values of the SDT parameters themselves,
+but in relationships between these parameters and additional measured
+or manipulated variables; a good example is the dependence of $d'$ on
+stimulus strength. Also, in a great majority of psychology studies in
+which classification tasks are used, the data have a hierarchical
+structure, i.e., there are repeated measures for each participant or
+item, and participants or items are only samples from the target
+population. A general-purpose method of inference for SDT models
+should accommodate both kinds of situations.
 
@@
 strictly speaking, it requires such demonstrations only in order to
-make a point about a specific case. Because SDT is a non-linear model,
-by definition, when estimates of its parameters are based on data
-aggregated over sampled factors, the expected values of these
-estimates are not in general true population averages. Such estimates
-are biased and inference about a target population based on them is
-simply not valid. To give a concrete example, consider two unbiased
-participants, one with $d'_1 = 2$ and one with $d'_2 = 4$. Their
-expected average accuracy is given by
+illustrate a mathematical truth in a specific case. Because SDT is a
+non-linear model, by definition, when estimates of its parameters are
+based on data aggregated over sampled factors (e.g., $d'$ estimated
+for hits and false alarms averaged over subjects), the expected values
+of these estimates (e.g., what the calculated $d'$ actually estimates)
+are not in general true population averages (e.g., true $d'$ in some
+condition). Such estimates are biased and inference about a target
+population based on them is simply not valid. To give a concrete
+example, consider two unbiased participants, one with $d'_1 = 2$ and
+one with $d'_2 = 4$. Their expected average accuracy is given by
 $(\Phi(d'_1/2) + \Phi(d'_2/2)) / 2 = .91$, which corresponds to
@@
 SDT parameters are estimated separately for each subject and
-condition, and only later regressed on predictors of interest, a
-number of additional issues may arise.
+condition, and only later are these estimates regressed on predictors
+of interest, a number of additional issues may arise.
 
@@
 The aforementioned problems can be dealt with by supplementing the SDT
-model with the general hierarchical linear regression structure. For
-reasons that we will now explain, this can only be achieved if the
-model is substantially reparametrized.
+model with the hierarchical linear regression structure. This can only
+be achieved if the model is substantially reparametrized.
 
@@
 criterion, the elements of the $\bm{c}$ vector are order restricted
-($\bm{c}_{i+1} > \bm{c}_i$). Because normal distribution is unbounded,
-the hierarchical linear regression structure in which random effects
-are normally distributed and fixed effects are represented by
-unconstrained parameters can only be defined on unconstrained
-parameters. We provide examples of the problems that may otherwise
-arise in the following summary of hierarchical SDT implementations.
+($\bm{c}_{i+1} > \bm{c}_i$). Hierarchical linear regression structure
+can only be defined on unconstrained parameters, because random
+effects are assumed to be normally distributed and normal distribution
+is unbounded and because fixed effects are represented by
+unconstrained parameters. We provide examples of the problems that may
+arise when the SDT model is not correctly reparametrized in the
+following summary of two hierarchical SDT implementations.
 
-We are aware of only two attempts at creating a general-purpose method
-of inference for hierarchical SDT models with ratings. One is the
-Gibbs sampler proposed by \citeA{morey2008problematic} and the other
-is the Hierarchical Meta-d' model (HMeta-d') proposed by
+We are aware of two failed attempts at creating a general-purpose
+method of inference for hierarchical SDT models with ratings. One is
+the Gibbs sampler proposed by \citeA{morey2008problematic} and the
+other is the Hierarchical Meta-d' model (HMeta-d') proposed by
 \citeA{hmetad}. The HMeta-d' model is a hierarchical version of the
@@
 of the criteria vector cannot be affected differently by the same
-sampled factor, which is an unrealistic restriction.
+sampled factor, which is an unrealistic restriction. Participants
+differ in how they place the criteria just as they differ in their
+sensitivity and it is quite impossible for all the participants to
+place the criteria in the same distance from each other.
 
 In HMeta-d' the hierarchical structure is restricted to normally
-distributed random intercepts of one sampled factor. Perhaps the most
-problematic aspect of the HMeta-d' model, apart from the fact that the
-$d'$ parameter is allowed to assume negative values in this model
-also, is the representation of the criteria. Each element of the
-criteria vector has an associated independent normal distribution,
-which does allow for criteria random effects, but does not enforce the
-necessary ordering restriction. The elements of this vector are sorted
-to obtain another criteria vector, and it is this sorted criteria
-vector that is used to model the conditional combined response
+distributed random intercepts of one sampled factor. In the HMeta-d'
+model the $d'$ parameter is allowed to assume negative values also,
+but the most problematic aspect of this implementation is again the
+representation of the criteria. Each element of the criteria vector
+has an associated independent normal distribution, which does allow
+for criteria random effects, but does not enforce the necessary
+ordering restriction. The elements of this vector are sorted to obtain
+another criteria vector, and it is this sorted criteria vector that is
+used to model the conditional combined response
 distributions. Consequently, the model does contain parameters
@@
 preventing the outermost areas under the mapping distribution curve
-from becoming very small, especially when the criteria are widely
-spread, as can happen for moderate to large $d'$ values. This feature
-is implemented in the \code{bhsdtr} package by introducing a scaling
-factor.
+from becoming very small relative to ares delineated by adjacent pairs
+of non-outermost criteria. This is especially important when the
+criteria are widely spread, as can happen for moderate to large $d'$
+values. This feature is implemented in the \code{bhsdtr} package by
+introducing a criteria scaling factor.
 
@@
 
-The \code{bhsdtr} package implements the model described in the
-previous section in the Stan modelling language because it uses a
-state-of-the-art adaptive Hamiltonian Monte Carlo sampling algorithm
-which often handles high-dimensional correlated posteriors better than
-a Gibbs sampler. Our package is essentially a collection of documented
-functions: The \code{aggregate\_responses} function aggregates data as
-much as possible for efficiency, but without distorting the
-hierarchical structure. The \code{make\_stan\_model} function creates a
-model definition in the Stan language. The Stan code produced by the
-\code{make\_stan\_model} function can be fitted as is or modified by the
-user if needed, e.g., to change the prior distributions or to drop the
-equal variance assumption. The \code{make\_stan\_data} function creates
-regression model matrices and other data structures required by the
-model created using the \code{make\_stan\_model} function. Finally, the
-\code{plot\_sdt\_fit} function can be used to visually assess the fit of
-the model by creating publication-ready ROC curve plots or response
-distribution plots with posterior predictive intervals calculated for
-the chosen $\alpha$ level.
+The \code{bhsdtr} package implements the model in the Stan modelling
+language because it uses a state-of-the-art adaptive Hamiltonian Monte
+Carlo sampling algorithm which often handles high-dimensional
+correlated posteriors better than a Gibbs sampler. Our package is
+essentially a collection of documented functions: The
+\code{aggregate\_responses} function aggregates data as much as
+possible for efficiency, but without distorting the hierarchical
+structure. The \code{make\_stan\_model} function creates a model
+definition in the Stan language. The Stan code produced by the
+\code{make\_stan\_model} function can be fitted as is or modified by
+the user if needed, e.g., to change the prior distributions or to drop
+the equal variance assumption. The \code{make\_stan\_data} function
+creates regression model matrices and other data structures required
+by the model created using the \code{make\_stan\_model}
+function. Finally, the \code{plot\_sdt\_fit} function can be used to
+visually assess the fit of the model by creating publication-ready ROC
+curve plots or response distribution plots with posterior predictive
+intervals calculated for the chosen $\alpha$ level.
 
@@
 
-To make full use of the bhsdtr package functionality, three
+To make full use of the \code{bhsdtr} package functionality, three
 non-standard R packages are required, namely \code{rstan},
@@
 \code{devtools} package to install the \code{bhsdtr} package directly
-from the github repository. This will automatically install any
-missing required packages:
+from the github repository. This will automatically install the
+package and any missing required packages:
 
@@
 other sampled factors present in the data (e.g., items, replications,
-etc.), then these factors also have to be specified at this stage to
-preserve the hierarchical data structure. The
-\code{aggregate\_responses} function creates a list with three
-components. The \code{data} component is a data frame containing
-additional preserved variables, the \code{stimulus} component is the
-stimulus class variable, and the \code{counts} component is an
-$N\times K$ matrix of combined response counts, where $N$ is the
-number of data points and $K$ is the number of possible combined
-response values.
+etc.) and the user decided to model the effects of these factors, then
+these factors also have to be specified at this stage to preserve the
+hierarchical data structure. The \code{aggregate\_responses} function
+creates a list with three components. The \code{data} component is a
+data frame containing additional preserved variables, the
+\code{stimulus} component is the stimulus class variable, and the
+\code{counts} component is an $N\times K$ matrix of combined response
+counts, where $N$ is the number of data points and $K$ is the number
+of possible combined response values.
 

### 2020-01-13 inst/preprint/paper.tex
po recenzjach 1, zmiana gabor.RData

@@
   as additional ratings and an arbitrary number of nested or crossed
-  random grouping factors. The SDT parameters can be regressed on
-  additional predictors within the same model via intermediate
-  unconstrained parameters, and the model can be extended by using
-  automatically generated human-readable Stan code as a template. In
-  the paper we explain how our method improves on other similar
-  available methods, we give an overview of the package, demonstrate
-  its ease of use by providing a real-study data analysis
-  walk-through, and show that the model successfully recovers known
-  parameter values when fitted to simulated data.}
+  random grouping factors associated with sensitivity or criteria
+  effects. The SDT parameters can be regressed on additional
+  predictors within the same model via intermediate unconstrained
+  parameters, and the model can be extended by using automatically
+  generated human-readable Stan code as a template. In the paper we
+  explain how our method improves on other similar available methods,
+  we give an overview of the package, demonstrate its ease of use by
+  providing a real-study data analysis walk-through, and show that the
+  model successfully recovers known parameter values when fitted to
+  simulated data.}
 
@@
 \begin{document}
-\maketitle
-Many tasks used in psychology studies are essentially classification
-tasks. For example, in a memory study participants may be required to
-decide if a given test item is old or new, or, in a perceptual study,
-an object may be either a letter or a digit. If a task requires
-classification, it is always possible that conclusions based on
-accuracy or percent correct are invalid because the ability to
-discriminate between stimulus classes (i.e., sensitivity) is
+\maketitle Many tasks used in psychology studies are essentially
+classification tasks. For example, in a memory study participants may
+be required to decide if a given test item is old or new, or, in a
+perceptual study, an object may be either a letter or a digit. If a
+task requires classification, it is always possible that conclusions
+based on accuracy or percent correct are invalid because the ability
+to discriminate between stimulus classes (i.e., sensitivity) is
 confounded with bias, which is a tendency to classify stimuli as
@@
 may reflect differences in sensitivity, bias, or both. Signal
-Detection Theory provides a simple and popular solution to this common
-and important problem.
+Detection Theory provides a popular solution to this common problem.
 
@@
 direct interest to the researcher such as participants or items has to
-be accounted for, otherwise the estimates of SDT parameters are
-biased. As we explain later in this paper, none of the available
-methods of inference for hierarchical SDT models that we are aware of
-addresses this problem correctly in it's generality. Hence, our main
-goal was to create a correct implementation of the general
-hierarchical linear regression structure defined on SDT parameters. We
-argue that the \code{bhsdtr} package for R \cite{rstatistical}
-provides exactly such an implementation and we have made it publicly
-available at \code{https://github.com/boryspaulewicz/bhsdtr}, together
-with the annotated source code that was used to perform all the
-analyses and produce all the figures presented in this paper.
+be accounted for. When they are not accounted for, e.g., by
+aggregating the data accros participants or stimuli, the estimates of
+SDT parameters are biased. As we explain later in this paper, none of
+the available methods of inference for hierarchical SDT models that we
+are aware of addresses this problem correctly in it's full generality,
+meaning that none of the available methods allows for fixed and random
+effects in both the sensitivity and the criteria parameters while
+restricting the parameters in accordance with SDT assumptions. Later
+in the paper we explain why the \code{bhsdtr} package for R
+\cite{rstatistical}, which we have made publicly available at
+\code{https://github.com/boryspaulewicz/bhsdtr}, provides a correct
+implementation of the general hierarchical linear regression structure
+defined on SDT parameters. The package repository also contains the
+annotated source code that was used to perform all the analyses and
+produce all the figures presented in this paper.
 
@@
 Fig.~\ref{basicsdt}, the two evidence distributions are normal with
-the same variance, which is usually fixed at unity to make the model
+the same variance, often fixed at unity to make the model
 identifiable. The distance $d'$ between the means of the evidence
@@
 class (e.g., a new item) if $s \geq c$. The location of the decision
-criterion represents bias.
+criterion may be interpreted in terms of bias.
 
@@
 percent correct scores; if the model is correct, the resulting
-performance measure is not contaminated by bias. Obviously, the model
-may not be correct, which is one of the reasons why we focus more on
-the generalized version shown in Fig.~\ref{sdtr} below.
+performance measure is not contaminated by bias. The model may not be
+correct, which in our view is the most important reason to focus more
+on the generalized version shown in Fig.~\ref{sdtr} below.
 
@@
 
-Ratings are accommodated by introducing additional criteria and
+Ratings are accommodated by introducing additional criteria and by
 modelling a combined response $y$, which represents both the binary
@@
 formally, if $K$ is the number of possible combined responses, then a
-subject is assumed to give response $k$ if
-$s \in (\bm{c}_{k-1},\bm{c}_k]$, where $\bm{c}$ are the decision
-criteria, with $\bm{c}_0$ and $\bm{c}_K$ fixed at $-\infty$ and
-$+\infty$ respectively.
+subject is assumed to give response $k$ if $s \in
+(\bm{c}_{k-1},\bm{c}_k]$, where $\bm{c}$ are the decision criteria,
+  with $\bm{c}_0$ and $\bm{c}_K$ fixed at $-\infty$ and $+\infty$
+  respectively.
 
@@
 SDT model from Fig.~\ref{sdtr}, even when neither the ratings nor the
-placement of criteria are relevant to the research problem. When
-$K = 2$ (no ratings), the SDT model fits perfectly, regardless of
-whether it is a reasonably good approximation to reality, because the
-data and the model have the same dimensionality. This makes the
-generalization to the $K > 2$ case particularly important, as it is
-only when $K > 2$ that the formal assumptions of the model (e.g.,
-equal or unequal variance) can be tested\footnote{In contrast to the
-  formal assumptions, a psychological interpretation of the SDT
-  parameters can be tested even when ratings are not available, e.g.,
-  by means of selective influence \cite{sternberg2001separate}}, which
-is often done by comparing observed and predicted ROC curves.
+placement of criteria are relevant to the research problem. When $K =
+2$ (no ratings), the SDT model fits perfectly, regardless of whether
+it is a reasonably good approximation to reality, because the data and
+the model have the same dimensionality. This makes the generalization
+to the $K > 2$ case (i.e., decision with ratings) particularly
+important, as it is only when $K > 2$ that the formal assumptions of
+the model (e.g., equal or unequal variance) can be tested\footnote{In
+  contrast to the formal assumptions, a psychological interpretation
+  of the SDT parameters can be tested even when ratings are not
+  available, e.g., by means of selective influence
+  \cite{sternberg2001separate}}, which is often done by comparing
+observed and predicted ROC curves.
 
@@
 simulational studies by \citeA{morey2008problematic}; however,
-strictly speaking, it requires such demonstrations only in order to
-illustrate a mathematical truth in a specific case. Because SDT is a
-non-linear model, by definition, when estimates of its parameters are
-based on data aggregated over sampled factors (e.g., $d'$ estimated
-for hits and false alarms averaged over subjects), the expected values
-of these estimates (e.g., what the calculated $d'$ actually estimates)
+strictly speaking, such demonstrations are irrelevant to prooving the
+invalidity in question. Because SDT is a non-linear model, by
+definition, when estimates of its parameters are based on data
+aggregated over sampled factors (e.g., $d'$ estimated for hits and
+false alarms averaged or summed over subjects), the expected values of
+these estimates (e.g., what the calculated $d'$ actually estimates)
 are not in general true population averages (e.g., true $d'$ in some
 condition). Such estimates are biased and inference about a target
-population based on them is simply not valid. To give a concrete
-example, consider two unbiased participants, one with $d'_1 = 2$ and
-one with $d'_2 = 4$. Their expected average accuracy is given by
-$(\Phi(d'_1/2) + \Phi(d'_2/2)) / 2 = .91$, which corresponds to
-$d' = 2.68$, whereas their true average $d'$ is $3$.
+population based on them is not valid. To give a concrete example,
+consider two unbiased participants, one with $d'_1 = 2$ and one with
+$d'_2 = 4$. Their expected average accuracy is given by $(\Phi(d'_1/2)
++ \Phi(d'_2/2)) / 2 = .91$, which corresponds to $d' = 2.68$, whereas
+their true average $d'$ is $3$.
 
@@
 assuming that random effects are samples from a common distribution
-may improve their estimates.
+may improve their estimates. The aforementioned problems can be dealt
+with by supplementing the SDT model with the hierarchical linear
+regression structure.
 
-The aforementioned problems can be dealt with by supplementing the SDT
-model with the hierarchical linear regression structure. This can only
-be achieved if the model is substantially reparametrized.
-
-\section{Hierarchical Signal Detection Theory in a constrained
-  parameter space}
+\section{Comparison with other software implementations}
+% \section{Hierarchical Signal Detection Theory in a constrained
+%   parameter space}
 
@@
 arise when the SDT model is not correctly reparametrized in the
-following summary of two hierarchical SDT implementations.
+following summary of two hierarchical SDT implementations. One is the
+Gibbs sampler proposed by \citeA{morey2008problematic} and the other
+is the Hierarchical Meta-d' model (HMeta-d') proposed by
+\citeA{hmetad}. After describing the problems associated with those
+two software implementations we will briefly explain how our method
+compares to what the otherwise excellent \code{brms} package has to
+offer.
 
-We are aware of two failed attempts at creating a general-purpose
-method of inference for hierarchical SDT models with ratings. One is
-the Gibbs sampler proposed by \citeA{morey2008problematic} and the
-other is the Hierarchical Meta-d' model (HMeta-d') proposed by
-\citeA{hmetad}. The HMeta-d' model is a hierarchical version of the
+The HMeta-d' model is a hierarchical version of the
 meta-d' model \cite{maniscalco2012signal}, which in turn is a
@@
 the first. The authors explicitly admit that their algorithm does not
-enforce this restriction because, as they claim in the paper, it would
-greatly complicate analysis. At the same time the authors fail to
-mention that an immediate consequence of allowing for negative $d'$
-values is that the resulting posterior distribution loses its intended
-interpretation since it can have a non-zero mass when $d' < 0$. The
-outermost criteria are fixed at $0$ and $1$, and the ordering
-restriction is enforced by assuming that the likelihood is $0$
-whenever $\bm{c}_{i+1} \leq \bm{c}_i$. As the authors explain, because
-a sampled factor can have independent random effects on the evidence
-distribution means, it can have an effect on all the criteria:
-shifting both means by the same amount in the same direction is
-equivalent to keeping the sensitivity intact, while shifting the
-criteria relative to the evidence distributions. However, the elements
-of the criteria vector cannot be affected differently by the same
-sampled factor, which is an unrealistic restriction. Participants
-differ in how they place the criteria just as they differ in their
-sensitivity and it is quite impossible for all the participants to
-place the criteria in the same distance from each other.
+enforce this restriction because, as they say in the paper, it would
+greatly complicate analysis, although nothing more is said about these
+complications. At the same time the authors fail to mention that an
+immediate consequence of allowing for negative $d'$ values is that the
+resulting posterior distribution loses its intended interpretation
+since it can have a non-zero mass when $d' < 0$. The outermost
+criteria are fixed at $0$ and $1$, and the ordering restriction is
+enforced by assuming that the likelihood is $0$ whenever $\bm{c}_{i+1}
+\leq \bm{c}_i$. As the authors explain, because a sampled factor can
+have independent random effects on the evidence distribution means, it
+can have an effect on all the criteria: shifting both means by the
+same amount in the same direction is equivalent to keeping the
+sensitivity intact, while shifting the criteria relative to the
+evidence distributions. However, the elements of the criteria vector
+cannot be affected differently by the same sampled factor, which is an
+unrealistic restriction: participants differ in how they place the
+criteria just as they differ in their sensitivity.
 
@@
 unrestricted criteria vectors that are associated with the
-hierarchical structure.
+hierarchical structure. This makes posterior distributions of random
+criteria effects uninterpretable.
+
+% PORÓWNANIE Z brms
+Some extensions of the SDT model can be fitted correctly using the
+excellent brms package, as described in \citeA{burkner2017brms}. The
+brms package is a flexible tool that shares some deep design
+similarities with our method. Both our package and the brms belong to
+a growing family of software tools that aim to provide a somewhat
+simplified and domain or application specific interface to one of the
+general purpose bayesian inference engines, in this case the stan
+modelling language \cite{carpenter2016stan}. The \code{brms} is a
+highly flexible, well documented package that offers an elegant
+interface for fitting a large class of mixed models. Among the models
+that can be fitted using this package are the ordinal regression
+models. The hierarchical SDT model with ratings is essentially a
+generalization of mixed ordinal regression, since ratings are an
+ordinal-scale variable. However, that does not mean that a general
+hierarchical SDT model with ratings can be fitted using the
+\code{brms} package. There are three categorical distributions
+available at present in brms, i.e., the cumulative model, the adjacent
+category model and the sequential model, each with it's own set of
+link functions. As the author of the package explains
+\cite{burkner2017brms}, the only model that respects the ordering of
+the thresholds is the cumulative model. The thresholds in the
+cumulative model cannot be regressed on additional variables,
+consequently they cannot be associated with random effects.
+
+Another important difference between our method and the \code{brms}
+package is that \code{brms} does not utilize the fact that for ordinal
+regression models the data can often be significantly aggregated
+without loosing any information. For example, if a basic SDT model
+with ratings was to be fitted to an arbitrary large dataset using our
+method the aggregated dataset would consist of a matrix of dimension
+$2\times K$. In that case each iteration of the sampling algorithm
+implemented in stan will consist of essentially 2 computations of the
+likelihood function, whereas without any aggregation step the number
+of such computations will be a linear function of the sample
+size. This is especially problematic for high-dimensional non-linear
+models such as hierarchical SDT with ratings, because they require
+relatively large datasets (i.e., many data points per participant
+$\times$ condition).
 
@@
 mapping device, not a third evidence distribution, and that, for the
-reasons that will soon become clear, it is wider than the two evidence
-distributions. The mapping expressed by Eq.~\ref{eq:1} is an
+reasons that will be explained later, it is wider than the two
+evidence distributions. The mapping expressed by Eq.~\ref{eq:1} is an
 isomorphism between the $R^{K-1}$ space and the space of
@@
 
-The formal definition of a Bayesian model is not complete without
-providing fixed values of all the parameters that define prior
-distributions. Specifying the priors on sensitivity effects does not
-pose any special difficulties. The sensitivity of an unbiased
-classifier given percent correct (pc) is given by
-$2 \Phi^{-1}(\text{pc})$. When $p(stim = 1) = 0.5$, the greater the
-bias, the lower the accuracy, meaning that an unbiased sensitivity is
-a lower bound on sensitivity given percent correct. Let us assume that
-the majority of subjects are expected to achieve percent correct
-within the $.51$ to $.99$ range, with negligible bias. Since
-$\ln(2\Phi^{-1}(.51)) = -2.99$ and $\ln(2 \Phi^{-1}(.99)) = 1.54$, a
-reasonable weakly informative prior on $\delta$ is normal with mean
-$(1.54 - 2.99) / 2$ and standard deviation $(1.54 + 2.99) / 2$, which
-is the default prior on delta effects in the \code{bhsdtr} package.
+% BOOTSTRAPOWANIE PRIORÓW NA PODST VARIACYJNEGO BAYESA
+
+A Bayesian model is not complete without providing fixed values of all
+the parameters that define prior distributions. Specifying the priors
+on sensitivity effects does not pose any special difficulties. The
+sensitivity of an unbiased classifier given percent correct (pc) is
+given by $2 \Phi^{-1}(\text{pc})$. When $p(stim = 1) = 0.5$, the
+greater the bias, the lower the accuracy, meaning that an unbiased
+sensitivity is a lower bound on sensitivity given percent correct. Let
+us assume that the majority of subjects are expected to achieve
+percent correct within the $.51$ to $.99$ range, with negligible
+bias. Since $\ln(2\Phi^{-1}(.51)) = -2.99$ and
+$\ln(2 \Phi^{-1}(.99)) = 1.54$, a reasonable weakly informative prior
+on $\delta$ is normal with mean $(1.54 - 2.99) / 2$ and standard
+deviation $(1.54 + 2.99) / 2$, which is the default prior on delta
+effects in the \code{bhsdtr} package.
 
@@
 The \code{bhsdtr} package implements the model in the Stan modelling
-language because it uses a state-of-the-art adaptive Hamiltonian Monte
-Carlo sampling algorithm which often handles high-dimensional
-correlated posteriors better than a Gibbs sampler. Our package is
-essentially a collection of documented functions: The
+language because the stan sampler uses a state-of-the-art adaptive
+Hamiltonian Monte Carlo algorithm which often handles high-dimensional
+correlated posteriors better than a Gibbs sampler.
+
+Our package is essentially a collection of documented functions: The
 \code{aggregate\_responses} function aggregates data as much as
@@
 
-We will now explain how to perform some of the essential steps of a
-typical data analysis process, which at the very least will usually
+The essential steps of a typical data analysis process will usually
 involve preparing the data, creating the model code, fitting the
@@
 by the model as well as a model specification expressed in the Stan
-language. Every model has some fixed effects structure since, even
-when there are no predictors, the model parameters can be expressed as
-regressed on a vector of ones (i.e., an intercept). However, many
-models also have a hierarchical structure and, if that is the case,
-this hierarchical structure has to be specified when using the
+language.
+
+Every model has some fixed effects structure since, even when there
+are no predictors, the model parameters can be expressed as regressed
+on a vector of ones (i.e., an intercept). However, many models also
+have a hierarchical structure and, if that is the case, this
+hierarchical structure has to be specified when using the
 \code{make\_stan\_model} function. This is done by providing a list of
@@
 effects are assumed to vary between the levels of this sampled
-factor. When \code{make\_stan\_model} is used without any arguments, it
-specifies a model without any random effects. Fixed effects model
+factor. When \code{make\_stan\_model} is used without any arguments,
+it specifies a model without any random effects. Fixed effects model
 matrices are specified by providing a list with at least two model
@@
 SDT parameters. However, a preliminary data analysis indicated that
-the barely noticeable 32 ms difference in duration seemed to affect
-only the sensitivity. Thus, it was assumed that $\delta$ may depend on
-duration and order, but $\gamma$ may only be affected by
-order. Because duration was a within-subjects variable, its effect on
-$\delta$ was assumed to vary between the participants, but the only
-random effect associated with $\gamma$ was the intercept:
+the 32 ms difference in duration seemed to affect only the
+sensitivity. Thus, it was assumed that $\delta$ may depend on duration
+and order, but $\gamma$ may only be affected by order. Because
+duration was a within-subjects variable, its effect on $\delta$ was
+assumed to vary between the participants, but the only random effect
+associated with $\gamma$ was the intercept.
 
@@
 model matrices based on the respective model formulae using dummy
-contrast coding. Note that the implicit intercept was suppressed for
-the $\delta$ model matrix. In this way, $\delta$ was estimated for
-every duration $\times$ order condition. The resulting separate
-intercepts and slopes parametrization makes it easier to calculate
-arbitrary contrasts on posterior samples. A more standard
-parametrization was used for the $\gamma$ parameter because it was
-initially assumed that the criteria depend only on order, and so there
-was only one contrast of interest for every element of the $\gamma$
-vector. On the other hand, in such cases nested parametrization (with
-separate intercepts and slopes for every condition) may be more
+contrast coding. Note that the implicit intercept was removed for the
+$\delta$ model matrix (the $-1$ term on the right hand side of the
+model formula). In this way, $\delta$ was estimated for every duration
+$\times$ order condition. The resulting separate intercepts and slopes
+parametrization makes it easier to calculate arbitrary contrasts on
+posterior samples. A more standard parametrization was used for the
+$\gamma$ parameter because it was initially assumed that the criteria
+depend only on order, which is a two-level factor, and so there was
+only one contrast of interest for every element of the $\gamma$
+vector. On the other hand, even in such cases the parametrization with
+separate intercepts and slopes for every condition may be more
 convenient if a researcher is interested in the actual criteria, as we
@@
 Note that since more than one sampled factor is allowed, the names of
-all the hierarchical parameters are indexed (e.g., \code{delta\_sd\_1},
-\code{delta\_random\_1}, \code{Corr\_delta\_1}). The name
-\code{counts\_new} refers to posterior predictive samples that are
-required by the \code{plot\_sdt\_fit} function. Names starting with
-\code{Corr} refer to random effects correlation matrices, which are
-computed from Cholesky decompositions.
+all the hierarchical parameters are indexed (e.g.,
+\code{delta\_sd\_1}, \code{delta\_random\_1},
+\code{Corr\_delta\_1}). The name \code{counts\_new} refers to
+posterior predictive samples that are required by the
+\code{plot\_sdt\_fit} function. Names starting with \code{Corr} refer
+to random effects correlation matrices.
 
@@
 
-Four chains of 8,000 iterations each were run simultaneously; the
-first half of the posterior samples, which served as a warm-up period
-for tuning the parameters of the sampling algorithm, was
-discarded. Part of the resulting Stan output is presented in
-Table~\ref{fitsum} below.
+As can be seen in the previous code fragment four chains of 8,000
+iterations each were run simultaneously; the first half of the
+posterior samples, which served as a warm-up period for tuning the
+parameters of the sampling algorithm, was discarded. Part of the
+resulting Stan output is presented in Table~\ref{fitsum} below.
 
@@
 
-As can be seen, given the complexity of the model, the chains
-exhibited good mixing and seemed to have converged; there were enough
-effective samples of the fixed effect parameters to estimate $95\%$
-credible intervals well and none of the Gelman-Rubin $\hat{R}$
-statistics crossed the conventional $1.01$ threshold, suggesting
-negligible sensitivity to the initial values.
+Given the complexity of the model the chains exhibited good mixing and
+seemed to have converged; there were enough effective posterior
+samples for the fixed effect parameters to estimate $95\%$ credible
+intervals well and none of the Gelman-Rubin $\hat{R}$ statistics
+crossed the conventional $1.01$ threshold, suggesting negligible
+sensitivity to the initial values.
 
@@
 quantile-quantile plots of random effects. The plots indicate that the
-normally distributions of random $\delta$ and $\gamma$ effects can be
-approximated by normal distributions and that these parameters seem to
-be good candidates for representing variability in the sensitivity and
-criteria parameters due to the sampled factors.
+distributions of random $\delta$ and $\gamma$ effects can be
+approximated by normal distributions and that -- at least in this
+particular example -- these parameters seem to be good candidates for
+representing variability in the sensitivity and criteria parameters
+due to the sampled factors.
 
@@
 parameters of interest, the inference process can be carried out by
-calculating credible intervals, HPD intervals, or Bayes factors for
-any function of the parameters. However, even when the stan output
-summary does not indicate sampler convergence issues, before drawing
-any further conclusions the researcher should first check if the model
-fits the data. The \code{plot\_sdt\_fit} function can be used for this
-purpose:
+calculating credible intervals or HPD intervals for any function of
+the parameters or Bayes factors for each parameter
+separately. However, even when the stan output summary does not
+indicate sampler convergence issues, before drawing any further
+conclusions the researcher should first check if the model fits the
+data. The \code{plot\_sdt\_fit} function can be used for this purpose:
 
@@
 how the data will be partitioned before plotting. We recommend
-assessing the fit at the individual level, but we did not include the
-participant identification number in the list of conditioning
-variables because the resulting plot would take up too much space.
+assessing the fit at the individual level by including the participant
+identification number in the list of conditioning variables, but we
+did not do it here because the resulting plot would take up too much
+space.
 
@@
 
-This kind of plot can be informative about the reasons why a model
-does not fit the data. In this particular case, the plot seems to
-suggest that it may be a good idea to inspect the fit at the
-individual level and see if there are some participants with unusual
-$p(y|stim=1)$ distributions in the RD $\times$ 64 ms condition. On the
-other hand, it is also possible that the lack of fit is mainly a
-consequence of the assumption that duration had zero effect on
-$\bm{\gamma}$, or that more substantial modifications are necessary,
-such as dropping the equal variance assumption.
+Both plots can be informative about the reasons why a model does not
+fit the data. In this particular case, the plot seems to suggest that
+it may be a good idea to inspect the fit at the individual level and
+see if there are some participants with unusual $p(y|stim=1)$
+distributions in the RD $\times$ 64 ms condition. On the other hand,
+it is also possible that the lack of fit is mainly a consequence of
+the assumption that duration had zero effect on $\bm{\gamma}$, or that
+more substantial modifications are necessary, such as dropping the
+equal variance assumption.
 
@@
 straightforward only when fixed effects represent average parameter
-values in separate conditions, not differences between conditions or
-regression slopes. In our example, because nested parametrization was
-used for the $\delta$ fixed effects model matrix, all four
-\code{delta\_fixed} parameters can easily be transformed to
-sensitivities by applying the exponential function. It is important to
-remember that because this is a non-linear transformation, the
+values in separate conditions, not when they represent differences
+between conditions or regression slopes. In our example, because
+separate intercepts and slopes parametrization was used for the
+$\delta$ fixed effects model matrix, all four \code{delta\_fixed}
+parameters represent condition averages and can easily be transformed
+to sensitivities by applying the exponential function. It is important
+to remember that because this is a non-linear transformation, the
 $\delta$ to $d'$ conversion step should be done first before applying
 any other transformations to the posterior samples; for example, the
-exponential of a point and interval estimate is not equal to the point
-and interval estimate calculated after transforming the $\delta$
-posterior samples to the $d'$ samples.
+exponential of a point and interval estimate of $\delta$ is not equal
+to the point and interval estimate calculated after first transforming
+the $\delta$ posterior samples to the $d'$ samples.
 
@@
 the \code{gamma\_to\_crit} function only for the first column of the
-\code{gamma\_fixed} matrix.
+\code{gamma\_fixed} matrix. That's because the second column
+represents the \emph{difference} in $\bm{\gamma}$ between conditions
+and a nonlinear transformation from $\bm{\gamma}$ to $\bm{c}$ will not
+give a correct difference in $\bm{c}$.
 
@@
 
-To test if the model correctly recovers known parameter values, we
-simulated the data from a hypothetical exact replication of the
+We simulated the data from a hypothetical exact replication of the
 previously described experiment using the point estimates from the
@@
 
-For illustration purposes, an SDT model that differed from the true
-model only in that it did not have any hierarchical structure was
-fitted to the same simulated dataset. Since the non-hierarchical model
-was much simpler and the data consisted of only eight vectors of
-response counts, the mixing of the chains was excellent.
+As we have repeatedly stressed the models that lack the necessary
+hierarchical structure may easily show reliable effects where none
+exist or fail to detect true differences. To illustrate this problem,
+an SDT model that differed from the true model only in that it did not
+have any hierarchical structure was fitted to the same simulated
+dataset. Tring to obtain general theoretical results on the typical
+extent of bias in models that ignore the hierarchical structure would
+be a serious undertaking, but even this simple simulational example
+shows how misleading can the results of such analyses be. Since the
+non-hierarchical model was much simpler and the data consisted of only
+eight vectors of response counts, the mixing of the chains was
+excellent.
 
@@
 values, but the estimates based on the simplified, non-hierarchical
-model were biased; the credible intervals were not only about three
-times shorter on average, but also failed to contain most of the true
-values.
+model were severely biased; the credible intervals were not only about
+three times shorter on average, but also failed to contain most of the
+true values.
 
@@
 
-However, as can be seen in Fig.~\ref{fig:8} below, in this case the
-observed ROC curves seemed to fit the simplified model's predictions
-quite well, giving a false impression of model validity.
+To avoid bias, the hierarchical structure present in the data can be
+reduced or ignored only if the fit of the full hierarchical model
+indicates that the random effect variances or correlations are likely
+to be equal to zero. Importantly, this rule is valid regardless of the
+apparent fit of the model, as illustrated by the contents of
+Fig.~\ref{fig:8} below; even though the simplified model provided
+strongly biased results the observed ROC curves seemed to fit the
+model's predictions quite well, giving a false impression of model
+validity.
 
@@
 
-The great importance of SDT to psychology stems from the fact that
-given weak assumptions about an underlying decision process, it
-promises to deconfound sensitivity from bias in arbitrary binary
-classification tasks. To the best of our knowledge, at present the
+The importance of SDT to psychology stems from the fact that given
+weak assumptions about an underlying decision process, it promises to
+deconfound sensitivity from bias in arbitrary binary classification
+tasks - a problem almost as common in psychology studies as the usage
+of classification tasks. To the best of our knowledge, at present the
 \code{bhsdtr} package provides the only method of Bayesian inference
@@
 by using automatically generated human-readable Stan code as a
-template. We hope that researchers with a basic understanding of
-Signal Detection Theory, Bayesian inference, and hierarchical
-modelling will find our package useful and adopt it as a method of
-analysis of binary classification performance in future studies.
+template.
+% We hope that researchers with a basic understanding of Signal
+% Detection Theory, Bayesian inference, and hierarchical modelling will
+% find our package useful and adopt it as a method of analysis of binary
+% classification performance in future studies.
 
@@
 \end{document}
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: "paper.toc"
+%%% End:

### 2020-01-13 R/aggregate_responses.R

@@ -39,7 +39,8 @@ aggregate_responses = function(data, stimulus, response, variables = NULL){
     res = plyr::ddply(data, unique(c(variables, stimulus[1])),
                       function(df)table(c(df[[response[1]]], 1:K)) - 1)
     counts = res[, c((ncol(res)-K+1):ncol(res))]
-    list(data = res[, setdiff(variables, c(stimulus[1], response[1]))],
+    ## drop = F is important for the single column data frame case
+    list(data = res[, setdiff(variables, c(stimulus[1], response[1])), drop = F],
          stimulus = res[[stimulus[1]]],
          counts = counts)
 }

### 2020-01-13 R/make_stan_data.R

@@ -132,7 +132,7 @@ make_stan_data = function(adata, fixed, random = list(), criteria_scale = 2, met
   K = ncol(adata$counts)
   if(length(random) > 0){
     for(l in 1:length(random)){
-      group.mm = stats::model.matrix(random[[l]]$group, adata$data)
+        group.mm = stats::model.matrix(random[[l]]$group, adata$data)
       if(ncol(group.mm) == 2){
         ## Probably a numeric variable
         random[[l]]$group = as.numeric(as.factor(as.character(group.mm[,2])))

### 2020-01-13 inst/preprint/analysis_script.R

@@ -6,17 +6,23 @@
 ## Change this to FALSE once all the stanfit objects are stored to
 ## save time and memory
 fresh_start = FALSE
-## Where do you want to store all these large stanfit objects?
+## Where do you want to store all the (large) stanfit objects? (I use
+## Linux so '~/path' makes sense to me)
 temp_path = '~/temp'
+
+## WARNING we are not setting the random seed here because 1) R and
+## Stan have separate seeds and 2) the user of this script may use a
+## different number of mcmc threads.
 
 library(rstan)
 library(ggplot2)
 library(Hmisc) # rMultinom for simulations
 library(bhsdtr)
+library(plyr)
 ## Standard Stan optimizations
 rstan_options(auto_write = TRUE)
 options(mc.cores = parallel::detectCores())
-library(plyr)
+data(gabor)
 
 ######################################################################
 ## Fitting the model to real study data
@@ -87,7 +93,7 @@ round(apply(crit, 2, mean), 2)
 ######################################################################
 ## Fitting the single criterion hierarchical SDT model
 
-fixed = list(delta = ~ -1 + duration:order, gamma = ~ order)
+xfixed = list(delta = ~ -1 + duration:order, gamma = ~ order)
 random = list(list(group = ~ id, delta = ~ -1 + duration, gamma = ~ 1))
 ## this time we ignore the ratings
 gabor$r = combined_response(gabor$stim, accuracy = gabor$acc)
@@ -156,47 +162,53 @@ data = aggregate_responses(gabor, 'stim', 'r', c('duration', 'id', 'order'))
 fixed = list(delta = ~ -1 + duration:order, gamma = ~ order)
 random = list(list(group = ~ id, delta = ~ -1 + duration, gamma = ~ 1))
 
-## Realistic known parameter values
-s = apply(as.data.frame(fit), 2, mean)
-sdata = make_stan_data(adata, fixed, random)
-gamma_fixed = matrix(nrow = ncol(sdata$X_gamma), ncol = sdata$K - 1)
-for(r in 1:nrow(gamma_fixed))
[…]

### 2020-01-14 R/gabor.R
after first round of reviews BRM

@@ -7,24 +7,26 @@
 #' clear image') of discrimination responses to gabor patches presented for 32ms
 #' or 64ms before or after the ratings were provided.
 #'
-#' On each trial participants had to classify the briefly presented gabor patch
-#' as either oriented to the left or to the right using the arrow keys. The
-#' participants were also asked to rate the stimuli on a 4 point Perceptual
-#' Awareness Scale presented at the bottom of the screen. The gabor patch was
-#' immediately followed by a mask. The PAS ratings ranged from "no experience"
-#' to "absolutely clear image" and were provided either before (RD condition) or
-#' after (DR condition) pressing the arrow keys. On each trial the gabor patch
-#' was equally likely to be presented for 32ms or 64ms. Decision-rating order
-#' was a between-subject variable and gabor patch presentation duration was a
-#' within-subject variable. There were 47 participants and 48 trials per
-#' condition.
+#' On each trial participants had to classify the briefly presented
+#' gabor patch as either oriented to the left or to the right using
+#' the arrow keys. The participants were also asked to rate the
+#' stimuli on a 4 point Perceptual Awareness Scale presented at the
+#' bottom of the screen. The gabor patch was immediately followed by a
+#' mask. The PAS ratings ranged from "no experience" to
+#' "absolutely clear image" and were provided either before
+#' (RATING-DECISION condition) or after (DECISION-RATING condition)
+#' pressing the arrow keys. On each trial the gabor patch was equally
+#' likely to be presented for 32ms or 64ms. Decision-rating order was
+#' a between-subject variable and gabor patch presentation duration
+#' was a within-subject variable. There were 47 participants and 48
+#' trials per condition.
 #'
 #' @format A data frame with 4509 rows and 10 variables: \describe{
 #'   \item{timestamp}{Time when response was recorded} \item{duration}{Stimulus
 #'   duration (ms)} \item{trial}{Trial number} \item{acc}{Accuracy: 1 = correct,
 #'   0 = incorrect} \item{id}{Unique participant identification number}
-#'   \item{order}{Between-subject order manipulation: DR = rating after
-#'   decision, RD = decision after rating} \item{age}{Participant's age at the
+#'   \item{order}{Between-subject order manipulation: DECISION-RATING = rating after
+#'   decision, RATING-DECISION = decision after rating} \item{age}{Participant's age at the
[…]

### 2020-01-14 R/plot_sdt_fit.R

@@ -85,7 +85,7 @@ plot_sdt_fit = function(fit, adata, variables = NULL, type = 'roc', alpha = .05,
             geom_errorbar(aes(ymin = stim2$cumfr.lo, ymax = stim2$cumfr.hi, x = cumfr.fit), width = 0.02) +
             geom_errorbarh(aes(xmin = cumfr.lo, xmax = cumfr.hi, y = stim2$cumfr.fit), height = 0.02) +
             geom_point() +
-            labs(x = 'p(Hit)', y = 'p(FA)') +
+            labs(x = 'p(F)', y = 'p(H)') +
             coord_fixed() +
             facet_wrap(~f)
         if(bw){

### 2020-01-14 inst/preprint/analysis_script.R

@@ -1,5 +1,8 @@
 ## -*- coding: utf-8 -*-
 
+## TODO Include the analysis for the nested model (comparing the true
+## hierarchical and non-hierarchical estimates)
+
 ######################################################################
 ## IMPORTANT GLOBAL VARIABLES
 ##
@@ -19,6 +22,7 @@ library(ggplot2)
 library(Hmisc) # rMultinom for simulations
 library(bhsdtr)
 library(plyr)
+library(gridExtra)
 ## Standard Stan optimizations
 rstan_options(auto_write = TRUE)
 options(mc.cores = parallel::detectCores())
@@ -64,6 +68,26 @@ if(fresh_start){
     load(paste(temp_path, 'fit', sep = '/'))
 }
 
+## if(fresh_start){
+##     fixed = list(delta = ~ -1 + duration:order, gamma = ~ -1 + order)
+##     random = list(list(group = ~ id, delta = ~ -1 + duration, gamma = ~ 1))
+##     model = make_stan_model(random)
+##     sdata = make_stan_data(adata, fixed, random)
+##     fit_nested = stan(model_code = model,
+##                       data = sdata,
+##                       pars = c('delta_fixed', 'gamma_fixed',
+##                                'delta_sd_1', 'gamma_sd_1',
+##                                'delta_random_1', 'gamma_random_1',
+##                                'Corr_delta_1', 'Corr_gamma_1',
+##                                ## we need counts_new for plotting
+##                                'counts_new'),
+##                       iter = 8000,
+##                       chains = 4)
+##     save(fit_nested, file = paste(temp_path, 'fit_nested', sep = '/'))
+## }else{
+##     load(paste(temp_path, 'fit_nested', sep = '/'))
+## }
+
 ## Model fit summary does not indicate any convergence issues
 print(fit, probs = c(.025, .957),
       pars = c('delta_fixed', 'gamma_fixed',
@@ -87,13 +111,13 @@ ggsave(p2, file = 'response_fit.pdf')
 ggsave(p2, file = 'response_fit.png')
 
 ## The criteria are placed fairly symmetrically
-crit = gamma_to_c(as.data.frame(fit))
+crit = gamma_to_crit(as.data.frame(fit))
 round(apply(crit, 2, mean), 2)
 
[…]

### 2020-02-11 R/make_stan_data.R
ordinal and uvordinal models, random effects need fixing

@@ -208,6 +208,21 @@ make_stan_data = function(adata, fixed, random = list(), criteria_scale = 2, gam
                                                                    default_prior[[par]][[par_type]],
                                                                    c(data[[sf('%s_size', par_type)]], ncol(data[[sf('X_%s', par_type)]])),
                                                                    sf('%s_%s', par_type, par))
+        data[[sf('%s_is_fixed', par_type)]] = matrix(0, nrow = data[[sf('%s_size', par_type)]], ncol = ncol(data[[sf('X_%s', par_type)]]))
+        data[[sf('%s_fixed_value', par_type)]] = matrix(0, nrow = data[[sf('%s_size', par_type)]], ncol = ncol(data[[sf('X_%s', par_type)]]))
+        ## In ordinal models the effect associated with the intercept is fixed at 0 for identifiability
+        if((model %in% c('ordinal', 'uvordinal')) & par_type == 'eta'){
+            ## the effect for the intercept is 0
+            intercept = FALSE
+            for(i in 1:ncol(data$X_eta))
+                if(all(data$X_eta[,i] == rep(1, nrow(data$X_eta)))){
+                    data$eta_is_fixed[,i] = 1
+                    data$eta_fixed_value[,i] = 0
+                    intercept = TRUE
+                }
+            if((!intercept) & all((data$X_eta %*% rep(1, ncol(data$X_eta))) != 0))
+                stop("Separate intercepts and slopes parametrization of eta fixed effects not allowed in ordinal models")
+        }
     }
     if((gamma_link == 'identity') & (is.null(fixed$gamma_mu))){
         for(i in 1:(data$X_gamma_ncol))
@@ -217,7 +232,7 @@ make_stan_data = function(adata, fixed, random = list(), criteria_scale = 2, gam
     if(length(random) > 0){ 
         ## converting the formulae (e.g., ~ id) to numeric indices (1, 2, 3, ..., max(id))        
         for(l in 1:length(random)){
-            ## Just the group indicator, make sure it is a column vector
+            ##! this is not pretty ! Just the group indicator, make sure it is a column vector
             group.mm = stats::model.matrix(random[[l]]$group, adata$data)
             if(ncol(group.mm) == 2){
                 ## Probably a numeric variable
[…]

### 2020-02-11 inst/preprint/analysis_script.R

@@ -394,6 +394,9 @@ df$stimulus = adata_sim$stimulus
 df$i = 1:nrow(df)
 df = ddply(df, c('order', 'duration', 'stimulus'), function(x)apply(adata_sim$counts[x$i,], 2, sum))
 data_sim_2 = list(data = df[,c('order', 'duration')], stimulus = df$stimulus, counts = df[, paste(1:8)])
+rbind(data_sim_2$counts[data_sim_2$data$order == 'DECISION-RATING' & data_sim_2$data$duration == '32 ms' & data_sim_2$stimulus == 1,],
+      apply(adata_sim$counts[adata_sim$data$order == 'DECISION-RATING' & adata_sim$data$duration == '32 ms' & adata_sim$stimulus == 1,], 2, sum))
+## Ok
 
 if(fresh_start){
     fit.aggr = stan(model_code = make_stan_model(),

### 2020-02-14 ordinal/ordinal.Rmd
ordinal works

```{r init, include=F}
knitr() ## Zdefiniowa³em w .Rprofile, ustawia domy¶lne opcje chunków

sdt_ml_plot = function(model, adata, sdata){
    sdt = !is.null(adata$stimulus)
    o = optimizing(stan_model(model_code = model), data = sdata, init = 0)
    mp = matrix(o$par[grep('multinomial_p', names(o$par))], nrow = nrow(adata$data))
    obs = t(apply(adata$counts, 1, function(x)(x / sum(x))))
    if(sdt)
        stim = as.factor(adata$stimulus)
    f = as.factor(adata$data[,1])
    if(ncol(adata$data) > 1)
        for(i in 2:ncol(adata$data))
            f = f:as.factor(adata$data[,i])
    if(sdt){
        ggplot(data.frame(f = rep(f, ncol(mp)), stim = rep(stim, ncol(mp)), th = rep(1:ncol(mp), each = length(f)),
                          mp = as.vector(mp), obs = as.vector(obs)),
               aes(th, obs, group = stim, color = stim)) +
        geom_line(aes(y = mp)) +
        geom_point() +
        labs(title = 'Subj x Condition panels, lines = model, points = observed response distribution') +
        xlab('response = decision + rating') +
        ylab('p(response)') +
        facet_wrap(~ f)
    }else{
        ggplot(data.frame(f = rep(f, ncol(mp)), th = rep(1:ncol(mp), each = length(f)),
                          mp = as.vector(mp), obs = as.vector(obs)),
               aes(th, obs)) +
        geom_line(aes(y = mp)) +
        geom_point() +
        labs(title = 'lines = fitted model, points = observed response distribution') +
        xlab('response = decision + rating') +
        ylab('p(response)') +
        facet_wrap(~ f)
    }
}
simulate_from_fit = function(fit, adata, fixed, random){
    ## Realistic known parameter values
    s = apply(as.data.frame(fit), 2, mean)
    sdata = make_stan_data(adata, fixed, random)
    gamma_fixed = matrix(nrow = ncol(sdata$X_gamma), ncol = sdata$K - 1)
    for(r in 1:nrow(gamma_fixed))
        gamma_fixed[r,] = s[paste('gamma_fixed[', 1:(sdata$K - 1), ',', r, ']', sep = '')]
    gamma_random = list()
    for(r in 1:sdata$group_1_size)
        gamma_random[[r]] = matrix(s[paste('gamma_random_1[', r, ',', 1:(sdata$K - 1), ',1]', sep = '')],
                                   nrow = sdata$K - 1, ncol = sdata$Z_gamma_ncol)
    delta_fixed = s[grep('delta_fixed', names(s))]
    delta_random = matrix(nrow = sdata$group_1_size, ncol = sdata$Z_delta_ncol_1)
    for(r in 1:nrow(delta_random))
        for(i in 1:ncol(delta_random))
            delta_random[r, i] = s[sprintf('delta_random_1[%d,%d,1]', r, i)]
    multinomial_p = matrix(ncol = sdata$K, nrow = sdata$N)
    thr = gamma = matrix(ncol = sdata$K - 1, nrow = sdata$N)
    dprim = rep(NA, sdata$N)
    for(n in 1:sdata$N){
        dprim[n] = exp(sdata$X_delta[n,] %*% delta_fixed + sdata$Z_delta_1[n,] %*% delta_random[sdata$group_1[n],])
        gamma[n,] = t(gamma_fixed) %*% sdata$X_gamma[n,] + gamma_random[[sdata$group_1[n]]] %*% sdata$Z_gamma_1[n,]
    }
    criteria = t(apply(gamma, 1, function(x){
        exp_x_0 = exp(c(x, 0))
        sdata$criteria_scale * qnorm(cumsum(exp_x_0 / sum(exp_x_0))[-sdata$K])
    }))
    distr_shift = -sdata$stim_sign * dprim / 2
    multinomial_p[, 1] = pnorm(criteria[, 1] + distr_shift)
    for(k in 2:(sdata$K - 1)){
        multinomial_p[, k] = pnorm(criteria[, k] + distr_shift) - pnorm(criteria[, k - 1] + distr_shift)
    }
    multinomial_p[, sdata$K] = pnorm(-(criteria[, sdata$K - 1] + distr_shift))
    data_sim = adata
    ## Now we can simulate
    for(r in 1:nrow(adata$counts))
        data_sim$counts[r,] = table(c(rMultinom(t(multinomial_p[r,]), sum(adata$counts[r,])), 1:sdata$K)) - 1
    data_sim$dprim = dprim
    data_sim$gamma = gamma
    data_sim$criteria = criteria
    data_sim
}

## Works only when there is just one grouping factor
simulate_from_fit = function(fit, adata, fixed, random, gamma_link){
    s = apply(as.data.frame(fit), 2, mean)
    sdata = make_stan_data(adata, fixed, random)
    gamma_fixed = matrix(nrow = ncol(sdata$X_gamma), ncol = sdata$K - 1)
    for(r in 1:nrow(gamma_fixed))
        gamma_fixed[r,] = s[paste('gamma_fixed[', 1:(sdata$K - 1), ',', r, ']', sep = '')]
    gamma_random = list()
    for(r in 1:sdata$group_1_size)
        gamma_random[[r]] = matrix(s[paste('gamma_random_1[', r, ',', 1:(sdata$K - 1), ',1]', sep = '')],
                                   nrow = sdata$K - 1, ncol = sdata$Z_gamma_ncol)
    delta_fixed = s[grep('delta_fixed\\[', names(s))]
    delta_random = matrix(nrow = sdata$group_1_size, ncol = sdata$Z_delta_ncol_1)
    for(r in 1:nrow(delta_random))
        for(i in 1:ncol(delta_random))
            delta_random[r, i] = s[sprintf('delta_random_1[%d,%d]', r, i)]
    multinomial_p = matrix(ncol = sdata$K, nrow = sdata$N)
    thr = gamma = matrix(ncol = sdata$K - 1, nrow = sdata$N)
    dprim = rep(NA, sdata$N)
    for(n in 1:sdata$N){
        dprim[n] = exp(sdata$X_delta[n,] %*% delta_fixed + sdata$Z_delta_1[n,] %*% delta_random[sdata$group_1[n],])
        gamma[n,] = t(gamma_fixed) %*% sdata$X_gamma[n,] + gamma_random[[sdata$group_1[n]]] %*% sdata$Z_gamma_1[n,]
    }
    criteria = t(apply(gamma, 1, function(x){
        exp_x_0 = exp(c(x, 0))
        sdata$criteria_scale * qnorm(cumsum(exp_x_0 / sum(exp_x_0))[-sdata$K])
    }))
    distr_shift = -sdata$stim_sign * dprim / 2
    multinomial_p[, 1] = pnorm(criteria[, 1] + distr_shift)
    for(k in 2:(sdata$K - 1)){
        multinomial_p[, k] = pnorm(criteria[, k] + distr_shift) - pnorm(criteria[, k - 1] + distr_shift)
    }
    multinomial_p[, sdata$K] = pnorm(-(criteria[, sdata$K - 1] + distr_shift))
    data_sim = adata
    ## Now we can simulate
    for(r in 1:nrow(adata$counts))
        data_sim$counts[r,] = table(c(rMultinom(t(multinomial_p[r,]), sum(adata$counts[r,])), 1:sdata$K)) - 1
    data_sim$dprim = dprim
    data_sim$gamma = gamma
    data_sim$criteria = criteria
    data_sim
}
```

It is now possible to fit hierarchical ordinal polytomous regression
models in bhsdtr by using order-preserving link functions for the
thresholds. This way not only the constant shift in the thresholds,
but also the differences in the pattern of the thresholds can be
accounted for in certain situations. To my knowledge this is a new
kind of hierarchical ordinal model that can be used to estimate
ordinal polytomous item-parameter variability in a way that until now
was not possible. If you have a bunch of Likert-type items, or
confidence ratings, or anything of this sort in your data, than you
may find this model useful.

Because by definition a polytomous item has a multinomial distribution
and the multinomial distribution does not have a separate parameter
for the residual variance, when item-parameter variability is not
accounted for by the model it can only be misplaced
(overdispersion). The correct way to account for item-parameter
variability in ordinal polytomous items is to allow for individual
threshold effects while respecting the ordering of the
thresholds. This can only be achieved by using the order-preserving
link functions.

Assume that the ordered responses ranging from 1 to K come from the
same general kind of process as the responses in an SDT model when the
stimulus is 1 ("noise"). For example, when John is faced with the
Likert-type item "How often did you cry during the last two weeks" an
unknown process produces some "internal experience" value. This latent
value manifests itself in the response "never" (= 1) if it is low
enough, "a few times" (= 2) if it is a bit higher than some latent
threshold, or in the response "frequently" (= 3) if it is higher than
some other, still higher, threshold. The thresholds are assumed to be
ordered, because otherwise the item would not have the intended
meaning and thus would not be a valid instrument of measuring
depression.

Now consider a participant who repeatedly responds to the same item
and who has the memory of this event wiped out after each
response. Despite the wiping out of memory we can still expect some
variability in the responses due to the instability of the complex
process which produces the internal value, as well as due to the
instability of the decision criteria. This variability is certainly
there, the question is if it is non-negligible, and if so, what can be
done about it. Unfortunately, without additional assumptions it is
impossible to model both kinds of variability at the same time even in
the hypothetical repeated measurement with memory wipeout scenario,
because every shift in the internal value is indistiguishible from a
uniform effect in the thresholds, i.e., a constant shift of all the
thresholds. As I will explain shortly, the required additional
assumptions that make such models identifiable under certain
conditions are necessarily *causal*.

In many situations this is both a statistical and a conceptual
problem. In particular, when the origin of the internal value is
unknown because the design is purely observational (e.g.,
questionnaire measurement), the way of describing the process of
responding to a Likert-type item in terms of internal values and
thresholds is simply redundant. As far as the design of the study is
concerned, the value of the internal evidence "comes from nowhere" -
all we can say about it is that it is relative to the thresholds - and
the position of each threshold is relative to the positions of other
thresholds and to the position of the internal value. In this case to
say that John thinks that he was not crying at all during the last two
weeks and to say that *John's* threshold for the response "a few
times" is above the amount of crying that John thinks he experienced,
or decided to reveal, is to say the same thing.

It follows that *in such situations we do not need the internal value
parameter at all*, unless there is some factor about which we can
perhaps assume that it can *only* produce a uniform effect in the
thresholds, such as the stimulus class ("noise" or "signal") in a
binary classification task, which can *perhaps* be assumed to induce a
change (*d'*) in the internal value only; In such cases the internal
value parameter may be justified and useful, because it does constrain
the model: in the repeated measurement case the model is no longer
saturated.

But what about the promise of Item Response Theory? Isn't it true that
we can separate the internal values (so called "person parameters")
from the thresholds (so called "item parameters") and place them on a
common scale? Once the participants leave the room the items retain
their estimated "difficulty levels", right? The item difficulty
(=thresholds) is not defined relative to some sample of participants
who filled the questionnaire. How is this possible if the two kinds of
effects are completely confounded? In particular, how does Item
Response Theory achieve when each person provides only a single
response to each item?

It is not an *achievment* of Item Response Theory, but it's
(unrealistic) *assumption*. The illusory nature of this achievement
becomes obvious as soon as we realize that without additional
(causal!)  assumptions it is impossible to distinguish between genuine
internal value effects and (dis)simulation (i.e., faking), which
conteptually corresponds to the change in the *participant*-specific,
not item-specific, decision thresholds, also known as
"item-parameters". There is nothing in the *desing* of the process of
measurement by a questionnaire that makes it possible to distinguish
genuine differences in internal values from the differences in the way
of responding. This "achievement" of Item Response Theory is a
consequence of treating the "item-parameters" as constant accross
participants when fitting an IRT model.

Because of random assignment of the stimulus class above chance
performance in a binary classification task cannot be faked, and so
the *d'* paramater definately captures at least part of the effect of
the stimulus on the internal value. On the other hand, every possible
pattern of responses to a questionnaire can be faked. The only way to
distinguish the "item" and "person" effects is to *assume selective
influence*, i.e., that there is some factor f (such as the stimulus
class in a binary classification task, or, in case of a depression
questionnaire, participants trait, or a therapeutic intervention)
which can *only* affect the internal value, thus producing a
*constant* shift in the thresholds. The assumption of selective
influence makes the model with the two kinds of parameters
identifiable if there are repeated measurements for at least one
participant and one item or a large number of participants provides
single responses to all of the items. In particular, in typical
applications of IRT it is simply *assumed* in the model that the
participants differ in the internal values, and the item parameters
are constant accross the sample of participants (selectivity).

In every book on Item Response Theory that I know of the issue of
measurement invariance due to the variability in the "item-parameters"
is reduced to the problem of Differential Item Functioning
(DIF). However, the variability in "item-parameters" that I talk about
here has nothing to do with DIF, because DIF is a *population* level
effect; An estimate of DIF is obtained when the estimates of
"item-parameters" obtained for two different (large) samples of
participants are compared. However, these estimates are obtained by
fixing the "item-parameters" within each sample. It does not take much
thought to see that the assumption of constant "item-parameters" at
the group/population/sample level is utterly unrealistic. It follows
that the achievement of separating participant and item effects in IRT
models is not an achievement at all, it is just wishful thinking
combined with aggressive marketing.

Some of the problems related to item-parameter variability can be
addressed *when additional causal assumptions of selective influence
are introduced and the design of the study makes the model
identifiable*. For example, we can use the data from a binary
classification task and model only the responses for one kind of
stimulus, e.g., "noise".

Here is a simple example which illustrates how we can fit a
hierarchical ordinal polytomous regression model using the bhsdtr
package. We will be using the data from one condition in the gabor
dataset:

```{r c1, echo=T}
library(ggplot2)
library(bhsdtr)
library(Hmisc)
library(rstan)
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)

## r = 1 ~ absolutely clear image left, r = 4 ~ no experience left,
## r = 5 ~ no experience right, ... r = 8 ~ absolutely clear image right
gabor$r = combined_response(gabor$stim, gabor$rating, gabor$acc)
d1 = gabor[gabor$duration == '32 ms' & gabor$order == 'DECISION-RATING',]
## Only the left-tilted gabor patches
d1 = d1[d1$stim == 0,]
```

To fit a hierachical model we have to aggregate the data (this makes
the process of sampling much faster), create the fixed- and
random-effects lists, and create the data structures and the model
code required by stan. Note that we have introduced a new parameter -
eta - which represents the internal value. As we already know, without
the selectivity assumption the difference between item and person
parameters is arbitrary and so in bhsdtr the first element of the
fixed effects vector for the eta parameter is automatically fixed at 0
eto make the model identifiable (this bahavior may change in the near
future):

```{r c2, echo = T}        
## This order-preserving link function represents the main criterion
## directly as a free paramter and all the other criteria as log of
## distance between the criterion and the criterion adjacent to it
## which is closer to the main criterion
link = 'log_distance'
adata = aggregate_responses(d1, response = 'r', variables = c('id'))
fixed = list(eta = ~ 1, gamma = ~ 1)
random = list(list(group = ~ id, gamma = ~ 1))
model = make_stan_model(random, model = 'ordinal', gamma_link = link)
sdata = make_stan_data(adata, fixed, random, model = 'ordinal', gamma_link = link)
```

To simplify the task we will use maximum likelihood optimization,
since the model is saturated anyway:

```{r c3, echo=T}
o = optimizing(stan_model(model_code = model), sdata, init = 0)
```

Unsuprisingly, the model fits perfectly for every participant:

```{r p1}
sdt_ml_plot(model, adata, sdata)
```

Now we will add the estimated gamma/criteria random effects to the
fixed effects to obtain participant-specific gamma/criteria estimates,
convert between the gamma and the criteria vectors, and take a look at
the estimated variability in the criteria:

```{r c4,echo=T}
gamma_fixed = o$par[grep('gamma_fixed\\[', names(o$par))]
gamma_random = matrix(o$par[grep('gamma_random_1', names(o$par))], nrow = sdata$group_1_size)
for(i in 1:nrow(gamma_random))
    gamma_random[i,] = gamma_random[i,] + gamma_fixed
criteria = gamma_to_crit(gamma_random, beta_index = NULL, gamma_link = link)
df = expand.grid(criterion = 1:7, id = 1:nrow(criteria))
df$estimate = as.vector(t(criteria))
ggplot(df, aes(id, estimate, group = id)) + geom_line() + geom_point()
```
The thresholds may look a bit weird, but we have to rememberd that
what we see here is *a mixture of d' and threshold effects*. The
estimates are somewhat similar to what we would have obtained if we
fitted an SDT model, but they clearly do not have the intended
interpretation; They do not deconfound the two things that an SDT
model is supposed to deconfound, namely sensitivity and bias. Contrary
to what seems to be commonly assumed, if we included the stimulus
class we would not solve the problem - for that we have to also
introduce the strong causal assumption of selective influence, which,
like any other strong assumption, can easily be false.

The justification of the selective influence assumption comes from a
psychological theory of the task, not from the design of the
study. All that is guaranteed by randomly assigning the stimulus class
(i.e., by choosing the stimulus class at random on every trial) is
that any statistical effect of the stimulus class is an unbiased
estimate of the total causal effect of the stimulus class. In order to
empirically justify the assumption of *selective* influence of the
stimulus class on the internal evidence value we would have to *test*
if the stimulus class does not also affect the thresholds.

We can do exactly this. We will use the 'log_distance' link function,
because when this link function is used one of the elements of the
gamma vector directly represents the position of the main decision
criterion, and all the other elements represent the pattern of the
distances between the thresholds. If the stimulus does not affect the
pattern of the thresholds we should observe an effect only for the
main criterion parameter. This model is not exactly equivalent to a
hierarchical SDT model without selective influence since we do not
constrain the effect of the stimulus class on the main criterion to be
non-negative, but there is enough data points per participant that we
can at least hope that this is not a serious problem.

KOD

There at least two important advantages that an SDT model fitted to
binary classification task data has over an IRT model fitted to
questionnaire data. One is that the causal assumption of selective
influence of the stimulus class on the internal values is much more
plausible than the - implicit in most applications of IRT - causal
assumption of selective influence of participants on the
"person-parameters". The other is that in a typical binary
classification task case the process of responding is repeated many
times, the task is relatively simple ("was the gabor patch tilted left
or right" vs "do you feel that other people are happier than you") and
the processes responsible for the internal values and for the
placement of the decision thresholds are likely to be much more
stable.

If we campared the estimates of the main decision criterion (i.e.,
bias) in SDT models fitted to two different large samples of
participants the difference would likely be negligible with very
narrow confidence/credible intervals. By the same token, instability
of item parameters in IRT models cannot be estimated by assuming
selective influence and comparing estimates of "item-parameters"
obtained in two different large samples, because large-sample average
estimates of item parameters are the best way to *hide or misplace*
(it becomes part of the variability in the "person-parameters") the
intra- or inter-individual variability in item parameters; By assuming
that internal values are participant-specific and item parameters are
item-specific we force the estimates of item parameters to become
estimates of average person-specific thresholds. This not only begs
the question, but also, because the model is non-linear, makes the
estimates of item-parameters *asymptotically biased*.

Back to our simulation. The variability due to *d'* is still here, we
just cannot deconfound it from response bias. This means that, just
like in an IRT model that does not fully account for true person- and
and item-parameter effects, there is *no way to correctly estimate
reliability of the estimates of person-parameters/internal values*. We
will now demonstrate this fact.

First, we will fit a hierarchical SDT model with multiple criteria and
an order-preserving link function. This will give us realistic
estimates of "person-" (*d'*) and "item"-parameters (the thresholds).

KOD 4

What will happen if we emulate a typical IRT analysis and fit a
hierarchical ordinal model in which the thresholds are assumed to be
constant accross participants? After all, we can just invoke the
magical power of wishful thinking and assume that the thresholds are
item-parameters, right? How well will the internal value estimates
based on multiple responses to a simple, non-personal question match
the estimates based on a model that accounts for participant specific
threshold patterns?

KOD 5

The answer is - the two kinds of estimates will match poorly. Here is
the comparison of point estimates:

and here is the comparison of interval estimates:

WYKRES 3

Let's see more directly how the lengths of the credible intervals
differ. If the interval estimate widths were comparable they should
all be scattered closely to the horizontal line at 1. Unsurprisingly,
this is now what we see here:

WYKRES

The credible intervals directly represent reliability. For the
interval estimates based on a simplified model to contain the great
majority of internal values *or* account for the variability in the
thresholds they would have to be wider. We do not need to look at
"item-parameter" estimates in the simplified model, because it is
already obvious that the point and interval estimates of the
thresholds based on this model have to be severily biased.

An ordinal model for ordered polytomous responses that does not
correctly account for the variability in the thresholds produces
biased point and interval estimates in *every* parameter, and the
estimates seem more reliable than they truly are. That's because it
assumes that the thresholds, which are determined both by the items
and by the properties of the participants, are constant accross the
participants. This assumption, which seems to be universally accepted
in Item Response Theory, is so ridiculous that it should be laughed at
on the street.

### 2020-02-14 R/acc_to_delta.R

@@ -8,12 +8,12 @@
 #'     converted to accuracy of an unbiased observer
 #' @return delta = log(d') or expected accuracy
 #' @export
-acc_to_delta = function(acc = .75, delta = NULL){
+acc_to_delta = function(acc = .75, delta = NULL, crit = 0){
     if(any((acc <= 0) || (acc >= 1)))
         stop('Some accuracy scores were to extreme')
     if(is.null(delta)){
         log(2 * stats::qnorm(acc))
     }else{
-        pnorm(exp(delta) / 2)
+        (pnorm(exp(delta) / 2 - crit) + pnorm(exp(delta) / 2 + crit)) / 2
     }
 }

### 2020-02-14 R/make_stan_data.R

@@ -138,7 +138,8 @@
 #' @param gamma_link either 'softmax' (described in the paper),
 #'     'log_distance' or 'log_ratio' (See the Readme file in the
 #'     github repository)
-#' @param model can be either 'sdt' (the default), 'uvsdt', or 'metad'
+#' @param model can be either 'sdt' (the default), 'uvsdt', 'metad', 'ordinal',
+#' or 'uvordinal'
 #' @return a list with response and stimulus data, model matrices,
 #'     prior parameter values, and other data required by the stan
 #'     model generated using \code{make_stan_model}.
@@ -153,8 +154,13 @@
 #' @export
 make_stan_data = function(adata, fixed, random = list(), criteria_scale = 2, gamma_link = 'softmax', model = 'sdt'){
     sf = sprintf
+    check_fixed(fixed)
+    check_random(random)
+    check_adata(adata)
     check_link(gamma_link)
     check_model(model)
+    if((model %in% c('ordinal', 'uvordinal')) & !(gamma_link %in% c('log_distance', 'twoparameter', 'parsimonious')))
+        stop('This link function is not implemented for ordinal models')
     default_prior = list(mu = list(delta = acc_to_delta(.75), theta = 0, gamma = 0, eta = 0),
                          sd = list(delta = .5 * (acc_to_delta(.99) - acc_to_delta(.51)), theta = log(2), eta = 5),
                          scale = list(delta = .5 * (acc_to_delta(.99) - acc_to_delta(.51)), theta = log(2), eta = 5))
@@ -167,7 +173,7 @@ make_stan_data = function(adata, fixed, random = list(), criteria_scale = 2, gam
     data = list(PRINT = 0,
                 N = nrow(adata$counts),
                 K = K,
-                Kb2 = K / 2,
+                Kb2 = round(K / 2),
                 fixed = fixed,
                 random = random,
                 gamma_link = gamma_link,
@@ -212,16 +218,10 @@ make_stan_data = function(adata, fixed, random = list(), criteria_scale = 2, gam
         data[[sf('%s_fixed_value', par_type)]] = matrix(0, nrow = data[[sf('%s_size', par_type)]], ncol = ncol(data[[sf('X_%s', par_type)]]))
         ## In ordinal models the effect associated with the intercept is fixed at 0 for identifiability
         if((model %in% c('ordinal', 'uvordinal')) & par_type == 'eta'){
-            ## the effect for the intercept is 0
-            intercept = FALSE
-            for(i in 1:ncol(data$X_eta))
[…]

### 2020-02-14 R/make_stan_model.R

@@ -14,7 +14,8 @@
 #' @param gamma_link can be either 'softmax' (described in the paper),
 #'     'log_distance', or 'log_ratio' (See the Readme file in the
 #'     github repository)
-#' @param model can be either 'sdt' (the default), 'uvsdt', or 'metad'
+#' @param model can be either 'sdt' (the default), 'uvsdt', 'metad',
+#'     'ordinal', or 'uvordinal'
 #' @return a string containing the full model definition in the stan
 #'     modelling language.
 #' @examples

### 2020-02-14 inst/preprint/analysis_script.R

@@ -257,14 +257,14 @@ simulate_from_fit = function(fit, adata, fixed, random){
     for(r in 1:nrow(gamma_fixed))
         gamma_fixed[r,] = s[paste('gamma_fixed[', 1:(sdata$K - 1), ',', r, ']', sep = '')]
     gamma_random = list()
-    for(r in 1:sdata$G_1)
+    for(r in 1:sdata$group_1_size)
         gamma_random[[r]] = matrix(s[paste('gamma_random_1[', r, ',', 1:(sdata$K - 1), ',1]', sep = '')],
                                    nrow = sdata$K - 1, ncol = sdata$Z_gamma_ncol)
     delta_fixed = s[grep('delta_fixed', names(s))]
-    delta_random = matrix(nrow = sdata$G_1, ncol = sdata$Z_delta_ncol_1)
+    delta_random = matrix(nrow = sdata$group_1_size, ncol = sdata$Z_delta_ncol_1)
     for(r in 1:nrow(delta_random))
         for(i in 1:ncol(delta_random))
-            delta_random[r, i] = s[sprintf('delta_random_1[%d,%d]', r, i)]
+            delta_random[r, i] = s[sprintf('delta_random_1[%d,%d,1]', r, i)]
     multinomial_p = matrix(ncol = sdata$K, nrow = sdata$N)
     thr = gamma = matrix(ncol = sdata$K - 1, nrow = sdata$N)
     dprim = rep(NA, sdata$N)

### 2020-02-14 ordinal/code.R

@@ -1,0 +1,183 @@
+ggs = function(f)ggsave(f, width = 10, height = 10, dpi = 200)
+
+######################################################################
+## KOD 1
+
+library(ggplot2)
+library(bhsdtr)
+library(Hmisc)
+library(rstan)
+options(mc.cores = parallel::detectCores())
+rstan_options(auto_write = TRUE)
+
+## r = 1 ~ absolutely clear image left, r = 4 ~ no experience left, r = 5 ~ no experience right, ... r = 8 ~  absolutely clear image right
+gabor$r = combined_response(gabor$stim, gabor$rating, gabor$acc)
+d1 = gabor[gabor$duration == '32 ms' & gabor$order == 'DECISION-RATING',]
+d1$r = combined_response()
+## Only the left-tilted gabor patches
+d1 = d1[d1$stim == 0,]
+
+######################################################################
+## KOD 2
+
+## This order-preserving link function represents the main criterion
+## directly as a free paramter and all the other criteria as log of
+## distance between the criterion and the criterion adjacent to it
+## which is closer to the main criterion
+
+adata = aggregate_responses(d1, response = 'r', variables = c('id'))
+fixed = list(eta = ~ 1, gamma = ~ 1)
+random = list(list(group = ~ id, gamma = ~ 1))
+model = make_stan_model(random, model = 'ordinal')
+sdata = make_stan_data(adata, fixed, random, model = 'ordinal')
+
+######################################################################
+## KOD 3
+
+o = optimizing(stan_model(model_code = model), sdata, init = 0)
+if(o$return_code)warning('not ok')
+
+######################################################################
+## Wykres
+
+sdt_ml_plot(model, adata, sdata)
+ggs('plot1.jpg')
+
+######################################################################
+## Wykres
+
+gamma_fixed = o$par[grep('gamma_fixed\\[', names(o$par))]
+gamma_random = matrix(o$par[grep('gamma_random_1', names(o$par))], nrow = sdata$group_1_size)
+for(i in 1:nrow(gamma_random))
+    gamma_random[i,] = gamma_random[i,] + gamma_fixed
[…]

### 2020-02-14 R/gamma_to_crit.R
analysis_script updated, ordinal progress

@@ -88,7 +88,8 @@ gamma_to_crit = function(samples, beta_index = 1, gamma_link = 'softmax', s = 2,
     }else{
         K = length(nms) + 1
     }
-    samples = samples[, nms]
+    ## as.matrix because if samples is a vector nrow is NULL
+    samples = as.matrix(samples[, nms])
     criteria = matrix(nrow = nrow(samples), ncol = K - 1)
     if(link == 'log_ratio'){
         criteria[, K / 2] = samples[, K / 2];

### 2020-02-14 R/make_stan_data.R

@@ -179,7 +179,7 @@ make_stan_data = function(adata, fixed, random = list(), criteria_scale = 2, gam
                 gamma_link = gamma_link,
                 model = model,
                 criteria_scale = criteria_scale,
-                unbiased = unbiased(K),
+                unbiased = fix_stan_dim(unbiased(K)),
                 eta_size = 1,
                 dprim_size = 1,
                 delta_size = 1,
@@ -212,6 +212,7 @@ make_stan_data = function(adata, fixed, random = list(), criteria_scale = 2, gam
         for(par in c('mu', 'sd'))
             data[[sf('%s_fixed_%s', par_type, par)]] = parse_prior(fixed[[sf('%s_%s', par_type, par)]],
                                                                    default_prior[[par]][[par_type]],
+                                                                   ## matrix dimensions
                                                                    c(data[[sf('%s_size', par_type)]], ncol(data[[sf('X_%s', par_type)]])),
                                                                    sf('%s_%s', par_type, par))
         data[[sf('%s_is_fixed', par_type)]] = matrix(0, nrow = data[[sf('%s_size', par_type)]], ncol = ncol(data[[sf('X_%s', par_type)]]))
@@ -256,25 +257,31 @@ make_stan_data = function(adata, fixed, random = list(), criteria_scale = 2, gam
                     mm = remove.zero.cols(stats::model.matrix(random[[l]][[par_type]], adata$data))
                     data[[sf('Z_%s_ncol_%d', par_type, l)]] = ncol(mm)
                     data[[sf('Z_%s_%d', par_type, l)]] = mm
-                    ## just the name of the variable
+                    ## just the name of the variable in the random effects specification list
+                    v = sf('%s_sd_scale', par_type)
+                    data[[sf('%s_sd_scale_%d', par_type, l)]] = parse_prior(random[[l]][[v]],
+                                                                            default_prior$scale[[par_type]],
+                                                                            ## matrix dimensions
+                                                                            c(data[[sf('%s_size', par_type)]], data[[sf('Z_%s_ncol_%d', par_type, l)]]),
+                                                                            sf('%s_sd_scale', par_type))
                     v = sf('%s_nu', par_type)
                     data[[sf('lkj_%s_nu_%d', par_type, l)]] = if(is.null(random[[l]][[v]])){ 1 }else{ random[[l]][[v]][1] }
-                    v = sf('%s_sd_scale', par_type)
[…]

### 2020-02-14 inst/preprint/analysis_script.R

@@ -4,12 +4,12 @@
 ## WARNING The interface has changed since the Behavior Research
 ## Methods paper was accepted for publication. In the current version
 ## of the package every parameter (delta, gamma, theta - in uv sdt
-## models) is represented by a matrix even when it is unidimensional
-## in a given model. So, for example, in an SDT model delta fixed
-## effects are represented by parameters delta_fixed[1,1], ...,
-## delta_fixed[1,n], where n is the number of delta fixed effects
-## (i.e., the number of columns in the fixed effects model matrix for
-## delta).
+## models, eta in ordinal models, delta_sd_scale_1, etc.) is
+## represented by a matrix even when it is unidimensional in a given
+## model. So, for example, in an SDT model delta fixed effects are
+## represented by parameters delta_fixed[1,1], ..., delta_fixed[1,n],
+## where n is the number of delta fixed effects (i.e., the number of
+## columns in the fixed effects model matrix for delta).
 
 ######################################################################
 ## IMPORTANT GLOBAL VARIABLES
@@ -108,18 +108,15 @@ data(gabor)
 
 ## Combined response has to be calculated for this dataset
 gabor$r = combined_response(gabor$stim, gabor$rating, gabor$acc)
-
 ## Aggregation without information loss - we keep keep the variability
 ## due to the participants and all the experimental conditions.
 adata = aggregate_responses(gabor, 'stim', 'r', c('duration', 'order', 'id'))
-
 ## Model specification
 fixed = list(delta = ~ -1 + duration:order, gamma = ~ order)
 random = list(list(group = ~ id, delta = ~ -1 + duration, gamma = ~ 1))
 model = make_stan_model(random)
 ## The model code is human-readable
 cat(model)
-
 ## Required Stan data structure
 sdata = make_stan_data(adata, fixed, random)
 
@@ -133,10 +130,10 @@ if(fresh_start){
[…]

### 2020-02-17 ordinal/ordinal.Rmd
ordinal tutorial

@@
-```{r init, include=F}
-knitr() ## Zdefiniowa³em w .Rprofile, ustawia domy¶lne opcje chunków
+```{r init, include=F,echo=F}
+knitr::opts_chunk$set(dev = 'svg', comment = NA, fig.path='Figs/',fig.width=8, fig.height=6,
+                      echo=TRUE, warning = FALSE, error = FALSE, message = FALSE,
+                      cache.path = '~/windows/temp/ordinal', cache=T)
 
@@
 }
-simulate_from_fit = function(fit, adata, fixed, random){
-    ## Realistic known parameter values
-    s = apply(as.data.frame(fit), 2, mean)
-    sdata = make_stan_data(adata, fixed, random)
-    gamma_fixed = matrix(nrow = ncol(sdata$X_gamma), ncol = sdata$K - 1)
-    for(r in 1:nrow(gamma_fixed))
-        gamma_fixed[r,] = s[paste('gamma_fixed[', 1:(sdata$K - 1), ',', r, ']', sep = '')]
-    gamma_random = list()
-    for(r in 1:sdata$group_1_size)
-        gamma_random[[r]] = matrix(s[paste('gamma_random_1[', r, ',', 1:(sdata$K - 1), ',1]', sep = '')],
-                                   nrow = sdata$K - 1, ncol = sdata$Z_gamma_ncol)
-    delta_fixed = s[grep('delta_fixed', names(s))]
-    delta_random = matrix(nrow = sdata$group_1_size, ncol = sdata$Z_delta_ncol_1)
-    for(r in 1:nrow(delta_random))
-        for(i in 1:ncol(delta_random))
-            delta_random[r, i] = s[sprintf('delta_random_1[%d,%d,1]', r, i)]
-    multinomial_p = matrix(ncol = sdata$K, nrow = sdata$N)
-    thr = gamma = matrix(ncol = sdata$K - 1, nrow = sdata$N)
-    dprim = rep(NA, sdata$N)
-    for(n in 1:sdata$N){
-        dprim[n] = exp(sdata$X_delta[n,] %*% delta_fixed + sdata$Z_delta_1[n,] %*% delta_random[sdata$group_1[n],])
-        gamma[n,] = t(gamma_fixed) %*% sdata$X_gamma[n,] + gamma_random[[sdata$group_1[n]]] %*% sdata$Z_gamma_1[n,]
+```
+
+# Hierarchical ordinal regression models in bhsdtr
+
+This document contains a short tutorial on fitting hierarchical
+ordinal polytomous regression models using the bhsdtr package, but
+there is more; It also contains some explanations, illustrated with
+annotated code, of several issues related to ordinal models in
+general, and Signal Detection Theory and Item Response Theory in
+particular. If you are interested in SDT, IRT, or ordinal models in
+general, perhaps you will find something useful, new, or even mildly
+provocative here.
+
+Anyway, it is now possible to fit hierarchical (more than two possible
+values) ordinal regression models in bhsdtr by using order-preserving
+link functions for the thresholds. If you have a bunch of Likert-type
+items, or confidence ratings, or PAS ratings, or anything of this sort
+in your data, then this model may suit your needs. To my knowledge
+this is a new kind of hierarchical ordinal model that can be used to
+estimate variability in the thresholds in a way that until recently
+was not possible.
+
+Ordinal polytomous variables are common in psychology, they are
+usually analyzed by fitting a linear model, and all sorts of bad
+things happen because of this, but only a few people seem to care. I
+for one do not always care about this. We will not address this kind
+of problems here, because we will analyze ordinal data using
+(hierarchical) ordinal models.
+
+We will use the following libraries and global variables:
+
+```{r echo=T,results="show"}
+knitr::opts_chunk$set(echo=TRUE,results="show")
+library(ggplot2)
+library(bhsdtr)
+library(Hmisc)
+library(coda)
+library(rstan)
+options(mc.cores = parallel::detectCores())
+rstan_options(auto_write = TRUE)
+## This variable indicates if the fitted models can be loaded from disk
+fresh_start = FALSE
+sumfun = function(x)c(quantile(x, c(.025, .975)), mean(x))[c(1,3,2)]
+```
+
+First, we will consider an ordinal model which assumes that there is
+an underlying normally distributed latent variable (e.g., an evidence
+distribution in Signal Detection Theory, or a latent trait in an IRT
+model) and when the value of this variable is above the latent
+threshold *i*, but below the threshold *i + i*, the response *i + 1*
+is generated. Threshold *0* is fixed at *-infinity* and threshold *K*
+is fixed at *+infinity*, which means that there are *K-1* "regular"
+thresholds and *K* possible responses. Since once we know the total
+number of observations and the number of occurrences of *K - 1*
+different responses, the number of occurrences of the only remaining
+kind of response is known, there are *K - 1* degrees of freedom in
+such data. The response variable is ordinal, and so the thresholds are
+ordered. This is how this model looks like when the thresholds are
+equidistant and centered at the mean of the latent variable
+distribution (they do not have to be equidistant nor do they have to
+be centered):
+
+```{r echo=F}
+x = seq(-3, 3, length.out = 1000)
+y = dnorm(x)
+p = ggplot(data.frame(x = x, y = y), aes(x, y)) + geom_line(lty = 2)
+crit = c(-2, -1, 0, 1, 2)
+for(i in 1:length(crit))
+    p = p + eval(substitute(geom_vline(aes(xintercept = cr)), list(cr = crit[i])))
+p = p + scale_x_continuous(breaks = crit, labels = 1:length(crit), name = 'threshold / criterion')
+for(i in 1:(length(crit) + 1))
+    p = p + eval(substitute(geom_text(aes(x, y, label = label, size = 20)),
+                            list(x = c(-3, crit)[i] + .5, y = max(y) / 2, label = paste(i))))
+p
+```
+
+The number of parameters in this model matches the degrees of freedom
+in the data, which means that the model is saturated, which means that
+it always fits the data, but it does not mean that there is no
+uncertainty associated with the estimates of its parameters.
+
+We will be using this dataset...
+
+```{r echo=T,results="show"}
+data(gabor)
+```
+
+... from which we will select two subsets of data for one condition,
+one subset (*d1*) with only the "noise" trials and the other (*d2*)
+with both the "noise" and the "signal" stimuli. We will also calculate
+the response variable by combining the information about the stimulus,
+the accuracy, and the rating.
+
+```{r echo=T,results="show"}
+## r = 1 ~ absolutely clear image left, r = 4 ~ no experience left, r
+## = 5 ~ no experience right, ... r = 8 ~ absolutely clear image right
+gabor$r = combined_response(gabor$stim, gabor$rating, gabor$acc)
+d1 = gabor[(gabor$duration == '32 ms') & (gabor$order == 'DECISION-RATING') & (gabor$stim == 0),]
+d2 = gabor[(gabor$duration == '32 ms') & (gabor$order == 'DECISION-RATING'),]
+```
+
+We cannot naively model the effects in the thresholds, because the
+thresholds have to stay ordered, and assuming, for example, that each
+threshold has a normal distribution due to the effect of participants
+would violate the ordering assumption.
+
+To supplement our model with a hierarchical linear regression
+structure the vector of the thresholds has to be translated to the
+unconstrained vector gamma. The 'log_distance' order-preserving link
+function is one solution to this problem, but other order-preserving
+link functions are also implemented in the bhsdtr package. The choice
+of the link function may sometimes be important, but, since we will
+not introduce numerical predictors for the thresholds, we can choose
+any link function which seems most convenient.
+
+The log_distance function is convenient for our purposes because it is
+simple, and because the elements of the gamma vector under this link
+function are easy to interpret and may be directly related to simple
+research questions. In particular, one of the elements of the gamma
+vector under the log_distance link function, i.e., the middle
+threshold, will soon become important. This link function represents
+the main threshold/criterion directly as a free parameter (here
+gamma[4], because K = 8, and 8 / 2 = 4); Every other criterion is
+represented as log of distance between the criterion and the one
+criterion adjacent to it which is closer to the main criterion. This
+way the order is preserved, and each threshold *i* is associated with
+the corresponding unconstrained element *i* of the gamma
+vector. Random and fixed effects in the thresholds are represented as
+(hierarchical) linear effects in gamma. The two representations are
+isomorphic, which means that we can always translate between them if
+we need to.
+
+```{r echo=T,results="show"}        
+link = 'log_distance'
+```
+
+To fit a model we have to aggregate the data (this makes the process
+of sampling much faster), create the fixed- and random-effects lists,
+and create the data structures and the model code required by
+stan. Note that here we introduce a new parameter - eta - which
+represents the average internal value. If we are going to model all
+the thresholds (gamma parameter vector) then this parameter is
+redundant, because a constant shift in the thresholds is equivalent to
+the opposite effect in eta. The bhsdtr package will automatically fix
+the first element (here the only element) of the eta fixed effects
+vector at 0.
+
+First, we will fit a model which assumes that the pattern of the
+thresholds is the same for all the participants, but the mean of the
+latent distribution may differ across participants. If this makes
+little sense to you please take a look at the picture of the ordinal
+model with ordered thresholds and imagine that the underlying
+distribution has a different mean for different participants.
+
+Note that the two assumptions, i.e., that the thresholds are constant
+across participants and latent means or samples are participant-, not
+item-specific seem to be accepted in the great majority of
+applications of Item Response Theory. This is reflected in the IRT
+terminology: thresholds are often called "item-parameters" and latent
+means or samples are often called "person-parameters". Some even say
+that IRT can provide separate estimates of the true item- and
+person-parameters.
+
+```{r echo=T,results="show"}
+fixed = list(eta = ~ 1, gamma = ~ 1)
+random = list(list(group = ~ id, eta = ~ 1))
+adata = aggregate_responses(d1, response = 'r', variables = 'id')
+model = make_stan_model(random, model = 'ordinal', gamma_link = link)
+sdata = make_stan_data(adata, fixed, random, model = 'ordinal', gamma_link = link)
+if(fresh_start){
+    fit = stan(model_code = model, data = sdata,
+                   pars = c('eta_fixed', 'gamma_fixed', 'eta_random_1', 'eta_sd_1'),
+                   init_r = .5,
+                   iter = 5000,
+                   chains = 4)
+    saveRDS(fit, '~/windows/temp/ordinal_fit.rds')
+}else{
+    fit = readRDS('~/windows/temp/ordinal_fit.rds')
 }
-    criteria = t(apply(gamma, 1, function(x){
-        exp_x_0 = exp(c(x, 0))
-        sdata$criteria_scale * qnorm(cumsum(exp_x_0 / sum(exp_x_0))[-sdata$K])
-    }))
-    distr_shift = -sdata$stim_sign * dprim / 2
-    multinomial_p[, 1] = pnorm(criteria[, 1] + distr_shift)
-    for(k in 2:(sdata$K - 1)){
-        multinomial_p[, k] = pnorm(criteria[, k] + distr_shift) - pnorm(criteria[, k - 1] + distr_shift)
-    }
-    multinomial_p[, sdata$K] = pnorm(-(criteria[, sdata$K - 1] + distr_shift))
-    data_sim = adata
-    ## Now we can simulate
-    for(r in 1:nrow(adata$counts))
-        data_sim$counts[r,] = table(c(rMultinom(t(multinomial_p[r,]), sum(adata$counts[r,])), 1:sdata$K)) - 1
-    data_sim$dprim = dprim
-    data_sim$gamma = gamma
-    data_sim$criteria = criteria
-    data_sim
-}
+print(fit, probs = c(.025, .975), pars = c('eta_fixed', 'gamma_fixed'))
+```
 
-## Works only when there is just one grouping factor
-simulate_from_fit = function(fit, adata, fixed, random, gamma_link){
-    s = apply(as.data.frame(fit), 2, mean)
-    sdata = make_stan_data(adata, fixed, random)
-    gamma_fixed = matrix(nrow = ncol(sdata$X_gamma), ncol = sdata$K - 1)
-    for(r in 1:nrow(gamma_fixed))
-        gamma_fixed[r,] = s[paste('gamma_fixed[', 1:(sdata$K - 1), ',', r, ']', sep = '')]
-    gamma_random = list()
-    for(r in 1:sdata$group_1_size)
-        gamma_random[[r]] = matrix(s[paste('gamma_random_1[', r, ',', 1:(sdata$K - 1), ',1]', sep = '')],
-                                   nrow = sdata$K - 1, ncol = sdata$Z_gamma_ncol)
-    delta_fixed = s[grep('delta_fixed\\[', names(s))]
-    delta_random = matrix(nrow = sdata$group_1_size, ncol = sdata$Z_delta_ncol_1)
-    for(r in 1:nrow(delta_random))
-        for(i in 1:ncol(delta_random))
-            delta_random[r, i] = s[sprintf('delta_random_1[%d,%d]', r, i)]
-    multinomial_p = matrix(ncol = sdata$K, nrow = sdata$N)
-    thr = gamma = matrix(ncol = sdata$K - 1, nrow = sdata$N)
-    dprim = rep(NA, sdata$N)
-    for(n in 1:sdata$N){
-        dprim[n] = exp(sdata$X_delta[n,] %*% delta_fixed + sdata$Z_delta_1[n,] %*% delta_random[sdata$group_1[n],])
-        gamma[n,] = t(gamma_fixed) %*% sdata$X_gamma[n,] + gamma_random[[sdata$group_1[n]]] %*% sdata$Z_gamma_1[n,]
-    }
-    criteria = t(apply(gamma, 1, function(x){
-        exp_x_0 = exp(c(x, 0))
-        sdata$criteria_scale * qnorm(cumsum(exp_x_0 / sum(exp_x_0))[-sdata$K])
-    }))
-    distr_shift = -sdata$stim_sign * dprim / 2
-    multinomial_p[, 1] = pnorm(criteria[, 1] + distr_shift)
-    for(k in 2:(sdata$K - 1)){
-        multinomial_p[, k] = pnorm(criteria[, k] + distr_shift) - pnorm(criteria[, k - 1] + distr_shift)
-    }
-    multinomial_p[, sdata$K] = pnorm(-(criteria[, sdata$K - 1] + distr_shift))
-    data_sim = adata
-    ## Now we can simulate
-    for(r in 1:nrow(adata$counts))
-        data_sim$counts[r,] = table(c(rMultinom(t(multinomial_p[r,]), sum(adata$counts[r,])), 1:sdata$K)) - 1
-    data_sim$dprim = dprim
-    data_sim$gamma = gamma
-    data_sim$criteria = criteria
-    data_sim
+The neff and Rhat columns look good. Here is how the model fits (this
+is a maximum likelihood fit):
+
+```{r echo=T,results="show"}
+sdt_ml_plot(model, adata, sdata)
+```
+
+So how do the "item-parameters" look like according to this model?
+
+```{r echo=T,results="show"}
+crit = gamma_to_crit(as.data.frame(fit))
+round(t(apply(crit, 2, sumfun)), 2)
+```
+
+Note that the main criterion (gamma[4,1]) is *positive*, which makes
+sense, since when the stimulus is of class "noise" the middle
+criterion / threshold / average position of the thresholds is *to the
+right* of the evidence distribution mean. Otherwise, as we will see
+later, in this particular case, and perhaps also in many other similar
+cases, these thresholds-as-"item-parameters" are probably meaningless.
+
+Now we will introduce the model which accounts for the
+inter-individual differences in the pattern of the thresholds by
+modelling the gamma random effects. We cannot simultaneously model the
+eta random effects in this model, since all the information that can
+be captured by the eta parameter (i.e., the constant shift in the
+thresholds) is already captured by the fixed and the random gamma effects.
+
+```{r echo=T,results="show"}                
+adata0 = aggregate_responses(d1, response = 'r', variables = c('id'))
+## The only element of the eta fixed effects vector will be automatically fixed at 0
+fixed0 = list(eta = ~ 1, gamma = ~ 1)
+## Each participant has a unique patterh of the thresholds
+random0 = list(list(group = ~ id, gamma = ~ 1))
+model0 = make_stan_model(random0, model = 'ordinal', gamma_link = link)
+sdata0 = make_stan_data(adata0, fixed0, random0, model = 'ordinal', gamma_link = link)
+if(fresh_start){
+    fit0 = stan(model_code = model0, data = sdata0,
+                pars = c('gamma_fixed', 'gamma_random_1', 'gamma_sd_1'),
+                init_r = .5,
+                iter = 5000,
+                chains = 4)
+    saveRDS(fit0, '~/windows/temp/ordinal_fit_0.rds')
+}else{
+    fit0 = readRDS('~/windows/temp/ordinal_fit_0.rds')
 }
@@
 
-It is now possible to fit hierarchical ordinal polytomous regression
-models in bhsdtr by using order-preserving link functions for the
-thresholds. This way not only the constant shift in the thresholds,
-but also the differences in the pattern of the thresholds can be
-accounted for in certain situations. To my knowledge this is a new
-kind of hierarchical ordinal model that can be used to estimate
-ordinal polytomous item-parameter variability in a way that until now
-was not possible. If you have a bunch of Likert-type items, or
-confidence ratings, or anything of this sort in your data, than you
-may find this model useful.
+```{r echo=T,results="show"}              
+print(fit0, probs = c(.025, .975), pars = c('gamma_fixed', 'gamma_sd_1'))
+## neff and Rhat look good
+```
 
-Because by definition a polytomous item has a multinomial distribution
-and the multinomial distribution does not have a separate parameter
-for the residual variance, when item-parameter variability is not
-accounted for by the model it can only be misplaced
+The neff and Rhat columns look good. Below we translate between the
+gamma and the criteria vectors (fixed effects = sample average):
+
+```{r echo=T,results="show"}                
+s0 = as.data.frame(fit0)
+crit0 = gamma_to_crit(s0, gamma_link = link)
+```
+
+Are the sample average thesholds different in the two models? Off
+course they are, since the models are non-linear and averaging over
+anything that affects any parameter of such models is a bad idea:
+
+```{r echo=T,results="show"}                
+crit_sm = as.data.frame(t(apply(crit, 2, sumfun)))
+crit0_sm = as.data.frame(t(apply(crit0, 2, sumfun)))
+crit_sm$model = 'fixed thresholds'
+crit0_sm$model = "random thresholds"
+crit_sm$threshold = 1:7 - .1
+crit0_sm$threshold = 1:7 + .1
+df = rbind(crit_sm, crit0_sm)
+names(df)[1:3] = c('lo', 'mean', 'hi')
+ggplot(df, aes(threshold, mean, group = model, color = model)) + geom_point() + geom_line() + geom_errorbar(aes(ymin = lo, ymax = hi))
+```
+
+The point and interval estimates of the thresholds based on the
+thresholds-as-"item-parameters" model are severely biased. In
+particular, the uncertainty in the estimates is severely
+underestimated in this model. Was it obvious to you that the model
+with no threshold random effects is such a bad idea?
+
+Now we will fit a model which is similar to an SDT model, but there
+will be two important differences. We will not assume that the answer
+"signal" cannot be less likely for the stimulus "signal" than for the
+stimulus "noise", i.e., we will not assume that d' must be
+non-negative, and we will *not* assume that the pattern of the
+thresholds is the same for "noise" and "signal" stimuli. If you know
+your Signal Detection Theory well, this may seem to you like a
+perverse idea. If that is the case, please bear with me.
+
+In this model each participant may have two different sets of
+thresholds, one for "noise" and one for "signal" stimuli. Since the
+thresholds are represented as random effects, we will account for the
+fact that "random" thresholds may be correlated, and they are - if
+some threshold is higher (lower) than the sample average, then the
+threshold above (below) it may also be higher (lower).
+
+Here we introduce the effect of the stimulus on the gamma / thresholds
+vector. We use separate intercepts and slopes parametrization (-1 +
+stim), i.e., the thresholds are to be estimated for each stimulus class
+instead of estimating them for "noise" stimuli and estimating the
+difference in the thresholds between "signal" and "noise".
+
+```{r echo=T,results="show"}                
+fixed1 = list(eta = ~ 1, gamma = ~ -1 + stim)
+random1 = list(list(group = ~ id, gamma = ~ -1 + stim))
+## We have to make sure that stim is a factor for the separate
+## intercepts and slopes parametrization to work as intended.
+d2$stim = as.factor(d2$stim)
+## Note that because for an ordinal model the stimulus variable (here
+## the 'stim' variable) is not special, we have to include it in the
+## additional variables vector
+adata1 = aggregate_responses(d2, response = 'r', variables = c('id', 'stim'))
+sdata1 = make_stan_data(adata1, fixed1, random1, model = 'ordinal', gamma_link = link)
+model1 = make_stan_model(random1, model = 'ordinal', gamma_link = link)
+if(fresh_start){
+    fit1 = stan(model_code = model1, data = sdata1,
+                pars = c('gamma_fixed', 'gamma_random_1', 'gamma_sd_1'),
+                init_r = .5,
+                iter = 5000,
+                chains = 4)
+    saveRDS(fit1, '~/windows/temp/ordinal_fit_1.rds')
+}else{
+    fit1 = readRDS('~/windows/temp/ordinal_fit_1.rds')
+}
+## We will not look at the eta parameter since it was fixed at 0
+print(fit1, probs = c(.025, .975), pars = c('gamma_fixed', 'gamma_sd_1'))
+## neff and Rhat look good
+```
+
+Now we can translate between the gamma and the criteria vectors (fixed
+effects = sample average) for each stimulus:
+
+```{r echo=T,results="show"}
+s1 = as.data.frame(fit1)
+crit1 = gamma_to_crit(s1, 1, gamma_link = link)
+crit2 = gamma_to_crit(s1, 1, beta_index = 2, gamma_link = link)
+sumfun = function(x)c(quantile(x, c(.025, .975)), mean(x))[c(1,3,2)]
+crit1_sm = apply(crit1, 2, sumfun)
+crit2_sm = apply(crit2, 2, sumfun)
+```
+
+The criteria for the stimulus class "noise" look like this:
+
+```{r echo=T,results="show"}
+round(t(crit1_sm), 2)
+```
+
+and the criteria for the stimulus class "signal" look like this:
+
+```{r echo=T,results="show"}
+round(t(crit2_sm), 2)
+```
+
+Note the difference between the main criterion (row 4). This
+difference approximates d'. Even though, as we will later see, this
+model does not correctly account for the effect of the stimulus, we can
+still recover this effect pretty well, because d' is equivalent to a
+constant shift of all the thresholds in one predefined
+direction. Shifting all the criteria to the right is equivalent to
+shifting the underlying distribution of "latent values" (or evidence)
+to the left, so we have to subtract the average position of the
+criteria for the "signal" stimuli from the average position of the
+criteria for the "noise" stimuli. Here is our first attempt:
+
+```{r echo=T,results="show"}
+dprim1 = sumfun(s1[, 'gamma_fixed[4,1]'] - s1[, 'gamma_fixed[4,2]'])
+round(dprim1, 2)
+```
+
+We will later see that this is our sample average *d'*, but it also
+isn't.
+
+In an SDT model we assume that the stimulus class affects *only* the
+d' parameter, and that both criteria (gamma) and *d'* may differ
+between the participants (or items, etc.). In bhsdtr d' is modelled as
+log(d') = delta, which forces it to be non-negative. This is an
+important assumption of Signal Detection Theory - if this assumption
+is violated, for example, because of the response reversal in some
+participants or conditions, then an SDT model is not valid and a
+non-trivial generalization of SDT should be used instead (see the
+bhsdtr preprint for a more detailed explanation and some
+references). Otherwise, this is the same general hierarchical ordinal
+regression model.
+
+```{r echo=T,results="show"}   
+fixed2 = list(delta = ~ 1, gamma = ~ 1)
+random2 = list(list(group = ~ id, delta = ~ 1, gamma = ~ 1))
+```
+
+Because of the importance of the stimulus class ("noise", or 1, and
+"signal", or 2) in SDT models we have to indicate which variable
+represents this information ( aggregate_responses(..., stimulus =
+'stim', ...) ).
+
+```{r echo=T,results="show"}
+adata2 = aggregate_responses(d2, stimulus = 'stim', response = 'r', variables = c('id'))
+sdata2 = make_stan_data(adata2, fixed2, random2, model = 'sdt', gamma_link = link)
+model2 = make_stan_model(random2, model = 'sdt', gamma_link = link)
+if(fresh_start){
+    fit2 = stan(model_code = model2, data = sdata2,
+                pars = c('delta_fixed', 'gamma_fixed', 'delta_sd_1', 'gamma_sd_1',
+                         'delta_random_1', 'gamma_random_1',
+                         ## We will need this for the plot_sdt_fit function
+                         'counts_new'),
+                init_r = .5,
+                iter = 5000,
+                chains = 4)
+    saveRDS(fit2, '~/windows/temp/ordinal_fit_2.rds')
+}else{
+    fit2 = readRDS('~/windows/temp/ordinal_fit_2.rds')
+}
+print(fit2, probs = c(.025, .975), pars = c('delta_fixed', 'gamma_fixed', 'delta_sd_1', 'gamma_sd_1'))
+## neff and Rhat look good
+```
+
+So how well do the two d' estimates match? We can translate between
+the delta and the d' parameters by using exponentiation. We have to do
+this first, before doing anything else with the posterior samples,
+because the exponential function is non-linear.
+
+```{r echo=T,results="show"}
+s2 = as.data.frame(fit2)
+dprim2 = sumfun(exp(s2[,'delta_fixed[1,1]']))
+round(cbind(dprim1, dprim2), 2)
+```
+
+The estimates match poorly. Or do they? What about the individual *d'*
+estimates? To calculate the individual shifts in the thresholds /
+criteria we have to add the random effects to the fixed effect:
+
+```{r echo=T,results="show"}   
+dprim1i = (s1[['gamma_fixed[4,1]']] + s1[, grep('gamma_random_1\\[.+,4,1]', names(s1))]) -
+    (s1[['gamma_fixed[4,2]']] + s1[, grep('gamma_random_1\\[.+,4,2]', names(s1))])
+dprim1i = as.data.frame(t(apply(dprim1i, 2, sumfun)))
+dprim2i = as.data.frame(t(apply(exp(s2[['delta_fixed[1,1]']] +
+                                    s2[, grep('delta_random_1', names(s2))]), 2, sumfun)))
+names(dprim1i) = names(dprim2i) = c('lo', 'mean', 'hi')
+dprim1i$model = 'ordinal'
+dprim1i$i = 1:nrow(dprim1i) - .1
+dprim2i$model = 'sdt'
+dprim2i$i = 1:nrow(dprim2i) + .1
+ggplot(rbind(dprim1i, dprim2i), aes(i, mean, group = model, color = model)) +
+    geom_point() + geom_errorbar(aes(ymin = lo, ymax = hi)) +
+    geom_abline(intercept = 0, slope = 0) +
+    ylab('d\'') +
+    labs(title = sprintf('Correlation between individual d\' estimates: r = %.2f',
+                         cor(dprim1i$mean, dprim2i$mean)))
+```
+
+The point and interval estimates match quite well, with some
+notable exceptions. According to the ordinal model some credible
+intervals around individual d' point estimates include 0 and one point
+estimate is even below 0, which contradicts an important assumption of
+the SDT model. We will address this issue later.
+
+If the participant-specific d' estimates match, the sample average d'
+should also match, right? Maybe the SDT model does not fit the data
+well?
+
+```{r echo=T,results="show"}   
+plot_sdt_fit(fit2, adata2, c('id'), type = 'response', bw = F, verbose = F)
+```
+
+The SDT model seems to fit very well in a sense that the observed
+distributions of responses are well within the 95% posterior
+predictive intervals for each participant. However, this does not mean
+much; an SDT model with multiple criteria is not saturated, so it does
+not have to fit the data, but it does not leave too many degrees of
+freedom in the data, so we can safely assume a priori that it will
+usually fit the data well, regardless if it is true or not.
+
+The discrepancy between the sample average *d'* estimates from both
+models is not really a consequence of allowing for negative *d'*
+values in the ordinal model. This discrepancy is a consequence of the
+assumed distribution of individual shifts / *d'* values. In the
+ordinal model we assumed that the shifts (*= d'*) are normally
+distributed, whereas in the SDT model we assumed that *log(d')* are
+normally distributed. We can see if the assumption that *log(d')* (or
+*d'*) are normally distributed is approximately correct in our
+dataset. First, let's take a look at participant-specific delta
+estimates in the SDT model:
+
+```{r echo=T,results="show"}
+ggplot(data.frame(delta = log(dprim2i$mean)), aes(sample = delta)) +
+    stat_qq() +
+    ylab('Sample quantile of log(d\') = delta random effects') +
+    xlab('Theoretical normal quantile') +
+    labs(title = 'SDT model')
+```
+
+The individual *delta = log(d')* estimates in the SDT model seem to be
+approximately normally distributed. We cannot plot log(d') estimates
+based on the ordinal model, because according to this model one
+participant has a negative *d'* value. However, we can see if the *d'*
+point estimates based on the ordinal model are approximately normally
+distributed:
+
+```{r echo=T,results="show"}   
+ggplot(data.frame(dprim = dprim1i$mean), aes(sample = dprim)) +
+    stat_qq() +
+    ylab('Sample quantile of d\' random effects') +
+    xlab('Theoretical normal quantile') +
+    labs('Ordinal model')
+```
+
+They seem to be approximately normally distributed as well. The
+average / fixed effect *d'* in each model is estimated as the mean of
+different distributions: in the ordinal model participant-specific
+shifts are assumed to be normally distributed and in the SDT model
+delta = log(d') random effects are assumed to be normally
+distributed. We can easily account for this by making use of the
+relationship between the parameters of the normal distribution and the
+associated log-normal distribution:
+
+```{r echo=T,results="show"}   
+round(cbind(sumfun(exp(s2[['delta_sd_1[1,1]']]^2 / 2 + s2[['delta_fixed[1,1]']])),
+            dprim1), 2)
+```
+
+Now we see that the two models give very similar point and interval
+estimates of sample-average *d'*.
+
+To summarize, the SDT model is much simpler (it has only one set of
+thresholds for each participant), it fits the data well at the
+individual level, and it correctly assumes that the individual *d'* /
+shift values are log-normally distributed. SDT is a clear winner,
+right? Not yet it isn't.
+
+An SDT model is not only a statistical model, it also has *causal*
+assumptions. Because the stimulus class was chosen at random on every
+trial, every estimate of the statistical effect of the stimulus class
+on any parameter of the model is an unbiased estimate of the total
+causal effect of the stimulus class on this parameter. Moreover, in
+such designs every statistical effect of the participant factor is an
+estimate of the property or the causal effect of the
+participants. This means that both the individual *d'* parameters and
+the individual thresholds are participant-, not item-, parameters in
+this model. So far so good.
+
+However, in the SDT model we have also assumed that the stimulus class
+can *only* affect the distance between the evidence distribution means
+(= *d'*), whereas in the ordinal model we allowed for the possibility
+of the stimulus class having an effect on the thresholds. It may seem
+odd to you to allow for this, but there is nothing in the design of
+the study that justifies the causal assumption of selective
+influence. The estimates of the thresholds are based on the data that
+were generated after the stimulus was presented, and so,
+theoretically, the *pattern* of the thresholds could be affected by
+the stimulus class as well.
+
+The selectivity assumption is central to Signal Detection Theory. If
+you know your Signal Detection Theory well, please take a moment and
+try to recall one instant of this important assumption being
+explicitly addressed in a paper or a book on Signal Detection Theory.
+
+It does not matter if an SDT model with selectivity assumption fits
+the data well, because if the effect of the stimulus class is not
+selective, the model is simply false. In particular, when the
+selectivity assumption is false, an SDT model cannot deconfound
+sensitivity / *d'* from bias / threshold / criteria effects.
+
+Before we go any further, just for fun, let's see what happens if we
+introduce yet another assumption of selective influence, one that is
+often accepted in Item Response Theory. We will fit a hierarchical SDT
+model in which we will call the thresholds "item-parameters" and we
+will call *d'* a "person-parameter".
+
+```{r echo=T,results="show"}
+fixed3 = list(delta = ~ 1, gamma = ~ 1)
+random3 = list(list(group = ~ id, delta = ~ 1))
+adata3 = aggregate_responses(d2, stimulus = 'stim', response = 'r', variables = c('id'))
+sdata3 = make_stan_data(adata3, fixed3, random3, model = 'sdt', gamma_link = link)
+model3 = make_stan_model(random3, model = 'sdt', gamma_link = link)
+if(fresh_start){
+    fit3 = stan(model_code = model3, data = sdata3,
+                pars = c('delta_fixed', 'delta_sd_1', 'gamma_fixed', 'delta_random_1',
+                         ## We will need this for the plot_sdt_fit function
+                         'counts_new'),
+                init_r = .5,
+                iter = 5000,
+                chains = 4)
+    saveRDS(fit3, '~/windows/temp/ordinal_fit_3.rds')
+}else{
+    fit3 = readRDS('~/windows/temp/ordinal_fit_3.rds')
+}
+print(fit3, probs = c(.025, .975), pars = c('delta_fixed', 'gamma_fixed', 'delta_sd_1'))
+## neff and Rhat look good
+```
+
+Does the model fit?
+
+```{r echo=T,results="show"}
+plot_sdt_fit(fit3, adata3, type = 'response', bw = F, verbose = F)
+```
+
+At the aggregated data level the fit is not perfect, but it is not
+that terrible either. What about the fit at the individual level?
+
+```{r echo=T,results="show"}
+plot_sdt_fit(fit3, adata3, 'id', type = 'response', bw = F, verbose = F)
+```
+
+A few points are outside the predictive intervals, but maybe that's
+ok? What if we compared the point and interval estimates of the so-called "person-parameters"?
+
+```{r echo=T,results="show"}   
+s3 = as.data.frame(fit3)
+dprim3i = as.data.frame(t(apply(exp(s3[['delta_fixed[1,1]']] + s3[, grep('delta_random_1', names(s3))]), 2, sumfun)))
+names(dprim3i) = c('lo', 'mean', 'hi')
+dprim3i$model = 'person- and item-parameters'
+dprim3i$i = 1:nrow(dprim3i) - .1
+ggplot(rbind(dprim2i, dprim3i), aes(i, mean, group = model, color = model)) +
+    geom_point() + geom_errorbar(aes(ymin = lo, ymax = hi)) +
+    geom_abline(intercept = 0, slope = 0) +
+    labs(title = sprintf('Correlation between individual d\' estimates: r = %.2f', cor(dprim2i$mean, dprim3i$mean)))    
+```
+
+This does not look good at all. In particular, some of the point and
+interval estimates are clearly biased. If we compare the average width
+of the intervals...
+
+```{r echo=T, results="show"}
+round(mean(dprim3i$hi - dprim3i$lo) / mean(dprim2i$hi - dprim2i$lo), 2)
+```
+
+... we see that the simplified model not only gives biased
+"person-parameter" estimates but the estimates seem more reliable than
+they truly are (here by about 20%). It does not make sense to
+look at the estimates of the "item-parameters" since we already know
+that they have little in common with the true thresholds.
+
+Now back to Signal Detection Theory and the assumption of selective
+influence. We can at least try to test this assumption. We will not
+alter the ordinal model code to account for the non-negativity of d'
+(and the log-normal distribution of random d' effects) just yet. As we
+already know, even without this correction the estimates of
+participants' shifts are similar to the *d'* estimates.
+
+Note that we do not use the separate intercepts and slopes
+parametrization here (i.e., ~ -1 + stim), because we are very much
+interested in the *effect* of the stimulus class on the *pattern* of
+the thresholds.
+
+```{r echo=T,results="show"}   
+fixed4 = list(eta = ~ 1, gamma = ~ stim)
+random4 = list(list(group = ~ id, gamma = ~ stim))
+adata4 = aggregate_responses(d2, response = 'r', variables = c('id', 'stim'))
+sdata4 = make_stan_data(adata4, fixed4, random4, model = 'ordinal', gamma_link = link)
+model4 = make_stan_model(random4, model = 'ordinal', gamma_link = link)
+if(fresh_start){
+    fit4 = stan(model_code = model4, data = sdata4,
+                pars = c('gamma_fixed', 'gamma_random_1', 'gamma_sd_1'),
+                init_r = .5,
+                iter = 5000,
+                chains = 4)
+    saveRDS(fit4, '~/windows/temp/ordinal_fit_4.rds')
+}else{
+    fit4 = readRDS('~/windows/temp/ordinal_fit_4.rds')
+}
+print(fit4, probs = c(.025, .975), pars = c('gamma_fixed', 'gamma_sd_1'))
+```
+
+Here we see the estimated average shift / *d'* thought of as the mean
+of normally distributed *d'* (not *delta = log(d')*) random effects
+again:
+
+```{r echo=T,results="show"}
+s4 = as.data.frame(fit4)
+round(cbind(sumfun(-s4[['gamma_fixed[4,2]']]), dprim1), 2)
+```
+
+Ok, we did simplify things significantly, but it still makes sense to
+take a look at the other gamma effects, because these other effects
+correspond to the variability in the *pattern* of the thresholds, and
+the estimates of the thresholds in our ordinal model seem perfectly
+fine.
+
+```{r echo=T,results="show"}   
+round(HPDinterval(as.mcmc(s4[, grep('gamma_fixed', names(s4))])), 2)
+```
+
+Among the parameters that represent the stimulus-induced average
+differences between the gamma parameters (i.e., gamma_fixed[.,2]) the
+gamma_fixed[4,2] parameter does not seem to be the only one that is
+non-zero. This indicates that the pattern of the thresholds - not only
+their average position - is affected by the stimulus
+class. Interestingly, the random effects' standard deviations tell a
+different story:
+
+```{r echo=T,results="show"}   
+round(HPDinterval(as.mcmc(s4[, grep('gamma_sd_1', names(s4))])), 2)
+```
+
+Note the clear pattern for the gamma_sd_1[.,2] parameters. Even though
+the *sample-average* pattern of the thresholds seems to be different
+for the two stimulus classes (gamma_fixed[.,2] except for
+gamma_fixed[4,2] ~ d'), this difference in pattern seems to be fairly
+constant across the participants: all the gamma_sd_1[.,2] parameters
+seem to be near-zero (the 95% credible intervals contain zero), except
+for the gamma_sd_1[4,2] parameter, which corresponds to the
+variability in *d'* values. This is *really* weird, since the pattern
+of the thresholds for the "noise" stimuli clearly varies with
+participants - none of the 95% HPD intervals for the gamma_sd_1[.,1]
+parameters include zero: in terms of standard deviations of posterior
+distributions of standard deviations (phew) most are in fact quite far
+from zero.
+
+If only we could do a better job and use a model which respects the
+non-negativity assumption. Is it really that complicated? Let's index
+the rows of our dataset by *i*. Because of the way the model was
+parametrized...
+
+```{r echo=T,results="show"}   
+head(sdata4$X_gamma)
+```
+
+when stim = "noise" = 0, gamma[4]_i = gamma_fixed[4,1] +
+gamma_random_1[g_i][4,1], where g_i is the participant index in row
+*i*. We can leave this part of the model intact, since this is the
+main criterion position, which is an unconstrained parameter, but
+also, as we can see below, the observed distribution of the main
+criterion random effects is approximately normal:
+
+```{r echo=T,results="show"}
+ggplot(data.frame(gamma_4 = apply(s4[['gamma_fixed[4,1]']] + s4[, grep('gamma_random_1\\[.+,4,1]', names(s4))], 2, mean)),
+       aes(sample = gamma_4)) +
+    stat_qq() +
+    ylab('Sample quantile of main criterion random effects') +
+    xlab('Theoretical normal quantile') +
+    labs('Ordinal model')
+```
+
+However, when stim = "signal" = 1, gamma[4]_i = gamma_fixed[4,1] +
+gamma_random_1[g][4,1] + gamma_fixed[4,2] + gamma_random_1[g][4,2]. We
+want this other part (the last sum which represents the difference in
+gamma[4] between the "noise" and "signal" stimuli) to represent log of
+average distance between the thresholds = shift = *d'*, so we have to
+subtract it from gamma[4]_i, exponentiate this sum, and add it again
+to gamma[4]_i.
+
+```{r echo=T,results="show"}   
+fixed5 = list(eta = ~ 1, gamma = ~ stim)
+random5 = list(list(group = ~ id, gamma = ~ stim))
+adata5 = aggregate_responses(d2, response = 'r', variables = c('id', 'stim'))
+sdata5 = make_stan_data(adata5, fixed5, random5, model = 'ordinal', gamma_link = link)
+model5_ = make_stan_model(random5, model = 'ordinal', gamma_link = link)
+## Here we introduce the necessary change in the model code. Kb2
+## stands for K by 2 = the middle criterion index.
+model5 = gsub('criteria\\[Kb2] = gamma\\[Kb2];',
+              'if(X_gamma[n,2] == 0){
+                  criteria[Kb2] = gamma[Kb2];
+               }else{ 
+                  criteria[Kb2] = gamma[Kb2] - (gamma_fixed[Kb2,2] + gamma_random_1[group_1[n]][Kb2,2]) - exp(gamma_fixed[Kb2,2] + gamma_random_1[group_1[n]][Kb2,2]);
+               }',
+              model5_)
+```
+
+This way the gamma[4,1] parameter represents the position of the main
+criterion and gamma[4,2] = delta = log(d').
+
+```{r echo=T,results="show"}   
+if(fresh_start){
+    fit5 = stan(model_code = model5, data = sdata5,
+                pars = c('gamma_fixed', 'gamma_random_1', 'gamma_sd_1'),
+                init_r = .5,
+                iter = 5000,
+                chains = 4)
+    saveRDS(fit5, '~/windows/temp/ordinal_fit_5.rds')
+}else{
+    fit5 = readRDS('~/windows/temp/ordinal_fit_5.rds')
+}
+print(fit5, probs = c(.025, .975), pars = c('gamma_fixed', 'gamma_sd_1'))
+```
+
+Now let's have a look at the *d'* estimate (fixed effect) and compare
+it with the *d'* estimate in the SDT model:
+
+```{r echo=T,results="show"}
+s5 = as.data.frame(fit5)
+round(cbind(sumfun(exp(s5[['gamma_fixed[4,2]']])), dprim2), 2)
+```
+
+This is close. Not the same, but the model is different. What about
+the bias = main criterion relative the the midpoint between the main
+criteria?
+```{r echo=T,results="show"}   
+round(cbind(sumfun(s5[['gamma_fixed[4,1]']] - exp(s5[['gamma_fixed[4,2]']]) / 2),
+      sumfun(s2[['gamma_fixed[4,1]']])), 2)
+```
+
+Also close.
+
+We have fitted a hierarchical SDT model which respects the
+non-negativity assumption, the ordering of the thresholds, accounts
+for the log-normal distribution of the *d'* random effects, allows for
+random effects in both the *d'* and the individual threshold
+parameters while still respecting the SDT assumptions (thanks to an
+order-preserving gamma link function and the delta log link function
+which respects the non-negativity assumption), and accounts for the
+correlations between the random effects. First such model was
+described in the bhsdtr preprint. This model, however, does not assume
+that the stimulus class affects only the *d'* parameter. We can now
+again see if there is any evidence of non-selectivity:
+
+```{r echo=T,results="show"}   
+round(HPDinterval(as.mcmc(s5[, grep('gamma_fixed', names(s5))])), 2)
+```
+
+```{r echo=T,results="show"}                
+round(HPDinterval(as.mcmc(s5[, grep('gamma_sd_1', names(s5))])), 2)
+```
+
+We see again that gamma_sd_1[.,2] parameters, except for
+gamma_sd_1[4,2] = log(d'), may be zero - the effect of the stimulus on
+all the log-distances seems to be more or less the same for each
+participant. What is going on here?
+
+Interaction is tricky, you see. For example, if you have an additive
+effect of two factors on the original scale of the dependent
+variable...
+
+```{r echo=T,results="show",fig.asp=.6}   
+df = expand.grid(f1 = 1:2, f2 = 1:2)
+df$y = df$f1 + df$f2
+df$f1 = as.factor(df$f1)
+df$f2 = as.factor(df$f2)
+ggplot(df, aes(f1, y, group = f2, color = f2)) + geom_line()
+```
+
+...then after applying a non-linear transformation to the dependent
+variable you may obtain a significant interacive effect:
+
+```{r echo=T,results="show",fig.asp=.6}
+ggplot(df, aes(f1, exp(y), group = f2, color = f2)) + geom_line()
+```
+
+This means that whenever you see an interactive effect it may be an
+artefact of the scale of the dependent variable! Every time I say it
+it makes me feel uneasy. So, instead of examining the effects in the
+gamma parameters. let's have a look at the average *criteria /
+thresholds*. Notice that we have to add the effect on gamma to gamma
+for stimulus = "noise" to obtain gamma for stimulus = "signal".
+
+```{r echo=T,results="show"}   
+crit1 = gamma_to_crit(s5, gamma_link = link)
+crit2 = gamma_to_crit(s5[, grep('gamma_fixed\\[.+,1]', names(s5))] + s5[, grep('gamma_fixed\\[.+,2]', names(s5))], gamma_link = link)
+```
+
+Now we can directly test if the *pattern* of the criteria may be
+different by comparing *centered* criteria vectors:
+```{r echo=T,results="show"}
+round(HPDinterval(as.mcmc((crit2 - apply(crit2, 1, mean)) - (crit1 - apply(crit1, 1, mean)))), 2)
+```
+
+When we center the criteria for the two classes of stimuli we no
+longer see any evidence of the differences in their relative
+positions. The only thing that seems to differ, and not by much, is
+the main criterion, but this is confounded with *d'* in our model (I
+think). Selectivity seems to hold, kind of.
+
+This is how you *begin* to test a hierarchical SDT model. An SDT model
+has other problematic assumptions. One is that the time-accuracy
+tradeoff can be ignored. If participants or conditions differ in the
+response time, then the differences in *d'* may not mean that
+sensitivity is different, because participants may simply take more
+time to respond in certain experimental conditions, which inflates the
+*d'* estimate. You need something like the diffusion model to deal
+with this problem. If you combine the diffusion model with ratings,
+you also need something like a hierarchical diffusion model with
+ratings and an order-preserving link function. Sounds like something
+that could perhaps be implemented in the bhsdtr package in the future.
+
+The other problems that need to be addressed have to do with
+non-linearity and aggregation. If you fit an SDT model, or any other
+non-linear model, to data aggregated over grouping factors, such as
+participants or items, because of non-linearity, which does not play
+well with aggregation, some or all of the point and interval estimates
+will be biased, perhaps severely so, and the inference will be
+invalid. You can see how horribly biased the point *and* interval
+estimates of every SDT parameter may become when you overly aggregate
+your data in the bhsdtr preprint. I am now in the process of examining
+the direction and magnitude of bias in SDT models fitted to
+overly-aggregated data by using the excellent Confidence database
+created by Doby Rahnev.
+
+All this means that every SDT analysis to this day that was done using
+data aggregated over participants is invalid. In particular, to my
+knowledge, to this day there is little *valid* evidence that the
+Unequal Variance SDT model fits the data better than the Equal
+Variance SDT model.
+
+Note also that here we were aggregating over trials, which means that,
+if the SDT parameters may differ between trials, and they certainly do
+differ, the only question is by how much, then all the SDT estimates
+that I have presented to you are biased to an unknown extent. There is
+a lot to be done before we can say how well does the humble SDT model
+deconfound sensitivity from bias in any particular case.
+
+What if we tried this trick with testing the selectivity assumption,
+but using the Unequal Variance SDT model, or the meta-d' model, or
+some other generalization of the SDT model? We cannot introduce any
+new parameters to our model, because it is already saturated. What we
+can do, for example, is fit an Unequal Variance SDT model, estimate
+the ratio of the variances, fix the ratio, introduce it to the
+saturated ordinal model, thus turning it into the Fixed Unequal
+Variance Ordinal Model. If there is evidence of non-selectivity in
+this model, we can keep trying with other modifications. If
+non-selectivity disappears, we have evidence that the selectivity
+assumption may be correct *and* that the Unequal Variance SDT model
+may be better than the Equal Variance SDT model.
+
+# A rant about Item Response Theory
+
+An SDT model is essentially an ordinal model with some additional
+assumptions (e.g., selectivity, non-negativity of d', no time-accuracy
+tradeoff issues, etc.). When there is more than one criterion /
+threshold this model represents the distribution of polytomous ordinal
+responses. Because by definition a polytomous item has a multinomial
+distribution and the multinomial distribution does not have a separate
+parameter for the residual variance, when item-parameter variability
+is not accounted for by the model it can only be hidden or misplaced
 (overdispersion). The correct way to account for item-parameter
@@
 some other, still higher, threshold. The thresholds are assumed to be
-ordered, because otherwise the item would not have the intended
+ordered, because otherwise, the item would not have the intended
 meaning and thus would not be a valid instrument of measuring
@@
 there, the question is if it is non-negligible, and if so, what can be
-done about it. Unfortunately, without additional assumptions it is
-impossible to model both kinds of variability at the same time even in
-the hypothetical repeated measurement with memory wipeout scenario,
-because every shift in the internal value is indistiguishible from a
+done about it.
+
+Unfortunately, without additional assumptions, it is impossible to
+model both kinds of variability at the same time even in the
+hypothetical repeated-measurement-with-memory-wipeout-scenario,
+because every shift in the internal value is indistinguishable from a
 uniform effect in the thresholds, i.e., a constant shift of all the
-thresholds. As I will explain shortly, the required additional
+thresholds. As I hope you already realize, the required additional
 assumptions that make such models identifiable under certain
@@
 
-In many situations this is both a statistical and a conceptual
-problem. In particular, when the origin of the internal value is
-unknown because the design is purely observational (e.g.,
-questionnaire measurement), the way of describing the process of
-responding to a Likert-type item in terms of internal values and
-thresholds is simply redundant. As far as the design of the study is
-concerned, the value of the internal evidence "comes from nowhere" -
-all we can say about it is that it is relative to the thresholds - and
-the position of each threshold is relative to the positions of other
-thresholds and to the position of the internal value. In this case to
-say that John thinks that he was not crying at all during the last two
-weeks and to say that *John's* threshold for the response "a few
-times" is above the amount of crying that John thinks he experienced,
-or decided to reveal, is to say the same thing.
+In many situations, this is both a statistical and a conceptual
+problem. In particular, when the design is purely observational -
+e.g., questionnaire measurement - without additional causal
+assumptions, the way of describing the process of responding to a
+Likert-type item in terms of internal values and thresholds is simply
+redundant. As far as the design of the study is concerned, all we can
+say about the internal value is that it is relative to the thresholds
+- and the position of each threshold is relative to the positions of
+the remaining thresholds and to the position of the internal
+value. Without the selectivity assumption, to say that John thinks
+that he was not crying at all during the last two weeks and to say
+that *John's* threshold for the response "a few times" is above the
+amount of crying that John thinks he experienced, or decided to
+reveal, is to say the same thing.
 
-It follows that *in such situations we do not need the internal value
+It follows that *in such situations, we do not need the internal value
 parameter at all*, unless there is some factor about which we can
@@
 change (*d'*) in the internal value only; In such cases the internal
-value parameter may be justified and useful, because it does constrain
-the model: in the repeated measurement case the model is no longer
-saturated.
+value parameter may be justified and useful because it does constrain
+the model. We have obtained evidence that this assumption perhaps
+holds for the analyzed dataset.
 
-But what about the promise of Item Response Theory? Isn't it true that
-we can separate the internal values (so called "person parameters")
-from the thresholds (so called "item parameters") and place them on a
-common scale? Once the participants leave the room the items retain
-their estimated "difficulty levels", right? The item difficulty
-(=thresholds) is not defined relative to some sample of participants
-who filled the questionnaire. How is this possible if the two kinds of
-effects are completely confounded? In particular, how does Item
-Response Theory achieve when each person provides only a single
-response to each item?
+Notice that an SDT model is just an Item Response Theory model with
+the cumulative distribution and a non-negative effect of the stimulus
+class on the latent mean. So what about the central promise of Item
+Response Theory?  Isn't it true that we can separate the internal
+values ("person parameters") from the thresholds ("item parameters")
+and place them on a common scale? Once the participants leave the room
+the items retain their estimated "difficulty levels", right? The item
+difficulty (=thresholds) is not defined relative to some sample of
+participants who filled the questionnaire. How is this possible if the
+two kinds of effects are completely confounded? In particular, how
+does Item Response Theory achieve this magical separation when *each
+person provides only a few responses, usually just one, to each item*?
 
-It is not an *achievment* of Item Response Theory, but it's
-(unrealistic) *assumption*. The illusory nature of this achievement
-becomes obvious as soon as we realize that without additional
-(causal!)  assumptions it is impossible to distinguish between genuine
+It is usually not an *achievement* of Item Response Theory, it is just
+an unrealistic *assumption*. The illusory nature of this achievement
+becomes obvious as soon as we realize that without additional strong
+causal assumptions it is impossible to distinguish between genuine
 internal value effects and (dis)simulation (i.e., faking), which
-conteptually corresponds to the change in the *participant*-specific,
+conceptually corresponds to the change in the *participant*-specific,
 not item-specific, decision thresholds, also known as
-"item-parameters". There is nothing in the *desing* of the process of
+"item-parameters". There is nothing in the *design* of the process of
 measurement by a questionnaire that makes it possible to distinguish
 genuine differences in internal values from the differences in the way
-of responding. This "achievement" of Item Response Theory is a
-consequence of treating the "item-parameters" as constant accross
-participants when fitting an IRT model.
+of responding.
 
@@
 performance in a binary classification task cannot be faked, and so
-the *d'* paramater definately captures at least part of the effect of
+the *d'* parameter definitely captures at least part of the effect of
 the stimulus on the internal value. On the other hand, every possible
@@
 identifiable if there are repeated measurements for at least one
-participant and one item or a large number of participants provides
-single responses to all of the items. In particular, in typical
-applications of IRT it is simply *assumed* in the model that the
-participants differ in the internal values, and the item parameters
-are constant accross the sample of participants (selectivity).
+participant and one item or many participants provide single responses
+to all of the items.
 
-In every book on Item Response Theory that I know of the issue of
-measurement invariance due to the variability in the "item-parameters"
-is reduced to the problem of Differential Item Functioning
+In the context of Item Response Theory the issue of measurement
+non-invariance due to the variability in the "item-parameters" is
+often reduced to the problem of Differential Item Functioning
 (DIF). However, the variability in "item-parameters" that I talk about
@@
 effect; An estimate of DIF is obtained when the estimates of
-"item-parameters" obtained for two different (large) samples of
+"item-parameters" obtained for two different (often large) samples of
 participants are compared. However, these estimates are obtained by
-fixing the "item-parameters" within each sample. It does not take much
-thought to see that the assumption of constant "item-parameters" at
-the group/population/sample level is utterly unrealistic. It follows
-that the achievement of separating participant and item effects in IRT
-models is not an achievement at all, it is just wishful thinking
-combined with aggressive marketing.
-
-Some of the problems related to item-parameter variability can be
-addressed *when additional causal assumptions of selective influence
-are introduced and the design of the study makes the model
-identifiable*. For example, we can use the data from a binary
-classification task and model only the responses for one kind of
-stimulus, e.g., "noise".
-
-Here is a simple example which illustrates how we can fit a
-hierarchical ordinal polytomous regression model using the bhsdtr
-package. We will be using the data from one condition in the gabor
-dataset:
-
-```{r c1, echo=T}
-library(ggplot2)
-library(bhsdtr)
-library(Hmisc)
-library(rstan)
-options(mc.cores = parallel::detectCores())
-rstan_options(auto_write = TRUE)
-
-## r = 1 ~ absolutely clear image left, r = 4 ~ no experience left,
-## r = 5 ~ no experience right, ... r = 8 ~ absolutely clear image right
-gabor$r = combined_response(gabor$stim, gabor$rating, gabor$acc)
-d1 = gabor[gabor$duration == '32 ms' & gabor$order == 'DECISION-RATING',]
-## Only the left-tilted gabor patches
-d1 = d1[d1$stim == 0,]
-```
-
-To fit a hierachical model we have to aggregate the data (this makes
-the process of sampling much faster), create the fixed- and
-random-effects lists, and create the data structures and the model
-code required by stan. Note that we have introduced a new parameter -
-eta - which represents the internal value. As we already know, without
-the selectivity assumption the difference between item and person
-parameters is arbitrary and so in bhsdtr the first element of the
-fixed effects vector for the eta parameter is automatically fixed at 0
-eto make the model identifiable (this bahavior may change in the near
-future):
-
-```{r c2, echo = T}        
-## This order-preserving link function represents the main criterion
-## directly as a free paramter and all the other criteria as log of
-## distance between the criterion and the criterion adjacent to it
-## which is closer to the main criterion
-link = 'log_distance'
-adata = aggregate_responses(d1, response = 'r', variables = c('id'))
-fixed = list(eta = ~ 1, gamma = ~ 1)
-random = list(list(group = ~ id, gamma = ~ 1))
-model = make_stan_model(random, model = 'ordinal', gamma_link = link)
-sdata = make_stan_data(adata, fixed, random, model = 'ordinal', gamma_link = link)
-```
-
-To simplify the task we will use maximum likelihood optimization,
-since the model is saturated anyway:
-
-```{r c3, echo=T}
-o = optimizing(stan_model(model_code = model), sdata, init = 0)
-```
-
-Unsuprisingly, the model fits perfectly for every participant:
-
-```{r p1}
-sdt_ml_plot(model, adata, sdata)
-```
-
-Now we will add the estimated gamma/criteria random effects to the
-fixed effects to obtain participant-specific gamma/criteria estimates,
-convert between the gamma and the criteria vectors, and take a look at
-the estimated variability in the criteria:
-
-```{r c4,echo=T}
-gamma_fixed = o$par[grep('gamma_fixed\\[', names(o$par))]
-gamma_random = matrix(o$par[grep('gamma_random_1', names(o$par))], nrow = sdata$group_1_size)
-for(i in 1:nrow(gamma_random))
-    gamma_random[i,] = gamma_random[i,] + gamma_fixed
-criteria = gamma_to_crit(gamma_random, beta_index = NULL, gamma_link = link)
-df = expand.grid(criterion = 1:7, id = 1:nrow(criteria))
-df$estimate = as.vector(t(criteria))
-ggplot(df, aes(id, estimate, group = id)) + geom_line() + geom_point()
-```
-The thresholds may look a bit weird, but we have to rememberd that
-what we see here is *a mixture of d' and threshold effects*. The
-estimates are somewhat similar to what we would have obtained if we
-fitted an SDT model, but they clearly do not have the intended
-interpretation; They do not deconfound the two things that an SDT
-model is supposed to deconfound, namely sensitivity and bias. Contrary
-to what seems to be commonly assumed, if we included the stimulus
-class we would not solve the problem - for that we have to also
-introduce the strong causal assumption of selective influence, which,
-like any other strong assumption, can easily be false.
-
-The justification of the selective influence assumption comes from a
-psychological theory of the task, not from the design of the
-study. All that is guaranteed by randomly assigning the stimulus class
-(i.e., by choosing the stimulus class at random on every trial) is
-that any statistical effect of the stimulus class is an unbiased
-estimate of the total causal effect of the stimulus class. In order to
-empirically justify the assumption of *selective* influence of the
-stimulus class on the internal evidence value we would have to *test*
-if the stimulus class does not also affect the thresholds.
-
-We can do exactly this. We will use the 'log_distance' link function,
-because when this link function is used one of the elements of the
-gamma vector directly represents the position of the main decision
-criterion, and all the other elements represent the pattern of the
-distances between the thresholds. If the stimulus does not affect the
-pattern of the thresholds we should observe an effect only for the
-main criterion parameter. This model is not exactly equivalent to a
-hierarchical SDT model without selective influence since we do not
-constrain the effect of the stimulus class on the main criterion to be
-non-negative, but there is enough data points per participant that we
-can at least hope that this is not a serious problem.
-
-KOD
-
-There at least two important advantages that an SDT model fitted to
-binary classification task data has over an IRT model fitted to
-questionnaire data. One is that the causal assumption of selective
-influence of the stimulus class on the internal values is much more
-plausible than the - implicit in most applications of IRT - causal
-assumption of selective influence of participants on the
-"person-parameters". The other is that in a typical binary
-classification task case the process of responding is repeated many
-times, the task is relatively simple ("was the gabor patch tilted left
-or right" vs "do you feel that other people are happier than you") and
-the processes responsible for the internal values and for the
-placement of the decision thresholds are likely to be much more
-stable.
+*fixing the "item-parameters" within each sample*.
 
@@
 participants the difference would likely be negligible with very
-narrow confidence/credible intervals. By the same token, instability
+narrow confidence / credible intervals. By the same token, instability
 of item parameters in IRT models cannot be estimated by assuming
@@
 
-Back to our simulation. The variability due to *d'* is still here, we
-just cannot deconfound it from response bias. This means that, just
-like in an IRT model that does not fully account for true person- and
-and item-parameter effects, there is *no way to correctly estimate
-reliability of the estimates of person-parameters/internal values*. We
-will now demonstrate this fact.
+It does not take much thought to see that the assumption of constant
+"item-parameters" at the group / sample / population level is utterly
+unrealistic. It follows that the achievement of separating participant
+and item effects in IRT models is often not an achievement at all, it
+is just wishful thinking.
 
-First, we will fit a hierarchical SDT model with multiple criteria and
-an order-preserving link function. This will give us realistic
-estimates of "person-" (*d'*) and "item"-parameters (the thresholds).
+The justification of the selective influence assumption comes from a
+psychological theory of the task, not from the design of the
+study. All that is guaranteed by randomly assigning the stimulus class
+(i.e., by choosing the stimulus class at random on every trial) is
+that any statistical effect of the stimulus class is an unbiased
+estimate of the total causal effect of the stimulus class. To
+empirically justify the assumption of *selective* influence of the
+stimulus class on the internal evidence value we have to *test* if the
+stimulus class does not also affect the thresholds.
 
-KOD 4
+There are a few important advantages that an SDT model fitted to
+binary classification task data has over an IRT model fitted to
+questionnaire data. One is that the causal assumption of selective
+influence of the stimulus class on the internal values is much more
+plausible than the - implicit in most applications of IRT - causal
+assumption of selective influence of participants on the
+"person-parameters". The other is that in a typical binary
+classification task case the process of responding is repeated many
+times, the task is relatively simple ("was the gabor patch tilted left
+or right" vs "do you wish you were as happy as other people seem to
+be?") and the processes responsible for the generation of internal
+values and for the placement of the decision thresholds are likely to
+be more stable.
 
-What will happen if we emulate a typical IRT analysis and fit a
-hierarchical ordinal model in which the thresholds are assumed to be
-constant accross participants? After all, we can just invoke the
-magical power of wishful thinking and assume that the thresholds are
-item-parameters, right? How well will the internal value estimates
-based on multiple responses to a simple, non-personal question match
-the estimates based on a model that accounts for participant specific
-threshold patterns?
-
-KOD 5
-
-The answer is - the two kinds of estimates will match poorly. Here is
-the comparison of point estimates:
-
-and here is the comparison of interval estimates:
-
-WYKRES 3
-
-Let's see more directly how the lengths of the credible intervals
-differ. If the interval estimate widths were comparable they should
-all be scattered closely to the horizontal line at 1. Unsurprisingly,
-this is now what we see here:
-
-WYKRES
-
-The credible intervals directly represent reliability. For the
-interval estimates based on a simplified model to contain the great
-majority of internal values *or* account for the variability in the
-thresholds they would have to be wider. We do not need to look at
-"item-parameter" estimates in the simplified model, because it is
-already obvious that the point and interval estimates of the
-thresholds based on this model have to be severily biased.
-
-An ordinal model for ordered polytomous responses that does not
-correctly account for the variability in the thresholds produces
-biased point and interval estimates in *every* parameter, and the
-estimates seem more reliable than they truly are. That's because it
-assumes that the thresholds, which are determined both by the items
-and by the properties of the participants, are constant accross the
-participants. This assumption, which seems to be universally accepted
-in Item Response Theory, is so ridiculous that it should be laughed at
-on the street.
+The thresholds are determined by the items, by the properties of the
+participants, and by the context in which the measurement takes
+place. The assumption that important latent aspects of the process of
+psychological measurement are constant across participants is so
+ridiculous it should be laughed at on the street.

### 2020-02-17 R/plot_sdt_fit.R

@@ -21,7 +21,7 @@
 #'     of the response distribution plot will be created.
 #' @return a plot object.
 #' @export
-plot_sdt_fit = function(fit, adata, variables = NULL, type = 'roc', alpha = .05, bw = TRUE){
+plot_sdt_fit = function(fit, adata, variables = NULL, type = 'roc', alpha = .05, bw = TRUE, verbose = T){
     s = as.data.frame(fit)
     rm(fit)
     cnt_new = t(s[,grep('counts_new', names(s))])
@@ -50,13 +50,13 @@ plot_sdt_fit = function(fit, adata, variables = NULL, type = 'roc', alpha = .05,
     }
     dfa = stats::aggregate(df[[cnt]] ~ df[[resp]] + df[[stim]] + f, FUN = sum)
     cnt_new_a = matrix(nrow = ncol(cnt_new), ncol = nrow(dfa))
-    pb = utils::txtProgressBar(min = 1, max = ncol(cnt_new), style = 3)
-    print('Aggregating posterior samples...')
+    if(verbose)pb = utils::txtProgressBar(min = 1, max = ncol(cnt_new), style = 3)
+    if(verbose)print('Aggregating posterior samples...')
     for(r in 1:ncol(cnt_new)){
         cnt_new_a[r,] = stats::aggregate(cnt_new[,r] ~ df[[resp]] + df[[stim]] + f, FUN = sum)[[4]]
-        utils::setTxtProgressBar(pb, r)
+        if(verbose)utils::setTxtProgressBar(pb, r)
     }
-    close(pb)
+    if(verbose)close(pb)
     rm(cnt_new)
     names(dfa) = c('response', 'stimulus', 'f', 'count')
     dfa$stimulus = as.factor(dfa$stimulus)
@@ -98,6 +98,7 @@ plot_sdt_fit = function(fit, adata, variables = NULL, type = 'roc', alpha = .05,
         dfa[,c('count.lo', 'count.hi', 'count.fit')] = cbind(apply(cnt_new_a, 2, function(x)stats::quantile(x, alpha / 2)),
                                                              apply(cnt_new_a, 2, function(x)stats::quantile(x, 1 - alpha / 2)),
                                                              apply(cnt_new_a, 2, mean))
+        dfa$response = dfa$response + (as.numeric(dfa$stimulus) - 1.5) / 4
         if(bw){
             ggplot(dfa, aes(response, count / n), group = stimulus) +
[…]

### 2020-02-17 ordinal/ordinal.Rmd
ordinal tutorial corrected

@@
 particular. If you are interested in SDT, IRT, or ordinal models in
-general, perhaps you will find something useful, new, or even mildly
-provocative here.
+general, perhaps you will find something useful here.
 
@@
 
-To fit a model we have to aggregate the data (this makes the process
+To fit a model we have to aggregate the data as much as possible
+without distorting the hierarchical structure (this makes the process
 of sampling much faster), create the fixed- and random-effects lists,
 and create the data structures and the model code required by
-stan. Note that here we introduce a new parameter - eta - which
-represents the average internal value. If we are going to model all
-the thresholds (gamma parameter vector) then this parameter is
-redundant, because a constant shift in the thresholds is equivalent to
-the opposite effect in eta. The bhsdtr package will automatically fix
-the first element (here the only element) of the eta fixed effects
-vector at 0.
+stan. Note that we introduce a new parameter - eta - which represents
+the average internal value. If we are going to model all the
+thresholds (gamma parameter vector) then this parameter is redundant,
+because a constant shift in the thresholds is equivalent to the
+opposite effect in eta. The bhsdtr package will automatically fix the
+first element (here the only element) of the eta fixed effects vector
+at 0, which will make this model identifiable.
 
@@
 across participants and latent means or samples are participant-, not
-item-specific seem to be accepted in the great majority of
-applications of Item Response Theory. This is reflected in the IRT
-terminology: thresholds are often called "item-parameters" and latent
-means or samples are often called "person-parameters". Some even say
-that IRT can provide separate estimates of the true item- and
-person-parameters.
+item-specific seem to be often accepted in applications of Item
+Response Theory. This is also reflected in the IRT terminology:
+thresholds are often called "item-parameters" and latent means or
+samples are often called "person-parameters".
 
@@
 
-The neff and Rhat columns look good. Here is how the model fits (this
-is a maximum likelihood fit):
+Here is how the model fits (this is a maximum likelihood fit):
 
@@
 inter-individual differences in the pattern of the thresholds by
-modelling the gamma random effects. We cannot simultaneously model the
-eta random effects in this model, since all the information that can
-be captured by the eta parameter (i.e., the constant shift in the
-thresholds) is already captured by the fixed and the random gamma effects.
+modelling the gamma random effects.
 
@@
 print(fit0, probs = c(.025, .975), pars = c('gamma_fixed', 'gamma_sd_1'))
-## neff and Rhat look good
-```
-
-The neff and Rhat columns look good. Below we translate between the
-gamma and the criteria vectors (fixed effects = sample average):
+```
+
+Below we translate between the gamma and the criteria vectors (fixed
+effects):
 
@@
 
-Are the sample average thesholds different in the two models? Off
-course they are, since the models are non-linear and averaging over
-anything that affects any parameter of such models is a bad idea:
+Are the average thesholds different in the two models? Off course they
+are, since the models are non-linear and averaging over anything that
+affects any parameter of such models is a bad idea:
 
@@
 "signal" cannot be less likely for the stimulus "signal" than for the
-stimulus "noise", i.e., we will not assume that d' must be
+stimulus "noise", i.e., we will not assume that *d'* must be
 non-negative, and we will *not* assume that the pattern of the
@@
 thresholds are represented as random effects, we will account for the
-fact that "random" thresholds may be correlated, and they are - if
-some threshold is higher (lower) than the sample average, then the
-threshold above (below) it may also be higher (lower).
+fact that "random" thresholds may be correlated; They have to be
+correlated, because if some threshold is higher (lower) than the
+sample average, then the threshold above (below) it may also be higher
+(lower).
 
@@
 print(fit1, probs = c(.025, .975), pars = c('gamma_fixed', 'gamma_sd_1'))
-## neff and Rhat look good
 ```
@@
 Note the difference between the main criterion (row 4). This
-difference approximates d'. Even though, as we will later see, this
-model does not correctly account for the effect of the stimulus, we can
-still recover this effect pretty well, because d' is equivalent to a
+difference approximates *d'*. Even though, as we will later see, this
+model does not correctly account for the effect of the stimulus
+(because it violates the non-negativity assumption), we can still
+recover this effect pretty well, because *d'* is equivalent to a
 constant shift of all the thresholds in one predefined
@@
 In an SDT model we assume that the stimulus class affects *only* the
-d' parameter, and that both criteria (gamma) and *d'* may differ
-between the participants (or items, etc.). In bhsdtr d' is modelled as
-log(d') = delta, which forces it to be non-negative. This is an
+*d'* parameter, and that both criteria (gamma) and *d'* may differ
+between the participants (or items, etc.). In bhsdtr *d'* is modelled
+as *log(d') = delta*, which forces it to be non-negative. This is an
 important assumption of Signal Detection Theory - if this assumption
@@
 print(fit2, probs = c(.025, .975), pars = c('delta_fixed', 'gamma_fixed', 'delta_sd_1', 'gamma_sd_1'))
-## neff and Rhat look good
-```
-
-So how well do the two d' estimates match? We can translate between
-the delta and the d' parameters by using exponentiation. We have to do
-this first, before doing anything else with the posterior samples,
+```
+
+So how well do the two *d'* estimates match? We can translate between
+the delta and the *d'* parameters by using exponentiation. We have to
+do this first, before doing anything else with the posterior samples,
 because the exponential function is non-linear.
@@
 
-The point and interval estimates match quite well, with some
-notable exceptions. According to the ordinal model some credible
-intervals around individual d' point estimates include 0 and one point
+The point and interval estimates match quite well, with some notable
+exceptions. According to the ordinal model some credible intervals
+around individual *d'* point estimates include 0 and one point
 estimate is even below 0, which contradicts an important assumption of
@@
 
-If the participant-specific d' estimates match, the sample average d'
-should also match, right? Maybe the SDT model does not fit the data
-well?
+If the participant-specific *d'* estimates match, the sample average
+*d'* should also match, right? Maybe the SDT model does not fit the
+data well?
 
@@
 ```{r echo=T,results="show"}
-ggplot(data.frame(delta = log(dprim2i$mean)), aes(sample = delta)) +
+delta2 = as.data.frame(t(apply(exp(s2[['delta_fixed[1,1]']] +
+                                   s2[, grep('delta_random_1', names(s2))]), 2, sumfun)))
+ggplot(data.frame(delta = delta2), aes(sample = delta)) +
     stat_qq() +
@@
 The individual *delta = log(d')* estimates in the SDT model seem to be
-approximately normally distributed. We cannot plot log(d') estimates
+approximately normally distributed. We cannot plot *log(d')* estimates
 based on the ordinal model, because according to this model one
-participant has a negative *d'* value. However, we can see if the *d'*
-point estimates based on the ordinal model are approximately normally
-distributed:
+participant has a negative *d'* value and even without this
+participant's data there are quite a few negative posterior *d'*
+samples. However, we can see if the *d'* point estimates based on the
+ordinal model are approximately normally distributed:
 
@@
 shifts are assumed to be normally distributed and in the SDT model
-delta = log(d') random effects are assumed to be normally
+*delta = log(d')* random effects are assumed to be normally
 distributed. We can easily account for this by making use of the
@@
 
-The selectivity assumption is central to Signal Detection Theory. If
-you know your Signal Detection Theory well, please take a moment and
-try to recall one instant of this important assumption being
-explicitly addressed in a paper or a book on Signal Detection Theory.
-
 It does not matter if an SDT model with selectivity assumption fits
@@
 
+The selectivity assumption is central to Signal Detection Theory. If
+you know your Signal Detection Theory well, please take a moment and
+try to recall one instant of this important assumption being
+explicitly addressed in a paper or a book on Signal Detection Theory.
+
 Before we go any further, just for fun, let's see what happens if we
@@
 often accepted in Item Response Theory. We will fit a hierarchical SDT
-model in which we will call the thresholds "item-parameters" and we
-will call *d'* a "person-parameter".
+model in which we will treat the thresholds as "item-parameters" and
+we will treat the *d'* parameter as a "person-parameter".
 
@@
 print(fit3, probs = c(.025, .975), pars = c('delta_fixed', 'gamma_fixed', 'delta_sd_1'))
-## neff and Rhat look good
 ```
@@
 
-At the aggregated data level the fit is not perfect, but it is not
-that terrible either. What about the fit at the individual level?
+At the aggregated data level the fit is not perfect, but maybe it is
+not that terrible either. What about the fit at the individual level?
 
@@
 A few points are outside the predictive intervals, but maybe that's
-ok? What if we compared the point and interval estimates of the so-called "person-parameters"?
+ok? What if we compared the point and interval estimates of the
+so-called "person-parameters"?
 
@@
 ... we see that the simplified model not only gives biased
-"person-parameter" estimates but the estimates seem more reliable than
-they truly are (here by about 20%). It does not make sense to
-look at the estimates of the "item-parameters" since we already know
-that they have little in common with the true thresholds.
+"person-parameter" estimates but the estimates seem more reliable
+(here by about 20%). It does not make sense to look at the estimates
+of the "item-parameters" since we already know that they have little
+in common with the true thresholds.
 
@@
 influence. We can at least try to test this assumption. We will not
-alter the ordinal model code to account for the non-negativity of d'
-(and the log-normal distribution of random d' effects) just yet. As we
-already know, even without this correction the estimates of
+alter the ordinal model code to account for the non-negativity of *d'*
+(and the log-normal distribution of random *d'* effects) just yet. As
+we already know, even without this correction the estimates of
 participants' shifts are similar to the *d'* estimates.
@@
 
-Ok, we did simplify things significantly, but it still makes sense to
-take a look at the other gamma effects, because these other effects
-correspond to the variability in the *pattern* of the thresholds, and
-the estimates of the thresholds in our ordinal model seem perfectly
-fine.
+Ok, we did simplify things, but it still makes sense to take a look at
+the other gamma effects, because these other effects correspond to the
+variability in the *pattern* of the thresholds, and the estimates of
+the thresholds in our ordinal model seem perfectly fine.
 
@@
 constant across the participants: all the gamma_sd_1[.,2] parameters
-seem to be near-zero (the 95% credible intervals contain zero), except
-for the gamma_sd_1[4,2] parameter, which corresponds to the
-variability in *d'* values. This is *really* weird, since the pattern
-of the thresholds for the "noise" stimuli clearly varies with
-participants - none of the 95% HPD intervals for the gamma_sd_1[.,1]
-parameters include zero: in terms of standard deviations of posterior
-distributions of standard deviations (phew) most are in fact quite far
-from zero.
-
-If only we could do a better job and use a model which respects the
-non-negativity assumption. Is it really that complicated? Let's index
-the rows of our dataset by *i*. Because of the way the model was
-parametrized...
+seem to be near-zero, except for the gamma_sd_1[4,2] parameter, which
+corresponds to the variability in *d'* values. This is weird, since
+the pattern of the thresholds for the "noise" stimuli clearly varies
+with participants - none of the 95% HPD intervals for the
+gamma_sd_1[.,1] parameters include zero: in terms of standard
+deviations of posterior distributions of standard deviations (phew)
+most are in fact quite far from zero.
+
+Now we will modify the stan model code to account for the
+non-negativity of *d'*. Let's index the rows of our dataset by
+*i*. Because of the way the model was parametrized...
 
@@
 gamma[4] between the "noise" and "signal" stimuli) to represent log of
-average distance between the thresholds = shift = *d'*, so we have to
-subtract it from gamma[4]_i, exponentiate this sum, and add it again
+average distance between the thresholds = log(shift) = log(d'), so we
+have to subtract it from gamma[4]_i, exponentiate this sum, and add it
 to gamma[4]_i.
@@
 
-Interaction is tricky, you see. For example, if you have an additive
-effect of two factors on the original scale of the dependent
-variable...
+Interaction is tricky. For example, if you have an additive effect of
+two factors on the original scale of the dependent variable...
 
@@
 This means that whenever you see an interactive effect it may be an
-artefact of the scale of the dependent variable! Every time I say it
-it makes me feel uneasy. So, instead of examining the effects in the
-gamma parameters. let's have a look at the average *criteria /
-thresholds*. Notice that we have to add the effect on gamma to gamma
-for stimulus = "noise" to obtain gamma for stimulus = "signal".
+artefact of the scale of the dependent variable. So, instead of
+examining the effects in the gamma parameters. let's have a look at
+the *average criteria / thresholds*. Notice that we have to add the
+effect on gamma to gamma for stimulus = "noise" to obtain gamma for
+stimulus = "signal".
 
@@
 
-Now we can directly test if the *pattern* of the criteria may be
+Now we can directly test if the average pattern of the criteria may be
 different by comparing *centered* criteria vectors:
+
 ```{r echo=T,results="show"}
@@
 positions. The only thing that seems to differ, and not by much, is
-the main criterion, but this is confounded with *d'* in our model (I
-think). Selectivity seems to hold, kind of.
-
-This is how you *begin* to test a hierarchical SDT model. An SDT model
-has other problematic assumptions. One is that the time-accuracy
-tradeoff can be ignored. If participants or conditions differ in the
-response time, then the differences in *d'* may not mean that
-sensitivity is different, because participants may simply take more
-time to respond in certain experimental conditions, which inflates the
-*d'* estimate. You need something like the diffusion model to deal
-with this problem. If you combine the diffusion model with ratings,
-you also need something like a hierarchical diffusion model with
-ratings and an order-preserving link function. Sounds like something
-that could perhaps be implemented in the bhsdtr package in the future.
+the main criterion, but this, I think, is confounded with *d'* in our
+model. Selectivity seems to hold, kind of.
+
+This is one way to *begin* to test an SDT model. An SDT model has
+other problematic assumptions. One is that the time-accuracy tradeoff
+can be ignored. If participants or conditions differ in the response
+time, then the differences in *d'* may not mean that sensitivity is
+different, because participants may simply take more time to respond
+in certain experimental conditions, which inflates the *d'*
+estimate. You need something like the diffusion model to deal with
+this problem. If you combine the diffusion model with ratings, you
+also need something like a hierarchical diffusion model with ratings
+and an order-preserving link function.
 
@@
 estimates of every SDT parameter may become when you overly aggregate
-your data in the bhsdtr preprint. I am now in the process of examining
-the direction and magnitude of bias in SDT models fitted to
-overly-aggregated data by using the excellent Confidence database
+your data in the bhsdtr preprint. By the way, I am now in the process
+of examining the direction and magnitude of bias in SDT models fitted
+to overly-aggregated data by using the excellent Confidence database
 created by Doby Rahnev.
@@
 assumptions that make such models identifiable under certain
-conditions are necessarily *causal*.
+conditions are necessarily causal.
 
@@
 redundant. As far as the design of the study is concerned, all we can
-say about the internal value is that it is relative to the thresholds
-- and the position of each threshold is relative to the positions of
-the remaining thresholds and to the position of the internal
+say about the internal value is that it is relative to the thresholds-
+and the position of each threshold is relative to the positions of the
+remaining thresholds and to the position of the internal
 value. Without the selectivity assumption, to say that John thinks
@@
 change (*d'*) in the internal value only; In such cases the internal
-value parameter may be justified and useful because it does constrain
-the model. We have obtained evidence that this assumption perhaps
-holds for the analyzed dataset.
+value parameter may be justified and useful. We have obtained evidence
+that this assumption perhaps holds for the analyzed dataset.
 
@@
 not item-specific, decision thresholds, also known as
-"item-parameters". There is nothing in the *design* of the process of
-measurement by a questionnaire that makes it possible to distinguish
-genuine differences in internal values from the differences in the way
-of responding.
-
-Because of random assignment of the stimulus class above chance
-performance in a binary classification task cannot be faked, and so
-the *d'* parameter definitely captures at least part of the effect of
-the stimulus on the internal value. On the other hand, every possible
-pattern of responses to a questionnaire can be faked. The only way to
-distinguish the "item" and "person" effects is to *assume selective
-influence*, i.e., that there is some factor f (such as the stimulus
-class in a binary classification task, or, in case of a depression
-questionnaire, participants trait, or a therapeutic intervention)
-which can *only* affect the internal value, thus producing a
-*constant* shift in the thresholds. The assumption of selective
-influence makes the model with the two kinds of parameters
-identifiable if there are repeated measurements for at least one
-participant and one item or many participants provide single responses
-to all of the items.
+"item-parameters".
+
+There is nothing in the *design* of the process of measurement by a
+questionnaire that makes it possible to distinguish genuine
+differences in internal values from the differences in the way of
+responding. Because of random assignment of the stimulus class above
+chance performance in a binary classification task cannot be faked,
+and so the *d'* parameter definitely captures at least part of the
+effect of the stimulus on the internal value. On the other hand, every
+possible pattern of responses to a questionnaire can be faked.
+
+The only way to (perhaps incorrectly) separate the "item" and "person"
+effects is to *assume selective influence*, i.e., that there is some
+factor f (such as the stimulus class in a binary classification task,
+or, in case of a depression questionnaire, participants trait, or a
+therapeutic intervention) which can only affect the internal value,
+thus producing a constant shift in the thresholds. The assumption of
+selective influence makes the model with the two kinds of parameters
+identifiable.
 
@@
 If we campared the estimates of the main decision criterion (i.e.,
-bias) in SDT models fitted to two different large samples of
+bias) in SDT models fitted to two different large *samples* of
 participants the difference would likely be negligible with very
-narrow confidence / credible intervals. By the same token, instability
-of item parameters in IRT models cannot be estimated by assuming
-selective influence and comparing estimates of "item-parameters"
-obtained in two different large samples, because large-sample average
-estimates of item parameters are the best way to *hide or misplace*
-(it becomes part of the variability in the "person-parameters") the
-intra- or inter-individual variability in item parameters; By assuming
-that internal values are participant-specific and item parameters are
-item-specific we force the estimates of item parameters to become
-estimates of average person-specific thresholds. This not only begs
-the question, but also, because the model is non-linear, makes the
-estimates of item-parameters *asymptotically biased*.
-
-It does not take much thought to see that the assumption of constant
+narrow confidence / credible intervals, even though *individual
+participants* clearly differ in how they use the main decision
+criterion. By the same token, instability of item parameters in IRT
+models cannot be estimated by assuming selective influence and
+comparing estimates of "item-parameters" obtained in two different
+large samples, because sample average estimates of item-parameters
+*hide or misplace* (overdispersion) the intra- or inter-individual
+variability in item parameters.
+
+By assuming that internal values are participant-specific and item
+parameters are item-specific we force the estimates of item parameters
+to become estimates of average person-specific thresholds. This not
+only begs the question, but also, because the model is non-linear,
+makes the estimates of item-parameters *asymptotically biased*. It
+does not take much thought to see that the assumption of constant
 "item-parameters" at the group / sample / population level is utterly
@@
 
-The justification of the selective influence assumption comes from a
-psychological theory of the task, not from the design of the
+The primary justification of the selective influence assumption comes
+from a psychological theory of the task, not from the design of the
 study. All that is guaranteed by randomly assigning the stimulus class
@@
 estimate of the total causal effect of the stimulus class. To
-empirically justify the assumption of *selective* influence of the
+*empirically* justify the assumption of *selective* influence of the
 stimulus class on the internal evidence value we have to *test* if the

### 2020-02-17 ordinal/ordinal.Rmd
ordinal html

@@
 ```{r echo=T,results="show"}
-delta2 = as.data.frame(t(apply(exp(s2[['delta_fixed[1,1]']] +
-                                   s2[, grep('delta_random_1', names(s2))]), 2, sumfun)))
+delta2 = apply(s2[['delta_fixed[1,1]']] + s2[, grep('delta_random_1', names(s2))], 2, mean)
 ggplot(data.frame(delta = delta2), aes(sample = delta)) +

### 2020-02-17 ordinal/ordinal.Rmd
ordinal github markdown readme

@@
+---
+title: "Ordinal bhsdtr tutorial"
+output:
+  md_document:
+    variant: markdown_github
+---
 ```{r init, include=F,echo=F}
@@
                       echo=TRUE, warning = FALSE, error = FALSE, message = FALSE,
-                      cache.path = '~/windows/temp/ordinal', cache=T)
+                      cache.path = '~/windows/temp/ordinal/', cache=T)
 
@@
 participants, and by the context in which the measurement takes
-place. The assumption that important latent aspects of the process of
-psychological measurement are constant across participants is so
+place. The assumption that an important latent aspect of the process
+of psychological measurement is constant across participants is so
 ridiculous it should be laughed at on the street.

### 2020-02-17 ordinal/code.R
an old R file removed

@@ -1,183 +1,0 @@
-ggs = function(f)ggsave(f, width = 10, height = 10, dpi = 200)
-
-######################################################################
-## KOD 1
-
-library(ggplot2)
-library(bhsdtr)
-library(Hmisc)
-library(rstan)
-options(mc.cores = parallel::detectCores())
-rstan_options(auto_write = TRUE)
-
-## r = 1 ~ absolutely clear image left, r = 4 ~ no experience left, r = 5 ~ no experience right, ... r = 8 ~  absolutely clear image right
-gabor$r = combined_response(gabor$stim, gabor$rating, gabor$acc)
-d1 = gabor[gabor$duration == '32 ms' & gabor$order == 'DECISION-RATING',]
-d1$r = combined_response()
-## Only the left-tilted gabor patches
-d1 = d1[d1$stim == 0,]
-
-######################################################################
-## KOD 2
-
-## This order-preserving link function represents the main criterion
-## directly as a free paramter and all the other criteria as log of
-## distance between the criterion and the criterion adjacent to it
-## which is closer to the main criterion
-
-adata = aggregate_responses(d1, response = 'r', variables = c('id'))
-fixed = list(eta = ~ 1, gamma = ~ 1)
-random = list(list(group = ~ id, gamma = ~ 1))
-model = make_stan_model(random, model = 'ordinal')
-sdata = make_stan_data(adata, fixed, random, model = 'ordinal')
-
-######################################################################
-## KOD 3
-
-o = optimizing(stan_model(model_code = model), sdata, init = 0)
-if(o$return_code)warning('not ok')
-
-######################################################################
-## Wykres
-
-sdt_ml_plot(model, adata, sdata)
-ggs('plot1.jpg')
-
-######################################################################
-## Wykres
-
-gamma_fixed = o$par[grep('gamma_fixed\\[', names(o$par))]
-gamma_random = matrix(o$par[grep('gamma_random_1', names(o$par))], nrow = sdata$group_1_size)
-for(i in 1:nrow(gamma_random))
-    gamma_random[i,] = gamma_random[i,] + gamma_fixed
[…]

### 2020-02-17 ordinal/ordinal.Rmd
"sample average" mistakes corrected

@@
 Now we can translate between the gamma and the criteria vectors (fixed
-effects = sample average) for each stimulus:
+effects) for each stimulus:
 
@@
 
-We will later see that this is our sample average *d'*, but it also
-isn't.
+We will later see that this is our average *d'*, but it also isn't.
 
@@
 
-If the participant-specific *d'* estimates match, the sample average
-*d'* should also match, right? Maybe the SDT model does not fit the
-data well?
+If the participant-specific *d'* estimates match, the average *d'*
+should also match, right? Maybe the SDT model does not fit the data
+well?
 
@@
 
-The discrepancy between the sample average *d'* estimates from both
-models is not really a consequence of allowing for negative *d'*
-values in the ordinal model. This discrepancy is a consequence of the
-assumed distribution of individual shifts / *d'* values. In the
-ordinal model we assumed that the shifts (*= d'*) are normally
-distributed, whereas in the SDT model we assumed that *log(d')* are
-normally distributed. We can see if the assumption that *log(d')* (or
-*d'*) are normally distributed is approximately correct in our
-dataset. First, let's take a look at participant-specific delta
-estimates in the SDT model:
+The discrepancy between the average *d'* estimates from both models is
+not really a consequence of allowing for negative *d'* values in the
+ordinal model. This discrepancy is a consequence of the assumed
+distribution of individual shifts / *d'* values. In the ordinal model
+we assumed that the shifts (*= d'*) are normally distributed, whereas
+in the SDT model we assumed that *log(d')* are normally
+distributed. We can see if the assumption that *log(d')* (or *d'*) are
+normally distributed is approximately correct in our dataset. First,
+let's take a look at participant-specific delta estimates in the SDT
+model:
 
@@
 Now we see that the two models give very similar point and interval
-estimates of sample-average *d'*.
+estimates of average *d'*.
 
@@
 Note the clear pattern for the gamma_sd_1[.,2] parameters. Even though
-the *sample-average* pattern of the thresholds seems to be different
+the *average* pattern of the thresholds seems to be different
 for the two stimulus classes (gamma_fixed[.,2] except for
@@
 comparing estimates of "item-parameters" obtained in two different
-large samples, because sample average estimates of item-parameters
+large samples, because estimates of sample average item-parameters
 *hide or misplace* (overdispersion) the intra- or inter-individual

### 2020-02-17 ordinal/Readme.Rmd
the third layer

--- 2020-02-17 ordinal/ordinal.Rmd
@@
 general, and Signal Detection Theory and Item Response Theory in
-particular. If you are interested in SDT, IRT, or ordinal models in
-general, perhaps you will find something useful here.
+particular. If you are interested in any of these topics, perhaps you
+will find something useful here.
 
@@
 usually analyzed by fitting a linear model, and all sorts of bad
-things happen because of this, but only a few people seem to care. I
+things happen because of this, but most people do not seem to care. I
 for one do not always care about this. We will not address this kind
@@
 rstan_options(auto_write = TRUE)
-## This variable indicates if the fitted models can be loaded from disk
+## This variable indicates if the fitted models can be loaded from
+## disk. I use it for caching
 fresh_start = FALSE
@@
 The number of parameters in this model matches the degrees of freedom
-in the data, which means that the model is saturated, which means that
-it always fits the data, but it does not mean that there is no
-uncertainty associated with the estimates of its parameters.
+in the data (if the data are aggregated), which means that the model
+is saturated, which means that it always fits the data, but it does
+not mean that there is no uncertainty associated with the estimates of
+its parameters.
 
@@
 threshold has a normal distribution due to the effect of participants
-would violate the ordering assumption.
-
-To supplement our model with a hierarchical linear regression
-structure the vector of the thresholds has to be translated to the
-unconstrained vector gamma. The 'log_distance' order-preserving link
-function is one solution to this problem, but other order-preserving
-link functions are also implemented in the bhsdtr package. The choice
-of the link function may sometimes be important, but, since we will
-not introduce numerical predictors for the thresholds, we can choose
-any link function which seems most convenient.
+would violate the ordering assumption. To supplement our model with a
+hierarchical linear regression structure the vector of the thresholds
+has to be translated to the unconstrained vector gamma. The
+'log_distance' order-preserving link function is one solution to this
+problem, but other order-preserving link functions are also
+implemented in the bhsdtr package. The choice of the link function may
+sometimes be important, but, since we will not introduce quantitative
+predictors for the thresholds, we can choose any link function which
+seems most convenient.
 
@@
 function are easy to interpret and may be directly related to simple
-research questions. In particular, one of the elements of the gamma
-vector under the log_distance link function, i.e., the middle
-threshold, will soon become important. This link function represents
-the main threshold/criterion directly as a free parameter (here
-gamma[4], because K = 8, and 8 / 2 = 4); Every other criterion is
-represented as log of distance between the criterion and the one
-criterion adjacent to it which is closer to the main criterion. This
-way the order is preserved, and each threshold *i* is associated with
-the corresponding unconstrained element *i* of the gamma
-vector. Random and fixed effects in the thresholds are represented as
-(hierarchical) linear effects in gamma. The two representations are
-isomorphic, which means that we can always translate between them if
-we need to.
+research questions. This link function represents the main threshold /
+criterion directly as a free parameter (here gamma[4], because K = 8,
+and 8 / 2 = 4); Every other criterion is represented as log of
+distance between the criterion and the one criterion adjacent to it
+which is closer to the main criterion. This way the order is
+preserved, and each threshold *i* is associated with the corresponding
+unconstrained element *i* of the gamma vector. Random and fixed
+effects in the thresholds are represented as (hierarchical) linear
+effects in gamma. The two representations are isomorphic, which means
+that we can always translate between the gamma and the ordered vectors
+if we need to.
 
@@
 stan. Note that we introduce a new parameter - eta - which represents
-the average internal value. If we are going to model all the
-thresholds (gamma parameter vector) then this parameter is redundant,
-because a constant shift in the thresholds is equivalent to the
-opposite effect in eta. The bhsdtr package will automatically fix the
-first element (here the only element) of the eta fixed effects vector
-at 0, which will make this model identifiable.
-
-First, we will fit a model which assumes that the pattern of the
+the average internal value. If we are going to model the effects in
+all the thresholds (gamma parameter vector) then this parameter is
+redundant, because a constant shift in the thresholds is equivalent to
+the opposite effect in eta. The bhsdtr package will automatically fix
+the first element (here the only element) of the eta fixed effects
+vector at 0, which will make this model identifiable.
+
+First, we will fit the model which assumes that the pattern of the
 thresholds is the same for all the participants, but the mean of the
@@
 Note that the two assumptions, i.e., that the thresholds are constant
-across participants and latent means or samples are participant-, not
-item-specific seem to be often accepted in applications of Item
-Response Theory. This is also reflected in the IRT terminology:
-thresholds are often called "item-parameters" and latent means or
-samples are often called "person-parameters".
+across participants and that the latent means or samples are
+participant-, not item-specific seem to be often accepted in
+applications of Item Response Theory. This is also reflected in the
+IRT terminology: thresholds are often called "item-parameters" and
+latent means or samples are often called
+"person-parameters".
 
@@
 ```{r echo=T,results="show"}
+## This function is not defined in the bhsdtr package
 sdt_ml_plot(model, adata, sdata)
@@
 fixed0 = list(eta = ~ 1, gamma = ~ 1)
-## Each participant has a unique patterh of the thresholds
+## Each participant has a unique patterh of the thresholds / gamma
 random0 = list(list(group = ~ id, gamma = ~ 1))
@@
 non-negative, and we will *not* assume that the pattern of the
-thresholds is the same for "noise" and "signal" stimuli. If you know
-your Signal Detection Theory well, this may seem to you like a
-perverse idea. If that is the case, please bear with me.
+thresholds is the same for the "noise" and the "signal" stimuli. If
+you know your Signal Detection Theory well, this may seem to you like
+a perverse idea. If that is the case, please bear with me.
 
 In this model each participant may have two different sets of
-thresholds, one for "noise" and one for "signal" stimuli. Since the
-thresholds are represented as random effects, we will account for the
-fact that "random" thresholds may be correlated; They have to be
-correlated, because if some threshold is higher (lower) than the
-sample average, then the threshold above (below) it may also be higher
-(lower).
+thresholds, one for the "noise" and one for the "signal" stimuli. We
+will account for the fact that participant specific threshold effects
+may be correlated; They are usually correlated, because if some
+threshold is higher (lower) than the sample average, then the
+threshold above (below) it may also be higher (lower).
 
@@
 
-Now we can translate between the gamma and the criteria vectors (fixed
-effects) for each stimulus:
+Now we can again translate between the gamma and the criteria vectors
+(fixed effects) for each stimulus:
 
@@
 direction. Shifting all the criteria to the right is equivalent to
-shifting the underlying distribution of "latent values" (or evidence)
-to the left, so we have to subtract the average position of the
-criteria for the "signal" stimuli from the average position of the
-criteria for the "noise" stimuli. Here is our first attempt:
+shifting the underlying distribution of latent values (or evidence) to
+the left, so we have to subtract the average position of the criteria
+for the "signal" stimuli from the average position of the criteria for
+the "noise" stimuli. Here is our first attempt:
 
@@
 
-In an SDT model we assume that the stimulus class affects *only* the
-*d'* parameter, and that both criteria (gamma) and *d'* may differ
-between the participants (or items, etc.). In bhsdtr *d'* is modelled
-as *log(d') = delta*, which forces it to be non-negative. This is an
-important assumption of Signal Detection Theory - if this assumption
-is violated, for example, because of the response reversal in some
-participants or conditions, then an SDT model is not valid and a
-non-trivial generalization of SDT should be used instead (see the
-bhsdtr preprint for a more detailed explanation and some
-references). Otherwise, this is the same general hierarchical ordinal
-regression model.
-
-```{r echo=T,results="show"}   
+In an SDT model we usually assume that the stimulus class affects
+*only* the *d'* parameter, and that both criteria (gamma) and *d'* may
+differ between the participants (or items, etc.). In bhsdtr *d'* is
+represented as *log(d') = delta*, which forces it to be
+non-negative. This is an important assumption of Signal Detection
+Theory - if this assumption is violated, for example, because of the
+response reversal in some participants or conditions, then an SDT
+model is not valid and a non-trivial generalization of SDT should be
+used instead (see the bhsdtr preprint for a more detailed explanation
+and some references).
+
+```{r echo=T,results="show"}   
+## We want one average d' / delta and one vector of average
+## thresholds / criteria / gammma
 fixed2 = list(delta = ~ 1, gamma = ~ 1)
+## Each participant has a unique d' / delta and a unique set of
+## thresholds / criteria / gamma
 random2 = list(list(group = ~ id, delta = ~ 1, gamma = ~ 1))
@@
 
-Because of the importance of the stimulus class ("noise", or 1, and
-"signal", or 2) in SDT models we have to indicate which variable
+Because of the importance of the stimulus class in SDT models
+("noise", or 1, and "signal", or 2) we have to indicate which variable
 represents this information ( aggregate_responses(..., stimulus =
-'stim', ...) ).
+'stim', ...) ) when calling the aggregate_responses function.
 
@@
 estimates? To calculate the individual shifts in the thresholds /
-criteria we have to add the random effects to the fixed effect:
+criteria in the more general ordinal model we have to add the random
+effects of participants to the fixed effect:
 
@@
 
-The point and interval estimates match quite well, with some notable
-exceptions. According to the ordinal model some credible intervals
-around individual *d'* point estimates include 0 and one point
-estimate is even below 0, which contradicts an important assumption of
-the SDT model. We will address this issue later.
+The point and interval estimates match quite well (*r* = .93 for the
+point estimates), with some notable exceptions. According to the
+ordinal model some credible intervals around individual *d'* point
+estimates include 0 and one point estimate is even below 0, which
+contradicts an important assumption of the SDT model. We will address
+this issue later.
 
@@
 predictive intervals for each participant. However, this does not mean
-much; an SDT model with multiple criteria is not saturated, so it does
+much; An SDT model with multiple criteria is not saturated, so it does
 not have to fit the data, but it does not leave too many degrees of
-freedom in the data, so we can safely assume a priori that it will
+freedom in the data either. We can safely assume a priori that it will
 usually fit the data well, regardless if it is true or not.
@@
 normally distributed is approximately correct in our dataset. First,
-let's take a look at participant-specific delta estimates in the SDT
-model:
-
-```{r echo=T,results="show"}
+let's take a look at the participant-specific delta estimates in the
+SDT model:
+
+```{r echo=T,results="show"}
+## Participant's delta = average delta + participant's effect
 delta2 = apply(s2[['delta_fixed[1,1]']] + s2[, grep('delta_random_1', names(s2))], 2, mean)
@@
 samples. However, we can see if the *d'* point estimates based on the
-ordinal model are approximately normally distributed:
+general ordinal model are approximately normally distributed:
 
@@
 They seem to be approximately normally distributed as well. The
-average / fixed effect *d'* in each model is estimated as the mean of
-different distributions: in the ordinal model participant-specific
-shifts are assumed to be normally distributed and in the SDT model
-*delta = log(d')* random effects are assumed to be normally
-distributed. We can easily account for this by making use of the
-relationship between the parameters of the normal distribution and the
-associated log-normal distribution:
+average *d'* in each model is estimated as a mean of different
+distributions: in the ordinal model participant-specific shifts are
+assumed to be normally distributed, but in the SDT model *delta =
+log(d')* random effects are assumed to be normally distributed. We can
+easily account for this by making use of the relationship between the
+parameters of the normal distribution and the associated log-normal
+distribution:
 
@@
 Now we see that the two models give very similar point and interval
-estimates of average *d'*.
+estimates of average *d'*. How cool is this?
 
@@
 
+# Causal assumptions of Signal Detection Theory
+
 An SDT model is not only a statistical model, it also has *causal*
-assumptions. Because the stimulus class was chosen at random on every
-trial, every estimate of the statistical effect of the stimulus class
-on any parameter of the model is an unbiased estimate of the total
-causal effect of the stimulus class on this parameter. Moreover, in
-such designs every statistical effect of the participant factor is an
-estimate of the property or the causal effect of the
-participants. This means that both the individual *d'* parameters and
-the individual thresholds are participant-, not item-, parameters in
-this model. So far so good.
-
-However, in the SDT model we have also assumed that the stimulus class
-can *only* affect the distance between the evidence distribution means
-(= *d'*), whereas in the ordinal model we allowed for the possibility
-of the stimulus class having an effect on the thresholds. It may seem
-odd to you to allow for this, but there is nothing in the design of
-the study that justifies the causal assumption of selective
-influence. The estimates of the thresholds are based on the data that
-were generated after the stimulus was presented, and so,
-theoretically, the *pattern* of the thresholds could be affected by
-the stimulus class as well.
+assumptions. To be clear, an SDT model *as a statistical model* does
+not have any causal assumptions. However, causal assumptions are
+present when the parameters of this model are interpreted as
+properties of a psychological process, or when it is used to
+*deconfound* (a causal notion) sensitivity (another extra-statistical
+notion) from bias (this term has a technical meaning in statistics,
+but it means something *psychological* in an SDT model - *a tendency
+to prefer one class of stimuli over the other*).
+
+Because the gabor dataset comes from a study in which the stimulus
+class was chosen at random on every trial, every estimate of the
+statistical effect of the stimulus class on any parameter of the model
+is an unbiased estimate of the total causal effect of the stimulus
+class on this parameter. Moreover, in such designs every statistical
+effect of the participant factor is an estimate of the property or the
+causal effect of the participants. This means that both the individual
+*d'* parameters and the individual thresholds are participant-, not
+item-, parameters in this model. So far so good.
+
+However, in the SDT model we also assumed that the stimulus class can
+*only* affect the distance between the evidence distribution means (=
+*d'*), whereas in the ordinal model we allowed for the possibility of
+the stimulus class having an effect on the pattern of the
+thresholds. It may seem odd to you to allow for this, but there is
+nothing in the design of the study that justifies the causal
+assumption of selective influence. The estimates of the thresholds are
+based on the data that were generated after the stimulus was
+presented, and so, theoretically, the pattern of the thresholds could
+be affected by the stimulus class as well.
 
@@
 selectivity assumption is false, an SDT model cannot deconfound
-sensitivity / *d'* from bias / threshold / criteria effects.
-
-The selectivity assumption is central to Signal Detection Theory. If
-you know your Signal Detection Theory well, please take a moment and
-try to recall one instant of this important assumption being
-explicitly addressed in a paper or a book on Signal Detection Theory.
+sensitivity / *d'* from bias / threshold / criteria effects. Once we
+drop the selectivity assumption there is no longer any *d'* / internal
+value parameter, only the thresholds, like in our general ordinal
+model. The selectivity assumption is central to Signal Detection
+Theory. If you know your Signal Detection Theory well, please take a
+moment and try to recall one instant of this important assumption
+being explicitly addressed in a paper or a book on Signal Detection
+Theory. If you manage to find one such example, please send me an
+email.
 
 Before we go any further, just for fun, let's see what happens if we
-introduce yet another assumption of selective influence, one that is
-often accepted in Item Response Theory. We will fit a hierarchical SDT
-model in which we will treat the thresholds as "item-parameters" and
-we will treat the *d'* parameter as a "person-parameter".
+introduce yet another assumption of selective influence. We will fit a
+hierarchical SDT model in which we will treat the thresholds as
+"item-parameters" and we will treat the *d'* parameter as a
+"person-parameter".
 
@@
 fixed3 = list(delta = ~ 1, gamma = ~ 1)
+## Each participant has a unique d', but not thresholds
 random3 = list(list(group = ~ id, delta = ~ 1))
@@
 "person-parameter" estimates but the estimates seem more reliable
-(here by about 20%). It does not make sense to look at the estimates
-of the "item-parameters" since we already know that they have little
-in common with the true thresholds.
+(they are shorter by about 20% on average). It does not make sense to
+look at the estimates of the "item-parameters" since we already know
+that they have little in common with the true thresholds.
 
@@
 Note that we do not use the separate intercepts and slopes
-parametrization here (i.e., ~ -1 + stim), because we are very much
-interested in the *effect* of the stimulus class on the *pattern* of
-the thresholds.
-
-```{r echo=T,results="show"}   
+parametrization here (we use ~ stim instead of ~ -1 + stim), because
+we are very much interested in the difference in the pattern of the
+thresholds caused by the stimulus class.
+
+```{r echo=T,results="show"}   
+## Separate thresholds for "signa" and "noise"
 fixed4 = list(eta = ~ 1, gamma = ~ stim)
+## Each participants has two unique sets of thresholds, one for "signa" and one for "noise".
 random4 = list(list(group = ~ id, gamma = ~ stim))
@@
 Here we see the estimated average shift / *d'* thought of as the mean
-of normally distributed *d'* (not *delta = log(d')*) random effects
-again:
+of the normally distributed *d'* (not *delta = log(d')*) random
+effects again:
 
@@
 
-Ok, we did simplify things, but it still makes sense to take a look at
-the other gamma effects, because these other effects correspond to the
-variability in the *pattern* of the thresholds, and the estimates of
-the thresholds in our ordinal model seem perfectly fine.
+Ok, we did simplify things a bit, but it still makes sense to take a
+look at the other gamma effects, because these other effects
+correspond to the variability in the *pattern* of the thresholds, and
+the estimates of the thresholds in our ordinal model seem perfectly
+fine.
 
@@
 non-zero. This indicates that the pattern of the thresholds - not only
-their average position - is affected by the stimulus
-class. Interestingly, the random effects' standard deviations tell a
-different story:
+their average position (= gamma_fixed[4,.]) - may be affected by the
+stimulus class. Interestingly, the random effects' standard deviations
+tell a different story:
 
@@
 
-Note the clear pattern for the gamma_sd_1[.,2] parameters. Even though
-the *average* pattern of the thresholds seems to be different
-for the two stimulus classes (gamma_fixed[.,2] except for
-gamma_fixed[4,2] ~ d'), this difference in pattern seems to be fairly
-constant across the participants: all the gamma_sd_1[.,2] parameters
-seem to be near-zero, except for the gamma_sd_1[4,2] parameter, which
-corresponds to the variability in *d'* values. This is weird, since
-the pattern of the thresholds for the "noise" stimuli clearly varies
-with participants - none of the 95% HPD intervals for the
+Note the clear pattern for the gamma_sd_1[.,2] parameters. The
+*average* pattern of the thresholds seems to be different for the two
+stimulus classes (gamma_fixed[.,2] except for gamma_fixed[4,2] ~ d'),
+and participants' thresholds for the "noise" stimuli seem to differ
+(gamma_sd_1[.,1]) - none of the 95% HPD intervals for the
 gamma_sd_1[.,1] parameters include zero: in terms of standard
 deviations of posterior distributions of standard deviations (phew)
-most are in fact quite far from zero.
-
-Now we will modify the stan model code to account for the
-non-negativity of *d'*. Let's index the rows of our dataset by
-*i*. Because of the way the model was parametrized...
+most are in fact quite far from zero. However, the difference in the
+pattern of the thresholds for the "signal" and the "noise" stimuli
+seems to be fairly constant across the participants: all the
+gamma_sd_1[.,2] parameters seem to be near-zero, except for the
+gamma_sd_1[4,2] parameter, which corresponds to the variability in the
+*d'* parameter. This is weird.
+
+We will modify the stan model code to account for the non-negativity
+of *d'*. Let's index the rows of our dataset by *i*. Because of the
+way the model was parametrized...
 
@@
 
-when stim = "noise" = 0, gamma[4]_i = gamma_fixed[4,1] +
+... when stim = "noise" = 0, gamma[4]_i = gamma_fixed[4,1] +
 gamma_random_1[g_i][4,1], where g_i is the participant index in row
@@
 
-However, when stim = "signal" = 1, gamma[4]_i = gamma_fixed[4,1] +
+When stim = "signal" = 1, gamma[4]_i = gamma_fixed[4,1] +
 gamma_random_1[g][4,1] + gamma_fixed[4,2] + gamma_random_1[g][4,2]. We
 want this other part (the last sum which represents the difference in
-gamma[4] between the "noise" and "signal" stimuli) to represent log of
-average distance between the thresholds = log(shift) = log(d'), so we
-have to subtract it from gamma[4]_i, exponentiate this sum, and add it
-to gamma[4]_i.
+gamma[4] between the "noise" and the "signal" stimuli) to represent
+log of average distance between the thresholds = log(shift) = log(d'),
+so we have to subtract it from gamma[4]_i, exponentiate this sum, and
+add it to gamma[4]_i.
 
@@
 criteria?
+
 ```{r echo=T,results="show"}   
@@
 
-We see again that gamma_sd_1[.,2] parameters, except for
+We see again that the gamma_sd_1[.,2] parameters, except for
 gamma_sd_1[4,2] = log(d'), may be zero - the effect of the stimulus on
@@
 ...then after applying a non-linear transformation to the dependent
-variable you may obtain a significant interacive effect:
+variable you may observe a significant interacive effect:
 
@@
 examining the effects in the gamma parameters. let's have a look at
-the *average criteria / thresholds*. Notice that we have to add the
+the average *criteria / thresholds*. Notice that we have to add the
 effect on gamma to gamma for stimulus = "noise" to obtain gamma for
@@
 Now we can directly test if the average pattern of the criteria may be
-different by comparing *centered* criteria vectors:
+different by comparing the *centered* criteria vectors:
 
@@
 the main criterion, but this, I think, is confounded with *d'* in our
-model. Selectivity seems to hold, kind of.
+model (this is only my guess). Selectivity seems to hold, kind of. To
+really justify the selectivity assumption we have to argue for the
+null, which, in a bayesian setting, usually means using the bayes
+factor. Even though I love bayesian inference, I do not like the bayes
+factor, and I am not alone in this.
 
@@
 
-The other problems that need to be addressed have to do with
-non-linearity and aggregation. If you fit an SDT model, or any other
-non-linear model, to data aggregated over grouping factors, such as
-participants or items, because of non-linearity, which does not play
-well with aggregation, some or all of the point and interval estimates
-will be biased, perhaps severely so, and the inference will be
-invalid. You can see how horribly biased the point *and* interval
+Among the other problems that need to be addressed some have to do
+with non-linearity and aggregation. If you fit an SDT model, or any
+other non-linear model, to data aggregated over grouping factors, such
+as participants or items, because of non-linearity, which does not
+play well with aggregation, some or all of the point and interval
+estimates may be biased, perhaps severely so, and the inference will
+be invalid. You can see how horribly biased the point *and* interval
 estimates of every SDT parameter may become when you overly aggregate
@@
 knowledge, to this day there is little *valid* evidence that the
-Unequal Variance SDT model fits the data better than the Equal
+Unequal Variance SDT model is closer to the true model than the Equal
 Variance SDT model.
@@
 saturated ordinal model, thus turning it into the Fixed Unequal
-Variance Ordinal Model. If there is evidence of non-selectivity in
-this model, we can keep trying with other modifications. If
-non-selectivity disappears, we have evidence that the selectivity
-assumption may be correct *and* that the Unequal Variance SDT model
-may be better than the Equal Variance SDT model.
-
-# A rant about Item Response Theory
+Variance Ordinal Model. I think this is a neat idea. If there is
+evidence of non-selectivity in this model, we can keep trying with
+other modifications. If non-selectivity disappears, we have evidence
+that the selectivity assumption may be correct *and* that the Unequal
+Variance SDT model may be better than the Equal Variance SDT model.
+
+# A rant about Item Response Theory (from someone who is relatively new to Item Response Theory)
 
@@
 variability in ordinal polytomous items is to allow for individual
-threshold effects while respecting the ordering of the
-thresholds. This can only be achieved by using the order-preserving
-link functions.
+threshold effects while respecting the ordering of the thresholds. In
+the ordinal model that we consider here this can only be achieved by
+using the order-preserving link functions.
 
@@
 It follows that *in such situations, we do not need the internal value
-parameter at all*, unless there is some factor about which we can
-perhaps assume that it can *only* produce a uniform effect in the
+parameter at all*, nor should we introduce it without a very good
+reason to do so. That is, unless there is some factor about which we
+can perhaps assume that it can *only* produce a uniform effect in the
 thresholds, such as the stimulus class ("noise" or "signal") in a
@@
 
-Notice that an SDT model is just an Item Response Theory model with
-the cumulative distribution and a non-negative effect of the stimulus
-class on the latent mean. So what about the central promise of Item
-Response Theory?  Isn't it true that we can separate the internal
-values ("person parameters") from the thresholds ("item parameters")
-and place them on a common scale? Once the participants leave the room
-the items retain their estimated "difficulty levels", right? The item
-difficulty (=thresholds) is not defined relative to some sample of
-participants who filled the questionnaire. How is this possible if the
-two kinds of effects are completely confounded? In particular, how
-does Item Response Theory achieve this magical separation when *each
-person provides only a few responses, usually just one, to each item*?
+So what about the central promise of Item Response Theory? Isn't it
+true that we can separate the internal values ("person parameters")
+from the thresholds ("item parameters") and place them on a common
+scale? Once the participants leave the room the items retain their
+estimated "difficulty levels", right? The item difficulty (=
+thresholds) is not defined relative to some sample of participants who
+filled the questionnaire. How is this separation possible if the two
+kinds of effects are completely confounded? In particular, how does
+Item Response Theory achieve this magical separation when *each person
+provides only a few responses, usually just one, to each item*?
 
 It is usually not an *achievement* of Item Response Theory, it is just
-an unrealistic *assumption*. The illusory nature of this achievement
-becomes obvious as soon as we realize that without additional strong
-causal assumptions it is impossible to distinguish between genuine
-internal value effects and (dis)simulation (i.e., faking), which
-conceptually corresponds to the change in the *participant*-specific,
-not item-specific, decision thresholds, also known as
-"item-parameters".
-
-There is nothing in the *design* of the process of measurement by a
-questionnaire that makes it possible to distinguish genuine
-differences in internal values from the differences in the way of
-responding. Because of random assignment of the stimulus class above
-chance performance in a binary classification task cannot be faked,
-and so the *d'* parameter definitely captures at least part of the
-effect of the stimulus on the internal value. On the other hand, every
-possible pattern of responses to a questionnaire can be faked.
-
-The only way to (perhaps incorrectly) separate the "item" and "person"
-effects is to *assume selective influence*, i.e., that there is some
-factor f (such as the stimulus class in a binary classification task,
-or, in case of a depression questionnaire, participants trait, or a
-therapeutic intervention) which can only affect the internal value,
-thus producing a constant shift in the thresholds. The assumption of
-selective influence makes the model with the two kinds of parameters
-identifiable.
+an unrealistic *causal assumption*. The illusory nature of this
+achievement becomes obvious as soon as we realize that without
+additional strong causal assumptions it is impossible to distinguish
+between genuine internal value effects and (dis)simulation (i.e.,
+faking), which conceptually seem to correspond to the change in the
+*participant*-specific, not item-specific, decision thresholds, also
+known as "item-parameters".
+
+There is nothing in the design of the process of measurement by a
+questionnaire that makes it possible to distinguish between the
+genuine differences in internal values and the differences in the way
+of responding. Because of random assignment of the stimulus class
+above chance performance in a binary classification task cannot be
+faked, and so the *d'* parameter definitely captures at least part of
+the effect of the stimulus on the internal value. On the other hand,
+every possible pattern of responses to a questionnaire can be faked.
 
@@
 "item-parameters" obtained for two different (often large) samples of
-participants are compared. However, these estimates are obtained by
-*fixing the "item-parameters" within each sample*.
+participants are compared. Moreover, these estimates are obtained by
+more or less arbitrarily fixing the "item-parameters" within each
+sample.
 
@@
 participants* clearly differ in how they use the main decision
-criterion. By the same token, instability of item parameters in IRT
-models cannot be estimated by assuming selective influence and
-comparing estimates of "item-parameters" obtained in two different
-large samples, because estimates of sample average item-parameters
+criterion. By the same token, variability in item parameters in IRT
+models cannot be correctly estimated just by assuming selective
+influence and comparing estimates of "item-parameters" obtained in two
+different samples, because estimates of sample average item-parameters
 *hide or misplace* (overdispersion) the intra- or inter-individual
@@
 stimulus class on the internal evidence value we have to *test* if the
-stimulus class does not also affect the thresholds.
+stimulus class does not also affect the thresholds. In principle, this
+means that we have to argue for the null.
 
@@
 influence of the stimulus class on the internal values is much more
-plausible than the - implicit in most applications of IRT - causal
+plausible than the - implicit in many applications of IRT - causal
 assumption of selective influence of participants on the
@@
 or right" vs "do you wish you were as happy as other people seem to
-be?") and the processes responsible for the generation of internal
+be?") and the processes responsible for the generation of the internal
 values and for the placement of the decision thresholds are likely to
@@
 ridiculous it should be laughed at on the street.
+
+# The missing part
+
+At the risk of sounding ridiculous, I will now make use of the fact
+that I can say whatever I want in an informal tutorial such as this
+one. Also, I just had a beer.
+
+It may seem that once (maybe in some distant future) we solve the
+statistical problems and the causal problems of testing an SDT model
+we can say that we are done. We have the great formal tools of
+statistical and causal inference at our disposal, and they are
+indispensable, but there is one last layer of abstraction that we did
+not explicitly identify.
+
+Let's assume that both the causal and the statistical assumptions of
+an SDT model are true. This means that *d'* represents the total
+causal effect of the stimulus class on the distribution of the
+responses. However, both kinds of assumptions may be true about an
+arbitrary physical process. For example. these assumptions may be true
+about some diagnostic device. We can say that this device is more or
+less sensitive, but the meaning of the term "sensitivity" is different
+in an important way than when we say that the participant *who
+performs the task* of binary classification is sensitive to the class
+of the stimulus.
+
+The meaning is different, because an arbitrary device is not yet an
+autonomous purposeful agent. In particular, the sense in which an
+arbitrary device "does" something i different than the sense in which
+a participant does something. To be a bit more precise, an arbitrary
+device does not *try* to do something, it just so happens that it was
+designed by an autonomous purposeful agent (one of us) to solve the
+problem faced by that agent, but it is the agent who "does all the
+trying". This layer of abstraction has to do with behavior, purpose,
+and function.
+
+It is only at this layer of abstraction that a mechanism becomes a
+*psychological* mechanism. It is my belief that it may be possible to
+formalize the process of inference at this level of abstraction. If
+such a formal theory of psychological inference was to be discovered,
+it would perhaps resemble an abstract theory of agents, tasks, and
+cognitive functions. Rational analysis (J.R. Anderson) and
+probabilistic models of cognition (Chater, Oaksford, Tenenbaum,
+Griffiths, etc) seem to be promising starting points, as does the
+Reinforcement Learning framework (Sutton and Barto) in which, to my
+knowledge, the first promising weak (a good thing), formal definition
+of purposeful action (i.e., of behavior) was proposed, inspired by the
+writings of Tolman and Perry.
+
+It is no accident that the internal values are often called "evidence
+samples" in Signal Detection Theory, or that the main criterion is
+interpreted in terms of bias, or that there are deep smilarities
+between the diffusion model and the process of sequential analysis. I
+hope I have made the point that explicitly addressing not only the
+statistical, but also the causal assumptions of Signal Detection
+Theory may dramatically improve our understanding of the process of
+fitting and interpreting the parameters of an SDT model. Who knows
+what we could achieve if the statistical, causal, *and* functional
+layers were explicitly combined.

### 2020-02-17 ordinal/Readme.Rmd
a note on free parameters

@@
 of purposeful action (i.e., of behavior) was proposed, inspired by the
-writings of Tolman and Perry.
+writings of Tolman and Perry. There were times when saw that the
+number of free parameters prevents us from answering certain valid and
+important questions. By identifying the limits of optimal purposeful
+action, this kind of analysis may sometimes significantly lower the
+number of free parameters in the models of psychological processes.
 

### 2020-02-18 ordinal/Readme.Rmd
grammar

@@
 autonomous purposeful agent. In particular, the sense in which an
-arbitrary device "does" something i different than the sense in which
+arbitrary device "does" something is different than the sense in which
 a participant does something. To be a bit more precise, an arbitrary
@@
 of purposeful action (i.e., of behavior) was proposed, inspired by the
-writings of Tolman and Perry. There were times when saw that the
+writings of Tolman and Perry. There were times when we saw that the
 number of free parameters prevents us from answering certain valid and

### 2020-02-18 ordinal/Readme.Rmd
wording

@@
 play well with aggregation, some or all of the point and interval
-estimates may be biased, perhaps severely so, and the inference will
-be invalid. You can see how horribly biased the point *and* interval
+estimates may be biased, perhaps severely so, and the inference may be
+invalid. You can see how horribly biased the point *and* interval
 estimates of every SDT parameter may become when you overly aggregate

### 2020-02-18 ordinal/Readme.Rmd
deleted the last part

@@
 be more stable.
-
-The thresholds are determined by the items, by the properties of the
-participants, and by the context in which the measurement takes
-place. The assumption that an important latent aspect of the process
-of psychological measurement is constant across participants is so
-ridiculous it should be laughed at on the street.
-
-# The missing part
-
-At the risk of sounding ridiculous, I will now make use of the fact
-that I can say whatever I want in an informal tutorial such as this
-one. Also, I just had a beer.
-
-It may seem that once (maybe in some distant future) we solve the
-statistical problems and the causal problems of testing an SDT model
-we can say that we are done. We have the great formal tools of
-statistical and causal inference at our disposal, and they are
-indispensable, but there is one last layer of abstraction that we did
-not explicitly identify.
-
-Let's assume that both the causal and the statistical assumptions of
-an SDT model are true. This means that *d'* represents the total
-causal effect of the stimulus class on the distribution of the
-responses. However, both kinds of assumptions may be true about an
-arbitrary physical process. For example. these assumptions may be true
-about some diagnostic device. We can say that this device is more or
-less sensitive, but the meaning of the term "sensitivity" is different
-in an important way than when we say that the participant *who
-performs the task* of binary classification is sensitive to the class
-of the stimulus.
-
-The meaning is different, because an arbitrary device is not yet an
-autonomous purposeful agent. In particular, the sense in which an
-arbitrary device "does" something is different than the sense in which
-a participant does something. To be a bit more precise, an arbitrary
-device does not *try* to do something, it just so happens that it was
-designed by an autonomous purposeful agent (one of us) to solve the
-problem faced by that agent, but it is the agent who "does all the
-trying". This layer of abstraction has to do with behavior, purpose,
-and function.
-
-It is only at this layer of abstraction that a mechanism becomes a
-*psychological* mechanism. It is my belief that it may be possible to
-formalize the process of inference at this level of abstraction. If
-such a formal theory of psychological inference was to be discovered,
-it would perhaps resemble an abstract theory of agents, tasks, and
-cognitive functions. Rational analysis (J.R. Anderson) and
-probabilistic models of cognition (Chater, Oaksford, Tenenbaum,
-Griffiths, etc) seem to be promising starting points, as does the
-Reinforcement Learning framework (Sutton and Barto) in which, to my
-knowledge, the first promising weak (a good thing), formal definition
-of purposeful action (i.e., of behavior) was proposed, inspired by the
-writings of Tolman and Perry. There were times when we saw that the
-number of free parameters prevents us from answering certain valid and
-important questions. By identifying the limits of optimal purposeful
-action, this kind of analysis may sometimes significantly lower the
-number of free parameters in the models of psychological processes.
-
-It is no accident that the internal values are often called "evidence
-samples" in Signal Detection Theory, or that the main criterion is
-interpreted in terms of bias, or that there are deep smilarities
-between the diffusion model and the process of sequential analysis. I
-hope I have made the point that explicitly addressing not only the
-statistical, but also the causal assumptions of Signal Detection
-Theory may dramatically improve our understanding of the process of
-fitting and interpreting the parameters of an SDT model. Who knows
-what we could achieve if the statistical, causal, *and* functional
-layers were explicitly combined.

### 2020-02-18 ordinal/Readme.Rmd
preprint links

@@
 model is not valid and a non-trivial generalization of SDT should be
-used instead (see the bhsdtr preprint for a more detailed explanation
-and some references).
+used instead (see the bhsdtr
+[preprint](http://dx.doi.org/10.23668/psycharchives.2725) for a more
+detailed explanation and some references).
 
@@
 correlations between the random effects. First such model was
-described in the bhsdtr preprint. This model, however, does not assume
-that the stimulus class affects only the *d'* parameter. We can now
-again see if there is any evidence of non-selectivity:
+described in the bhsdtr
+[preprint](http://dx.doi.org/10.23668/psycharchives.2725). This model,
+however, does not assume that the stimulus class affects only the *d'*
+parameter. We can now again see if there is any evidence of
+non-selectivity:
 
@@
 estimates of every SDT parameter may become when you overly aggregate
-your data in the bhsdtr preprint. By the way, I am now in the process
-of examining the direction and magnitude of bias in SDT models fitted
-to overly-aggregated data by using the excellent Confidence database
-created by Doby Rahnev.
+your data in the bhsdtr
+[preprint](http://dx.doi.org/10.23668/psycharchives.2725). By the way,
+I am now in the process of examining the direction and magnitude of
+bias in SDT models fitted to overly-aggregated data by using the
+excellent Confidence database created by Doby Rahnev.
 

### 2020-02-18 ordinal/Readme.Rmd
a remark about additive item and person effects

@@
 
-Unfortunately, without additional assumptions, it is impossible to
-model both kinds of variability at the same time even in the
-hypothetical repeated-measurement-with-memory-wipeout-scenario,
-because every shift in the internal value is indistinguishable from a
-uniform effect in the thresholds, i.e., a constant shift of all the
-thresholds. As I hope you already realize, the required additional
-assumptions that make such models identifiable under certain
-conditions are necessarily causal.
+Unfortunately, without additional assumptions (e.g., additivity of
+item and participant effects), it is impossible to model both kinds of
+variability at the same time even in the hypothetical
+repeated-measurement-with-memory-wipeout-scenario, because every shift
+in the internal value is indistinguishable from a uniform effect in
+the thresholds, i.e., a constant shift of all the thresholds. As I
+hope you already realize, the required additional assumptions that
+make such models identifiable under certain conditions are necessarily
+causal.
 
